### 2009-06-22 VerboseNotes.tex
QFTの素材など色々追加。
git-svn-id: https://dav.misho-web.com/svn/cyghome/trunk/QFT/CheatSheet@181 c04c31ef-c4f7-4b37-ba94-86ebe2b9c820

%#!latexmk CheatSheet.tex
%%% Time-Stamp: <2009-06-22 08:11:22>

\section{Verbose Notes}
\subsection{Polarization Sum}\label{Sec:Verbose:PolarizationSum}
Firstly we focus on the single photon case $M=\epsilon_\mu^*(k)\epsilon_\nu'^*(k')M^{\mu\nu}$.
Here we set $k=(E,0,0,E)$, and $\epsilon=(0,1,0,0)\oplus(0,0,1,0)$. Then
\begin{equation}
  \sum\s{pol.}\left|M\right|^2
= \sum\s{pol.}\epsilon_\mu^*(k)\epsilon_\nu(k)M^{\mu}M^{\nu*}
= |M^1|^2+|M^2|^2,
\end{equation}
while
\begin{equation}
 \Hmnd M^{\mu}M^{\nu*} = |M^1|^2+|M^2|^2
\end{equation}
for Ward identity $k_\mu M^\mu=0$. Therefore the replacement
\begin{equation}
 \sum\s{pol.}\epsilon_\mu\epsilon'_\nu \to \Hmnd
\end{equation}
is valid.

Secondly we think about the double photons case
\footnote{This part is derived from ‡_å˚çKàÍ's notebook.}
$M=\epsilon_\mu^*(k)\epsilon_\nu'^*(k')M^{\mu\nu}$.
Here we set
\begin{align}
 k &=(E,0,0, E) & \epsilon &=(0,1,0,0)\oplus(0,0,1,0)\\
 k'&=(E,0,0,-E) & \epsilon'&=(0,\cos\theta,\sin\theta,0)\oplus(0,-\sin\theta,\cos\theta,0).
\end{align}
Then doing some simple calculations, we can get
\begin{equation}
  \sum\s{pol.}\left|M\right|^2 = |M^{11}|^2+|M^{12}|^2+|M^{21}|^2+|M^{22}|^2.
\end{equation}
Nevertheless, na\"ive replacement does not work, because our Ward identities
\begin{equation}
 k_\mu\epsilon_\nu'^*(k')M^{\mu\nu} =
 \epsilon_\mu^*(k)k'_\nu M^{\mu\nu} = 0 \label{eq:WardIdentities}
\end{equation}
obviously does not help us. If we can omit $\epsilon$s from these
identities, that is if
\begin{equation}
 k_\mu M^{\mu\nu} = k'_\nu M^{\mu\nu} = 0,\label{eq:TrueIdentities}
\end{equation}
we can recover validity of the replacement:
\begin{align}
 \Hmrd\Hnsd M^{\mu\nu}M^{\rho\sigma*}
&= -\Hnsd\left(
M^{1\nu}M^{1\sigma*}+M^{2\nu}M^{2\sigma*}
\right)\\
&= |M^{11}|^2+|M^{12}|^2+|M^{21}|^2+|M^{22}|^2.
\end{align}

Then what's happening? Why this replacement is not valid?
Actually our new conditions \eqref{eq:TrueIdentities} seem to guarantee
that we are summing not only ``physical'' but also ``unphysical'' polarizations.
Meanwhile if we use some physical condition such as $\epsilon\cdot k=0$,
\eqref{eq:TrueIdentities} break down while Ward identities
\eqref{eq:WardIdentities} are still valid.

Now let's check what is happening from another viewpoint. First we suppose
$M$ satisfies our new conditions \eqref{eq:TrueIdentities}, and define
$\tilde M^{\mu\nu}$ and $\tilde M$ as
\begin{align}
 \tilde M^{\mu\nu}& := M^{\mu\nu} + k^\mu p^\nu + p'^\mu k'^\nu,\\
 \tilde M         & := \epsilon_\mu^*(k)\epsilon_\nu'^*(k')\tilde M^{\mu\nu}.
\end{align}
This alternative amplitude satisfies Ward identities (since photon is
massless and $\epsilon\cdot k=0$), and furthermore $\tilde M = M$.
Therefore $\tilde M$ is physically identical to $M$.
However technically these are very different, just because we cannot
perform our ``na\"ive replacement'' for this $\tilde M$:
\begin{align}
 \Hmrd\Hnsd \tilde M^{\mu\nu}\tilde M^{\rho\sigma*}
 &=  \Hmrd\Hnsd
\left(M^{\mu\nu} + k^\mu p^\nu + p'^\mu k'^\nu \right)
\left(M^{\rho\sigma*} + k^{\rho} p^{\sigma*} + p'^{\rho*}k'^{\sigma} \right)\\
 &= \sum\s{pol.}|M|^2 + \left[(k\cdot p'^*)(k'\cdot p) + \Hc\right].
\end{align}
After all, we have obtained following expression:
\begin{align}
 \sum\s{pol.}|\tilde M|^2 &=
 \sum\s{pol.}|M|^2\qquad \text{(Furthermore $\tilde M=M$)}\notag\\
&= \sum\s{pol.}|\epsilon_\mu^*(k)\epsilon_\nu'^*(k')M^{\mu\nu}|^2
 = \sum\s{pol.}|\epsilon_\mu^*(k)\epsilon_\nu'^*(k')\tilde M^{\mu\nu}|^2\\
&= \Hmrd\Hnsd M^{\mu\nu}M^{\rho\sigma*}\notag\\
&\ne \Hmrd\Hnsd \tilde M^{\mu\nu}\tilde M^{\rho\sigma*}
 = \sum\s{pol.}|\tilde M|^2 + \left[(k\cdot p'^*)(k'\cdot p) + \Hc\right].\notag
\end{align}

%%% Local Variables:
%%% TeX-master: CheatSheet.tex
%%% End:

### 2010-03-22 VerboseNotes.tex
Cheatsheetをだいぶ整理＆修正
git-svn-id: https://dav.misho-web.com/svn/cyghome/trunk/QFT/CheatSheet@380 c04c31ef-c4f7-4b37-ba94-86ebe2b9c820

@@
 %#!latexmk CheatSheet.tex
-%%% Time-Stamp: <2009-06-22 08:11:22>
+%%% Time-Stamp: <2010-03-23 00:38:46 misho>
 
@@
 
+\subsection{Phantom Terms in the Gauge Theory}
+\label{sec:no-other-term}
+You may think we forget to introduce
+$\ol\psi\G5\psi$ÅC
+$\ol\psi\G5\slashed D\psi$ÅC
+$\epsilon^{\mu\nu\rho\sigma}F^a_{\mu\nu}F^a_{\rho\sigma}$ÅC
+$\epsilon^{\mu\nu\rho\sigma}D_\mu D_\nu F^a_{\rho\sigma}$
+terms, but being a bit careful,
+\begin{itemize}
+ \item the first two terms are nonsense, for now we use $\PL$ and $\PR$,
+ \item the last term is equivalent to the third term as\\[.5zw]\qquad
+$\epsilon^{\mu\nu\rho\sigma}D_\mu D_\nu F^a_{\rho\sigma}
+=\epsilon^{\mu\nu\rho\sigma}\dfrac12[D_\mu,D_\nu]F^a_{\rho\sigma}
+=\dfrac12\epsilon^{\mu\nu\rho\sigma}F^a_{\mu\nu}F^a_{\rho\sigma}$.
+\end{itemize}\vspace{.5zw}
+Therefore, we have to discuss only the $\epsilon FF$ terms.
+If the gauge group is simple, we can take the structure constant as totally antisymmetric, which leads these terms to fall into surface terms as:
+\begin{align}
+ \epsilon^{\mu\nu\rho\sigma}f^{abc}f^{ade}A^b_\mu A^c_\nu A^d_\rho A^e_\sigma
+ &= \epsilon^{\mu\nu\rho\sigma}\left(-f^{acd}f^{abe}-f^{adb}f^{ace}\right)
+     A^b_\mu A^c_\nu A^d_\rho A^e_\sigma\notag\\
+ &= -2\epsilon^{\mu\nu\rho\sigma}f^{abc}f^{ade}A^b_\mu A^c_\nu A^d_\rho A^e_\sigma\\
+ &=  0,\notag
+\end{align}
+\begin{align}
+ \therefore\ \epsilon^{\nu\nu\rho\sigma}F_{\mu\nu}F_{\rho\sigma}
+&= 4\epsilon^{\mu\nu\rho\sigma}\Pm A^a_\nu\Pr A^a_\sigma
+ + 4g\epsilon^{\mu\nu\rho\sigma}f^{abc}A^a_\mu A^b_\nu\Pr A^c_\sigma\notag\\
+&= 2\Pm G^\mu,&
+\end{align}
+where $G^\mu$ is the Chern--Simons term which is defined as
+\begin{equation}
+  G^\mu
+:= 2\epsilon^{\mu\nu\rho\sigma}
+\left( A^a_\nu\Pr A^a_\sigma+\frac13gf^{abc}A^a_\nu A^b_\rho A^c_\sigma \right)
+= \epsilon^{\mu\nu\rho\sigma}
+\left( A^a_\nu F^a_{\rho\sigma}-\frac13gf^{abc}A^a_\nu A^b_\rho A^c_\sigma \right).
+\end{equation}
+See Appendix~\ref{sec:cs-instanton} for the instanton effect.
+
+\subsection{Instanton}
+\label{sec:cs-instanton}
+\TODO{Postpone...}
 %%% Local Variables:
-%%% TeX-master: CheatSheet.tex
+%%% TeX-master: "CheatSheet.tex"
 %%% End:

### 2010-03-28 VerboseNotes.tex
CheatSheet
git-svn-id: https://dav.misho-web.com/svn/cyghome/trunk/QFT/CheatSheet@381 c04c31ef-c4f7-4b37-ba94-86ebe2b9c820

@@
 %#!latexmk CheatSheet.tex
-%%% Time-Stamp: <2010-03-23 00:38:46 misho>
+%%% Time-Stamp: <2010-03-25 01:33:22 misho>
+\section{Verbose Notes}
+\subsection{Spinor Fields}
+\subsubsection{Lorentz group and Lorentz algebra}
+\begin{tabular}{l@{ :\ \ \ }l}
+Metric &  $\eta^\% = \diag(+1, -1, -1, -1),\qquad \eta^\# = \diag(-1,+1,+1,+1).$\\
+Lorentz transf. in $\mathbb R^{1,3}$ & Linear transf. $x^\mu \mapsto \Lambda\T^\mu_\nu x^\nu$ which conserve $x^2$.\\
+ & \then $\Hrsd = \Hmnd \Lambda\T^\mu_\rho \Lambda\T^\nu_\sigma.$
+ and form a group $L$.\\[.8zw]
 
-\section{Verbose Notes}
+Disconnected parts of $L$ &
+  $L_0 := \left\{ \det\Lambda = +1 \land \Lambda\T_0^0 >0\right\} \quad
+   L_P := \left\{ \det\Lambda = -1 \land \Lambda\T_0^0 >0\right\}$\\
+& $L_T := \left\{ \det\Lambda = +1 \land \Lambda\T_0^0 <0\right\} \quad
+   L_{PT} := \left\{ \det\Lambda = -1 \land \Lambda\T_0^0 <0\right\}$\\
+& \qquad ($L_0$ is identical with $\%\gSO(1,3) / \#\gSO(3,1)$. )\\[1zw]
+
+Infinitesimal one in $L_0$& $\Lambda\T^\mu_\nu = \delta^\mu_\nu + \epsilon\T^\mu_\nu$ where $\epsilon_{\mu\nu}=-\epsilon_{\nu\mu}$ (for $\eta=\eta\Lambda\Lambda$)\\[1zw]
+
+Generators &
+ $x'^\mu =  x^\mu + \epsilon\T^\mu_\nu x^\nu
+   =: \left(\delta^\mu_\nu\mp\frac\ii2\epsilon_{\rho\sigma}(J^{\rho\sigma})\T^\mu_\nu\right)x^\nu$
+   \quad with $J^{\rho\sigma} = -J^{\sigma\rho}$\\
+&\then 
+$ (J_{\rho\sigma})\T^\mu_\nu = \mp\ii\left(
+                                     \delta^\mu_\rho\Hsnd-\delta^\mu_\sigma\Hrnd
+                                    \right),$\\
+Lorentz algebra & therefore
+ $[J_{\mu\nu},J_{\rho\sigma}] = \pm\ii(
+    \Hmrd J_{\nu\sigma} + \Hnsd J_{\mu\rho} - \Hmsd J_{\nu\rho} - \Hnrd J_{\mu\sigma})$\\
+\end{tabular}
+
+\vspace{1zw}
+Here we physicists derive the algebra from the property of the Lorentz transformation, but we can start from the albegra like mathematicians.
+
+\TODO{Jの符号確認}
+
+\subsubsection{Lorentz group and $\gSL(2,\Complex)$}
+Here we will explain that $L_0$ is isomorphic to $\gSL(2,\Complex)/\gZ_2$, which is defined as
+\begin{align}
+ \aSL(2,\Complex) :=& \{a\in \aGL(2,\Complex)\ |\ \Tr(a)=0\},&
+ \gSL(2,\Complex) :=& \{g\in \gGL(2,\Complex) |\ \det(g)=1\}.
+\end{align}
+
+We can build a one-to-one correspondent between $x^\mu$ and an Hermitian matrix
+\begin{equation}
+ A := t\cdot1+\vipro x \sigma = \pmat{t+z&x-\ii y\\x+\ii y&t-z}.
+\end{equation}
+Then, we can define Lorentz transformations $f^g$ from $g\in\gSL(2,\Complex)$ as
+\begin{equation}
+ f^g: A\mapsto g^\dagger Ag,
+\end{equation}
+and thus we can induce Lorentz transformations $\Lambda(g)$ from $g$.
+
+Actually, $\Lambda(g)$ covers only $L_0$, not whole $L$. I do not provide the proof, but since $\gSL(2,\Complex)$ is a connected group,
+it is clear that it can cover only the connected part of Lorentz group.
+\begin{rightnote}
+I omit the proof here, but in principle, it can be done by direct calculation. For instance, $\Lambda\T^0_0>0$ can be proved as follows:
+
+Since $t = \frac12\Tr(t\cdot1+\vipro x\sigma)$,
+$\Lambda\T^0_0$ is nothing but
+$\frac1{2t}\Tr\left[g^\dagger\left({t\atop0}{0\atop t}\right)g\right]$.
+Therefore
+\begin{equation}
+ \Lambda\T^0_0
+ = \frac12\Tr\left(g^\dagger g\right)
+ = \frac12\left(|g_{11}|+|g_{12}|+|g_{21}|+|g_{22}|\right)>0.
+\end{equation}
+This shows $\Lambda(g)$ are orthochronous.
+\end{rightnote}
+
+As an element of $\aSU(2)$ can be expressed by the 3 bases known as Pauli matrices,
+
+The Lorentz group has 6 generators $J_{\rho\sigma}$, and these corresponds to
+the 6 elements of $\aSL(2,\Complex)$.
+These can be broken down into three ``rotation generators'' $\vc J$, and three ``boost generators'' $\vc K$ as follows:
+\begin{align}
+ J_i &:= (J_{23}, J_{31}, J_{12}) &  K_i &:= (J_{10},J_{20},J_{30}) &\for i=1,2,3.\footnotemark
+\end{align}
+\footnotetext{For your information, $J_{23} = J^{23}$ and $K_{01} = -K^{01}$, regardless of the definition of the metic.}
+With this definition, the algebra is, regardless of signature of the metric, as follows:
+\begin{align}
+ [J_i, J_j]&= \ii\epsilon_{ijk}J_k,&
+ [J_i, K_j]&= \ii\epsilon_{ijk}K_k,&
+ [K_i, K_j]&= -\ii\epsilon_{ijk}J_k.
+\end{align}
+
+Then, we define $\vc A$ and $\vc B$ as
+\begin{align}
+ \vc A&:=\frac12(\vc J+\ii\vc K), & \vc B &:= \frac12(\vc J-\ii\vc K).
+\end{align}
+Now the commutation relations are
+\begin{align}
+ [A_i, A_j] &= \ii\epsilon_{ijk}A_k,&
+ [B_i, B_j] &= \ii\epsilon_{ijk}B_k,&
+ [A_i, B_j] &= 0,
+\end{align}
+which means we have reduced the Lorentz group into $\gSU(2)\times\gSU(2)$.
+\starline
+
+In some neighborhood of the identity, the elements $\Lambda$ of Lorentz group can be described as
+\begin{equation}
+ \Lambda = \exp(\mp\frac\ii2\epsilon_{\rho\sigma}J^{\rho\sigma}),
+\end{equation}
+using the (antisymmetric) generators $J^{\rho\sigma}$.
+Here, defining $\vc \theta$ and $\vc \beta$ as
+\begin{align}
+ \pm\theta_i&:= (\epsilon_{23},\epsilon_{31},\epsilon_{12})&
+ \pm\beta_i &:= (\epsilon_{10},\epsilon_{20},\epsilon_{30})&
+ \for i=1,2,3,
+\end{align}
+the element is denoted as:
+\begin{equation}
+ \Lambda = \exp\left[\ii\left(\vipro\theta J + \vipro \omega K\right)\right].
+\end{equation}
+
+\subsection{Weyl Spinor}
+Now we introduce Weyl spinors. First, I de
+%%% Local Variables:
+%%% TeX-master: t
+%%% End:
+
 \subsection{Polarization Sum}\label{Sec:Verbose:PolarizationSum}
@@
 Secondly we think about the double photons case
-\footnote{This part is derived from ‡_å˚çKàÍ's notebook.}
+\footnote{This part is derived from 濱口幸一's notebook.}
 $M=\epsilon_\mu^*(k)\epsilon_\nu'^*(k')M^{\mu\nu}$.
@@
 You may think we forget to introduce
-$\ol\psi\G5\psi$ÅC
-$\ol\psi\G5\slashed D\psi$ÅC
-$\epsilon^{\mu\nu\rho\sigma}F^a_{\mu\nu}F^a_{\rho\sigma}$ÅC
+$\ol\psi\G5\psi$，
+$\ol\psi\G5\slashed D\psi$，
+$\epsilon^{\mu\nu\rho\sigma}F^a_{\mu\nu}F^a_{\rho\sigma}$，
 $\epsilon^{\mu\nu\rho\sigma}D_\mu D_\nu F^a_{\rho\sigma}$

### 2010-04-27 VerboseNotes.tex
cheatsheet拡充
git-svn-id: https://dav.misho-web.com/svn/cyghome/trunk/QFT/CheatSheet@383 c04c31ef-c4f7-4b37-ba94-86ebe2b9c820

@@
 %#!latexmk CheatSheet.tex
-%%% Time-Stamp: <2010-03-25 01:33:22 misho>
+%%% Time-Stamp: <2010-04-27 22:42:12 misho>
 \section{Verbose Notes}
@@
 \begin{tabular}{l@{ :\ \ \ }l}
-Metric &  $\eta^\% = \diag(+1, -1, -1, -1),\qquad \eta^\# = \diag(-1,+1,+1,+1).$\\
+Metric &  ${\RED\eta} = \diag(+1, -1, -1, -1),\qquad {\BLUE \eta} = \diag(-1,+1,+1,+1).$\\
 Lorentz transf. in $\mathbb R^{1,3}$ & Linear transf. $x^\mu \mapsto \Lambda\T^\mu_\nu x^\nu$ which conserve $x^2$.\\
@@
    L_{PT} := \left\{ \det\Lambda = -1 \land \Lambda\T_0^0 <0\right\}$\\
-& \qquad ($L_0$ is identical with $\%\gSO(1,3) / \#\gSO(3,1)$. )\\[1zw]
+& \qquad ($L_0$ is identical with ${\RED\gSO(1,3)} / {\BLUE\gSO(3,1)}$. )\\[1zw]
 
@@
 &\then 
-$ (J_{\rho\sigma})\T^\mu_\nu = \mp\ii\left(
+$ (J_{\rho\sigma})\T^\mu_\nu = \pm\ii\left(
                                      \delta^\mu_\rho\Hsnd-\delta^\mu_\sigma\Hrnd
@@
 Lorentz algebra & therefore
- $[J_{\mu\nu},J_{\rho\sigma}] = \pm\ii(
+ $[J_{\mu\nu},J_{\rho\sigma}] = \mp\ii(
     \Hmrd J_{\nu\sigma} + \Hnsd J_{\mu\rho} - \Hmsd J_{\nu\rho} - \Hnrd J_{\mu\sigma})$\\
@@
 \subsection{Weyl Spinor}
-Now we introduce Weyl spinors. First, I de
+
 %%% Local Variables:

### 2010-04-28 VerboseNotes.tex
git-svn-id: https://dav.misho-web.com/svn/cyghome/trunk/QFT/CheatSheet@384 c04c31ef-c4f7-4b37-ba94-86ebe2b9c820

--- 2010-03-22 VerboseNotes.tex
@@
 %#!latexmk CheatSheet.tex
-%%% Time-Stamp: <2010-03-23 00:38:46 misho>
+%%% Time-Stamp: <2010-04-28 20:33:15 misho>
+\section{Verbose Notes}
+\subsection{Spinor Fields}
+\subsubsection{Lorentz group and Lorentz algebra}
+\begin{tabular}{l@{ :\ \ \ }l}
+Metric &  ${\RED\eta} = \diag(+1, -1, -1, -1),\qquad {\BLUE \eta} = \diag(-1,+1,+1,+1).$\\
+Lorentz transf. in $\mathbb R^{1,3}$ & Linear transf. $x^\mu \mapsto \Lambda\T^\mu_\nu x^\nu$ which conserve $x^2$.\\
+ & \then $\Hrsd = \Hmnd \Lambda\T^\mu_\rho \Lambda\T^\nu_\sigma.$
+ and form a group $L$.\\
+ & \qquad
+   ($\then\quad
+     (\Lambda^{-1})\T^\mu_\nu=\eta_{\nu\alpha}\eta^{\mu\beta}\Lambda\T^\alpha_\beta=:\Lambda\T_\nu^\mu\quad\then\quad\Lambda\T^\mu_\nu\Lambda\T_\mu^\rho=\delta^\rho_\nu$)\\[.8zw]
 
-\section{Verbose Notes}
+Disconnected parts of $L$ &
+  $L_0 := \left\{ \det\Lambda = +1 \land \Lambda\T_0^0 >0\right\} \quad
+   L_P := \left\{ \det\Lambda = -1 \land \Lambda\T_0^0 >0\right\}$\\
+& $L_T := \left\{ \det\Lambda = +1 \land \Lambda\T_0^0 <0\right\} \quad
+   L_{PT} := \left\{ \det\Lambda = -1 \land \Lambda\T_0^0 <0\right\}$\\
+& \qquad ($L_0$ is identical with ${\RED\gSO(1,3)} / {\BLUE\gSO(3,1)}$. )\\[1zw]
+Infinitesimal one in $L_0$& $\Lambda\T^\mu_\nu = \delta^\mu_\nu + \epsilon\T^\mu_\nu$ where $\epsilon_{\mu\nu}=-\epsilon_{\nu\mu}$ (for $\eta=\eta\Lambda\Lambda$)\\[1zw]
+\end{tabular}
+
+\vspace{1zw}
+
+ここで微小変換から生成子を
+ $x'^\mu =  x^\mu + \epsilon\T^\mu_\nu x^\nu
+   =: \left(\delta^\mu_\nu\mp\frac\ii2\epsilon_{\rho\sigma}(J^{\rho\sigma})\T^\mu_\nu\right)x^\nu$
+  ただし$J^{\rho\sigma} = -J^{\sigma\rho}$と定義すると，
+$ (J_{\rho\sigma})\T^\mu_\nu =
+\pm\ii\left(\delta^\mu_\rho\Hsnd-\delta^\mu_\sigma\Hrnd\right)$
+となり，交換関係が得られ，これが閉じているのでLie代数であることもわかる：
+\begin{equation}
+ [J_{\mu\nu},J_{\rho\sigma}] = \mp\ii(
+    \Hmrd J_{\nu\sigma} + \Hnsd J_{\mu\rho} - \Hmsd J_{\nu\rho} - \Hnrd J_{\mu\sigma}).
+\end{equation}
+生成子は
+$
+ \displaystyle J\T_1_0^\mu_\nu=({\RED \pm\ii},{\BLUE \mp\ii})
+ \spmat{0&1&0&0\\1&0&0&0\\0&0&0&0\\0&0&0&0}_{\mu\nu},
+ \displaystyle J\T_2_3^\mu_\nu=({\RED \pm\ii},{\BLUE \mp\ii})
+ \spmat{0&0&0&0\\0&0&0&0\\0&0&0&-1\\0&0&1&0}_{\mu\nu}
+$となる。
+
+\subsubsection{Lorentz group and $\gSL(2,\Complex)$}
+次に，連結Lie群$L_0$が，連結Lie群$\gSL(2,\Complex)/\gZ_2$と同型であることを見る：
+\begin{align}
+ \aSL(2,\Complex) :=& \{a\in \aGL(2,\Complex)\ |\ \Tr(a)=0\},&
+ \gSL(2,\Complex) :=& \{g\in \gGL(2,\Complex) |\ \det(g)=1\}.
+\end{align}
+まず,$\Sm$を（極めて一般的に） $\Sm:=(\alpha1,\beta\vc\sigma)$と定義する($\alpha=\beta=\pm1$)。
+$x^2 = ({\RED+}{\BLUE-})\det(x_\mu\Sm)$なので
+\begin{equation}
+ f^g: (x_\mu\Sm)\mapsto g (x_\mu\Sm) g^\dagger;\quad g\in\gSL(2,\Complex)
+\end{equation}
+は$x^2$を保存する。よってLorentz変換であり，生成子を比べることで局所同型だとわかる：
+\begin{rightnote}
+$\gSL(2,\Complex)\ni g = \exp(-\ii a)$として
+$x_\mu (g \Sm g^\dagger) = \Lambda\T_\mu^\nu x_\nu \Sm$を微小展開すると
+\begin{equation}
+\Lambda\T^\mu_\nu\Sn=g^{-1}\Sm (g^{-1})^\dagger\quad\then\quad
+\mp\frac12\epsilon_{\rho\sigma}(J^{\rho\sigma})\T^\mu_\nu\Sn=a\Sm-\Sm a^\dagger
+\end{equation}
+$\epsilon^{10}=\epsilon$のとき
+$a = ({\RED-}{\BLUE+})\cdot\dfrac\ii2\alpha\beta\sigma_x$
+であり，$\epsilon^{23}=\epsilon$のとき
+$a = ({\RED+}{\BLUE-})\cdot\dfrac12\sigma_x$である。
+\end{rightnote}
+
+このことを別の観点から見る。Lorentz群の生成子の交換関係を見ると，
+\begin{align}
+ J_i &:= (J_{23}, J_{31}, J_{12}) &  K_i &:= (J_{10},J_{20},J_{30}) &\for i=1,2,3.\footnotemark
+\end{align}
+に対して\footnotetext{For your information, $J_{23} = J^{23}$ and $K_{01} = -K^{01}$, regardless of the definition of the metic.}
+\begin{align}
+ [J_i, J_j]&= ({\RED\pm\ii},{\BLUE\mp\ii})\epsilon_{ijk}J_k,&
+ [J_i, K_j]&= ({\RED\pm\ii},{\BLUE\mp\ii})\epsilon_{ijk}K_k,&
+ [K_i, K_j]&= -({\RED\pm\ii},{\BLUE\mp\ii})\epsilon_{ijk}J_k
+\end{align}
+となるので，
+\begin{align}
+ \vc A&:=\frac{P}2(\vc J+\ii\vc K), & \vc B &:= \frac{P}2(\vc J-\ii\vc K).
+\end{align}
+と定義すると
+\begin{align}
+ [A_i, A_j] &= \ii\epsilon_{ijk}A_k,&
+ [B_i, B_j] &= \ii\epsilon_{ijk}B_k,&
+ [A_i, B_j] &= 0,
+\end{align}
+となり，Lorentz群が$\gSU(2)\times\gSU(2)$に分解できる。ここで$P=({\RED\pm1},{\BLUE\mp1})$であるので，$P=+1$となるように，複号の$({\RED 上},{\BLUE 下})$側を取ったことにする。
+
+\subsection{Weyl Spinor}
+\TODO{この部分は$+---$の場合の定義。$-+++$のときの右左の定義は？}
+左巻きspinor $\xi$ と右巻きspinor $\bar\xi$ を，
+\begin{align}
+ \xi&\mapsto\left(1-\frac\ii2\vipro\theta\sigma-\frac12\vipro\beta\sigma\right)\xi&
+ \bar\xi&\mapsto\left(1-\frac\ii2\vipro\theta\sigma+\frac12\vipro\beta\sigma\right)\bar\xi
+\end{align}
+と変換するものとして定義する。
+$\gSU(2)_A\times\gSU(2)_B$に対して左巻きは$(1/2,0)$表現，右巻きは$(0,1/2)$表現となっている。
+
+$g$を左巻きspinorの変換子として定義すると，$\vc J\to\frac12\vc \sigma$および$\vc K\to-\frac\ii2\alpha\beta\vc\sigma$の対応があったことを思い出すと
+$-\ii(\vipro\theta J+\vipro\beta K)=-\frac\ii2\vipro\theta\sigma-\frac12\alpha\beta\vipro\beta\sigma$であるので，$\alpha\beta=1$が分かる。
+\begin{align}
+  \xi_\alpha      &\mapsto g\T_\alpha^\beta\xi_\beta,
+& \bar\xi_\dalpha &\mapsto \bar\xi_\dbeta(g^\dagger)\T^\dbeta_\dalpha
+\end{align}
+であり，$\bar\xi=\xi^\dagger$，つまり$\bar\xi_\dalpha = (\xi_\alpha)^*$である。
+
+次に，$\xi^\alpha\chi_\alpha$および$\bar\xi_\dalpha\bar\chi^\dalpha$がscalarとなるようにしたい。
+$E=\spmat{0&1\\-1&0}$としておくと，$-E\trans{g}E=g^{-1}$であるので，
+\begin{equation}
+ (\xi')^\alpha=\xi^\beta(g^{-1})\T_\beta^\alpha
+              =-\xi^\beta(E_{\beta A}g\T_B^AE_{B\alpha})\qquad\therefore (-E_{\gamma\alpha})(\xi')^\alpha = -g\T_\gamma^\beta(E_{\beta\delta}\xi^\delta).
+\end{equation}
+よって，$\epsilon^{12}=\epsilon_{21}=1$として
+$\xi^\alpha:=\epsilon^{\alpha\beta}\xi_\beta$，$\xi_\alpha=\epsilon_{\alpha\beta}\xi^\beta$
+とすれば良い。
+
+同様に，$(\bar\xi^\dalpha E_{\dalpha\dbeta})'=(\bar\xi^\dalpha E_{\dalpha\dbeta})'g^\dagger\T^\dgamma_\dbeta$……としても良かったが，左からの演算，および$\epsilon_{12}=\epsilon_{\dot1\dot2}$の性質を重視して$(\epsilon_{\dalpha\dbeta}\bar\xi^\dbeta)'=(\epsilon_{\dgamma\dbeta}\bar\xi^\dbeta)'g^\dagger\T^\dgamma_\dbeta$とするのが一般的である。
+
+$x_\mu\Sm\mapsto x_\mu(g\Sm g^\dagger)$であるので，
+$x_\mu(\xi^\alpha \Sm \bar\chi^\dalpha)$はscalarである。よって，$\Sm_{\alpha\dalpha}$のように書ける。
+更にここで$x_\mu(\bar\xi_\dalpha\bSm{}^{\dalpha\alpha}\chi_\alpha)$もscalarとなるように$\bar\sigma$を定めよう。
+\begin{equation}
+ \bar\xi_\dalpha\bSm{}^{\dalpha\alpha}\chi_\alpha=-\epsilon_{\alpha\beta}\epsilon_{\dalpha\dbeta}\chi^\beta\bSm{}^{\dalpha\alpha}\bar\xi^\dbeta
+\qquad\therefore \Sm_{\beta\dbeta}\propto \epsilon_{\alpha\beta}\epsilon_{\dalpha\dbeta}\bSm{}^{\dalpha\alpha}
+\end{equation}
+であり，あとはconventionである。
+
+\starline
+
+\begin{align}
+  \epsilon^{12}=\epsilon_{21}=\epsilon^{\dot1\dot2}=\epsilon_{\dot2\dot1}&=1,&
+\xi^\alpha&:=\epsilon^{\alpha\beta}\xi_\beta,& \xi_\alpha&=\epsilon_{\alpha\beta}\xi^\beta,&
+\bar\xi^\dalpha&:=\epsilon^{\dalpha\dbeta}\bar\xi_\dbeta,& \bar\xi_\dalpha&=\epsilon_{\dalpha\dbeta}\bar\xi^\dbeta
+\end{align}
+\begin{align}
+  \xi_\alpha      &\mapsto g\T_\alpha^\beta\xi_\beta,
+& \xi^\alpha      &\mapsto \xi^\beta (g^{-1})\T_\beta^\alpha,
+& \bar\xi_\dalpha &\mapsto \bar\xi_\dbeta(g^\dagger)\T^\dbeta_\dalpha
+& \bar\xi^\dalpha &\mapsto (g^\dagger)^{-1}\T^\dalpha_\dbeta\bar\xi^\dbeta
+\end{align}
+\begin{align}
+  \bSm{}^{\dalpha\alpha} &:= \epsilon^{\alpha\beta}\epsilon^{\dalpha\dbeta}\Sm_{\beta\dbeta}&
+ \Sm_{\alpha\dalpha}&= \epsilon_{\alpha\beta}\epsilon_{\dalpha\dbeta}\bSm{}^{\dbeta\beta}&
+ \therefore \Sm:=(1,\vc\sigma),\quad\bSm:=(1,-\vc\sigma)
+\end{align}
+
 \subsection{Polarization Sum}\label{Sec:Verbose:PolarizationSum}

### 2010-04-28 VerboseNotes.tex
何もわからない……；；
git-svn-id: https://dav.misho-web.com/svn/cyghome/trunk/QFT/CheatSheet@385 c04c31ef-c4f7-4b37-ba94-86ebe2b9c820

@@
 %#!latexmk CheatSheet.tex
-%%% Time-Stamp: <2010-04-28 20:33:15 misho>
+%%% Time-Stamp: <2010-04-28 22:13:13 misho>
 \section{Verbose Notes}
@@
 \end{equation}
+
+%\vspace{5zw}
+%$x$方向のBoost演算子$K_x$と，$x$軸周りの回転演算子$J_x$は，それぞれ
+%\begin{align}
+%  \pmat{m\\p}&\longrightarrow\pmat{1&\beta\\\beta&1}\pmat{m\\p},
+%& \pmat{y\\z}&\longrightarrow\pmat{1&-\theta\\\theta&1}\pmat{y\\z},
+%&\therefore 
+%\epsilon&=\beta\spmat{0&1&0&0\\1&0&0&0\\0&0&0&0\\0&0&0&0},
+%         \theta\spmat{0&0&0&0\\0&0&0&0\\0&0&0&-1\\0&0&1&0}
+%\end{align}
+%に対応するものである。$\vc J$と$\vc K$を
+%\begin{align}
+% J_i &:= (J_{23}, J_{31}, J_{12}) &  K_i &:= (J_{10},J_{20},J_{30})
+%\end{align}
+%のように定めることにすると，複号は({\RED 上},{\BLUE 下})側を選ばねばならない。
+%\vspace{5zw}
+
 生成子は
@@
 $となる。
+ただし，$x$方向のboostは$({\RED \epsilon^{10}},{\BLUE\epsilon^{01}})$
+が引き起こし，$x$軸周りの回転は$({\RED \epsilon^{23}},{\BLUE\epsilon^{32}})$が引き起こすことに注意。
 
@@
 \end{equation}
-$\epsilon^{10}=\epsilon$のとき
-$a = ({\RED-}{\BLUE+})\cdot\dfrac\ii2\alpha\beta\sigma_x$
-であり，$\epsilon^{23}=\epsilon$のとき
-$a = ({\RED+}{\BLUE-})\cdot\dfrac12\sigma_x$である。
+$({\RED\epsilon^{10}},{\BLUE\epsilon^{01}})=1$のとき
+$a = -\dfrac\ii2\alpha\beta\sigma_x$
+であり，$({\RED\epsilon^{23}},{\BLUE\epsilon^{32}})=1$のとき
+$a = \dfrac12\sigma_x$である。
 \end{rightnote}
@@
 \end{align}
-となり，Lorentz群が$\gSU(2)\times\gSU(2)$に分解できる。ここで$P=({\RED\pm1},{\BLUE\mp1})$であるので，$P=+1$となるように，複号の$({\RED 上},{\BLUE 下})$側を取ったことにする。
+となり，Lorentz群が$\gSU(2)\times\gSU(2)$に分解できる。ここで$P=({\RED\pm1},{\BLUE\mp1})$であるので，$P=+1$となるように，複号の$({\RED 上},{\BLUE 下})$側を取ったことにする。（$\vc J$と$\vc K$がそのような場合に対してのものである。）
 
-\subsection{Weyl Spinor}
-\TODO{この部分は$+---$の場合の定義。$-+++$のときの右左の定義は？}
+\subsection{Weyl Spinor for $\eta=(+,-,-,-)$}
 左巻きspinor $\xi$ と右巻きspinor $\bar\xi$ を，
@@
 $g$を左巻きspinorの変換子として定義すると，$\vc J\to\frac12\vc \sigma$および$\vc K\to-\frac\ii2\alpha\beta\vc\sigma$の対応があったことを思い出すと
-$-\ii(\vipro\theta J+\vipro\beta K)=-\frac\ii2\vipro\theta\sigma-\frac12\alpha\beta\vipro\beta\sigma$であるので，$\alpha\beta=1$が分かる。
+$-\ii(\vipro\theta J+\vipro\beta K)=-\frac\ii2\vipro\theta\sigma-\frac12\alpha\beta\vipro\beta\sigma$であるので，$\alpha\beta=1$が分かる。また，記号を
+\begin{align}
+  \xi_\alpha      &\mapsto g\T_\alpha^\beta\xi_\beta,
+& \bar\xi^\dalpha &\mapsto (g^\dagger)^{-1}\T^\dalpha_\dbeta\bar\xi^\dbeta
+\end{align}
+と定義しておく。
+
+次に，$\xi^\alpha\chi_\alpha$および$\bar\xi_\dalpha\bar\chi^\dalpha$がscalarとなるようにしたい。
+$E=\spmat{0&1\\-1&0}$としておくと，$-E\trans{g}E=g^{-1}$であるので，
+\begin{equation}
+ (\xi')^\alpha=\xi^\beta(g^{-1})\T_\beta^\alpha
+              =-\xi^\beta(E_{\beta A}g\T_B^AE_{B\alpha})\qquad\therefore (-E_{\gamma\alpha})(\xi')^\alpha = -g\T_\gamma^\beta(E_{\beta\delta}\xi^\delta).
+\end{equation}
+よって，$\epsilon^{12}=\epsilon_{21}=1$として
+$\xi^\alpha:=\epsilon^{\alpha\beta}\xi_\beta$，$\xi_\alpha=\epsilon_{\alpha\beta}\xi^\beta$
+とすれば良い。
+
+同様に，$\bar\xi'_\dalpha=-E(g^\dagger)^{-1}E\bar\xi_\dbeta$から
+$(E^{\dalpha\dbeta}\bar\xi_\dbeta)'=(g^\dagger)^{-1}{}\T^\dalpha_\dbeta (E^{\dbeta\dgamma}\bar\xi_\dgamma)$となる。$\epsilon_{12}=\epsilon_{\dot1\dot2}$として$\bar\xi_\dalpha:=\epsilon_{\dalpha\dbeta}\bar\xi^\dalpha$とするのが一般的である。
+また，このことから$(\xi_a)^*=\bar\xi_\dalpha$（或いは$\xi^\dagger=\bar\xi$）が分かる。
+
+$x_\mu\Sm\mapsto x_\mu(g\Sm g^\dagger)$であるので，
+$x_\mu(\xi^\alpha \Sm \bar\chi^\dalpha)$はscalarである。よって，$\Sm_{\alpha\dalpha}$のように書ける。
+更にここで$x_\mu(\bar\xi_\dalpha\bSm{}^{\dalpha\alpha}\chi_\alpha)$もscalarとなるように$\bar\sigma$を定めよう。
+\begin{equation}
+ \bar\xi_\dalpha\bSm{}^{\dalpha\alpha}\chi_\alpha=-\epsilon_{\alpha\beta}\epsilon_{\dalpha\dbeta}\chi^\beta\bSm{}^{\dalpha\alpha}\bar\xi^\dbeta
+\qquad\therefore \Sm_{\beta\dbeta}\propto \epsilon_{\alpha\beta}\epsilon_{\dalpha\dbeta}\bSm{}^{\dalpha\alpha}
+\end{equation}
+であり，あとはconventionである。
+
+\starline
+
+\begin{align}
+  \epsilon^{12}=\epsilon_{21}=\epsilon^{\dot1\dot2}=\epsilon_{\dot2\dot1}&=1,&
+\xi^\alpha&:=\epsilon^{\alpha\beta}\xi_\beta,& \xi_\alpha&=\epsilon_{\alpha\beta}\xi^\beta,&
+\bar\xi^\dalpha&:=\epsilon^{\dalpha\dbeta}\bar\xi_\dbeta,& \bar\xi_\dalpha&=\epsilon_{\dalpha\dbeta}\bar\xi^\dbeta
+\end{align}
+\begin{align}
+  \xi_\alpha      &\mapsto g\T_\alpha^\beta\xi_\beta,
+& \xi^\alpha      &\mapsto \xi^\beta (g^{-1})\T_\beta^\alpha,
+& \bar\xi_\dalpha &\mapsto \bar\xi_\dbeta(g^\dagger)\T^\dbeta_\dalpha
+& \bar\xi^\dalpha &\mapsto (g^\dagger)^{-1}\T^\dalpha_\dbeta\bar\xi^\dbeta
+\end{align}
+\begin{align}
+  \bSm{}^{\dalpha\alpha} &:= \epsilon^{\alpha\beta}\epsilon^{\dalpha\dbeta}\Sm_{\beta\dbeta}&
+ \Sm_{\alpha\dalpha}&= \epsilon_{\alpha\beta}\epsilon_{\dalpha\dbeta}\bSm{}^{\dbeta\beta}&
+ \therefore \Sm:=(1,\vc\sigma),\quad\bSm:=(1,-\vc\sigma)
+\end{align}
+
+\subsection{Weyl Spinor for $\eta=(-,+,+,+)$}
+左巻き／右巻きの定義は同じであり，それぞれやはり$(1/2,0)$表現と$(0,1/2)$表現である：
+\begin{align}
+ \xi&\mapsto\left(1-\frac\ii2\vipro\theta\sigma-\frac12\vipro\beta\sigma\right)\xi&
+ \bar\xi&\mapsto\left(1-\frac\ii2\vipro\theta\sigma+\frac12\vipro\beta\sigma\right)\bar\xi
+\end{align}
+
+$g$を左巻きspinorの変換子として定義すると，$\vc J\to\frac12\vc \sigma$および$\vc K\to-\frac\ii2\alpha\beta\vc\sigma$の対応があったことを思い出すと
+$-\ii(\vipro\theta J+\vipro\beta K)=\frac\ii2\vipro\theta\sigma+\frac12\alpha\beta\vipro\beta\sigma$であるので，$\alpha\beta=1$が分かる。
 \begin{align}

### 2010-04-30 VerboseNotes.tex
CheatSheet Update
git-svn-id: https://dav.misho-web.com/svn/cyghome/trunk/QFT/CheatSheet@386 c04c31ef-c4f7-4b37-ba94-86ebe2b9c820

@@
 %#!latexmk CheatSheet.tex
-%%% Time-Stamp: <2010-04-28 22:13:13 misho>
+%%% Time-Stamp: <2010-04-29 16:59:55 misho>
 \section{Verbose Notes}
 \subsection{Spinor Fields}
+\def\redblue#1#2{({\RED{#1}},{\BLUE{#2}})}
+
 \subsubsection{Lorentz group and Lorentz algebra}
@@
    L_{PT} := \left\{ \det\Lambda = -1 \land \Lambda\T_0^0 <0\right\}$\\
-& \qquad ($L_0$ is identical with ${\RED\gSO(1,3)} / {\BLUE\gSO(3,1)}$. )\\[1zw]
+& \qquad ($L_0$ is identical with $\redblue{\gSO(1,3)}{\gSO(3,1)}$. )\\[1zw]
 Infinitesimal one in $L_0$& $\Lambda\T^\mu_\nu = \delta^\mu_\nu + \epsilon\T^\mu_\nu$ where $\epsilon_{\mu\nu}=-\epsilon_{\nu\mu}$ (for $\eta=\eta\Lambda\Lambda$)\\[1zw]
@@
 
-ここで微小変換から生成子を
- $x'^\mu =  x^\mu + \epsilon\T^\mu_\nu x^\nu
-   =: \left(\delta^\mu_\nu\mp\frac\ii2\epsilon_{\rho\sigma}(J^{\rho\sigma})\T^\mu_\nu\right)x^\nu$
-  ただし$J^{\rho\sigma} = -J^{\sigma\rho}$と定義すると，
+微小変換は$\epsilon\T^\mu_\nu=\spmat{
+0&\beta_x&\beta_y&\beta_z\\
+\beta_x & 0 & -\theta_z & \theta_y\\
+\beta_y & \theta_z & 0 & -\theta_x\\
+\beta_y & -\theta_y & \theta_x & 0}$の形となっているので，回転生成子$\vc J$と加速生成子$\vc K$は
+\begin{equation}
+ \Lambda = \exp\left[\kappa\ii(\vipro\theta J +\vipro\beta K)\right]
+\quad\then\quad
+J_x = -\kappa\ii\spmat{0&0&0&0\\0&0&0&0\\0&0&0&-1\\0&0&1&0},\quad
+K_x = -\kappa\ii\spmat{0&1&0&0\\1&0&0&0\\0&0&0&0\\0&0&0&0}\label{eq:explicit_JK}
+\end{equation}
+の形である。ここで$\kappa=\pm1$はnotationである。
+
+一方，微小変換から生成子を
+ $\epsilon\T^\mu_\nu  =: \mp\frac\ii2\epsilon^{\rho\sigma}(J_{\rho\sigma})\T^\mu_\nu$
+と定義すると，計量によらずに$\epsilon_{\mu\nu}$は反対称となり，従って$J^{\rho\sigma}$も反対称。
 $ (J_{\rho\sigma})\T^\mu_\nu =
@@
 
-%\vspace{5zw}
-%$x$方向のBoost演算子$K_x$と，$x$軸周りの回転演算子$J_x$は，それぞれ
-%\begin{align}
-%  \pmat{m\\p}&\longrightarrow\pmat{1&\beta\\\beta&1}\pmat{m\\p},
-%& \pmat{y\\z}&\longrightarrow\pmat{1&-\theta\\\theta&1}\pmat{y\\z},
-%&\therefore 
-%\epsilon&=\beta\spmat{0&1&0&0\\1&0&0&0\\0&0&0&0\\0&0&0&0},
-%         \theta\spmat{0&0&0&0\\0&0&0&0\\0&0&0&-1\\0&0&1&0}
-%\end{align}
-%に対応するものである。$\vc J$と$\vc K$を
-%\begin{align}
-% J_i &:= (J_{23}, J_{31}, J_{12}) &  K_i &:= (J_{10},J_{20},J_{30})
-%\end{align}
-%のように定めることにすると，複号は({\RED 上},{\BLUE 下})側を選ばねばならない。
-%\vspace{5zw}
-
-生成子は
+生成子の具体形は計量に依存し，
 $
- \displaystyle J\T_1_0^\mu_\nu=({\RED \pm\ii},{\BLUE \mp\ii})
+ \displaystyle J\T_1_0^\mu_\nu=\redblue{\pm\ii}{\mp\ii}
  \spmat{0&1&0&0\\1&0&0&0\\0&0&0&0\\0&0&0&0}_{\mu\nu},
- \displaystyle J\T_2_3^\mu_\nu=({\RED \pm\ii},{\BLUE \mp\ii})
+ \displaystyle J\T_2_3^\mu_\nu=\redblue{\pm\ii}{\mp\ii}
  \spmat{0&0&0&0\\0&0&0&0\\0&0&0&-1\\0&0&1&0}_{\mu\nu}
 $となる。
-ただし，$x$方向のboostは$({\RED \epsilon^{10}},{\BLUE\epsilon^{01}})$
-が引き起こし，$x$軸周りの回転は$({\RED \epsilon^{23}},{\BLUE\epsilon^{32}})$が引き起こすことに注意。
+よって，ここでの複号の取り方と$\kappa$および計量の定義によって，$\vc J$・$\vc K$と$J_{\rho\sigma}$の対応が定まることになる。
+\begin{equation}
+  \vc\theta = \redblue+-(\epsilon^{23},\epsilon^{31},\epsilon^{12}),\quad
+  \vc\beta  = \redblue+-(\epsilon^{10},\epsilon^{20},\epsilon^{30}).
+\end{equation}
+\starline
 
+以下では$\kappa=-1$にとり，複号の$({\RED 上},{\BLUE 下})$側を取ることで
+\begin{equation}
+  \vc J     = (J_{23}, J_{31}, J_{12}),\quad  \vc K = (J_{10},J_{20},J_{30});\qquad
+\end{equation}
+とする：
+\begin{equation}
+ \Lambda = \exp\epsilon = \exp\left[-\ii(\vipro\theta J +\vipro\beta K)\right]
+         = \exp\left[\redblue{-\ii}{+\ii}(\epsilon^{\rho\sigma}J_{\rho\sigma})/2\right].
+\end{equation}
+
+\newpage
 \subsubsection{Lorentz group and $\gSL(2,\Complex)$}
@@
 \end{equation}
+\TODO{正しく複号をとって再チェック！}
+
 $({\RED\epsilon^{10}},{\BLUE\epsilon^{01}})=1$のとき
@@
 
-このことを別の観点から見る。Lorentz群の生成子の交換関係を見ると，
-\begin{align}
- J_i &:= (J_{23}, J_{31}, J_{12}) &  K_i &:= (J_{10},J_{20},J_{30}) &\for i=1,2,3.\footnotemark
-\end{align}
-に対して\footnotetext{For your information, $J_{23} = J^{23}$ and $K_{01} = -K^{01}$, regardless of the definition of the metic.}
-\begin{align}
- [J_i, J_j]&= ({\RED\pm\ii},{\BLUE\mp\ii})\epsilon_{ijk}J_k,&
- [J_i, K_j]&= ({\RED\pm\ii},{\BLUE\mp\ii})\epsilon_{ijk}K_k,&
- [K_i, K_j]&= -({\RED\pm\ii},{\BLUE\mp\ii})\epsilon_{ijk}J_k
-\end{align}
+このことを別の観点から見る。Lorentz群の生成子の交換関係を見ると，（正しく複号を取った場合
+\begin{equation}
+  [J_i, J_j]= \ii\epsilon_{ijk}J_k,\qquad
+ [J_i, K_j]= \ii\epsilon_{ijk}K_k,\qquad
+ [K_i, K_j]= -\ii\epsilon_{ijk}J_k
+\end{equation}
 となるので，
-\begin{align}
- \vc A&:=\frac{P}2(\vc J+\ii\vc K), & \vc B &:= \frac{P}2(\vc J-\ii\vc K).
-\end{align}
+\begin{equation}
+  \vc A:=\frac{1}2(\vc J+\ii\vc K), \qquad \vc B := \frac{1}2(\vc J-\ii\vc K).
+\end{equation}
 と定義すると
-\begin{align}
- [A_i, A_j] &= \ii\epsilon_{ijk}A_k,&
- [B_i, B_j] &= \ii\epsilon_{ijk}B_k,&
- [A_i, B_j] &= 0,
-\end{align}
-となり，Lorentz群が$\gSU(2)\times\gSU(2)$に分解できる。ここで$P=({\RED\pm1},{\BLUE\mp1})$であるので，$P=+1$となるように，複号の$({\RED 上},{\BLUE 下})$側を取ったことにする。（$\vc J$と$\vc K$がそのような場合に対してのものである。）
+\begin{equation}
+  [A_i, A_j] = \ii\epsilon_{ijk}A_k,\qquad
+ [B_i, B_j] = \ii\epsilon_{ijk}B_k,\qquad
+ [A_i, B_j] = 0,
+\end{equation}
+となり，Lorentz群が$\gSU(2)\times\gSU(2)$に分解できる。
 

### 2010-07-28 VerboseNotes.tex
Cheatsheet更新，RpVDecayRate作成，その他色々
git-svn-id: https://dav.misho-web.com/svn/cyghome/trunk/QFT/CheatSheet@392 c04c31ef-c4f7-4b37-ba94-86ebe2b9c820

@@
 %#!latexmk CheatSheet.tex
-%%% Time-Stamp: <2010-04-29 16:59:55 misho>
+%%% Time-Stamp: <2010-05-01 20:39:42 misho>
 \section{Verbose Notes}
@@
 \subsubsection{Lorentz group and Lorentz algebra}
+\label{sec:verbose-lorentz-group}
 \begin{tabular}{l@{ :\ \ \ }l}
@@
  $\epsilon\T^\mu_\nu  =: \mp\frac\ii2\epsilon^{\rho\sigma}(J_{\rho\sigma})\T^\mu_\nu$
-と定義すると，計量によらずに$\epsilon_{\mu\nu}$は反対称となり，従って$J^{\rho\sigma}$も反対称。
+と定義すると，計量によらずに$\epsilon_{\mu\nu}$は反対称となり，
+\begin{align*}
+  \vc\theta &= \redblue+-(\epsilon^{23},\epsilon^{31},\epsilon^{12}),&
+  \vc\beta  &=
+\redblue+-(\epsilon^{10},\epsilon^{20},\epsilon^{30}).\\
+&=\redblue+-(\epsilon_{23},\epsilon_{31},\epsilon_{12})&
+&=\redblue-+(\epsilon_{10},\epsilon_{20},\epsilon_{30})
+\end{align*}
+で，従って$J^{\rho\sigma}$も反対称。
 $ (J_{\rho\sigma})\T^\mu_\nu =
@@
 よって，ここでの複号の取り方と$\kappa$および計量の定義によって，$\vc J$・$\vc K$と$J_{\rho\sigma}$の対応が定まることになる。
-\begin{equation}
-  \vc\theta = \redblue+-(\epsilon^{23},\epsilon^{31},\epsilon^{12}),\quad
-  \vc\beta  = \redblue+-(\epsilon^{10},\epsilon^{20},\epsilon^{30}).
-\end{equation}
 \starline
 
-以下では$\kappa=-1$にとり，複号の$({\RED 上},{\BLUE 下})$側を取ることで
+$\kappa$と複号について$({\RED -,上})$,$({\RED +,下})$,$({\BLUE -,下})$,$({\BLUE +,上})$を取れば
 \begin{equation}
@@
 \end{equation}
-とする：
+となる：
 \begin{equation}
- \Lambda = \exp\epsilon = \exp\left[-\ii(\vipro\theta J +\vipro\beta K)\right]
-         = \exp\left[\redblue{-\ii}{+\ii}(\epsilon^{\rho\sigma}J_{\rho\sigma})/2\right].
+ \Lambda = \exp\epsilon = \exp\left[\kappa\ii(\vipro\theta J +\vipro\beta K)\right]
+         = \exp\left[\mp\ii(\epsilon^{\rho\sigma}J_{\rho\sigma})/2\right].
 \end{equation}
@@
 \Lambda\T^\mu_\nu\Sn=g^{-1}\Sm (g^{-1})^\dagger\quad\then\quad
-\mp\frac12\epsilon_{\rho\sigma}(J^{\rho\sigma})\T^\mu_\nu\Sn=a\Sm-\Sm a^\dagger
+\epsilon\T^\mu_\nu\Sn=\ii(a\Sm-\Sm a^\dagger)
 \end{equation}
-\TODO{正しく複号をとって再チェック！}
-
-$({\RED\epsilon^{10}},{\BLUE\epsilon^{01}})=1$のとき
-$a = -\dfrac\ii2\alpha\beta\sigma_x$
-であり，$({\RED\epsilon^{23}},{\BLUE\epsilon^{32}})=1$のとき
-$a = \dfrac12\sigma_x$である。
+であり，ここから$g$がわかる：
+\begin{equation}
+ g=\exp\left(-\frac{\ii}2\vipro\theta\sigma-\frac{\alpha\beta}2\vipro\beta\sigma\right).
+\end{equation}
 \end{rightnote}
 
-このことを別の観点から見る。Lorentz群の生成子の交換関係を見ると，（正しく複号を取った場合
+このことを別の観点から見る。Lorentz群の生成子の交換関係を見ると，（正しく複号を取った場合）
 \begin{equation}
@@
 
-\subsection{Weyl Spinor for $\eta=(+,-,-,-)$}
-左巻きspinor $\xi$ と右巻きspinor $\bar\xi$ を，
+\subsection{Weyl Spinor}
+$\gSU(2)_A\times\gSU(2)_B$に対して$(1/2,0)$表現を為すものを左巻きspinor $\xi$，$(0,1/2)$表現を為すものを右巻きspinor $\bar\xi$と定義する。
 \begin{align}
  \xi&\mapsto\left(1-\frac\ii2\vipro\theta\sigma-\frac12\vipro\beta\sigma\right)\xi&
- \bar\xi&\mapsto\left(1-\frac\ii2\vipro\theta\sigma+\frac12\vipro\beta\sigma\right)\bar\xi
+ \bar\xi&\mapsto\left(1-\frac\ii2\vipro\theta\sigma+\frac12\vipro\beta\sigma\right)\bar\xi.
 \end{align}
-と変換するものとして定義する。
-$\gSU(2)_A\times\gSU(2)_B$に対して左巻きは$(1/2,0)$表現，右巻きは$(0,1/2)$表現となっている。
 
-$g$を左巻きspinorの変換子として定義すると，$\vc J\to\frac12\vc \sigma$および$\vc K\to-\frac\ii2\alpha\beta\vc\sigma$の対応があったことを思い出すと
-$-\ii(\vipro\theta J+\vipro\beta K)=-\frac\ii2\vipro\theta\sigma-\frac12\alpha\beta\vipro\beta\sigma$であるので，$\alpha\beta=1$が分かる。また，記号を
-\begin{align}
-  \xi_\alpha      &\mapsto g\T_\alpha^\beta\xi_\beta,
-& \bar\xi^\dalpha &\mapsto (g^\dagger)^{-1}\T^\dalpha_\dbeta\bar\xi^\dbeta
-\end{align}
-と定義しておく。
+$\alpha\beta=1$とすると$g$は左巻きspinorの変換子となる。記号を
+$\xi_\alpha\mapsto g\T_\alpha^\beta\xi_\beta$,
+$\bar\xi^\dalpha\mapsto (g^\dagger)^{-1}\T^\dalpha_\dbeta\bar\xi^\dbeta$
+と定義する。
 
 次に，$\xi^\alpha\chi_\alpha$および$\bar\xi_\dalpha\bar\chi^\dalpha$がscalarとなるようにしたい。
-$E=\spmat{0&1\\-1&0}$としておくと，$-E\trans{g}E=g^{-1}$であるので，
+$E=\spmat{0&1\\-1&0}$としておくと，$-E\trans{g}E=g^{-1}$より
 \begin{equation}
@@
 
-\starline
-
-\begin{align}
+\begin{conclusion}{}
+\begin{align*}
   \epsilon^{12}=\epsilon_{21}=\epsilon^{\dot1\dot2}=\epsilon_{\dot2\dot1}&=1,&
@@
 \bar\xi^\dalpha&:=\epsilon^{\dalpha\dbeta}\bar\xi_\dbeta,& \bar\xi_\dalpha&=\epsilon_{\dalpha\dbeta}\bar\xi^\dbeta
-\end{align}
-\begin{align}
+\end{align*}\vspace{-2.8zw}
+\begin{align*}
   \xi_\alpha      &\mapsto g\T_\alpha^\beta\xi_\beta,
@@
 & \bar\xi^\dalpha &\mapsto (g^\dagger)^{-1}\T^\dalpha_\dbeta\bar\xi^\dbeta
-\end{align}
-\begin{align}
+\end{align*}\vspace{-2.8zw}
+\begin{align*}
   \bSm{}^{\dalpha\alpha} &:= \epsilon^{\alpha\beta}\epsilon^{\dalpha\dbeta}\Sm_{\beta\dbeta}&
  \Sm_{\alpha\dalpha}&= \epsilon_{\alpha\beta}\epsilon_{\dalpha\dbeta}\bSm{}^{\dbeta\beta}&
- \therefore \Sm:=(1,\vc\sigma),\quad\bSm:=(1,-\vc\sigma)
-\end{align}
-
-\subsection{Weyl Spinor for $\eta=(-,+,+,+)$}
-左巻き／右巻きの定義は同じであり，それぞれやはり$(1/2,0)$表現と$(0,1/2)$表現である：
-\begin{align}
- \xi&\mapsto\left(1-\frac\ii2\vipro\theta\sigma-\frac12\vipro\beta\sigma\right)\xi&
- \bar\xi&\mapsto\left(1-\frac\ii2\vipro\theta\sigma+\frac12\vipro\beta\sigma\right)\bar\xi
-\end{align}
-
-$g$を左巻きspinorの変換子として定義すると，$\vc J\to\frac12\vc \sigma$および$\vc K\to-\frac\ii2\alpha\beta\vc\sigma$の対応があったことを思い出すと
-$-\ii(\vipro\theta J+\vipro\beta K)=\frac\ii2\vipro\theta\sigma+\frac12\alpha\beta\vipro\beta\sigma$であるので，$\alpha\beta=1$が分かる。
-\begin{align}
-  \xi_\alpha      &\mapsto g\T_\alpha^\beta\xi_\beta,
-& \bar\xi_\dalpha &\mapsto \bar\xi_\dbeta(g^\dagger)\T^\dbeta_\dalpha
-\end{align}
-であり，$\bar\xi=\xi^\dagger$，つまり$\bar\xi_\dalpha = (\xi_\alpha)^*$である。
-
-次に，$\xi^\alpha\chi_\alpha$および$\bar\xi_\dalpha\bar\chi^\dalpha$がscalarとなるようにしたい。
-$E=\spmat{0&1\\-1&0}$としておくと，$-E\trans{g}E=g^{-1}$であるので，
-\begin{equation}
- (\xi')^\alpha=\xi^\beta(g^{-1})\T_\beta^\alpha
-              =-\xi^\beta(E_{\beta A}g\T_B^AE_{B\alpha})\qquad\therefore (-E_{\gamma\alpha})(\xi')^\alpha = -g\T_\gamma^\beta(E_{\beta\delta}\xi^\delta).
-\end{equation}
-よって，$\epsilon^{12}=\epsilon_{21}=1$として
-$\xi^\alpha:=\epsilon^{\alpha\beta}\xi_\beta$，$\xi_\alpha=\epsilon_{\alpha\beta}\xi^\beta$
-とすれば良い。
-
-同様に，$(\bar\xi^\dalpha E_{\dalpha\dbeta})'=(\bar\xi^\dalpha E_{\dalpha\dbeta})'g^\dagger\T^\dgamma_\dbeta$……としても良かったが，左からの演算，および$\epsilon_{12}=\epsilon_{\dot1\dot2}$の性質を重視して$(\epsilon_{\dalpha\dbeta}\bar\xi^\dbeta)'=(\epsilon_{\dgamma\dbeta}\bar\xi^\dbeta)'g^\dagger\T^\dgamma_\dbeta$とするのが一般的である。
-
-$x_\mu\Sm\mapsto x_\mu(g\Sm g^\dagger)$であるので，
-$x_\mu(\xi^\alpha \Sm \bar\chi^\dalpha)$はscalarである。よって，$\Sm_{\alpha\dalpha}$のように書ける。
-更にここで$x_\mu(\bar\xi_\dalpha\bSm{}^{\dalpha\alpha}\chi_\alpha)$もscalarとなるように$\bar\sigma$を定めよう。
-\begin{equation}
- \bar\xi_\dalpha\bSm{}^{\dalpha\alpha}\chi_\alpha=-\epsilon_{\alpha\beta}\epsilon_{\dalpha\dbeta}\chi^\beta\bSm{}^{\dalpha\alpha}\bar\xi^\dbeta
-\qquad\therefore \Sm_{\beta\dbeta}\propto \epsilon_{\alpha\beta}\epsilon_{\dalpha\dbeta}\bSm{}^{\dalpha\alpha}
-\end{equation}
-であり，あとはconventionである。
-
-\starline
-
-\begin{align}
-  \epsilon^{12}=\epsilon_{21}=\epsilon^{\dot1\dot2}=\epsilon_{\dot2\dot1}&=1,&
-\xi^\alpha&:=\epsilon^{\alpha\beta}\xi_\beta,& \xi_\alpha&=\epsilon_{\alpha\beta}\xi^\beta,&
-\bar\xi^\dalpha&:=\epsilon^{\dalpha\dbeta}\bar\xi_\dbeta,& \bar\xi_\dalpha&=\epsilon_{\dalpha\dbeta}\bar\xi^\dbeta
-\end{align}
-\begin{align}
-  \xi_\alpha      &\mapsto g\T_\alpha^\beta\xi_\beta,
-& \xi^\alpha      &\mapsto \xi^\beta (g^{-1})\T_\beta^\alpha,
-& \bar\xi_\dalpha &\mapsto \bar\xi_\dbeta(g^\dagger)\T^\dbeta_\dalpha
-& \bar\xi^\dalpha &\mapsto (g^\dagger)^{-1}\T^\dalpha_\dbeta\bar\xi^\dbeta
-\end{align}
-\begin{align}
-  \bSm{}^{\dalpha\alpha} &:= \epsilon^{\alpha\beta}\epsilon^{\dalpha\dbeta}\Sm_{\beta\dbeta}&
- \Sm_{\alpha\dalpha}&= \epsilon_{\alpha\beta}\epsilon_{\dalpha\dbeta}\bSm{}^{\dbeta\beta}&
- \therefore \Sm:=(1,\vc\sigma),\quad\bSm:=(1,-\vc\sigma)
-\end{align}
+ \therefore \Sm:=(\alpha1,\beta\vc\sigma),\quad\bSm:=(\alpha1,-\beta\vc\sigma)
+\end{align*}
+\end{conclusion}
 

### 2010-11-20 VerboseNotes.tex
ページの節約
git-svn-id: https://dav.misho-web.com/svn/cyghome/trunk/QFT/CheatSheet@422 c04c31ef-c4f7-4b37-ba94-86ebe2b9c820
ミス修正
git-svn-id: https://dav.misho-web.com/svn/cyghome/trunk/QFT/CheatSheet@423 c04c31ef-c4f7-4b37-ba94-86ebe2b9c820
Cheatsheet update
git-svn-id: https://dav.misho-web.com/svn/cyghome/trunk/QFT/CheatSheet@435 c04c31ef-c4f7-4b37-ba94-86ebe2b9c820

@@
 %#!latexmk CheatSheet.tex
-%%% Time-Stamp: <2010-05-01 20:39:42 misho>
+%%% Time-Stamp: <2010-11-20 21:54:56 misho>
+%%% •¶ŽšƒR[ƒh‚ÍShift-JIS‚Å‚·‚æ‚Ë
+
 \section{Verbose Notes}
@@
 
-微小変換は$\epsilon\T^\mu_\nu=\spmat{
+$BHy>.JQ49$O(B$\epsilon\T^\mu_\nu=\spmat{
 0&\beta_x&\beta_y&\beta_z\\
@@
 \beta_y & \theta_z & 0 & -\theta_x\\
-\beta_y & -\theta_y & \theta_x & 0}$の形となっているので，回転生成子$\vc J$と加速生成子$\vc K$は
+\beta_y & -\theta_y & \theta_x & 0}$$B$N7A$H$J$C$F$$$k$N$G!$2sE>@8@.;R(B$\vc J$$B$H2CB.@8@.;R(B$\vc K$$B$O(B
 \begin{equation}
@@
 \end{equation}
-の形である。ここで$\kappa=\pm1$はnotationである。
+$B$N7A$G$"$k!#$3$3$G(B$\kappa=\pm1$$B$O(Bnotation$B$G$"$k!#(B
 
-一方，微小変換から生成子を
+$B0lJ}!$Hy>.JQ49$+$i@8@.;R$r(B
  $\epsilon\T^\mu_\nu  =: \mp\frac\ii2\epsilon^{\rho\sigma}(J_{\rho\sigma})\T^\mu_\nu$
-と定義すると，計量によらずに$\epsilon_{\mu\nu}$は反対称となり，
+$B$HDj5A$9$k$H!$7WNL$K$h$i$:$K(B$\epsilon_{\mu\nu}$$B$OH?BP>N$H$J$j!$(B
 \begin{align*}
@@
 \end{align*}
-で，従って$J^{\rho\sigma}$も反対称。
+$B$G!$=>$C$F(B$J^{\rho\sigma}$$B$bH?BP>N!#(B
 $ (J_{\rho\sigma})\T^\mu_\nu =
 \pm\ii\left(\delta^\mu_\rho\Hsnd-\delta^\mu_\sigma\Hrnd\right)$
-となり，交換関係が得られ，これが閉じているのでLie代数であることもわかる：
+$B$H$J$j!$8r494X78$,F@$i$l!$$3$l$,JD$8$F$$$k$N$G(BLie$BBe?t$G$"$k$3$H$b$o$+$k!'(B
 \begin{equation}
@@
 
-生成子の具体形は計量に依存し，
+$B@8@.;R$N6qBN7A$O7WNL$K0MB8$7!$(B
 $
@@
  \spmat{0&0&0&0\\0&0&0&0\\0&0&0&-1\\0&0&1&0}_{\mu\nu}
-$となる。
-よって，ここでの複号の取り方と$\kappa$および計量の定義によって，$\vc J$・$\vc K$と$J_{\rho\sigma}$の対応が定まることになる。
+$$B$H$J$k!#(B
+$B$h$C$F!$$3$3$G$NJ#9f$N<h$jJ}$H(B$\kappa$$B$*$h$S7WNL$NDj5A$K$h$C$F!$(B$\vc J$$B!&(B$\vc K$$B$H(B$J_{\rho\sigma}$$B$NBP1~$,Dj$^$k$3$H$K$J$k!#(B
 \starline
 
-$\kappa$と複号について$({\RED -,上})$,$({\RED +,下})$,$({\BLUE -,下})$,$({\BLUE +,上})$を取れば
+$\kappa$$B$HJ#9f$K$D$$$F(B$({\RED -,$B>e(B})$,$({\RED +,$B2<(B})$,$({\BLUE -,$B2<(B})$,$({\BLUE +,$B>e(B})$$B$r<h$l$P(B
 \begin{equation}
@@
 \end{equation}
-となる：
+$B$H$J$k!'(B
 \begin{equation}
@@
 \subsubsection{Lorentz group and $\gSL(2,\Complex)$}
-次に，連結Lie群$L_0$が，連結Lie群$\gSL(2,\Complex)/\gZ_2$と同型であることを見る：
+$B<!$K!$O"7k(BLie$B72(B$L_0$$B$,!$O"7k(BLie$B72(B$\gSL(2,\Complex)/\gZ_2$$B$HF17?$G$"$k$3$H$r8+$k!'(B
 \begin{align}
@@
 \end{align}
-まず,$\Sm$を（極めて一般的に） $\Sm:=(\alpha1,\beta\vc\sigma)$と定義する($\alpha=\beta=\pm1$)。
-$x^2 = ({\RED+}{\BLUE-})\det(x_\mu\Sm)$なので
+$B$^$:(B,$\Sm$$B$r!J6K$a$F0lHLE*$K!K(B $\Sm:=(\alpha1,\beta\vc\sigma)$$B$HDj5A$9$k(B($\alpha=\beta=\pm1$)$B!#(B
+$x^2 = ({\RED+}{\BLUE-})\det(x_\mu\Sm)$$B$J$N$G(B
 \begin{equation}
@@
 \end{equation}
-は$x^2$を保存する。よってLorentz変換であり，生成子を比べることで局所同型だとわかる：
+$B$O(B$x^2$$B$rJ]B8$9$k!#$h$C$F(BLorentz$BJQ49$G$"$j!$@8@.;R$rHf$Y$k$3$H$G6I=jF17?$@$H$o$+$k!'(B
 \begin{rightnote}
-$\gSL(2,\Complex)\ni g = \exp(-\ii a)$として
-$x_\mu (g \Sm g^\dagger) = \Lambda\T_\mu^\nu x_\nu \Sm$を微小展開すると
+$\gSL(2,\Complex)\ni g = \exp(-\ii a)$$B$H$7$F(B
+$x_\mu (g \Sm g^\dagger) = \Lambda\T_\mu^\nu x_\nu \Sm$$B$rHy>.E83+$9$k$H(B
 \begin{equation}
@@
 \end{equation}
-であり，ここから$g$がわかる：
+$B$G$"$j!$$3$3$+$i(B$g$$B$,$o$+$k!'(B
 \begin{equation}
@@
 
-このことを別の観点から見る。Lorentz群の生成子の交換関係を見ると，（正しく複号を取った場合）
+$B$3$N$3$H$rJL$N4QE@$+$i8+$k!#(BLorentz$B72$N@8@.;R$N8r494X78$r8+$k$H!$!J@5$7$/J#9f$r<h$C$?>l9g!K(B
 \begin{equation}
@@
 \end{equation}
-となるので，
+$B$H$J$k$N$G!$(B
 \begin{equation}
@@
 \end{equation}
-と定義すると
+$B$HDj5A$9$k$H(B
 \begin{equation}
@@
 \end{equation}
-となり，Lorentz群が$\gSU(2)\times\gSU(2)$に分解できる。
+$B$H$J$j!$(BLorentz$B72$,(B$\gSU(2)\times\gSU(2)$$B$KJ,2r$G$-$k!#(B
 
 \subsection{Weyl Spinor}
-$\gSU(2)_A\times\gSU(2)_B$に対して$(1/2,0)$表現を為すものを左巻きspinor $\xi$，$(0,1/2)$表現を為すものを右巻きspinor $\bar\xi$と定義する。
+$\gSU(2)_A\times\gSU(2)_B$$B$KBP$7$F(B$(1/2,0)$$BI=8=$r0Y$9$b$N$r:84,$-(Bspinor $\xi$$B!$(B$(0,1/2)$$BI=8=$r0Y$9$b$N$r1&4,$-(Bspinor $\bar\xi$$B$HDj5A$9$k!#(B
 \begin{align}
@@
 
-$\alpha\beta=1$とすると$g$は左巻きspinorの変換子となる。記号を
+$\alpha\beta=1$$B$H$9$k$H(B$g$$B$O:84,$-(Bspinor$B$NJQ49;R$H$J$k!#5-9f$r(B
 $\xi_\alpha\mapsto g\T_\alpha^\beta\xi_\beta$,
 $\bar\xi^\dalpha\mapsto (g^\dagger)^{-1}\T^\dalpha_\dbeta\bar\xi^\dbeta$
-と定義する。
+$B$HDj5A$9$k!#(B
 
-次に，$\xi^\alpha\chi_\alpha$および$\bar\xi_\dalpha\bar\chi^\dalpha$がscalarとなるようにしたい。
-$E=\spmat{0&1\\-1&0}$としておくと，$-E\trans{g}E=g^{-1}$より
+$B<!$K!$(B$\xi^\alpha\chi_\alpha$$B$*$h$S(B$\bar\xi_\dalpha\bar\chi^\dalpha$$B$,(Bscalar$B$H$J$k$h$&$K$7$?$$!#(B
+$E=\spmat{0&1\\-1&0}$$B$H$7$F$*$/$H!$(B$-E\trans{g}E=g^{-1}$$B$h$j(B
 \begin{equation}
@@
 \end{equation}
-よって，$\epsilon^{12}=\epsilon_{21}=1$として
-$\xi^\alpha:=\epsilon^{\alpha\beta}\xi_\beta$，$\xi_\alpha=\epsilon_{\alpha\beta}\xi^\beta$
-とすれば良い。
+$B$h$C$F!$(B$\epsilon^{12}=\epsilon_{21}=1$$B$H$7$F(B
+$\xi^\alpha:=\epsilon^{\alpha\beta}\xi_\beta$$B!$(B$\xi_\alpha=\epsilon_{\alpha\beta}\xi^\beta$
+$B$H$9$l$PNI$$!#(B
 
-同様に，$\bar\xi'_\dalpha=-E(g^\dagger)^{-1}E\bar\xi_\dbeta$から
-$(E^{\dalpha\dbeta}\bar\xi_\dbeta)'=(g^\dagger)^{-1}{}\T^\dalpha_\dbeta (E^{\dbeta\dgamma}\bar\xi_\dgamma)$となる。$\epsilon_{12}=\epsilon_{\dot1\dot2}$として$\bar\xi_\dalpha:=\epsilon_{\dalpha\dbeta}\bar\xi^\dalpha$とするのが一般的である。
-また，このことから$(\xi_a)^*=\bar\xi_\dalpha$（或いは$\xi^\dagger=\bar\xi$）が分かる。
+$BF1MM$K!$(B$\bar\xi'_\dalpha=-E(g^\dagger)^{-1}E\bar\xi_\dbeta$$B$+$i(B
+$(E^{\dalpha\dbeta}\bar\xi_\dbeta)'=(g^\dagger)^{-1}{}\T^\dalpha_\dbeta (E^{\dbeta\dgamma}\bar\xi_\dgamma)$$B$H$J$k!#(B$\epsilon_{12}=\epsilon_{\dot1\dot2}$$B$H$7$F(B$\bar\xi_\dalpha:=\epsilon_{\dalpha\dbeta}\bar\xi^\dalpha$$B$H$9$k$N$,0lHLE*$G$"$k!#(B
+$B$^$?!$$3$N$3$H$+$i(B$(\xi_a)^*=\bar\xi_\dalpha$$B!J0?$$$O(B$\xi^\dagger=\bar\xi$$B!K$,J,$+$k!#(B
 
-$x_\mu\Sm\mapsto x_\mu(g\Sm g^\dagger)$であるので，
-$x_\mu(\xi^\alpha \Sm \bar\chi^\dalpha)$はscalarである。よって，$\Sm_{\alpha\dalpha}$のように書ける。
-更にここで$x_\mu(\bar\xi_\dalpha\bSm{}^{\dalpha\alpha}\chi_\alpha)$もscalarとなるように$\bar\sigma$を定めよう。
+$x_\mu\Sm\mapsto x_\mu(g\Sm g^\dagger)$$B$G$"$k$N$G!$(B
+$x_\mu(\xi^\alpha \Sm \bar\chi^\dalpha)$$B$O(Bscalar$B$G$"$k!#$h$C$F!$(B$\Sm_{\alpha\dalpha}$$B$N$h$&$K=q$1$k!#(B
+$B99$K$3$3$G(B$x_\mu(\bar\xi_\dalpha\bSm{}^{\dalpha\alpha}\chi_\alpha)$$B$b(Bscalar$B$H$J$k$h$&$K(B$\bar\sigma$$B$rDj$a$h$&!#(B
 \begin{equation}
@@
 \end{equation}
-であり，あとはconventionである。
+$B$G$"$j!$$"$H$O(Bconvention$B$G$"$k!#(B
 
@@
 Secondly we think about the double photons case
-\footnote{This part is derived from 濱口幸一's notebook.}
+\footnote{This part is derived from $B_@8}9,0l(B's notebook.}
 $M=\epsilon_\mu^*(k)\epsilon_\nu'^*(k')M^{\mu\nu}$.
@@
 You may think we forget to introduce
-$\ol\psi\G5\psi$，
-$\ol\psi\G5\slashed D\psi$，
-$\epsilon^{\mu\nu\rho\sigma}F^a_{\mu\nu}F^a_{\rho\sigma}$，
+$\ol\psi\G5\psi$$B!$(B
+$\ol\psi\G5\slashed D\psi$$B!$(B
+$\epsilon^{\mu\nu\rho\sigma}F^a_{\mu\nu}F^a_{\rho\sigma}$$B!$(B
 $\epsilon^{\mu\nu\rho\sigma}D_\mu D_\nu F^a_{\rho\sigma}$
@@
 
+\subsection{$BML(B-Mills Theory}\label{sec:yang-mills-theory}
+
+\subsubsection{General Gauge Theory}\vskip-4pt
+For any Lie group $G$, we can consider ``gauge transformation'' $\phi(x) \mapsto V(x)\phi(x)$, where $V:\Real^{1,3}\to G$. Also we can define a ``connection field'' $A_\mu(x)$ as:
+\begin{align}
+ \phi_\parallel (x+\dd x):= \phi(x)+\ii g A_\mu(x)\phi(x)\dd x^\mu
+\qquad\text{s.t.}\quad \phi_\parallel(x+\dd x)\mapsto V(x+\dd x)\phi_\parallel(x+\dd x).
+\end{align}
+Then the covariant derivative $\Dm$ can be defined as
+\begin{equation}
+ \Dm\phi(x)\dd x^\mu := \Delta_{\dd x}\phi(x):= \phi(x+\dd x)-\phi_\parallel(x+\dd x)\qquad
+\therefore  \Dm := \Pm - \ii g A_\mu.
+\end{equation}
+Note that $\Delta_{\dd x}\phi(x)\mapsto V(x+\dd x)\Dm\phi(x)\dd x^\mu$, which means
+$\Dm\phi(x)\mapsto V(x)\Dm\phi(x)$.
+Now we can see
+\begin{align}
+ \phi&\mapsto V\phi,&
+  A_\mu&\mapsto V\left(A_\mu + \frac\ii g \partial_\mu\right)V^{-1},&
+ \Dm&\mapsto V\Dm V^{-1}.\label{eq:gaugetransf}
+\end{align}
+We can do another discussion: we can define $\Dm$ as a kind of derivative which satisfies \eqref{eq:gaugetransf}{}.
+
+Next we introduce the curvature tensor, or ``field strength'' as
+\begin{align}
+& \Delta\phi(x):=\phi_{\parallel}^{xy}(x+\dd x+\dd y)-\phi_{\parallel}^{yx}(x+\dd x+\dd y)
+= \left[\Dm,\Dn\right]\phi(x)\dd x^\mu\dd y^\nu
+=: -\ii g F_{\mu\nu}\phi(x)\dd x^\mu\dd y^\nu;\\
+&F_{\mu\nu}(x):=\frac{\ii}g\left[\Dm,\Dn\right]
+ = \Pm A_\nu(x)-\Pn A_\mu(x)-\ii g\left[A_\mu(x),A_\nu(x)\right].
+\end{align}
+$\Delta\phi(x)$ is transformed in terms of $V(x+\dd x+\dd y)\simeq V(x)$, thus $F_{\mu\nu}(x)\mapsto V(x)F_{\mu\nu}(x)V^{-1}(x)$.
+
+\subsubsection{Compact Gauge Theory}
+\subparagraph{Generators}
+If the gauge group $G$ is {\bf compact}, it has a finite-dimensional
+unitary representation. The generators $T_a$ can be taken to be Hermitian, and $V(x) = \exp\left[\ii g\theta^a(x) T^a\right]$ for $\theta^a(x)\in\Real$;
+\begin{align}
+ [T^a,T^b] &= \ii f\T^a^b_c T^c\quad(f\in\Real)&
+ 0 &=f\T^D_a_b f\T^E_D_c + f\T^D_c_a f\T^E_D_b  + f\T^D_b_c f\T^E_D_a
+\end{align}
+
+For the sake of the compactness Killing form is positive-definite, where we can normalize the generators as $\Tr T^aT^b=\frac12\delta^{ab}$, and the structure constant $f^{abc}$ would be totally antisymmetric.
+
+\subparagraph{Adjoint Representations}
+\begin{equation}
+ [\tilde T^a]\T_i^j:=-\ii f^{aij};\qquad
+[\tilde\Dm]\T_i^j:=\delta_i^j\Pm+gf^{iaj}A^a_\mu.
+\end{equation}
+
+\subparagraph{Field Expansion}
+In this {\em normalized Hermitian} basis, the relations would be\footnote{We can expand $A_\mu$ in $T^a$-basis, because it is induced by the gauge transformation.}
+\begin{align*}
+\phi'&=\ee^{\ii gT^a\theta^a}\phi;&
+F_{\mu\nu}^a  &= \Pm A_\nu^a-\Pn A_\mu^a + gf^{abc}A_\mu^bA_\nu^c\\
+A_\mu^a{}'&\simeq A_\mu^a+\Pm\theta^a+gf^{abc}A_\mu^b\theta^c&
+F^a_{\mu\nu}{}'&=[\ee^{\ii g\theta^c\tilde T^c}]^{ab}F^b_{\mu\nu}\\
+         &=A_\mu^a+(\tilde\Dm\theta)^a,&
+&\simeq F_{\mu\nu}^a+gf^{abc}F^b_{\mu\nu}\theta^c.
+\end{align*}
+
+\subparagraph{Covariant Derivative}
+For a field $\lambda^a$ under the adjoint representation,
+\begin{equation}
+ (\Dm\lambda)^a   = \Pm\lambda^a    + g f^{abc} A_\mu^b \lambda^c\qquad\text{or}\quad
+  \Dm\lambda^aT^a = \Pm\lambda^aT^a - \ii g [A_\mu^bT^b,\lambda^aT^a]\ \footnotemark
+\end{equation}
+\footnotetext{Note that we can use any representation $T^a$ but must the same ones for $A^a_\mu T^a$ and $\lambda^a T^a$.}
+
+\subparagraph{Bianchi Equation}
+\begin{equation}
+ \epsilon^{\mu\nu\rho\sigma}\left[\Dn,\left[\Dr,\Ds\right]\right] = \epsilon^{\mu\nu\rho\sigma}\Dn F_{\rho\sigma} = 0.
+\end{equation}
+
+\newpage
+
+\subsection{Spinor}
+\Paragraph{$\eta^{\mu\nu}=(-,+,+,+)$ case}
+\begin{tabular}[t]{l@{\ :\ }l}
+ Grassmann Number
+& $(ab)^\dagger=b^\dagger a^\dagger$ for $a,b\in\Grassmann$\\
+& \then for $a,b\in\RealGrassmann$, $ab\in\ii\RealGrassmann$\\
+ $\gamma$ matrix
+& $\displaystyle\{\Gm,\Gn\}=2\minkow\mu\nu\cdot\vc1$\\
+& $\displaystyle\G{\mu\nu} = \frac12\left(\Gm\Gn-\Gn\Gm\right)$
+  \quad etc...\\
+& $\displaystyle(\ii\G0)^\dagger:=\ii\G0,\quad\G i^\dagger:=\G i$\\
+ Dirac Conjugate
+& $\bar\psi=\ii\psi^\dagger\G0$
+\end{tabular}
+
+\TODO{SPINOR}
 \subsection{Instanton}
 \label{sec:cs-instanton}
-\TODO{Postpone...}
+\TODO{INSTANTON}
+
 %%% Local Variables:

### 2011-05-04 VerboseNotes.tex
headerに日付とrevを追加するようにした。
git-svn-id: https://dav.misho-web.com/svn/cyghome/trunk/QFT/CheatSheet@586 c04c31ef-c4f7-4b37-ba94-86ebe2b9c820
CheatSheet: YaTeXの都合でシェバングを変更；compileのアップデート
git-svn-id: https://dav.misho-web.com/svn/cyghome/trunk/QFT/CheatSheet@593 c04c31ef-c4f7-4b37-ba94-86ebe2b9c820
VerboseNotesの更新
git-svn-id: https://dav.misho-web.com/svn/cyghome/trunk/QFT/CheatSheet@594 c04c31ef-c4f7-4b37-ba94-86ebe2b9c820

@@
-%#!latexmk CheatSheet.tex
-%%% Time-Stamp: <2010-11-20 21:54:56 misho>
+%#!./compile CheatSheet
+%%% Time-Stamp: <2011-05-04 22:23:09 Misho>
 %%% •¶ŽšƒR[ƒh‚ÍShift-JIS‚Å‚·‚æ‚Ë
@@
 \subsection{Polarization Sum}\label{Sec:Verbose:PolarizationSum}
-Firstly we focus on the single photon case $M=\epsilon_\mu^*(k)\epsilon_\nu'^*(k')M^{\mu\nu}$.
-Here we set $k=(E,0,0,E)$, and $\epsilon=(0,1,0,0)\oplus(0,0,1,0)$. Then
+Firstly we focus on the single photon case $M=\epsilon_\mu^*(k)M^{\mu}$.
+In this case, the replacement
+\begin{equation}
+ \sum\s{pol.}\epsilon_\mu\epsilon'_\nu \to \Hmnd
+\end{equation}
+is valid. Let us prove this validity. First we set $k=(E,0,0,E)$, and $\epsilon=(0,1,0,0)\oplus(0,0,1,0)$. Then
 \begin{equation}
@@
 \end{equation}
-for Ward identity $k_\mu M^\mu=0$. Therefore the replacement
-\begin{equation}
- \sum\s{pol.}\epsilon_\mu\epsilon'_\nu \to \Hmnd
-\end{equation}
-is valid.
+for Ward identity $k_\mu M^\mu=0$. Now we can see the validity easily.
 
-Secondly we think about the double photons case
+Next we think about the double photons case
 \footnote{This part is derived from $B_@8}9,0l(B's notebook.}
@@
 Then doing some simple calculations, we can get
-\begin{equation}
-  \sum\s{pol.}\left|M\right|^2 = |M^{11}|^2+|M^{12}|^2+|M^{21}|^2+|M^{22}|^2.
-\end{equation}
-Nevertheless, na\"ive replacement does not work, because our Ward identities
+\begin{align}
+  \sum\s{pol.}\left|M\right|^2
+&= \sum\s{pol.}\epsilon_\mu^*(k)\epsilon_\nu(k)\epsilon_\rho^*(k')\epsilon_\sigma(k')
+M^{\mu\rho}M^{\nu\sigma*}\\
+&= |M^{11}|^2+|M^{12}|^2+|M^{21}|^2+|M^{22}|^2\\
+&\stackrel{?}{=} \Hmn\Hrs M^{\mu\rho}M^{\nu\sigma*}.
+\end{align}
+Badly, our Ward identities
 \begin{equation}
@@
 \end{equation}
-obviously does not help us. If we can omit $\epsilon$s from these
-identities, that is if
+do not help us; now we want to be given
 \begin{equation}
@@
 \end{equation}
-we can recover validity of the replacement:
-\begin{align}
- \Hmrd\Hnsd M^{\mu\nu}M^{\rho\sigma*}
-&= -\Hnsd\left(
-M^{1\nu}M^{1\sigma*}+M^{2\nu}M^{2\sigma*}
-\right)\\
-&= |M^{11}|^2+|M^{12}|^2+|M^{21}|^2+|M^{22}|^2.
-\end{align}
+to recover validity of the replacement.
 
-Then what's happening? Why this replacement is not valid?
-Actually our new conditions \eqref{eq:TrueIdentities} seem to guarantee
-that we are summing not only ``physical'' but also ``unphysical'' polarizations.
-Meanwhile if we use some physical condition such as $\epsilon\cdot k=0$,
-\eqref{eq:TrueIdentities} break down while Ward identities
-\eqref{eq:WardIdentities} are still valid.
+The difference between \eqref{eq:WardIdentities} and \eqref{eq:TrueIdentities} is that the former considers (and tries to sum up) all polarizations but the latter does only physical ones.
+Actually, as long as we are summing up all polarizations, the replacement is still valid; these two conditions are equivalent because of cancellation of unphysical polarizations.
+However, once we restrict the polarizations (for example with using a relation $\epsilon\cdot k=0$),
+we can no more use \eqref{eq:TrueIdentities} and thus the replacement becomes invalid.
 
 Now let's check what is happening from another viewpoint. First we suppose
-$M$ satisfies our new conditions \eqref{eq:TrueIdentities}, and define
+$M$ satisfies our latter conditions \eqref{eq:TrueIdentities}, and define
 $\tilde M^{\mu\nu}$ and $\tilde M$ as
@@
 \end{align}
-This alternative amplitude satisfies Ward identities (since photon is
-massless and $\epsilon\cdot k=0$), and furthermore $\tilde M = M$.
-Therefore $\tilde M$ is physically identical to $M$.
-However technically these are very different, just because we cannot
-perform our ``na\"ive replacement'' for this $\tilde M$:
+Here $\tilde M^{\mu\nu}\neq M^{\mu\nu}$ but $\tilde M=M$; thus $\tilde M$ satisfies Ward identities
+(since photon is massless and $\epsilon\cdot k=0$).
+However, we cannot utilize the replacement for $\tilde M$, while it is valid for $M$.
+If you did the replacement, a wrong result would come out, like
 \begin{align}
@@
 \begin{align}
- \sum\s{pol.}|\tilde M|^2 &=
- \sum\s{pol.}|M|^2\qquad \text{(Furthermore $\tilde M=M$)}\notag\\
+ \sum\s{pol.}|M|^2
 &= \sum\s{pol.}|\epsilon_\mu^*(k)\epsilon_\nu'^*(k')M^{\mu\nu}|^2
- = \sum\s{pol.}|\epsilon_\mu^*(k)\epsilon_\nu'^*(k')\tilde M^{\mu\nu}|^2\\
-&= \Hmrd\Hnsd M^{\mu\nu}M^{\rho\sigma*}\notag\\
-&\ne \Hmrd\Hnsd \tilde M^{\mu\nu}\tilde M^{\rho\sigma*}
- = \sum\s{pol.}|\tilde M|^2 + \left[(k\cdot p'^*)(k'\cdot p) + \Hc\right].\notag
+ = \Hmrd\Hnsd M^{\mu\nu}M^{\rho\sigma*}
+\notag\\
+ = \sum\s{pol.}|\tilde M|^2
+& = \sum\s{pol.}|\epsilon_\mu^*(k)\epsilon_\nu'^*(k')\tilde M^{\mu\nu}|^2
+\ne \Hmrd\Hnsd \tilde M^{\mu\nu}\tilde M^{\rho\sigma*}
+ = \sum\s{pol.}|\tilde M|^2 + \left[(k\cdot p'^*)(k'\cdot p) + \Hc\right].
 \end{align}
+
+To check the Ward identity always helps us!
 

### 2013-01-09 VerboseNotes.tex
Cheatsheet UTF-8に変更
git-svn-id: https://dav.misho-web.com/svn/cyghome/trunk/QFT/CheatSheet@947 c04c31ef-c4f7-4b37-ba94-86ebe2b9c820

@@
-%#!./compile CheatSheet
-%%% Time-Stamp: <2011-05-04 22:23:09 Misho>
-%%% •¶ŽšƒR[ƒh‚ÍShift-JIS‚Å‚·‚æ‚Ë
+%#!platexmake CheatSheet
+%%% Time-Stamp: <>
+%%% 一部で日本語が使用されています。
 
@@
 
-$BHy>.JQ49$O(B$\epsilon\T^\mu_\nu=\spmat{
+微小変換は$\epsilon\T^\mu_\nu=\spmat{
 0&\beta_x&\beta_y&\beta_z\\
@@
 \beta_y & \theta_z & 0 & -\theta_x\\
-\beta_y & -\theta_y & \theta_x & 0}$$B$N7A$H$J$C$F$$$k$N$G!$2sE>@8@.;R(B$\vc J$$B$H2CB.@8@.;R(B$\vc K$$B$O(B
+\beta_y & -\theta_y & \theta_x & 0}$の形となっているので，回転生成子$\vc J$と加速生成子$\vc K$は
 \begin{equation}
@@
 \end{equation}
-$B$N7A$G$"$k!#$3$3$G(B$\kappa=\pm1$$B$O(Bnotation$B$G$"$k!#(B
+の形である。ここで$\kappa=\pm1$はnotationである。
 
-$B0lJ}!$Hy>.JQ49$+$i@8@.;R$r(B
+一方，微小変換から生成子を
  $\epsilon\T^\mu_\nu  =: \mp\frac\ii2\epsilon^{\rho\sigma}(J_{\rho\sigma})\T^\mu_\nu$
-$B$HDj5A$9$k$H!$7WNL$K$h$i$:$K(B$\epsilon_{\mu\nu}$$B$OH?BP>N$H$J$j!$(B
+と定義すると，計量によらずに$\epsilon_{\mu\nu}$は反対称となり，
 \begin{align*}
@@
 \end{align*}
-$B$G!$=>$C$F(B$J^{\rho\sigma}$$B$bH?BP>N!#(B
+で，従って$J^{\rho\sigma}$も反対称。
 $ (J_{\rho\sigma})\T^\mu_\nu =
 \pm\ii\left(\delta^\mu_\rho\Hsnd-\delta^\mu_\sigma\Hrnd\right)$
-$B$H$J$j!$8r494X78$,F@$i$l!$$3$l$,JD$8$F$$$k$N$G(BLie$BBe?t$G$"$k$3$H$b$o$+$k!'(B
+となり，交換関係が得られ，これが閉じているのでLie代数であることもわかる：
 \begin{equation}
@@
 
-$B@8@.;R$N6qBN7A$O7WNL$K0MB8$7!$(B
+生成子の具体形は計量に依存し，
 $
@@
  \spmat{0&0&0&0\\0&0&0&0\\0&0&0&-1\\0&0&1&0}_{\mu\nu}
-$$B$H$J$k!#(B
-$B$h$C$F!$$3$3$G$NJ#9f$N<h$jJ}$H(B$\kappa$$B$*$h$S7WNL$NDj5A$K$h$C$F!$(B$\vc J$$B!&(B$\vc K$$B$H(B$J_{\rho\sigma}$$B$NBP1~$,Dj$^$k$3$H$K$J$k!#(B
+$となる。
+よって，ここでの複号の取り方と$\kappa$および計量の定義によって，$\vc J$・$\vc K$と$J_{\rho\sigma}$の対応が定まることになる。
 \starline
 
-$\kappa$$B$HJ#9f$K$D$$$F(B$({\RED -,$B>e(B})$,$({\RED +,$B2<(B})$,$({\BLUE -,$B2<(B})$,$({\BLUE +,$B>e(B})$$B$r<h$l$P(B
+$\kappa$と複号について$({\RED -,上})$,$({\RED +,下})$,$({\BLUE -,下})$,$({\BLUE +,上})$を取れば
 \begin{equation}
@@
 \end{equation}
-$B$H$J$k!'(B
+となる：
 \begin{equation}
@@
 \subsubsection{Lorentz group and $\gSL(2,\Complex)$}
-$B<!$K!$O"7k(BLie$B72(B$L_0$$B$,!$O"7k(BLie$B72(B$\gSL(2,\Complex)/\gZ_2$$B$HF17?$G$"$k$3$H$r8+$k!'(B
+次に，連結Lie群$L_0$が，連結Lie群$\gSL(2,\Complex)/\gZ_2$と同型であることを見る：
 \begin{align}
@@
 \end{align}
-$B$^$:(B,$\Sm$$B$r!J6K$a$F0lHLE*$K!K(B $\Sm:=(\alpha1,\beta\vc\sigma)$$B$HDj5A$9$k(B($\alpha=\beta=\pm1$)$B!#(B
-$x^2 = ({\RED+}{\BLUE-})\det(x_\mu\Sm)$$B$J$N$G(B
+まず,$\Sm$を（極めて一般的に） $\Sm:=(\alpha1,\beta\vc\sigma)$と定義する($\alpha=\beta=\pm1$)。
+$x^2 = ({\RED+}{\BLUE-})\det(x_\mu\Sm)$なので
 \begin{equation}
@@
 \end{equation}
-$B$O(B$x^2$$B$rJ]B8$9$k!#$h$C$F(BLorentz$BJQ49$G$"$j!$@8@.;R$rHf$Y$k$3$H$G6I=jF17?$@$H$o$+$k!'(B
+は$x^2$を保存する。よってLorentz変換であり，生成子を比べることで局所同型だとわかる：
 \begin{rightnote}
-$\gSL(2,\Complex)\ni g = \exp(-\ii a)$$B$H$7$F(B
-$x_\mu (g \Sm g^\dagger) = \Lambda\T_\mu^\nu x_\nu \Sm$$B$rHy>.E83+$9$k$H(B
+$\gSL(2,\Complex)\ni g = \exp(-\ii a)$として
+$x_\mu (g \Sm g^\dagger) = \Lambda\T_\mu^\nu x_\nu \Sm$を微小展開すると
 \begin{equation}
@@
 \end{equation}
-$B$G$"$j!$$3$3$+$i(B$g$$B$,$o$+$k!'(B
+であり，ここから$g$がわかる：
 \begin{equation}
@@
 
-$B$3$N$3$H$rJL$N4QE@$+$i8+$k!#(BLorentz$B72$N@8@.;R$N8r494X78$r8+$k$H!$!J@5$7$/J#9f$r<h$C$?>l9g!K(B
+このことを別の観点から見る。Lorentz群の生成子の交換関係を見ると，（正しく複号を取った場合）
 \begin{equation}
@@
 \end{equation}
-$B$H$J$k$N$G!$(B
+となるので，
 \begin{equation}
@@
 \end{equation}
-$B$HDj5A$9$k$H(B
+と定義すると
 \begin{equation}
@@
 \end{equation}
-$B$H$J$j!$(BLorentz$B72$,(B$\gSU(2)\times\gSU(2)$$B$KJ,2r$G$-$k!#(B
+となり，Lorentz群が$\gSU(2)\times\gSU(2)$に分解できる。
 
 \subsection{Weyl Spinor}
-$\gSU(2)_A\times\gSU(2)_B$$B$KBP$7$F(B$(1/2,0)$$BI=8=$r0Y$9$b$N$r:84,$-(Bspinor $\xi$$B!$(B$(0,1/2)$$BI=8=$r0Y$9$b$N$r1&4,$-(Bspinor $\bar\xi$$B$HDj5A$9$k!#(B
+$\gSU(2)_A\times\gSU(2)_B$に対して$(1/2,0)$表現を為すものを左巻きspinor $\xi$，$(0,1/2)$表現を為すものを右巻きspinor $\bar\xi$と定義する。
 \begin{align}
@@
 
-$\alpha\beta=1$$B$H$9$k$H(B$g$$B$O:84,$-(Bspinor$B$NJQ49;R$H$J$k!#5-9f$r(B
+$\alpha\beta=1$とすると$g$は左巻きspinorの変換子となる。記号を
 $\xi_\alpha\mapsto g\T_\alpha^\beta\xi_\beta$,
 $\bar\xi^\dalpha\mapsto (g^\dagger)^{-1}\T^\dalpha_\dbeta\bar\xi^\dbeta$
-$B$HDj5A$9$k!#(B
+と定義する。
 
-$B<!$K!$(B$\xi^\alpha\chi_\alpha$$B$*$h$S(B$\bar\xi_\dalpha\bar\chi^\dalpha$$B$,(Bscalar$B$H$J$k$h$&$K$7$?$$!#(B
-$E=\spmat{0&1\\-1&0}$$B$H$7$F$*$/$H!$(B$-E\trans{g}E=g^{-1}$$B$h$j(B
+次に，$\xi^\alpha\chi_\alpha$および$\bar\xi_\dalpha\bar\chi^\dalpha$がscalarとなるようにしたい。
+$E=\spmat{0&1\\-1&0}$としておくと，$-E\trans{g}E=g^{-1}$より
 \begin{equation}
@@
 \end{equation}
-$B$h$C$F!$(B$\epsilon^{12}=\epsilon_{21}=1$$B$H$7$F(B
-$\xi^\alpha:=\epsilon^{\alpha\beta}\xi_\beta$$B!$(B$\xi_\alpha=\epsilon_{\alpha\beta}\xi^\beta$
-$B$H$9$l$PNI$$!#(B
+よって，$\epsilon^{12}=\epsilon_{21}=1$として
+$\xi^\alpha:=\epsilon^{\alpha\beta}\xi_\beta$，$\xi_\alpha=\epsilon_{\alpha\beta}\xi^\beta$
+とすれば良い。
 
-$BF1MM$K!$(B$\bar\xi'_\dalpha=-E(g^\dagger)^{-1}E\bar\xi_\dbeta$$B$+$i(B
-$(E^{\dalpha\dbeta}\bar\xi_\dbeta)'=(g^\dagger)^{-1}{}\T^\dalpha_\dbeta (E^{\dbeta\dgamma}\bar\xi_\dgamma)$$B$H$J$k!#(B$\epsilon_{12}=\epsilon_{\dot1\dot2}$$B$H$7$F(B$\bar\xi_\dalpha:=\epsilon_{\dalpha\dbeta}\bar\xi^\dalpha$$B$H$9$k$N$,0lHLE*$G$"$k!#(B
-$B$^$?!$$3$N$3$H$+$i(B$(\xi_a)^*=\bar\xi_\dalpha$$B!J0?$$$O(B$\xi^\dagger=\bar\xi$$B!K$,J,$+$k!#(B
+同様に，$\bar\xi'_\dalpha=-E(g^\dagger)^{-1}E\bar\xi_\dbeta$から
+$(E^{\dalpha\dbeta}\bar\xi_\dbeta)'=(g^\dagger)^{-1}{}\T^\dalpha_\dbeta (E^{\dbeta\dgamma}\bar\xi_\dgamma)$となる。$\epsilon_{12}=\epsilon_{\dot1\dot2}$として$\bar\xi_\dalpha:=\epsilon_{\dalpha\dbeta}\bar\xi^\dalpha$とするのが一般的である。
+また，このことから$(\xi_a)^*=\bar\xi_\dalpha$（或いは$\xi^\dagger=\bar\xi$）が分かる。
 
-$x_\mu\Sm\mapsto x_\mu(g\Sm g^\dagger)$$B$G$"$k$N$G!$(B
-$x_\mu(\xi^\alpha \Sm \bar\chi^\dalpha)$$B$O(Bscalar$B$G$"$k!#$h$C$F!$(B$\Sm_{\alpha\dalpha}$$B$N$h$&$K=q$1$k!#(B
-$B99$K$3$3$G(B$x_\mu(\bar\xi_\dalpha\bSm{}^{\dalpha\alpha}\chi_\alpha)$$B$b(Bscalar$B$H$J$k$h$&$K(B$\bar\sigma$$B$rDj$a$h$&!#(B
+$x_\mu\Sm\mapsto x_\mu(g\Sm g^\dagger)$であるので，
+$x_\mu(\xi^\alpha \Sm \bar\chi^\dalpha)$はscalarである。よって，$\Sm_{\alpha\dalpha}$のように書ける。
+更にここで$x_\mu(\bar\xi_\dalpha\bSm{}^{\dalpha\alpha}\chi_\alpha)$もscalarとなるように$\bar\sigma$を定めよう。
 \begin{equation}
@@
 \end{equation}
-$B$G$"$j!$$"$H$O(Bconvention$B$G$"$k!#(B
+であり，あとはconventionである。
 
@@
 Next we think about the double photons case
-\footnote{This part is derived from $B_@8}9,0l(B's notebook.}
+\footnote{This part is derived from 濱口幸一's notebook.}
 $M=\epsilon_\mu^*(k)\epsilon_\nu'^*(k')M^{\mu\nu}$.
@@
 You may think we forget to introduce
-$\ol\psi\G5\psi$$B!$(B
-$\ol\psi\G5\slashed D\psi$$B!$(B
-$\epsilon^{\mu\nu\rho\sigma}F^a_{\mu\nu}F^a_{\rho\sigma}$$B!$(B
+$\ol\psi\G5\psi$，
+$\ol\psi\G5\slashed D\psi$，
+$\epsilon^{\mu\nu\rho\sigma}F^a_{\mu\nu}F^a_{\rho\sigma}$，
 $\epsilon^{\mu\nu\rho\sigma}D_\mu D_\nu F^a_{\rho\sigma}$
@@
 
-\subsection{$BML(B-Mills Theory}\label{sec:yang-mills-theory}
+\subsection{楊-Mills Theory}\label{sec:yang-mills-theory}
 

### 2013-01-09 VerboseNotes.tex
cheatsheat to MyCommand4
git-svn-id: https://dav.misho-web.com/svn/cyghome/trunk/QFT/CheatSheet@949 c04c31ef-c4f7-4b37-ba94-86ebe2b9c820

@@
 Lorentz transf. in $\mathbb R^{1,3}$ & Linear transf. $x^\mu \mapsto \Lambda\T^\mu_\nu x^\nu$ which conserve $x^2$.\\
- & \then $\Hrsd = \Hmnd \Lambda\T^\mu_\rho \Lambda\T^\nu_\sigma.$
+ & $\Longrightarrow \Hrsd = \Hmnd \Lambda\T^\mu_\rho \Lambda\T^\nu_\sigma.$
  and form a group $L$.\\
  & \qquad
-   ($\then\quad
-     (\Lambda^{-1})\T^\mu_\nu=\eta_{\nu\alpha}\eta^{\mu\beta}\Lambda\T^\alpha_\beta=:\Lambda\T_\nu^\mu\quad\then\quad\Lambda\T^\mu_\nu\Lambda\T_\mu^\rho=\delta^\rho_\nu$)\\[.8zw]
+   ($\Longrightarrow\quad
+     (\Lambda^{-1})\T^\mu_\nu=\eta_{\nu\alpha}\eta^{\mu\beta}\Lambda\T^\alpha_\beta=:\Lambda\T_\nu^\mu\quad\Longrightarrow\quad\Lambda\T^\mu_\nu\Lambda\T_\mu^\rho=\delta^\rho_\nu$)\\[.8zw]
 
@@
  \Lambda = \exp\left[\kappa\ii(\vipro\theta J +\vipro\beta K)\right]
-\quad\then\quad
+\quad\Longrightarrow\quad
 J_x = -\kappa\ii\spmat{0&0&0&0\\0&0&0&0\\0&0&0&-1\\0&0&1&0},\quad
@@
 \begin{equation}
-\Lambda\T^\mu_\nu\Sn=g^{-1}\Sm (g^{-1})^\dagger\quad\then\quad
+\Lambda\T^\mu_\nu\Sn=g^{-1}\Sm (g^{-1})^\dagger\quad\Longrightarrow\quad
 \epsilon\T^\mu_\nu\Sn=\ii(a\Sm-\Sm a^\dagger)
@@
 You may think we forget to introduce
-$\ol\psi\G5\psi$，
-$\ol\psi\G5\slashed D\psi$，
+$\overline\psi\G5\psi$，
+$\overline\psi\G5\slashed D\psi$，
 $\epsilon^{\mu\nu\rho\sigma}F^a_{\mu\nu}F^a_{\rho\sigma}$，
@@
 \subsection{Spinor}
-\Paragraph{$\eta^{\mu\nu}=(-,+,+,+)$ case}
+\paragraph{$\eta^{\mu\nu}=(-,+,+,+)$ case}
 \begin{tabular}[t]{l@{\ :\ }l}
@@
 & $(ab)^\dagger=b^\dagger a^\dagger$ for $a,b\in\Grassmann$\\
-& \then for $a,b\in\RealGrassmann$, $ab\in\ii\RealGrassmann$\\
+& $\Longrightarrow$ for $a,b\in\RealGrassmann$, $ab\in\ii\RealGrassmann$\\
  $\gamma$ matrix

### 2014-01-01 VerboseNotes.tex
SU(2)などいろいろ追加
git-svn-id: https://dav.misho-web.com/svn/cyghome/trunk/QFT/CheatSheet@1044 c04c31ef-c4f7-4b37-ba94-86ebe2b9c820

@@
 %#!platexmake CheatSheet
-%%% Time-Stamp: <>
+%%% Time-Stamp: <2014-01-02 02:19:22 misho>
 %%% 一部で日本語が使用されています。
@@
 
+\subparagraph{Conjugate representation}
+Consider a representation $r$ and its generators $\{T_r^a\}$.
+Then $\{-(T_r^a)^*\}$ also satisfies the group structure.
+The corresponding representation is called the {\em conjugate representation} $\overline r$ of $r$:
+\begin{equation}
+ T_{\overline r}^a:=-(T_r^a)^*.
+\end{equation}
+If $\overline r$ is equivalent to $r$, i.e. there exists an unitary matrix $U$ such that
+$T_{\overline r}^a=U^\dagger(T_r^a)U$, the representation $r$ is called a {\em real representation}.
+This $U$ is unique up to a trivial scale factor, and either symmetric or anti-symmetric.
+If $U$ is symmetric, the representation is called {\em real-positive}; otherwise {\em real-negative} or {\em pseudo-real}.
+
+The $\boldsymbol2$ representation of $\gSU(2)$ is pseudo-real because
+\begin{equation}
+ T_{\overline {{\boldsymbol 2}}}^a \equiv -(\sigma^a/2)^* = \sigma^2(\sigma^a/2)\sigma^2.
+\end{equation}
+
+If a field $X$ transforms under a representation $r$, its complex conjugate does under $\overline r$:
+\begin{equation}
+  \delta X=\ii g\theta^a(T_r^a X)
+\quad\Longrightarrow\quad
+ \delta X^*
+=-\ii g\theta^a(T_r^A X)^*
+=\ii g\theta^a(T_{\overline r}^AX^*)
+\end{equation}
+
 \newpage

### 2014-01-09 VerboseNotes.tex
Feynman rules 改訂
git-svn-id: https://dav.misho-web.com/svn/cyghome/trunk/QFT/CheatSheet@1046 c04c31ef-c4f7-4b37-ba94-86ebe2b9c820

@@
 %#!platexmake CheatSheet
-%%% Time-Stamp: <2014-01-02 02:19:22 misho>
+%%% Time-Stamp: <2014-01-09 21:34:25 misho>
 %%% 一部で日本語が使用されています。
@@
 
+\subsection{Note on the symmetry factor in the Feynman Rules}
+\label{sec:note-symmetry-factor}
+If a vertex consists of different particles, the Feynman rule for the vertex is defined as $\ii\lambda$, where $\lambda$ is the coupling constant for the vertex in the Lagrangian.
+However, if a particle appears $n$-times in the coupling, we define the vertex as $(n!)\times\ii\lambda$.
+
+This factor $n!$ corresponds to the freedom of contraction, which we encounter when contracting the fields.
+For example, $\bra{\phi_{k_1}\phi_{k_2}\phi_{k_3}}(\lambda\phi^3\psi)\ket{\psi_p}$ has $3!$ choice of the contraction; thus we define the vertex rule as $(3!)\ii\lambda$, and the amplitude is evaluated as $6\ii\lambda$.
+
+For this multiplication, we have to consider a symmetry factor.
+$\begC1{\bra{\phi}}\conC{(\lambda}\endC1{\phi}\begC1{{}^4}\conC{)(\lambda'}\endC1{\phi^3}\begC1{\chi}\conC{)}\endC1{\ket{\chi}}$ has $4\times3!$ freedom, but the rule is already multiplied by $4!\times3!$; thus we have to divide the amplitude by the symmetry factor ($=6$) to obtain $24\ii\lambda\lambda'$.
+Similarly, in
+$\begC1{\bra{\phi}}\conC{\frac12(\lambda}\endC1{\phi}\begC1{{}^4}\conC{)(\lambda}\endC1{\phi}\begC1{{}^4}\conC{)}\endC1{\ket{\phi}}+\begC2{\bra{\phi}}\conC{\frac12(\lambda}\begC3{\phi}\begC1{{}^4}\conC{)(\lambda}\endC1{\phi}\endC2{{}^4}\conC{)}\endC3{\ket{\phi}}
+=
+\begC1{\bra{\phi}}\conC{(\lambda}\endC1{\phi}\begC1{{}^4}\conC{)(\lambda}\endC1{\phi}\begC1{{}^4}\conC{)}\endC1{\ket{\phi}}$,
+ the symmetry factor $6=(4!)^2/(4\cdot4\cdot3!)$ appears.
+(Note that the $1/2$ from the Fourier expansion cancels with the freedom of operators' order.)
+
 %%% Local Variables:

### 2019-04-01 mathematics.tex
feat: start upgrading the document (in English)

\documentclass[CheatSheet]{subfiles}
\begin{document}

\detailstyle
\section{Mathematics}
\subsection{Matrix exponential}
Excerpted from \S2 and \S5 of Hall 2015 \cite{hall2015}:
\begin{alignat}{2}
 &\ee^X := \sum_{m=0}^\infty \frac{X^m}{m!} \text{~~(converges for any $X$)},
\quad&
 &\log X := \sum_{m=1}^\infty (-1)^{m+1}\frac{(A-1)^m}{m} \text{~~(conv.~if $\|A-I\|<1$)}.
\end{alignat}
\begin{alignat}{2}
 &\ee^{\log A} = A \text{~~(if $\|A-I\|<1$)},
\quad&
 &\log \ee^X = X \text{~and~} \|\ee^X-1\| < 1 \text{~~(if $\|X\|<\log2$).}
\end{alignat}
\begin{equation}
 \text{Hilbert-Schmidt norm}:~\|X\|^2 := \sum_{i,j}|X_{ij}|^2 = {\Tr X^\dagger X}.
\end{equation}
Properties:
\begin{alignat*}{4}
 \ee^{(X^\TT)} &= (\ee^X)^\TT,
&\qquad
 \ee^{(X^*)} &= (\ee^X)^*,
&\qquad
 (\ee^X)^{-1} &= \ee^{-X},
&\qquad
 \ee^{YXY^{-1}} &= Y\ee^X Y^{-1},
\end{alignat*}
\begin{alignat*}{3}
 \det \exp X &= \exp\Tr X,
&\qquad
 \frac{\dd}{\dd t}\ee^{tX} &= X\ee^{t X} = \ee^{tX} X
&\qquad
 \ee^{(\alpha+\beta)X} &= \ee^{\alpha X}\ee^{\beta X} \text{~for $\alpha, \beta\in \mathbb C$};
\end{alignat*}
Baker-Campbell-Hausdorff:
\begin{align}
  \ee^X Y \ee^{-X}
&= Y + [X, Y] + \frac1{2!}[X, [X, Y]] + \frac1{3!}[X, [X, [X, Y]]] + \cdots;
\\
  \log(\ee^X \ee^Y)
&= X + \int_0^1\!\dd t\, g(\ee^{[X,}\ee^{t[Y,})Y
\qquad\Bigl[
  g(z) = \frac{\log z}{1-z^{-1}} = 1+\sum_{n=1}^{\infty}\frac{(-1)^{n+1}(z-1)^n}{n(n+1)}
\Bigr]\\
&= X + Y + \frac12[X,Y] + \frac1{12}[X, [X,Y]] - \frac1{12}[Y, [X,Y]] + \cdots
\quad\text{(Baker-Campbell-Hausdorff).}
\end{align}

\end{document}

### 2019-04-01 spinors.tex
feat: start upgrading the document (in English)

\documentclass[CheatSheet]{subfiles}
\begin{document}

\changefontsizes{10pt}
\section{Spinors}

\changefontsizes{9pt}
\begin{equation}
 (\overline{\psi_1}\psi_2)^* = (\psi_2)^\dagger(\overline\psi_1)^\dagger = \overline{\psi_2}\psi_1.
\end{equation}

\end{document}

### 2019-04-01 calculator/convert_to_tex.pl

@@ -1,0 +1,126 @@
+#!/usr/bin/perl
+
+use strict;
+use warnings;
+
+my $scalar = {
+  "Hu" => "\\hu_{SUB}^{SUP}",
+  "Hd" => "\\hd_{SUB}^{SUP}",
+  "Q"  => "\\tqL[SUB]^{SUP}",
+  "bU" => "\\tuR[SUB]^{SUP*}",
+  "bD" => "\\tdR[SUB]^{SUP*}",
+  "L"  => "\\tlL[SUB]^{SUP}",
+  "bE" => "\\teR[SUB]^{SUP*}"
+};
+
+my $param = {
+  "yu" => "\\yu[SUB]^{SUP}",
+  "yd" => "\\yd[SUB]^{SUP}",
+  "ye" => "\\ye[SUB]^{SUP}",
+  "\\[Mu]" => "\\mu^{SUP}_{SUB}",
+  "\\[Mu]p" => "\\mu'^{SUP}_{SUB}",
+  "\\[Lambda]" => "\\lambda^{SUP}_{SUB}",
+  "\\[Lambda]p" => "\\lambda'^{SUP}_{SUB}",
+  "\\[Lambda]pp" => "\\lambda''^{SUP}_{SUB}",
+  "yudagyu" => "(\\yu^\\dagger\\yu)^{SUP}_{SUB}",
+  "yddagyd" => "(\\yd^\\dagger\\yd)^{SUP}_{SUB}",
+  "yedagye" => "(\\ye^\\dagger\\ye)^{SUP}_{SUB}",
+  "yuyudag" => "(\\yu\\yu^\\dagger)^{SUP}_{SUB}",
+  "ydyddag" => "(\\yd\\yd^\\dagger)^{SUP}_{SUB}",
+  "yeyedag" => "(\\ye\\ye^\\dagger)^{SUP}_{SUB}",
+};
+
+my $scalar_conj = {
+  "Hu" => "\\hu_{SUB}^{SUP*}",
+  "Hd" => "\\hd_{SUB}^{SUP*}",
+  "Q"  => "\\tqL[SUB]^{SUP*}",
+  "bU" => "\\tuR[SUB]^{SUP}",
+  "bD" => "\\tdR[SUB]^{SUP}",
+  "L"  => "\\tlL[SUB]^{SUP*}",
+  "bE" => "\\teR[SUB]^{SUP}"
+};
+
+my $param_conj = {
[…]

### 2019-08-05 mathematics.tex
feat: add a formula for Matrix exponential
feat: mathematics exp formula
feat: complete SUSY section

@@
   \ee^X Y \ee^{-X}
-&= Y + [X, Y] + \frac1{2!}[X, [X, Y]] + \frac1{3!}[X, [X, [X, Y]]] + \cdots;
+&= Y + [X, Y] + \frac1{2!}[X, [X, Y]] + \frac1{3!}[X, [X, [X, Y]]] + \cdots = \ee^{[X,]}Y;
+\\
+  \ee^X \ee^Y \ee^{-X}
+&= \sum_{n=0}^\infty\frac{1}{n!}(\ee^X Y \ee^{-X})^n
+ = \exp\left(\ee^{[X,]}Y\right);
 \\
@@
 \qquad\Bigl[
-  g(z) = \frac{\log z}{1-z^{-1}} = 1+\sum_{n=1}^{\infty}\frac{(-1)^{n+1}(z-1)^n}{n(n+1)}
+  g(z) = \frac{\log z}{1-z^{-1}} = 1-\sum_{n=1}^{\infty}\frac{(1-z)^n}{n(n+1)};
+\quad
+g(\ee^y)=\sum_{n=0}^{\infty}\frac{B_ny^n}{n!}
 \Bigr]\\
@@
 \end{align}
+\begin{align}
+   \log(\ee^X\ee^Y)
+&=\sum_{k=1}^\infty \frac{(-1)^{k-1}}{k}\left(\sum_{m,n=0}^{\infty}\frac{X^mY^n}{m!n!}-1\right)^k
+&=\sum_{k=1}^\infty
+   \sum_{m_1+n_1>0}\cdots\sum_{m_k+n_k>0}\frac{(-1)^{k-1}}{k}
+\frac{X^{m_1}Y^{n_1}\cdots X^{m_k}Y^{n_k}}{m_1!n_1!\cdots m_k!n_k!}
+\end{align}
+\begin{equation}
+   \log(\ee^X\ee^Y)=\sum_{k=1}^{\infty}\sum_{m_1+n_1>0}\cdots\sum_{m_k+n_k>0}
+\frac{(-1)^{k-1}}{k\sum_{i=1}^{k}(m_i+n_i)}
+\frac{
+\Bigl([X,\Bigr)^{m_1}
+\Bigl([Y,\Bigr)^{n_1}\cdots
+\Bigl([X,\Bigr)^{m_k}
+\Bigl([Y,\Bigr)^{n_k}]\cdots]
+}{m_1!n_1!\cdots m_k!n_k!}
+\end{equation}
+with $[X]:=X$ understood.
 
+If matrices $t^a$ satisfies $[t^a,t^b]=\ii f^{abc} t^c$ with totally-antisymmetric $f^{abc}\in\mathbb R$,
+\begin{equation}
+  \left[\ee^{\theta^at^a} t_b \ee^{-\theta^ct^c}\right]_{ij}
+=  \left[\ee^{\theta^a[t^a,}t_b\right]_{ij}
+= \left[\ee^{\ii \theta^af^{a}}\right]^{bc} t^c_{ij}
+\end{equation}
+holds for $\theta^a\in\mathbb C$, where $[f^a]_{bc}=f^{abc}$. \TODO{needs verification, generalization/restriction, and a nice proof or reference.}
 \end{document}

### 2019-08-07 calculator/mssm/convert_to_tex.pl
feat: add fermion part of superpotential

@@ -85,6 +85,17 @@ my $param_sq = {
   "g3" => "g_3^2",
 };
 
+my $fermion = {
+  "Hu" => "\\thu_{SUB}^{SUP}",
+  "Hd" => "\\thd_{SUB}^{SUP}",
+  "Q"  => "\\qL[SUB]^{SUP}",
+  "bU" => "\\uR[SUB]^{\\cc SUP}",
+  "bD" => "\\dR[SUB]^{\\cc SUP}",
+  "L"  => "\\lL[SUB]^{SUP}",
+  "bE" => "\\eR[SUB]^{\\cc SUP}"
+};
+
+
 sub fieldname_to_tex{
   my ($type, $name, $subscript, $superscript) = @_;
   my $tex = "";
@@ -102,6 +113,8 @@ sub fieldname_to_tex{
     $tex = $scalar_4->{$name} or die "undefined scalar $name";
   }elsif($type eq "P2"){
     $tex = $param_sq->{$name} or die "undefined param $name";
+  }elsif($type eq "F"){
+    $tex = $fermion->{$name} or die "undefined scalar $name";
   }
   $tex =~ s/SUB/$subscript/;
   $tex =~ s/SUP/$superscript/;

### 2019-08-07 calculator/mssm/convert_to_tex.pl
feat: use kappa instead of mu'

@@ -18,7 +18,7 @@ my $param = {
   "yd" => "\\yd[SUB]^{SUP}",
   "ye" => "\\ye[SUB]^{SUP}",
   "\\[Mu]" => "\\mu^{SUP}_{SUB}",
-  "\\[Mu]p" => "\\mu'^{SUP}_{SUB}",
+  "\\[Kappa]" => "\\kappa^{SUP}_{SUB}",
   "\\[Lambda]" => "\\lambda^{SUP}_{SUB}",
   "\\[Lambda]p" => "\\lambda'^{SUP}_{SUB}",
   "\\[Lambda]pp" => "\\lambda''^{SUP}_{SUB}",
@@ -45,7 +45,7 @@ my $param_conj = {
   "yd" => "\\yd[SUB]^{SUP*}",
   "ye" => "\\ye[SUB]^{SUP*}",
   "\\[Mu]" => "\\mu^{SUP*}_{SUB}",
-  "\\[Mu]p" => "\\mu'^{SUP*}_{SUB}",
+  "\\[Kappa]" => "\\Kappa^{SUP*}_{SUB}",
   "\\[Lambda]" => "\\lambda^{SUP*}_{SUB}",
   "\\[Lambda]p" => "\\lambda'^{SUP*}_{SUB}",
   "\\[Lambda]pp" => "\\lambda''^{SUP*}_{SUB}",
@@ -76,7 +76,7 @@ my $param_sq = {
   "yd" => "|\\yd[SUB]^{SUP}|^2",
   "ye" => "|\\ye[SUB]^{SUP}|^2",
   "\\[Mu]" => "|\\mu^{SUP}_{SUB}|^2",
-  "\\[Mu]p" => "|\\mu'^{SUP}_{SUB}|^2",
+  "\\[Kappa]" => "|\\kappa^{SUP}_{SUB}|^2",
   "\\[Lambda]" => "|\\lambda^{SUP}_{SUB}|^2",
   "\\[Lambda]p" => "|\\lambda'^{SUP}_{SUB}|^2",
   "\\[Lambda]pp" => "|\\lambda''^{SUP}_{SUB}|^2",

### 2019-08-07 calculator/mssm/convert_to_tex.pl
feat: add Lagrangian of MSSM

@@ -10,7 +10,15 @@ my $scalar = {
   "bU" => "\\tuR[SUB]^{SUP*}",
   "bD" => "\\tdR[SUB]^{SUP*}",
   "L"  => "\\tlL[SUB]^{SUP}",
-  "bE" => "\\teR[SUB]^{SUP*}"
+  "bE" => "\\teR[SUB]^{SUP*}",
+  "HuP" => "\\hup_{SUB}^{SUP}",
+  "HuZ" => "\\huz_{SUB}^{SUP}",
+  "HdZ" => "\\hdz_{SUB}^{SUP}",
+  "HdM" => "\\hdm_{SUB}^{SUP}",
+  "UL"  => "\\tuL[SUB]^{SUP}",
+  "DL"  => "\\tdL[SUB]^{SUP}",
+  "NUL" => "\\tnuL[SUB]^{SUP}",
+  "EL"  => "\\teL[SUB]^{SUP}",
 };
 
 my $param = {
@@ -28,6 +36,16 @@ my $param = {
   "yuyudag" => "(\\yu\\yu^\\dagger)^{SUP}_{SUB}",
   "ydyddag" => "(\\yd\\yd^\\dagger)^{SUP}_{SUB}",
   "yeyedag" => "(\\ye\\ye^\\dagger)^{SUP}_{SUB}",
+  "gY" => "g_Y^{SUP}_{SUB}",
+  "g2" => "g_2^{SUP}_{SUB}",
+  "g3" => "g_3^{SUP}_{SUB}",
+  "ee" => "|e|^{SUP}_{SUB}",
+  "gZ" => "g_Z^{SUP}_{SUB}",
+  "partial" => "\\partial^{SUP}_{SUB}",
+  "\\[Sigma]bmu" => "\\bsigma^{\\mu SUP}_{SUB}",
+  "sw" => "\\si{\\mathrm w}^{SUP}_{SUB}",
+  "cw" => "\\co{\\mathrm w}^{SUP}_{SUB}",
+  "\\tau" => "\\tau^{SUP}_{SUB}",
 };
 
 my $scalar_conj = {
@@ -37,7 +55,15 @@ my $scalar_conj = {
   "bU" => "\\tuR[SUB]^{SUP}",
   "bD" => "\\tdR[SUB]^{SUP}",
   "L"  => "\\tlL[SUB]^{SUP*}",
-  "bE" => "\\teR[SUB]^{SUP}"
+  "bE" => "\\teR[SUB]^{SUP}",
[…]

### 2019-08-17 calculator/mssm/convert_to_tex.pl
feat: tweak lagrangian

@@ -46,6 +46,7 @@ my $param = {
   "sw" => "\\si{\\mathrm w}^{SUP}_{SUB}",
   "cw" => "\\co{\\mathrm w}^{SUP}_{SUB}",
   "\\tau" => "\\tau^{SUP}_{SUB}",
+  "\\[CapitalTheta]G" => "\\Theta_G{}^{SUP}_{SUB}",
 };
 
 my $scalar_conj = {
@@ -218,6 +219,7 @@ foreach my $line(<>){
   $line =~ s/\$/\@/g;
   $line =~ s/\\left\(\s*\(-1\)/\\left\( - /g;
   $line =~ s/\+\s*\(-1\)\s*/- /g;
+  $line =~ s/\^\{2\}/^2/g;
   $line =~ s/\\frac\{(1)\}\{(\d+)\}\s*(g_Y\^2|g_2\^2|g_3\^2)/\\frac{$3}{$2}/g;
   $line =~ s/\\frac\{(\d+)\}\{(\d+)\}\s*(g_Y\^2|g_2\^2|g_3\^2)/\\frac{$1$3}{$2}/g;
   #

### 2019-08-17 calculator/mssm/convert_to_tex.pl
feat: start discussion on MSSM masses

@@ -47,6 +47,25 @@ my $param = {
   "cw" => "\\co{\\mathrm w}^{SUP}_{SUB}",
   "\\tau" => "\\tau^{SUP}_{SUB}",
   "\\[CapitalTheta]G" => "\\Theta_G{}^{SUP}_{SUB}",
+  "M1" => "M_1^{SUP}_{SUB}",
+  "M2" => "M_2^{SUP}_{SUB}",
+  "M3" => "M_3^{SUP}_{SUB}",
+  "c\\[Beta]" => "\\co{\\beta}^{SUP}_{SUB}",
+  "s\\[Beta]" => "\\si{\\beta}^{SUP}_{SUB}",
+  "c2\\[Beta]" => "\\co{2\\beta}^{SUP}_{SUB}",
+  "s2\\[Beta]" => "\\si{2\\beta}^{SUP}_{SUB}",
+  "mZ" => "m_Z^{SUP}_{SUB}",
+  "b" => "b^{SUP}_{SUB}",
+  "au" => "a^{SUP}_{{\\mathrm u} SUB}",
+  "ad" => "a^{SUP}_{{\\mathrm d} SUB}",
+  "ae" => "a^{SUP}_{{\\mathrm e} SUB}",
+  "mHd2" => "m_{\\Hd}^2^{SUP}_{SUB}",
+  "mHu2" => "m_{\\Hu}^2^{SUP}_{SUB}",
+  "mUc2" => "[m_{U^\\cc}^2]^{SUP}_{SUB}",
+  "mDc2" => "[m_{D^\\cc}^2]^{SUP}_{SUB}",
+  "mEc2" => "[m_{E^\\cc}^2]^{SUP}_{SUB}",
+  "mQ2" => "[m_{Q}^2]^{SUP}_{SUB}",
+  "mL2" => "[m_{L}^2]^{SUP}_{SUB}",
 };
 
 my $scalar_conj = {
@@ -76,6 +95,13 @@ my $param_conj = {
   "\\[Lambda]" => "\\lambda^{SUP*}_{SUB}",
   "\\[Lambda]p" => "\\lambda'^{SUP*}_{SUB}",
   "\\[Lambda]pp" => "\\lambda''^{SUP*}_{SUB}",
+  "M1" => "M_1^{*SUP}_{SUB}",
+  "M2" => "M_2^{*SUP}_{SUB}",
+  "M3" => "M_3^{*SUP}_{SUB}",
[…]

### 2019-08-22 calculator/mssm/convert_to_tex.pl
feat: show fermionic composition of MSSM

@@ -19,6 +19,12 @@ my $scalar = {
   "DL"  => "\\tdL[SUB]^{SUP}",
   "NUL" => "\\tnuL[SUB]^{SUP}",
   "EL"  => "\\teL[SUB]^{SUP}",
+  "HuZR" => "\\hu_{SUB}^{0RSUP}",
+  "HdZR" => "\\hd_{SUB}^{0RSUP}",
+  "HP" => "H_{SUB}^{+SUP}",
+  "G0" => "G_{SUB}^{0SUP}",
+  "GP" => "G_{SUB}^{+SUP}",
+  "A0" => "A_{SUB}^{0SUP}",
 };
 
 my $param = {
@@ -66,6 +72,8 @@ my $param = {
   "mEc2" => "[m_{E^\\cc}^2]^{SUP}_{SUB}",
   "mQ2" => "[m_{Q}^2]^{SUP}_{SUB}",
   "mL2" => "[m_{L}^2]^{SUP}_{SUB}",
+  "vu" => "\\vu",
+  "vd" => "\\vd",
 };
 
 my $scalar_conj = {
@@ -84,6 +92,8 @@ my $scalar_conj = {
   "DL"  => "\\tdL[SUB]^{SUP*}",
   "NUL" => "\\tnuL[SUB]^{SUP*}",
   "EL"  => "\\teL[SUB]^{SUP*}",
+  "HP" => "H_{SUB}^{-SUP}",
+  "GP" => "G_{SUB}^{-SUP}",
 };
 
 my $param_conj = {
@@ -164,6 +174,9 @@ my $fermion = {
   "DL"  => "\\dL[SUB]^{SUP}",
   "NUL" => "\\nuL[SUB]^{SUP}",
   "EL"  => "\\eL[SUB]^{SUP}",
+  "Neut"  => "\\neut[SUB]^{SUP}",
+  "CharP"  => "\\charP[SUB]^{SUP}",
+  "CharM"  => "\\charM[SUB]^{SUP}",
 };
 
 my $vector = {
@@ -199,6 +212,9 @@ my $fermion_conj = {
   "DL"  => "\\bdL[SUB]^{SUP}",
   "NUL" => "\\bnuL[SUB]^{SUP}",
   "EL"  => "\\beL[SUB]^{SUP}",
+  "Neut"  => "\\bar\\neut[SUB]^{SUP}",
+  "CharP"  => "\\bar\\charP[SUB]^{SUP}",
[…]

### 2019-09-12 spinors.tex
feat: (WIP) copy updates held on old note

@@
 
+\subsection{Convention \TODO{WIP!}}
+First we prepare a vector $x^\mu$ and a symmetric matrix $\eta^{\mu\nu}$, which we call ``contravariant vector'' $x^\mu$ and the metric $\eta^{\mu\nu}$.
+Then we perform a Lorentz transformation on $x^\mu$ to obtain $(x')^\mu$, with which we can define a matrix $\Lambda(\vc va,\vc \theta)^{\mu}_{\nu}$ through 
+$\Lambda{}^\mu{}_\nu x^{\nu}=(x')^{\mu}$.
+
+We then consider $\Lambda$s for infinitesimal transformations and define $\vc S$, $\vc J$, and $\vc K$ by
+\begin{align}
+ \Lambda^\mu{}_\nu
+  \simeq \delta^\mu_\nu-\ii(\vc\theta\cdot\vc J^\mu{}_\nu+\vc\beta\cdot\vc K^\mu{}_\nu)
+  \simeq \delta^\mu_\nu-\frac{\ii}{2}  \left[\Lambda^{\alpha\beta}S_{\alpha\beta}\right]^\mu{}_\nu
+\end{align}
+
+Imposing ``Lorentz condition'' (\TODO{what?}), we get the expression for $S=\ii(\delta\cdots)$ and $[J^i,J^j]=\cdots$; further, we get $\Lambda^{\mu}_{\nu}=\exp(-\ii\vc\theta\cdot\vc J -\ii \vc\xi\cdot\vc K)$, $\vc\theta=(\theta_{23},\theta_{31}, \theta_{12})$, $\vc\xi = \hat{\vc v}\tanh^{-1}\|\vc v\|=(\theta^{10},\theta^{20},\theta^{30})$; $J=(S_{23},S_{31},S_{12})$, $K=(S^{01}, S^{02}, S^{03})$....?
+
+Lorentz transformation with a rotation $\theta$ around an axis $\hat{\vc\theta}$ and a boost $\vc v$ are given by
+\begin{equation}
+ \Lambda = \exp\left[-\ii(\vc\theta\cdot\vc J+\vc \beta\cdot \vc K)\right];
+\qquad
+\vc\theta:=\theta\hat{\vc\theta}, \quad \vc\beta := \hat{\vc v}\tanh^{-1}\|\vc v\|,
+\end{equation}
+\TODO{check!}
+
+Lorentz transformation (infinitesimal):
+$\Lambda=\pmat{
+  0 &   & \vc{\beta}^\TT & \\
+    & 0 &-\theta_z & \theta_y\\
+  \vc{\beta} & \theta_z & 0 & -\theta_x \\
+   & -\theta_y & \theta_x & 0
+}$
+
+$[J_{\mu\nu}]^{\alpha}{}_{\beta}=
+ \ii(\delta^\alpha_\mu\eta_{\nu\beta}-\delta^\alpha_\nu\eta_{\mu\beta})
+$
+
+\vspace{2em}
+
+Lorentz tensor $
+M^{\mu_1\mu_2\cdots\mu_n}\propto
+\bsigma^{\mu_1\dbeta_1\alpha_1}\cdots M_{\alpha_1\cdots\dbeta_1\cdots}
+$
+
+Especially
+$V^\mu =: \frac12 \bsigma^{\mu\dbeta\alpha}V_{\alpha\dbeta}$,
+$V_{\alpha\dbeta}=V^\mu\sigma_{\mu\alpha\dbeta}$; hermite $V_{\alpha\dbeta}$ $\Leftrightarrow$ $real V^\mu$.
+
+\begin{align}
+ &(V^\TT)_{\alpha\dbeta} = V_{\beta\dalpha},\text{\TODO{(correct? possibly wrong dot-positions?)}}\\
+ &(V^*)_{\dalpha\beta}:= (V_{\alpha\dbeta})^*,\\
+ &(V^\dagger)_{\alpha\dbeta} := (V_{\beta\dalpha})^* = (V^*)_{\dbeta\alpha}
+\end{align}
+\TODO{anyway not very sure about the reasoning; though my old note says like this...}
+
+In general, metric is symmetric.
+
+\begin{equation}
+ (\Lambda^{-1})^\mu{}_\nu = \eta_{\nu\rho}\Lambda^{\rho}{}_\sigma(\eta^{-1})^{\sigma\mu}=:\Lambda_\nu{}^\mu
+\end{equation}
+
 \end{document}

### 2020-03-18 cosmology.tex
add one cosmology formula

\documentclass[CheatSheet]{subfiles}
\newcommand{\ddP}[2][3]{\frac{\dd^{#1}\vc#2}{(2\pi)^{#1}}}

\begin{document}

\section{Cosmology}
\detailstyle
\subsection{Cosmology formulae}
Assuming that $\sigma$ depends only on $s=(p_A+p_B)^2$, \TODO{(or always true?)}
\begin{align}
& \int\ddP{p_A}\,\,\ddP{p_B}\ee^{-E_A/T}\ee^{-E_B/T}\sigma(s)v_{\text{M\o l}}
\\&=4\int\dd\Pi_A\dd\Pi_B\, \ee^{-(E_A+E_B)/T}\sigma(s)\sqrt{(p_A\cdot p_B)^2-m_A^2m_B^2}
\\&=
\frac{1}{32\pi^4}
\int_{(m_A+m_B)^2}^\infty\dd s
\int_{\sqrt{s}}^\infty\dd E_+\left[\frac{2\sqrt{E_+^2-s}}{s}\mathop{\lambda^{1/2}}(s,m_A^2,m_B^2)\right]
\ee^{-E_+/T}\sigma(s)\left[\frac12\mathop{\lambda^{1/2}}(s,m_A^2,m_B^2)\right]
\\&=
\frac{T}{32\pi^4}
\int_{(m_A+m_B)^2}^\infty\dd s\,s^{3/2} \mathop{K_1}\left(\frac{\sqrt{s}}{T}\right)
\mathop{\lambda}\left(1,\frac{m_A^2}s,\frac{m_B^2}s\right)\sigma(s).
\end{align}

\end{document}

### 2020-05-30 mathematics.tex
feat: add general exp of unitary matrix
some updates in matrix-exp derivative

@@
 \end{alignat*}
-\begin{alignat*}{3}
+\begin{alignat*}{2}
  \det \exp X &= \exp\Tr X,
-&\qquad
- \frac{\dd}{\dd t}\ee^{tX} &= X\ee^{t X} = \ee^{tX} X
 &\qquad
@@
 \end{equation}
-with $[X]:=X$ understood.
+\hfill with $[X]$ being $X$.
+\\
+Derivative:
+\begin{alignat}{1}
+ &\frac{\dd}{\dd t}\ee^{tX} = X\ee^{t X} = \ee^{tX} X\\
+ &\ee^{-X(t)}\left(\frac{\dd}{\dd t}\ee^{X(t)}\right) =
+    \frac{I-\ee^{-\ad_X}}{\ad_X}\left(\frac{\dd X}{\dd t}\right)
+  = X' + \frac{[-X,X']}{2!} + \frac{[-X,[-X,X']]}{3!} + \cdots
+\\
+ &\left(\frac{\dd}{\dd t}\ee^{X(t)}\right)\ee^{-X(t)}
+= X' + \frac{[X,X']}{2!} + \frac{[X,[X,X']]}{3!} + \cdots
+\end{alignat}
+\hfill where $X'=\dd X/\dd t$ and $\ad_X(Y)=[X,Y]$ is the adjoint action of a Lie algebra. Thus, explicitly,
+\begin{alignat}{1}
+\frac{\dd}{\dd t}\ee^{aX(t)}
+  = \ee^{aX}\left\{\sum_{n=0}^\infty\frac{a^{n+1}}{(n+1)!}\Bigl([-X,\Bigr)^n X']\right\}
+  = \left\{\sum_{n=0}^\infty\frac{a^{n+1}}{(n+1)!}\Bigl([X,\Bigr)^n X']\right\}\ee^{aX}
+\end{alignat}
 
-If matrices $t^a$ satisfies $[t^a,t^b]=\ii f^{abc} t^c$ with totally-antisymmetric $f^{abc}\in\mathbb R$,
+\vspace{1em}
+
+\noindent
+Component:\quad If matrices $t^a$ satisfies $[t^a,t^b]=\ii f^{abc} t^c$ with totally-antisymmetric $f^{abc}\in\mathbb R$,
 \begin{equation}
@@
 holds for $\theta^a\in\mathbb C$, where $[f^a]_{bc}=f^{abc}$. \TODO{needs verification, generalization/restriction, and a nice proof or reference.}
+
+\subsection{General unitary matrix}
+\begin{equation}
+ U_2 =
+\pmat{1&0\\0&\ee^{\ii\alpha}}\pmat{\co\theta&\si\theta\\-\si\theta&\co\theta}\pmat{\ee^{\ii\beta}&0\\0&\ee^{\ii\gamma}}
+=
+\pmat{\phantom{-}
+\co\theta\ee^{\ii\beta} & \si\theta\ee^{\ii\gamma}\\
+-\si\theta\ee^{\ii(\alpha+\beta)} & \co\theta\ee^{\ii(\alpha+\gamma)}
+}
+\end{equation}
+\begin{align}
+ U_3 &=
+\pmat{1&&\\&\ee^{\ii a}&\\&&\ee^{\ii b}}
+ \pmat{1&&\\&\co{23}&\si{23}\\&-\si{23}&\co{23}}
+ \pmat{\co{13}&&\si{13}\ee^{-\ii\delta}\\&1&\\-\si{13}\ee^{\ii\delta}&&\co{13}}
+ \pmat{\co{12}&\si{12}&\\-\si{12}&\co{12}&\\&&1}
+\pmat{\ee^{\ii c}&&\\&\ee^{\ii d}&\\&&\ee^{\ii e}}
+\\&=
+\pmat{1&&\\&\ee^{\ii a}&\\&&\ee^{\ii b}}
+ \pmat{
+ \co{12} \co{13} & \si{12} \co{13} & \si{13} \ee^{-\ii\delta}\\
+ -\si{12} \co{23} - \co{12} \si{23} \si{13} \ee^{\ii\delta}& \co{12} \co{23} - \si{12}\si{23}\si{13} \ee^{\ii\delta}& \si{23}\co{13}\\
+  \si{12}\si{23} - \co{12} \co{23} \si{13}\ee^{\ii\delta} & -\co{12}\si{23}-\si{12}\co{23}\si{13}\ee^{\ii\delta} & \co{23} \co{13}}
+\pmat{\ee^{\ii c}&&\\&\ee^{\ii d}&\\&&\ee^{\ii e}}
+\end{align}
+(see, e.g., Ref.~\cite{Rasin:1997pn}).
+
 \end{document}

### 2020-08-07 Calculation.tex
write a bit on polarization sum

\documentclass[CheatSheet]{subfiles}

\begin{document}
\summarystyle

\section{Calculation techniques}
\paragraph{Polarization sum}
\begin{alignat}{2}
\text{massless:}\quad
\sum_{\pm}\epsilon^*_\mu(k) \epsilon_\nu(k)
&=
-\eta_{\mu\nu} - \frac{n^2 k_\mu k_\nu}{(n\cdot k)^2} + \frac{n_\mu k_\nu + n_\nu k_\mu}{n\cdot k}
\quad\Bigl(\leadsto -\eta_{\mu\nu}\quad\text{with Ward id.}\Bigr),
\\
\text{massive:}\quad
\sum_{\pm,0}\epsilon^*_\mu(k) \epsilon_\nu(k)
&=
-\eta_{\mu\nu}+\frac{k_\mu k_\nu}{m^2},
\end{alignat}
where $\epsilon\cdot k = 0$ is assumed and $n^\mu$ should satisfy $\epsilon\cdot n=0$, and $k\cdot n\neq 0$ (usually $n^\mu=(1,0,0,0)^\mu$).
\TODO{derivation}
\end{document}

### 2020-08-07 spinors.tex
change the font and style

@@
 
-\subsection{Convention \TODO{WIP!}}
+\subsection[Convention]{Convention \TODO{WIP!}}
 First we prepare a vector $x^\mu$ and a symmetric matrix $\eta^{\mu\nu}$, which we call ``contravariant vector'' $x^\mu$ and the metric $\eta^{\mu\nu}$.

### 2020-08-07 Calculation.tex
update many things

@@
 \TODO{derivation}
+
+\clearpage
+\detailstyle
+
+\subsection{Casas--Ibarra parameterization}
+\paragraph{Set up}
+We start from the neutrino mass matrix
+\begin{equation}
+-\mathcal L\supset \frac12
+ \pmat{\nu & n} \pmat{0 & v Y^\TT/\sqrt2 \\ v Y/\sqrt2 & M_N} \pmat{\nu \\ n}
+= \frac12\pmat{\nu_i & n_a} \pmat{0 & v (Y^\TT)_{ib}/\sqrt2 \\ v Y_{aj}/\sqrt2 & M_{Nab}} \pmat{\nu_j \\ n_b}
+\end{equation}
+(and its Hermitian conjugate), which can be obtained, e.g., from
+\begin{equation}
+ \mathcal L \supset \overline{N}Y H L - \frac12\overline{N}M_N N^\cc + \text{h.c.},
+\qquad
+\vev{H}=\pmat{0\\v/\sqrt2}\text{~with~}v\sim246\GeV,
+\end{equation}
+where the four spinors should be decomposed into Weyl spinors as
+$L^1=\spmat{\nu\\0}$,
+$N=\spmat{0\\\bar n}$.
+
+\paragraph{Generalized CIP}
+The mass matrix can be AT-factorized to yield 6 neutrino masses.
+This ATF can be separated into two steps:
+\begin{equation}
+ \pmat{0 & v Y^\TT/\sqrt2 \\ v Y/\sqrt2 & M_N}
+\longrightarrow
+\pmat{M\w L & 0 \\ 0 & M\w H}
+\stackrel{\text{ATF}}\longrightarrow
+\pmat{M\w L^{\text{diag}} & 0 \\ 0 & M\w H^{\text{diag}}},
+\end{equation}
+where $M\w L$ and $M\w H$ are complex symmetric matrices.
+Explicitly, with unitary matrices  $U_1$, $U_2$, and $U_3$,
+\begin{align}
+\pmat{M\w L & 0 \\ 0 & M\w H}
+&=
+ U_1^\TT \pmat{0 & v Y^\TT/\sqrt2 \\ v Y/\sqrt2 & M_N} U_1,
+&M\w L^\text{diag} &= U_2^\TT M\w L U_2,
+&M\w H^\text{diag} &= U_3^\TT M\w H U_3.
+\end{align}
+The first equation is calculated as an expansion in $v/M_N$:
+\begin{align}
+ U_1  &\simeq \pmat{1 & v Y^\dagger (M_N^*)^{-1}/\sqrt2 \\ -v M_N^{-1}Y/\sqrt2 & 1},
+&M\w L&\simeq -\frac{v^2}{2}Y^\TT M_N^{-1} Y,
+&M\w H&\simeq M_N.
+\end{align}
+These expressions lead to
+\begin{equation}
+  M\w L^{\text{diag}}
+\simeq \frac{-v^2}{2} U_2^\TT Y^\TT U_3(M_H^{\text{diag}})^{-1}U_3^\TT YU_2
+=\frac{-v^2}{2}
+\Bigl[(M_H^{\text{diag}})^{-1/2}U_3^\TT YU_2\Bigr]^\TT
+\Bigl[(M_H^{\text{diag}})^{-1/2}U_3^\TT YU_2\Bigr].
+\end{equation}
+From this expression, one can derive Casas--Ibarra parameterization~\cite{Casas:2001sr}, depending on the scenario considered.
+
+\paragraph{Standard parameterization}
+It is useful to assume $M_N$ is positive diagonal and $Y$ is in the charged-lepton diagonal basis:\footnote{One can reach this basis by ATF of $M_N$ and SVD of $Y^e$, which fixes the basis of $L$ and $\bar N$.}
+\begin{equation}
+-\mathcal L\supset \overline{E}_i Y^e_i H^\dagger \PL L_i - \overline N_a Y_{ai}H\PL L_i + \frac{M_{Na}}2\overline{N}_a\PL N^\cc_a - \frac{g_2}{\sqrt2}W_\mu^+\overline{\nu}_i\gamma^\mu e_{{\mathrm L}i}+\text{h.c.}
+\end{equation}
+Then obviously $U_3=1$, while
+\begin{equation}
+-\mathcal L\supset
+ \frac12 \pmat{\nu & n} \pmat{0 & v Y^\TT/\sqrt2 \\ v Y/\sqrt2 & M_N} \pmat{\nu \\ n}
+= \frac12 \pmat{\nu & n} U_1^*\pmat{U_2^*M\w L^{\text{diag}}U_2^\dagger & 0\\0&M_H} U_1^\dagger\pmat{\nu \\ n},
+\end{equation}
+thus $\nu^{\text{mass}}\simeq U_2^\dagger \nu$, and
+\begin{equation}
+-\mathcal L\supset - \frac{g_2}{\sqrt2}W_\mu^+\overline{\nu}_i\gamma^\mu e_{{\mathrm L}i}
++\text{h.c.}
+=- \frac{g_2}{\sqrt2}W_\mu^+\overline{\nu}^{\text{mass}} U_2^\dagger\gamma^\mu e\w L
++\text{h.c.}
+=
+- \frac{g_2}{\sqrt2}W_\mu^-
+\bar e\w L U_2\gamma^\mu {\nu}^{\text{mass}}+\text{h.c.}
+\end{equation}
+means, comparing with (14.25) of \cite{PDG2020}, $U_2=U\w{PMNS}$.
+Therefore, in this basis, the CIP is constructed from
+\begin{equation}
+  M\w L^{\text{diag}}
+=\frac{-v^2}{2}
+\Bigl[(M_H^{\text{diag}})^{-1/2}YU\w{PMNS}\Bigr]^\TT
+\Bigl[(M_H^{\text{diag}})^{-1/2}YU\w{PMNS}\Bigr].
+\end{equation}
+
+\paragraph{Examples}
+For models with three right-handed neutrinos, if we assume all neutrinos are massive, the matrix
+\begin{equation}
+ R:=\frac{-\ii v}{\sqrt2}(M\w H^\text{diag})^{-1/2}U_3^\TT Y U_2(M\w L^\text{diag})^{-1/2}
+\end{equation}
+satisfies $R^\TT R=1$. Conversely, with a matrix $R$ satisfying $R^\TT R=1$, the Yukawa matrix is given by
+\begin{equation}
+ Y = \frac{\ii\sqrt2}{v} U_3^*(M\w H^\text{diag})^{1/2}R(M\w L^\text{diag})^{1/2}U_2^\dagger.
+\label{eq:CIP}
+\end{equation}
+
+For models with two right-handed neutrinos, one neutrino is massless.
+The same parameterization \cref{eq:CIP} works with\footnote{See, e.g., Ref.~\cite{Brdar:2019iem}. Sho also thanks Kai Schmitz for his note.}
+\begin{align}
+ R_{\text{normal hierarchy}}&=\pmat{0 & \cos z & \zeta\sin z \\ 0 & -\sin z & \zeta \cos z},&
+ R_{\text{inverse hierarchy}}=\pmat{\cos z & \zeta\sin z &0 \\ -\sin z & \zeta \cos z & 0},
+\end{align}
+where $z\in\mathbb C$ and $\zeta=\pm1$.
+Note that, in general set-up, $R R^\TT\neq1$.
+
 \end{document}

### 2020-08-07 cosmology.tex
update many things

@@
+% Time-Stamp: <2020-08-12 13:07:42 misho>
+
 \documentclass[CheatSheet]{subfiles}
@@
 \begin{document}
+\summarystyle
+\section{Cosmology}
+\paragraph{FLRW metric} With a scale factor normalized by $a(t_0)=1$,\\
+\begin{minipage}{0.6\textwidth}
+\begin{equation}
+  \dd s^2 = \dd t^2-a^2(t)\left[
+  \frac{\dd r^2}{1-K r^2}+r^2\dd\theta^2+r^2\sin^2\theta\dd\phi^2
+\right]
+\end{equation}\vspace{-2em}
+\begin{align*}
+\text{comoving coordinate~~}& \vc r=(r\sin\theta\cos\phi,r\sin\theta\cos\phi,r\cos\theta),\\
+\text{proper coordinate~~}& \vc x(t)=a(t)\vc r,\\
+\text{comoving distance~~}& \chi_{AB}=\int_{r_A}^{r_B}\frac{\dd r}{\sqrt{1-Kr^2}},\\
+\text{proper distance~~}& d_{AB}(t)=a(t)\chi_{AB}.
+\end{align*}
+\end{minipage}
+\begin{minipage}{0.395\textwidth}
+\hfill\includegraphics[width=0.9\textwidth]{figs/expansion.pdf}
+\end{minipage}
 
-\section{Cosmology}
+\vspace{0.5em}
+
+Ricci tensor and scalar are given by
+\begin{alignat}{3}
+ &R_{00}=\tensor{R}{^0_0}=\frac{3\ddot a}{a}, \qquad&
+ &R_{0i}=R_{i0}=\tensor{R}{^0_i} = \tensor{R}{^i_0}=0, \qquad&
+ &R_{ij}\neq 0,\notag
+\\
+ &\tensor{R}{^i_j} = \delta^i_j\left(\frac{\ddot a}{a} + \frac{2\dot a^2}{a^2} + \frac{2K}{a^2}\right);\qquad&
+ &R = 6\left(\frac{\ddot a}{a} + \frac{\dot a^2}{a^2} + \frac{K}{a^2}\right).
+\end{alignat}
+
+\paragraph{Particle density}For a massless particle, with $L^\pm_n=\pm\mathop{\text{\texttt{PolyLog}}}(n,\pm\ee^{\mu/T})$ and arrows denoting $\mu\to0$,
+\begin{alignat}{4}
+ n\w{MB}    &=\frac{\ee^{\mu/T}}{\pi^2}gT^3 &\quad&\to\frac{1}{\pi^2}gT^3,&\qquad
+ \rho\w{MB} &=3Tn\w{MB}&\quad& \to \frac{3}{\pi^2}gT^4,\\
+ n\w{BE}&=\frac{L^+_3}{\pi^2}gT^3 &&\to \frac{\zeta_3}{\pi^2}gT^3,&
+ \rho\w{BE}&=\frac{3L^+_4}{\pi^2}gT^4&&\to\frac{\pi^2}{30}gT^4,\\
+ n\w{FD}&=\frac{L^-_3}{\pi^2}gT^3&&\to\frac34\frac{\zeta_3}{\pi^2}gT^3,&
+ \rho\w{FD}&=\frac{3L^-_4}{\pi^2}gT^4&&\to\frac78\frac{\pi^2}{30}gT^4,
+\end{alignat}
+
+For massive particle, with $x=m/T$ and $K_n(x)=\mathop{\text{\texttt{BesselK}}}(n,x)$,
+\begin{alignat}{3}
+ n\w{MB}&=g\ee^{\mu/T}\cdot\frac{T^3}{2\pi^2}x^2K_2(x)\qquad
+&\stackrel{x\gg1}\longrightarrow&~
+g\ee^{\mu/T}\frac{T^3}{(2\pi)^{3/2}}x^{3/2}\ee^{-x},\\
+\rho\w{MB}&=\left(3+\frac{xK_1(x)}{K_2(x)}\right)Tn\w{MB}
+&\stackrel{x\gg1}\longrightarrow&~
+\left(m+\frac{3}{2}T+\frac{15T^2}{8m}\right)n\w{MB},\qquad&
+ p\w{MB}&=T n\w{MB}.
+\end{alignat}
+
 \detailstyle
-\subsection{Cosmology formulae}
-Assuming that $\sigma$ depends only on $s=(p_A+p_B)^2$, \TODO{(or always true?)}
+
+\clearpage
+
+\subsection{FLRW metric}
+Two conventions are known for FLRW \GRAY{(\russian{Фр\'{и}дман}-Lema\^{i}tre-Robertson-Walker)} metric:
 \begin{align}
-& \int\ddP{p_A}\,\,\ddP{p_B}\ee^{-E_A/T}\ee^{-E_B/T}\sigma(s)v_{\text{M\o l}}
-\\&=4\int\dd\Pi_A\dd\Pi_B\, \ee^{-(E_A+E_B)/T}\sigma(s)\sqrt{(p_A\cdot p_B)^2-m_A^2m_B^2}
-\\&=
-\frac{1}{32\pi^4}
-\int_{(m_A+m_B)^2}^\infty\dd s
-\int_{\sqrt{s}}^\infty\dd E_+\left[\frac{2\sqrt{E_+^2-s}}{s}\mathop{\lambda^{1/2}}(s,m_A^2,m_B^2)\right]
-\ee^{-E_+/T}\sigma(s)\left[\frac12\mathop{\lambda^{1/2}}(s,m_A^2,m_B^2)\right]
-\\&=
-\frac{T}{32\pi^4}
-\int_{(m_A+m_B)^2}^\infty\dd s\,s^{3/2} \mathop{K_1}\left(\frac{\sqrt{s}}{T}\right)
-\mathop{\lambda}\left(1,\frac{m_A^2}s,\frac{m_B^2}s\right)\sigma(s).
+  \dd s^2
+&= \dd t^2-a^2(t)\left[\frac{\dd r^2}{1- K r^2}+ r^2\dd\theta^2+ r^2\sin^2\theta\dd\phi^2\right]
+&&\text{$[r]=\text{(length)}$, $a$ is unitless with $a(t_0)=1$}
+\\
+&= \dd t^2-R^2(t)\left[\frac{\dd \tilde r^2}{1-\tilde K \tilde r^2}+\tilde r^2\dd\theta^2+\tilde r^2\sin^2\theta\dd\phi^2\right]
+&&\text{$[R]=\text{(length)}$, $\tilde r$ is unitless, $\tilde K=\{0,\pm1\}$}
+\end{align}
+related by a rescaling, $R(t)/a(t)=R(t_0)\equiv R_0$, i.e., 
+$r = \tilde r R_0$ and $K = \tilde K/R_0^2$.
+The curvature radius is given by $6K/a^2$ and a spherical, flat, and hyperspherical universe are respectively given by $K>0$, $K=0$, and $K<0$.
+
+FLRW metric can have several forms. For $\{K>0,K=0,K<0\}$,
+\begin{align}
+\dd s^2
+&= \dd t^2-a^2(t)\left(\frac{\dd r^2}{1- K r^2}+ r^2\dd\Omega\right)
+&\dd\Omega &= \dd\theta^2+\sin^2\theta\dd\phi^2,
+\\&= \dd t^2-a^2(t)\left[\dd\vc r^2 + \frac{K(\vc r\cdot\dd\vc r)^2}{1-K\|\vc r\|^2}\right]
+&\vc r&=(r\sin\theta\cos\phi,\sin\theta\sin\phi,\cos\phi)
+\\&= \dd t^2-\left[\frac{a(t)}{1 + (K/4)\rho^2}\right]^2(\dd\rho^2 + \rho^2\dd\Omega)
+     &\rho&=R_0 \tilde \rho:=\frac{2r}{1+\sqrt{1-Kr^2}}=\frac{2\tilde r R_0}{1+\sqrt{1-\tilde K\tilde r^2}}
+\\&= \dd t^2-\left[\frac{R(t)}{1 + (\tilde K/4){\tilde\rho}^2}\right]^2(\dd{\tilde\rho}^2 + {\tilde\rho}^2\dd\Omega)
+\\
+&=
+\dd t^2-R^2(t)\Bigl(\dd{\tilde\chi}^2 + \{\sin\tilde\chi,\tilde\chi,\sinh\tilde\chi\}^2\dd\Omega\Bigr)
+&\dd\chi&=R_0\dd\tilde\chi = \frac{\dd r'}{\sqrt{1-K r'^2}}
+\quad\text{[comoving distance]}
+\\
+&=a^2(t)\Bigl(
+\dd\eta^2-\dd{\chi}^2 - R_0^2\{\sin\tilde\chi,\tilde\chi,\sinh\tilde\chi\}^2\dd\Omega
+\Bigr)
+&\dd\eta& := \frac{\dd t'}{a(t')}
+\quad\text{[conformal time]}.
+\end{align}
+Explicitly, $\chi$ is given by
+\begin{equation}
+ \chi
+  = \int_0^r\frac{\dd r'}{\sqrt{1-K r'^2}}
+  = \int_0^{\tilde r}\frac{R_0\dd \tilde r'}{\sqrt{1-\tilde K \tilde r'^2}}
+  = R_0\{\sin^{-1}\tilde r,\tilde r,\sinh^{-1}\tilde r\}
+  = 2R_0\left\{\tan^{-1}\frac{\tilde \rho}{2}, \frac{\tilde\rho}{2},\tanh^{-1}\frac{\tilde \rho}{2}\right\}.
+\end{equation}
+
+The Christofffel symbol, Riemann tensor, Ricci tensor, and Ricci scalar are given by
+\begin{align*}
+ \tensor{\Gamma}{^n_i_j} &= \frac{g^{nk}}{2}(g_{jk,i}+g_{ik,j}-g_{ij,k}),
+ &\tensor{R}{_i_j_k^l} &=
+ \tensor{\Gamma}{^l_{jk,i}}
+ -\tensor{\Gamma}{^l_{ik,j}}
+ +\tensor{\Gamma}{^a_j_k}\tensor\Gamma{^l_a_i}
+ -\tensor{\Gamma}{^a_i_k}\tensor\Gamma{^l_a_j}\footnotemark,
+& {R}_{ij}&=\tensor{R}{_i_k_j^k},
+& R &= g^{ij}R_{ij}.
+\end{align*}
+\footnotetext{Overall sign is convention-dependent.}
+
+\subsection{Particle cosmology}
+The particle number density, pressure, and energy density are calculated from distribution functions:
+\begin{alignat}{3}
+ f\w{MB}(\vc k)&=\frac{g}{\ee^{(E-\mu)/T}},\qquad&
+ f\w{BE}(\vc k)&=\frac{g}{\ee^{(E-\mu)/T}-1},\qquad&
+ f\w{FD}(\vc k)&=\frac{g}{\ee^{(E-\mu)/T}+1};
+\end{alignat}
+\begin{alignat}{3}
+ n   &=\int\ddP{k}f(\vc k),\qquad&
+ \rho&=\int\ddP{k}Ef(\vc k),\qquad&
+ p   &=\int\ddP{k}k_zv_z f(\vc k)
+      =\int\ddP{k}\frac{k^2\cos^2\theta}{E} f(\vc k).
+\end{alignat}
+Note the pressure is (momentum)$\times$(flux per time) on a ``wall''; assuming MB, $p=\rho/3$ for $m\ll T$ and $p=T\rho/m$ for $m\gg T$.
+
+A thermal average of a cross section $\sigma(s)$ is schematically given by
+\begin{equation}
+ \vev{\sigma v}_{AB\to12\cdots n}(T)
+=\frac{1}{n_An_B}\int\ddP{k_A}\ddP{k_B}\left(
+f_A f_B\right)\Bigl\{\phi_1 \cdots \phi_n\sigma(s)\Bigr\}
+v_{\text{M\o l}};\qquad \phi_X = \ee^{(E-\mu)/T}f_X/g,
+\end{equation}
+Here, the final state statistical factor $\phi_1\cdots\phi_n$ are subject to the phase space integral of the calculation of $\sigma(s)$.
+They are specifically given by $\phi\w{MB}=1$, $\phi\w{BE}=1+f\w{BE}/g$, and $\phi\w{FD}=1-f\w{FD}/g$.
+Similarly, a thermal averaged decay rate is given by
+\begin{equation}
+ \vev{\Gamma}_{A\to12\cdots n}
+=\frac{1}{n_A}\int\ddP{k_A}
+f_A \left\{\phi_1 \cdots \phi_n
+\frac{m_A}{E_A}\Gamma\right\}.
+\end{equation}
+
+With MB approximation,
+\begin{align}
+ \vev{\sigma v}
+&=
+\frac{g_Ag_B}{n_An_B}
+\ee^{(\mu_A+\mu_B)/T}
+\int\ddP{k_A}\ddP{k_B}
+\ee^{-(E_A+E_B)/T}
+\sigma(s)v_{\text{M\o l}}
+\\
+&=
+\int\frac{\dd s\dd E_+ \dd E_-}{32m_A^2 m_B^2 T^2 K_2(m_A/T)K_2(m_B/T)}
+\ee^{-E_+/T}
+4E_AE_B\sigma(s)v_{\text{M\o l}}\quad(\text{$\times1/2$ if $A=B$)}
+\\
+&=
+\frac{1}{16m_A^2 m_B^2 T K_2(m_A/T)K_2(m_B/T)}
+\int
+\frac{K_1(\sqrt{s}/T) \dd s}{\sqrt{s}}
+\sqrt{\Kallen(s,m_A^2,m_B^2)}
+\cdot2E_A2E_Bv_{\text{M\o l}}\sigma(s)
+\quad(\times1/2),
+\\
+ \vev{\Gamma}
+ &=\frac{K_1(m_A/T)}{K_2(m_A/T)}\Gamma.
 \end{align}

### 2020-08-07 mathematics.tex
update many things

@@
 
+\subsection{Matrix diagonalization}
+In this section, $\mathbb K=\mathbb{R}$ or $\mathbb{C}$ and $\mathbb U_{\mathbb K}^{n}\subset \mathbb K^{n\times n}$ is the set of the unitary matrices.
+
+\paragraph{Diagonalization}
+A matrix $M\in\mathbb K^{n\times n}$ is called diagonalizable if $\exists P$ and $\exists D$ s.t.
+\begin{equation}
+ M=PDP^{-1};\qquad
+ P\in\mathbb{K}^{n\times n},\quad
+ D:\text{diagonal matrix}~(D_{ii}\in\mathbb{K}).
+\end{equation}
+In particular,
+\begin{equation}
+ \text{$M$ is normal} \stackrel{\text{def}}\iff M^\dagger M = M M^\dagger \iff
+ \exists P\in\mathbb{U}_{\mathbb K}^{n} \text{~s.t.~} M=PDP^{-1}.
+\end{equation}
+
+\paragraph{Singular value decomposition}
+Any $M\in\mathbb{K}^{m\times n}$ can be singular-value decomposed as
+\begin{equation}
+ M=UDV^\dagger;\qquad
+ U\in\mathbb{U}_{\mathbb K}^m,\quad
+ V\in\mathbb{U}_{\mathbb K}^n,\quad
+ D: \text{non-negative real diagonal matrix}~(D_{ii}\ge0).
+\end{equation}
+Here, the matrix $U$ ($V$) diagonalizes $MM^\dagger$ ($M^\dagger M$) and $(D_{ii})^2$ are the eigenvalues of $MM^\dagger$ (and $M^\dagger M$).
+
+\paragraph{Autonne-Takagi factorization}
+If $M\in\mathbb{C}^{n\times n}$ is symmetric, it can be decomposed as
+\begin{equation}
+ M=RDR^\TT;\qquad
+ R\in\mathbb{U}_{\mathbb C}^n,\quad
+ D: \text{non-negative real diagonal matrix}~(D_{ii}\ge0).
+\end{equation}
+Real symmetric matrices are normal and thus do not need this factorization; we can apply the above ``diagonalization'' method.
+
 \end{document}

### 2020-09-25 mathematics.tex
add codes for diagonalization

@@
 
-\subsection{Matrix diagonalization}
+\subsection{Matrix diagonalization}\label{app:diagonalization}
 In this section, $\mathbb K=\mathbb{R}$ or $\mathbb{C}$ and $\mathbb U_{\mathbb K}^{n}\subset \mathbb K^{n\times n}$ is the set of the unitary matrices.
@@
 
+The calculation on Mathematica is straightforward for this convention:
+\begin{minted}{wolfram}
+{u, d, v} = SingularValueDecomposition[M]
+\end{minted}
+
 \paragraph{Autonne-Takagi factorization}
@@
 
+Sample Mathematica code to calculate $\{D,R\}$ (with ordering, if specified) is:
+\begin{minted}{wolfram}
+AutonneTakagi[M_, order_ : None] := Module[{v, p, ord, R, D},
+  ord = If[order === None, Range[Length[M]], order];
+  v = Transpose[SingularValueDecomposition[M][[3]]];
+  p = DiagonalMatrix[(#/Abs[#])^(-1/2) & /@ Diagonal[v.M.Transpose[v]]];
+  R = ConjugateTranspose[Reverse[p.v][[ord]]];
+  D = ConjugateTranspose[R].M.Conjugate[R];
+  {D, R} ];
+\end{minted}
 \end{document}

### 2020-09-25 Neutrino.tex
add neutrino.tex

\documentclass[CheatSheet]{subfiles}

\begin{document}
\summarystyle
\section{Neutrino}

(summary page)

\clearpage
\detailstyle

\subsection{Casas--Ibarra parameterization}
\paragraph{Set up}
We start from the neutrino mass matrix
\begin{equation}
-\mathcal L\supset \frac12
 \pmat{\nu & n} \pmat{0 & v Y^\TT/\sqrt2 \\ v Y/\sqrt2 & M_N} \pmat{\nu \\ n}
= \frac12\pmat{\nu_i & n_a} \pmat{0 & v (Y^\TT)_{ib}/\sqrt2 \\ v Y_{aj}/\sqrt2 & M_{Nab}} \pmat{\nu_j \\ n_b}
\end{equation}
(and its Hermitian conjugate), which can be obtained, e.g., from
\begin{equation}
 \mathcal L \supset \overline{N}Y H L - \frac12\overline{N}M_N N^\cc + \text{h.c.},
\qquad
\vev{H}=\pmat{0\\v/\sqrt2}\text{~with~}v\sim246\GeV,
\end{equation}
where the four spinors should be decomposed into Weyl spinors as
$L^1=\spmat{\nu\\0}$,
$N=\spmat{0\\\bar n}$.

\paragraph{Generalized CIP}
The mass matrix can be AT-factorized to yield 6 neutrino masses.
This ATF can be separated into two steps:
\begin{equation}
 \pmat{0 & v Y^\TT/\sqrt2 \\ v Y/\sqrt2 & M_N}
\longrightarrow
\pmat{M\w L & 0 \\ 0 & M\w H}
\stackrel{\text{ATF}}\longrightarrow
\pmat{M\w L^{\text{diag}} & 0 \\ 0 & M\w H^{\text{diag}}},
\end{equation}
where $M\w L$ and $M\w H$ are complex symmetric matrices.
Explicitly, with unitary matrices  $U_1$, $U_2$, and $U_3$,
\begin{align}
\pmat{M\w L & 0 \\ 0 & M\w H}
&=
 U_1^\TT \pmat{0 & v Y^\TT/\sqrt2 \\ v Y/\sqrt2 & M_N} U_1,
&M\w L^\text{diag} &= U_2^\TT M\w L U_2,
&M\w H^\text{diag} &= U_3^\TT M\w H U_3.
\end{align}
The first equation is calculated as an expansion in $v/M_N$:
\begin{align}
 U_1  &\simeq \pmat{1 & v Y^\dagger (M_N^*)^{-1}/\sqrt2 \\ -v M_N^{-1}Y/\sqrt2 & 1},
&M\w L&\simeq -\frac{v^2}{2}Y^\TT M_N^{-1} Y,
&M\w H&\simeq M_N.
\end{align}
These expressions lead to
\begin{equation}
  M\w L^{\text{diag}}
\simeq \frac{-v^2}{2} U_2^\TT Y^\TT U_3(M_H^{\text{diag}})^{-1}U_3^\TT YU_2
=\frac{-v^2}{2}
\Bigl[(M_H^{\text{diag}})^{-1/2}U_3^\TT YU_2\Bigr]^\TT
\Bigl[(M_H^{\text{diag}})^{-1/2}U_3^\TT YU_2\Bigr].
\end{equation}
From this expression, one can derive Casas--Ibarra parameterization~\cite{Casas:2001sr}, depending on the scenario considered.

\paragraph{Standard parameterization}
It is useful to assume $M_N$ is positive diagonal and $Y$ is in the charged-lepton diagonal basis:\footnote{One can reach this basis by ATF of $M_N$ and SVD of $Y^e$, which fixes the basis of $L$ and $\bar N$.}
\begin{equation}
-\mathcal L\supset \overline{E}_i Y^e_i H^\dagger \PL L_i - \overline N_a Y_{ai}H\PL L_i + \frac{M_{Na}}2\overline{N}_a\PL N^\cc_a - \frac{g_2}{\sqrt2}W_\mu^+\overline{\nu}_i\gamma^\mu e_{{\mathrm L}i}+\text{h.c.}
\end{equation}
Then obviously $U_3=1$, while
\begin{equation}
-\mathcal L\supset
 \frac12 \pmat{\nu & n} \pmat{0 & v Y^\TT/\sqrt2 \\ v Y/\sqrt2 & M_N} \pmat{\nu \\ n}
= \frac12 \pmat{\nu & n} U_1^*\pmat{U_2^*M\w L^{\text{diag}}U_2^\dagger & 0\\0&M_H} U_1^\dagger\pmat{\nu \\ n},
\end{equation}
thus $\nu^{\text{mass}}\simeq U_2^\dagger \nu$, and
\begin{equation}
-\mathcal L\supset - \frac{g_2}{\sqrt2}W_\mu^+\overline{\nu}_i\gamma^\mu e_{{\mathrm L}i}
+\text{h.c.}
=- \frac{g_2}{\sqrt2}W_\mu^+\overline{\nu}^{\text{mass}} U_2^\dagger\gamma^\mu e\w L
+\text{h.c.}
=
- \frac{g_2}{\sqrt2}W_\mu^-
\bar e\w L U_2\gamma^\mu {\nu}^{\text{mass}}+\text{h.c.}
\end{equation}
means, comparing with (14.25) of \cite{PDG2020}, $U_2=U\w{PMNS}$.
Therefore, in this basis, the CIP is constructed from
\begin{equation}
  M\w L^{\text{diag}}
=\frac{-v^2}{2}
\Bigl[(M_H^{\text{diag}})^{-1/2}YU\w{PMNS}\Bigr]^\TT
\Bigl[(M_H^{\text{diag}})^{-1/2}YU\w{PMNS}\Bigr].
\end{equation}

\paragraph{Examples}
For models with three right-handed neutrinos, if we assume all neutrinos are massive, the matrix
\begin{equation}
 R:=\frac{-\ii v}{\sqrt2}(M\w H^\text{diag})^{-1/2}U_3^\TT Y U_2(M\w L^\text{diag})^{-1/2}
\end{equation}
satisfies $R^\TT R=1$. Conversely, with a matrix $R$ satisfying $R^\TT R=1$, the Yukawa matrix is given by
\begin{equation}
 Y = \frac{\ii\sqrt2}{v} U_3^*(M\w H^\text{diag})^{1/2}R(M\w L^\text{diag})^{1/2}U_2^\dagger.
\label{eq:CIP}
\end{equation}

For models with two right-handed neutrinos, one neutrino is massless.
The same parameterization \cref{eq:CIP} works with\footnote{See, e.g., Ref.~\cite{Brdar:2019iem}. Sho also thanks Kai Schmitz for his note.}
\begin{align}
 R_{\text{normal hierarchy}}&=\pmat{0 & \cos z & \zeta\sin z \\ 0 & -\sin z & \zeta \cos z},&
 R_{\text{inverse hierarchy}}=\pmat{\cos z & \zeta\sin z &0 \\ -\sin z & \zeta \cos z & 0},
\end{align}
where $z\in\mathbb C$ and $\zeta=\pm1$.
Note that, in general set-up, $R R^\TT\neq1$.

\end{document}

### 2020-09-25 mathematics.tex
debug AT-diagonalize code

@@
 \begin{minted}{wolfram}
-AutonneTakagi[M_, order_ : None] := Module[{v, p, ord, R, D},
+AutonneTakagi[M_, order_ : None] := Module[{v0, v, p, ord, R, D},
   ord = If[order === None, Range[Length[M]], order];
-  v = Transpose[SingularValueDecomposition[M][[3]]];
-  p = DiagonalMatrix[(#/Abs[#])^(-1/2) & /@ Diagonal[v.M.Transpose[v]]];
-  R = ConjugateTranspose[Reverse[p.v][[ord]]];
+  v0 = Eigenvectors[Conjugate[M].M];
+  v = Eigenvectors[v0.M.Transpose[v0]].v0; (*resolve degenerate eigenvalues*)
+  p = DiagonalMatrix[If[Abs[#] > 0, (#/Abs[#])^(-1/2), 1] & /@ Diagonal[v.M.Transpose[v]]];
+  R = ConjugateTranspose[Reverse[p.v][[ord]] // Orthogonalize];
   D = ConjugateTranspose[R].M.Conjugate[R];
-  {D, R} ];
+  {D, R}];
 \end{minted}

### 2020-09-28 mathematics.tex
add code for U3

@@
 \pmat{\ee^{\ii c}&&\\&\ee^{\ii d}&\\&&\ee^{\ii e}}
+\label{eq:GeneralUnitary33}
 \end{align}
 (see, e.g., Ref.~\cite{Rasin:1997pn}).
+
+\begin{minted}{wolfram}
+U3 = Dot[
+  DiagonalMatrix[Exp[I {0, a, b}]],
+  RotationMatrix[\[Theta]23, {-1, 0, 0}],
+  DiagonalMatrix[Exp[I {0,0,+\[Delta]}]],
+  RotationMatrix[\[Theta]13, {0, 1, 0}],
+  DiagonalMatrix[Exp[I {0,0,-\[Delta]}]],
+  RotationMatrix[\[Theta]12, {0, 0, -1}],
+  DiagonalMatrix[Exp[I {c, d, e}]]
+]
+\end{minted}
 

### 2020-10-02 Neutrino.tex
rewrite CIP

@@
 \documentclass[CheatSheet]{subfiles}
+
+\newcommand\MD[1][]{M_{\mathrm{D}#1}}
+\newcommand\MN[1][]{M_{N#1}}
 
@@
 
+\detailstyle
 \clearpage
-\detailstyle
+\subsection{Convention}
+We extend the SM Yukawa \eqref{eq:SMyukawa} to include the neutrino mass terms:
+\begin{align}
+   \mathcal L\w{Y+\nu}&=
+  \overline{U}\Yu H \PL Q
+- \overline{D}\Yd H^\dagger \PL Q
++ \overline{N}\Yn H \PL L
+- \overline{E}\Ye H^\dagger \PL L
+- \frac{1}{2}\overline{N}\MN N^\cc
++ \text{h.c.}
+\\&=
+- \overline{Q^a}\Yu^\dagger \epsilon^{ab}H^{b*} \PR U
+- \overline{Q^a}\Yd^\dagger H^{a} \PR D
+- \overline{L^a}\Yn^\dagger \epsilon^{ab}H^{b*} \PR N
+- \overline{L^a}\Ye^\dagger H^{a} \PR E
+- \frac{1}{2}\overline{N}\MN N^\cc + \text{h.c.},
+\end{align}
+where $\MN$, a \emph{complex symmetric} Majorana mass matrix, may be absent.
+
+As we explicitly describe in \cref{sec:PMNS}, we use the standard convention for the PMNS matrix \GRAY{(Pontecorvo--\japanese{牧}--\japanese{中川}--\japanese{坂田})}:
+\begin{equation}
+ \ket{\nu_\alpha^{\text{flavor}}} = [U\w{PMNS}]_{\alpha i}^*\ket{\nu_i^{\text{mass}}},
+\qquad
+ \nu^\text{flavor} = U\w{PMNS}\nu^{\text{mass}}, \qquad[\alpha=e,\mu,\tau,\mathrm{r}_1,\mathrm{r}_2,\dots]\label{eq:PMNS-defining}
+\end{equation}
+where the ``flavor basis'' is defined by the charged lepton mass basis for first three elements, while in most cases by the basis that diagonalizes Majorana mass term for the rest.
+
+\paragraph{Dirac neutrino}
+If $\MN$ is absent, neutrinos become Dirac fermions and the discussion goes parallel to the CKM matrix:
+\begin{equation}
+ \Yn = U_n Y_n\wdiag V_n^\dagger,\qquad
+\{\nuL, \eL, \overline N, \overline E\}^{\text{mass basis}} = \{V_n^\dagger L^1, V_e^\dagger L^2, \overline N U_n, \overline E U_e\}
+\end{equation}
+and
+\begin{align}
+ \mathcal L&\supset\overline{L}\ii\gamma^\mu(-\ii g_2 W_\mu)\PL L
+\supset
+\frac{g_2}{\sqrt2}\left[
+\overline{L^1} \slashed W^+\PL L^2
++\overline{L^2}\slashed W^-\PL L^1
+\right]
+\\&=
+\frac{g_2}{\sqrt2}\left[
+\overline{(L^1)^{\text{mass}}} V_n^\dagger V_e\slashed W^+\PL (L^2)^{\text{mass}}
++\overline{(L^2)^{\text{mass}}}V_e^\dagger V_n\slashed W^-\PL(L^1)^{\text{mass}}
+\right]
+\\&=
+\frac{g_2}{\sqrt2}\left[
+\overline{\nuL^{\text{mass}}} [U\w{PMNS}]^\dagger\slashed W^+\eL^{\text{mass}}
++\overline{\eL^{\text{mass}}}[U\w{PMNS}] \slashed W^-\nuL^{\text{mass}}
+\right].
+\end{align}
+The mass eigenstate Dirac field is given by
+\begin{equation}
+ \nu^{\text{mass}} = \pmat{\nuL^{\text{mass}}\\N^{\text{mass}}} = 
+\nuL^{\text{mass}}+N^{\text{mass}}=
+V_n^\dagger L^1 + U_n^\dagger N
+\qquad
+ \Bigl(L^1 = \PL V_n \nu^{\text{mass}},~
+ N = \PR U_n \nu^{\text{mass}}\Bigr).
+\end{equation}
+The PMNS matrix is given in \emph{the opposite manner} to the CKM matrix:
+\begin{equation}
+ U\w{PMNS}^{\text{(Dirac)}}=V_e^\dagger V_n.
+\end{equation}
+In general, $\Yn\in\mathbb C^{n\times3}$, $V_n\in\mathbb U_{\mathbb C}^3$, $U_n\in\mathbb U_{\mathbb C}^n$, and the Dirac-PMNS matrix is also a $3\times 3$ unitary matrix.
+Models with $n<3$ yields $3-n$ massless left-handed neutrinos, while $n>3$ results in $n-3$ massless right-handed neutrinos.
+
+\paragraph{Majorana neutrino}
+Models with $\MN\neq0$ generate so-called Majorana neutrino masses.
+If $n<3$, the model has $3-n$ massless neutrinos and $2n$ massive neutrinos, while all neutrinos are massive if $n\ge 3$.
+
+Using Weyl spinor,
+\begin{align}
+& L^1 = \pmat{\nuL\\0},\quad
+ N   = \pmat{0\\\bnR},\quad
+ \MD := \frac{v}{\sqrt2}Y_n,\quad
+ \MD\wdiag := U_n \MD Y_n^\dagger,\\
+&\label{eq:MajoranaNuLag}
+ \mathcal L\w{Y+\nu} 
+\supset
+ -\frac{v}{\sqrt2}\overline{N}Y_n \PL L^1 - \frac{1}{2}\overline{N}\MN N^\cc + \text{h.c.}
+=
+ -\frac{v}{\sqrt2}\nR Y_n \nuL - \frac{1}{2}\nR \MN\nR + \text{h.c.}\\
+&\phantom{ \mathcal L\w{Y+\nu}}
+=-\frac12\pmat{\nuL & \nR}\pmat{0 & \MD^\TT \\ \MD & \MN} \pmat{\nuL \\ \nR} + \text{h.c.} =: -\frac12\tilde\nu^\TT \tilde M\tilde \nu+\text{h.c.}
+\end{align}
+As $\tilde M$ is a complex symmetric $(3+n)\times(3+n)$ matrix, it can be AT-diagonalized:
+\begin{equation}
+  \tilde M = \tilde R \tilde M\wdiag \tilde R^\TT;
+\qquad
+-\mathcal L\w{Y+\nu}\supset
+\frac12\tilde \nu^\TT \tilde M\tilde\nu
+=\frac12(\tilde\nu^{\text{mass}})^\TT \tilde M\wdiag \tilde\nu^{\text{mass}};\qquad \tilde\nu^{\text{mass}} =\tilde R^\TT\tilde\nu
+\end{equation}
+and $\tilde M$ gives the neutrino masses.
+The neutrino mixing is then given by
+\begin{align}
+ \pmat{\nuL\\\nR} = \tilde R^*\pmat{\nu_{1\TO3} \\ \nu_{4\TO}}
+\end{align}
+and find that the PMNS matrix is now extended to a $3\times n$ matrix $V_e^\dagger [\tilde R^*]\w{upper}$.
+
+Usually the discussion should be start from the basis in which $Y_e$ and $M_N$ are positive diagonal with increasing diagonal elements (``charged lepton mass basis'' combined with ``Majorana mass basis'').
+Then
+\begin{equation}
+\tilde M = \pmat{0 & \MD^\TT \\ \MD & \MN\wdiag},
+\quad
+\pmat{\nuL\\\nR}
+=
+\pmat{\begin{smallmatrix}\nu_e\\\nu_\mu\\\nu_\tau\end{smallmatrix}\\\nu_{\text{sterile};i}}
+=\tilde R^*\pmat{\nu_{1\TO3}\\\nu_{4\TO}}.
+\end{equation}
+The PMNS matrix appears as a submatrices, which is no longer unitary:
+\begin{equation}
+\tilde R^*
+=:
+\pmat{ R_{11}^* & R_{12}^* \\ R_{21}^* & R_{22}^*}
+=:
+\pmat{U\w{PMNS}^{\text{Majorana}} & U_{\text{active-heavy}}\\
+U_{\text{sterile-light}} & U_{\text{sterile-heavy}}
+}.
+\end{equation}
+
+\paragraph{Connection of the above two formalism}
+If we apply AT-diagonalization to the Dirac case (with $n=3$), we obtain
+\begin{align}
+\tilde M = \frac12\pmat{0 & (U_n\MD\wdiag V_n^\dagger)^\TT \\ U_n\MD\wdiag V_n^\dagger & 0},\quad
+  \tilde R=\pmat{V_n^* & \ii V_n^* \\ U_n & -\ii U_n},\quad
+  \tilde M\wdiag=\pmat{\MD\wdiag & 0 \\ 0 & \MD\wdiag}
+\end{align}
+and find that three pairs of degenerate Weyl fermions form three Dirac neutrinos.
+
+\paragraph{PMNS matrix}\label{sec:PMNS}
+In the Dirac neutrino models the PMNS matrix is unitary, which we parameterize
+\begin{equation}
+\begin{split}
+   U\w{PMNS}^{\text{Dirac}}&=V_e^\dagger V_n
+ =\pmat{
+ U_{e1} & U_{e2} & U_{e3}\\
+ U_{\mu1} & U_{\mu2} & U_{\mu3}\\
+ U_{\tau1} & U_{\tau2} & U_{\tau3}
+ }
+ =
+ %
+ \pmat{1&&\\&\co{23}&\si{23}\\&-\si{23}&\co{23}}
+ \pmat{\co{13}&&\si{13}\ee^{-\ii\delta\w{CP}}\\&1&\\-\si{13}\ee^{\ii\delta\w{CP}}&&\co{13}}
+ \pmat{\co{12}&\si{12}&\\-\si{12}&\co{12}&\\&&1}
+ %
+ \\&=
+ \pmat{
+ \co{12} \co{13} & \si{12} \co{13} & \si{13} \ee^{-\ii\delta\w{CP}}\\
+ -\si{12} \co{23} - \co{12} \si{23} \si{13} \ee^{\ii\delta\w{CP}}& \co{12} \co{23} - \si{12}\si{23}\si{13} \ee^{\ii\delta\w{CP}}& \si{23}\co{13}\\
+  \si{12}\si{23} - \co{12} \co{23} \si{13}\ee^{\ii\delta\w{CP}} & -\co{12}\si{23}-\si{12}\co{23}\si{13}\ee^{\ii\delta\w{CP}} & \co{23} \co{13}
+},
+\end{split}
+\end{equation}
+where $\theta_{ij}\in[0,\pi/2]$ and $\delta\w{CP}\in[0,2\pi]$ as shown in \cref{eq:GeneralUnitary33} and the five phases are removed as done in \cref{sec:SM-CKM}.
+Note that the discussion in \cref{sec:SM-FFdual} holds and $\Theta_W W\tilde W$ term is rotated away.
+
+If the Majorana mass terms are present, $U\w{PMNS}^{\text{(Majorana)}}$, which is defined by \cref{eq:PMNS-defining}, is no longer unitary.
+However, if $\MN\gg\MD$, it is approximately unitary:
+\begin{equation}
+   U\w{PMNS}^{\text{Majorana}}=
+\pmat{
+ U_{e1} & U_{e2} & U_{e3}\\
+ U_{\mu1} & U_{\mu2} & U_{\mu3}\\
+ U_{\tau1} & U_{\tau2} & U_{\tau3}
+ }
+\approx
+U\w{PMNS}^{\text{Dirac}}
+\pmat{\ee^{\ii\eta_1} \\ & \ee^{\ii\eta_2}\\&&1}.
+\end{equation}
+Here additional two phases are introduced because we can no longer rotate $\nR$; three among $(a,b,c,d,e)$ in \cref{eq:GeneralUnitary33} are removed by rotating $L_i$, and two remains.
+
+Current experiments measure the value of the matrix and the above parameterization still works well (cf.~Ref.~\cite{NUFIT}).
+Thus we hereafter identify the Dirac PMNS matrix as $U\w{PMNS}$:
+\begin{align}
+ &U\w{PMNS}:=U\w{PMNS}^{\text{(Dirac)}}=
+ \pmat{
+ \co{12} \co{13} & \si{12} \co{13} & \si{13} \ee^{-\ii\delta\w{CP}}\\
+ -\si{12} \co{23} - \co{12} \si{23} \si{13} \ee^{\ii\delta\w{CP}}& \co{12} \co{23} - \si{12}\si{23}\si{13} \ee^{\ii\delta\w{CP}}& \si{23}\co{13}\\
+  \si{12}\si{23} - \co{12} \co{23} \si{13}\ee^{\ii\delta\w{CP}} & -\co{12}\si{23}-\si{12}\co{23}\si{13}\ee^{\ii\delta\w{CP}} & \co{23} \co{13}
+};\\
+ &U\w{PMNS}^{\text{(Majorana)}} = U\w{PMNS}\pmat{\ee^{\ii\eta_1} \\ & \ee^{\ii\eta_2}\\&&1}
++\Order\left(\frac{\MD}{\MN}\right).
+\end{align}
+Accordingly, the components $U_{\alpha i}$ is in general slightly different from $(\alpha,i)$ component of $U\w{PMNS}$, e.g.,
+\begin{equation*}
+ U_{e1}\approx \co{12}\co{13}
+\end{equation*}
+with the exactness recovered in the Dirac case.
+
+\paragraph{PDG and NuFIT convention} The convention agrees with PDG \cite[\S14]{PDG2020} and NuFIT~\cite[v5.0]{NUFIT}.
+Compared with PDG,\footnote{Sho thinks Eq.~(14.9) of PDG2020 lacks 1/2 in the right-most term.}
+\begin{align*}
+& \mathcal L
+\supset
+ -\CONV{\bar\nu_s} \CONV{M_D} \nu_L
+ -\frac12\CONV{\bar\nu_s} \CONV{M_N} \CONV{\nu^c_s}
+ -\overline{L^2}\CONV{M_l}\PR E\quad\text{(14.6)+(14.27)},\\
+&
+\Bigl\{\CONV{(V^{\nu})^\TT \spmat{0&M_D^\TT \\ M_D &M_N} V^\nu}\text{~or~}\CONV{V_R^{\nu\dagger}M_D V^\nu}\Bigr\}=\diag(m_i)\quad
+\text{(Majorana; 14.9)+(Dirac; 14.15)},\\
+&\nu_{L}=\PL\CONV{V^\nu}\nu^{\text{mass}}\quad\text{(14.14)\&(14.18)},\qquad
+\CONV{V^{l\dagger}M_l V^l_R}=\diag(m_e,m_\mu,m_\tau)\quad\text{(14.31)}.
+\end{align*}
+These leads to $\CONV{V^\nu_R}=U_n$ and $\CONV{V^\nu}=V_n$ in Dirac case, $\CONV{M_D} = \Yn v/\sqrt2=\MD$, $\CONV{M_N}=M_N$, and $\CONV{V^\nu}=R^*$ in Majorana case, and $\CONV{V^{l}}=V_e$ and $\CONV{V_R^l}=U_e$ (note $\CONV{M_l}=Y_e^\dagger v/\sqrt{2}$).
+So
+\begin{equation*}
+ \CONV{U_{ij}} = \text{(diagonal phases)}\times\CONV{V^{l\dagger}}\CONV{V^{\nu}}\times\text{(diagonal phases)}=\text{(phases)}V_e^\dagger \{V_\nu\text{~or~}R^*\}\text{(phases)}= U\w{PMNS}^{\text{(Dirac/Majorana)}}
+\end{equation*}
 
 \subsection{Casas--Ibarra parameterization}
-\paragraph{Set up}
+\paragraph{General basis}
 We start from the neutrino mass matrix
@@
 -\mathcal L\supset \frac12
- \pmat{\nu & n} \pmat{0 & v Y^\TT/\sqrt2 \\ v Y/\sqrt2 & M_N} \pmat{\nu \\ n}
-= \frac12\pmat{\nu_i & n_a} \pmat{0 & v (Y^\TT)_{ib}/\sqrt2 \\ v Y_{aj}/\sqrt2 & M_{Nab}} \pmat{\nu_j \\ n_b}
+ \pmat{\nuL & \nR} \pmat{0 & \MD^\TT \\ \MD & \MN} \pmat{\nuL \\ \nR}+\text{h.c.}
+= \frac12\pmat{\nuL[i] & \nR[a]}
+\pmat{0_{ij} & (\MD^\TT)_{ib} \\ \MD[aj] & \MN[ab]}
+\pmat{\nuL[j] \\ \nR[b]}+\text{h.c.}
 \end{equation}
-(and its Hermitian conjugate), which can be obtained, e.g., from
+The AT-factorization can be separated into two steps:
 \begin{equation}
- \mathcal L \supset \overline{N}Y H L - \frac12\overline{N}M_N N^\cc + \text{h.c.},
-\qquad
-\vev{H}=\pmat{0\\v/\sqrt2}\text{~with~}v\sim246\GeV,
-\end{equation}
-where the four spinors should be decomposed into Weyl spinors as
-$L^1=\spmat{\nu\\0}$,
-$N=\spmat{0\\\bar n}$.
-
-\paragraph{Generalized CIP}
-The mass matrix can be AT-factorized to yield 6 neutrino masses.
-This ATF can be separated into two steps:
-\begin{equation}
- \pmat{0 & v Y^\TT/\sqrt2 \\ v Y/\sqrt2 & M_N}
+ \tilde M=\pmat{0 & \MD^\TT \\ \MD & \MN}
 \longrightarrow
@@
 \stackrel{\text{ATF}}\longrightarrow
-\pmat{M\w L^{\text{diag}} & 0 \\ 0 & M\w H^{\text{diag}}},
+\pmat{M\w L\wdiag & 0 \\ 0 & M\w H\wdiag}=\tilde M\wdiag,
 \end{equation}
-where $M\w L$ and $M\w H$ are complex symmetric matrices.
-Explicitly, with unitary matrices  $U_1$, $U_2$, and $U_3$,
+where intermediate matrices $M\w L$ and $M\w H$ are complex symmetric.
+This is expressed with unitary matrices $U$, $U_1$, and $U_2$:
+\begin{equation}
+ \tilde M =
+U \pmat{M\w L & 0 \\ 0 & M\w H} U_1^\TT =
+U \pmat{U_1 M\w L\wdiag U_1^\TT & 0 \\ 0 & U_2 M\w H\wdiag U_2^\TT} U^\TT,
+\end{equation}
+following the convention in \cref{eq:ATF}.
+The first equality is calculated as an expansion in $\MD/\MN$ once we assume the see-saw mechanism:
 \begin{align}
-\pmat{M\w L & 0 \\ 0 & M\w H}
-&=
- U_1^\TT \pmat{0 & v Y^\TT/\sqrt2 \\ v Y/\sqrt2 & M_N} U_1,
-&M\w L^\text{diag} &= U_2^\TT M\w L U_2,
-&M\w H^\text{diag} &= U_3^\TT M\w H U_3.
+ U&\simeq \pmat{1 & \MD^\TT\MN^{-1} \\ -\MN^{-1}\MD^* & 1}
+ - \frac12\pmat{\MD^\TT\MN^{-2}\MD^* & 0 \\ 0 & \MN^{-1}\MD^*\MD^\TT \MN^{-1}} + \cdots,\\
+ M\w L&\simeq -\MD^\TT \MN^{-1} \MD+\cdots,\\
+ M\w H&\simeq \MN + \frac12 (\MD\MD^\dagger\MN^{-1}+\MN^{-1}\MD^*\MD^\TT)+\cdots,
 \end{align}
-The first equation is calculated as an expansion in $v/M_N$:
-\begin{align}
- U_1  &\simeq \pmat{1 & v Y^\dagger (M_N^*)^{-1}/\sqrt2 \\ -v M_N^{-1}Y/\sqrt2 & 1},
-&M\w L&\simeq -\frac{v^2}{2}Y^\TT M_N^{-1} Y,
-&M\w H&\simeq M_N.
-\end{align}
-These expressions lead to
+At the leading order,
+\begin{equation*}
+ U_1 M\w L\wdiag U_1^\TT = [U^\dagger \tilde M U^*]_{\text{upper left}} \approx -\MD^\TT \MN^{-1}\MD
+\approx -\MD^\TT M\w H^{-1}\MD = -\MD^\TT (U_2 M\w H\wdiag U_2^\TT)^{-1}\MD,
+\end{equation*}
+or
+\begin{equation*}
+ -M\w L\wdiag \approx U_1^\dagger \MD^\TT U_2^*(M\w H\wdiag)^{-1}U_2^\dagger\MD U_1^*.
+\end{equation*}
+This is decomposed to
 \begin{equation}
-  M\w L^{\text{diag}}
-\simeq \frac{-v^2}{2} U_2^\TT Y^\TT U_3(M_H^{\text{diag}})^{-1}U_3^\TT YU_2
-=\frac{-v^2}{2}
-\Bigl[(M_H^{\text{diag}})^{-1/2}U_3^\TT YU_2\Bigr]^\TT
-\Bigl[(M_H^{\text{diag}})^{-1/2}U_3^\TT YU_2\Bigr].
+ [\ii M\w L\wdiag]^{1/2}[\ii M\w L\wdiag]^{1/2} = 
+[(M\w H\wdiag)^{-1/2}U_2^\dagger\MD U_1^*]^\TT
+[(M\w H\wdiag)^{-1/2}U_2^\dagger\MD U_1^*].
 \end{equation}
-From this expression, one can derive Casas--Ibarra parameterization~\cite{Casas:2001sr}, depending on the scenario considered.
+This is the master equation for Casas-Ibarra parameterization~\cite{Casas:2001sr}.
 
-\paragraph{Standard parameterization}
-It is useful to assume $M_N$ is positive diagonal and $Y$ is in the charged-lepton diagonal basis:\footnote{One can reach this basis by ATF of $M_N$ and SVD of $Y^e$, which fixes the basis of $L$ and $\bar N$.}
+\paragraph{``Standard'' parameterization}
+Let us assume that we started from the above-discussed $(Y_e,\MN)$-diagonal basis.
+Then, noting
 \begin{equation}
--\mathcal L\supset \overline{E}_i Y^e_i H^\dagger \PL L_i - \overline N_a Y_{ai}H\PL L_i + \frac{M_{Na}}2\overline{N}_a\PL N^\cc_a - \frac{g_2}{\sqrt2}W_\mu^+\overline{\nu}_i\gamma^\mu e_{{\mathrm L}i}+\text{h.c.}
+ \tilde R=U\pmat{U_1\\&U_2}\approx \pmat{U_1 & \MD^\TT\MN^{-1}U_2 \\ -\MN^{-1}\MD^* U_1 & U_2},
 \end{equation}
-Then obviously $U_3=1$, while
+we can identify $U_1^*\approx U^{\text{(Majorana)}}\w{PMNS}$ and $U_2\approx 1$; the master equation now becomes
 \begin{equation}
--\mathcal L\supset
- \frac12 \pmat{\nu & n} \pmat{0 & v Y^\TT/\sqrt2 \\ v Y/\sqrt2 & M_N} \pmat{\nu \\ n}
-= \frac12 \pmat{\nu & n} U_1^*\pmat{U_2^*M\w L^{\text{diag}}U_2^\dagger & 0\\0&M_H} U_1^\dagger\pmat{\nu \\ n},
+ [\ii M\w L\wdiag]^{1/2}[\ii M\w L\wdiag]^{1/2} =
+[(M\w H\wdiag)^{-1/2}\MD U^{\text{(Majorana)}}\w{PMNS}]^\TT
+[(M\w H\wdiag)^{-1/2}\MD U^{\text{(Majorana)}}\w{PMNS}].
 \end{equation}
-thus $\nu^{\text{mass}}\simeq U_2^\dagger \nu$, and
+We will use this parameterization below.
+
+\paragraph{Example: three right-handed neutrinos}
+Let us assume all the neutrinos are massive thanks to three right-handed neutrinos. Then $M\w L\wdiag$ is invertible and
 \begin{equation}
--\mathcal L\supset - \frac{g_2}{\sqrt2}W_\mu^+\overline{\nu}_i\gamma^\mu e_{{\mathrm L}i}
-+\text{h.c.}
-=- \frac{g_2}{\sqrt2}W_\mu^+\overline{\nu}^{\text{mass}} U_2^\dagger\gamma^\mu e\w L
-+\text{h.c.}
-=
-- \frac{g_2}{\sqrt2}W_\mu^-
-\bar e\w L U_2\gamma^\mu {\nu}^{\text{mass}}+\text{h.c.}
+ R:=-\ii (M\w H\wdiag)^{-1/2}\MD  U^{\text{(Majorana)}}\w{PMNS}  (M\w L^\text{diag})^{-1/2}
+\quad\Longrightarrow\quad R^\TT R=1.
 \end{equation}
-means, comparing with (14.25) of \cite{PDG2020}, $U_2=U\w{PMNS}$.
-Therefore, in this basis, the CIP is constructed from
+Conversely, with a matrix $R$ satisfying $R^\TT R=1$, the ``Yukawa matrix'' is given by
 \begin{equation}
-  M\w L^{\text{diag}}
-=\frac{-v^2}{2}
-\Bigl[(M_H^{\text{diag}})^{-1/2}YU\w{PMNS}\Bigr]^\TT
-\Bigl[(M_H^{\text{diag}})^{-1/2}YU\w{PMNS}\Bigr].
+ \MD = \ii \sqrt{M\w H^\text{diag}}R\sqrt{M\w L^\text{diag}}(U^{\text{(Majorana)}}\w{PMNS} )^\dagger.
 \end{equation}
+Now we successfully parameterized $Y_n$ by a ``complex orthogonal'' matrix $R$.
 
-\paragraph{Examples}
-For models with three right-handed neutrinos, if we assume all neutrinos are massive, the matrix
+Such $R$ is given by\footnote{For $w\in\mathbb C$, $\sin z_1=w$ and $\cos z_2=w$ always have solutions $z_{1,2}\in\mathbb C$.
+Meanwhile, $\tan z=w$ has no solution if and only if $w=\pm \ii$.
+Then, using this fact, one first expresses $R_{i3}$ components by $\zeta\w{A,B,C}=\pm1$ and $\theta\w{A,B}\in\mathbb C$, restricting $0\le\Re\theta\w{A,B}\le\pi/2$ ($\Leftrightarrow\Re{\sin{\theta}}\ge0\land\Re\cos\theta\ge0$), and then get an expression of $R$ with three angles and six signs.
+Five signs are absorbed by enlarging $\Re\theta$ and one sign remains, which is $\zeta$.}
 \begin{equation}
- R:=\frac{-\ii v}{\sqrt2}(M\w H^\text{diag})^{-1/2}U_3^\TT Y U_2(M\w L^\text{diag})^{-1/2}
+ R=\pmat{
+ \co{12} \co{13} & \si{12} \co{13} & \si{13} \\
+ -\zeta \si{12} \co{23}-\co{12} \si{23} \si{13} & \zeta \co{12} \co{23} -\si{12} \si{23} \si{13} & \si{23} \co{13} \\
+  \zeta \si{12} \si{23}-\co{12} \co{23} \si{13} & -\zeta\co{12} \si{23} -\si{12} \co{23} \si{13} & \co{23} \co{13}
+},
 \end{equation}
-satisfies $R^\TT R=1$. Conversely, with a matrix $R$ satisfying $R^\TT R=1$, the Yukawa matrix is given by
+where $\co{12}\equiv \cos{\theta_{12}}$ etc.\ and
 \begin{equation}
- Y = \frac{\ii\sqrt2}{v} U_3^*(M\w H^\text{diag})^{1/2}R(M\w L^\text{diag})^{1/2}U_2^\dagger.
-\label{eq:CIP}
+ \zeta=\pm1; \qquad
+(\theta_{12},\theta_{23},\theta_{13})\in\mathbb C,\quad
+|\Re\theta_{12}|\le\pi, \quad |\Re\theta_{23}|\le\pi,  \quad |\Re\theta_{13}|\le\frac{\pi}{2}.
 \end{equation}
+This $R$ satisfies $RR^\TT=1$, which however is not general (as in the next example).
 
-For models with two right-handed neutrinos, one neutrino is massless.
-The same parameterization \cref{eq:CIP} works with\footnote{See, e.g., Ref.~\cite{Brdar:2019iem}. Sho also thanks Kai Schmitz for his note.}
+\paragraph{Example: two right-handed neutrinos}
+For models with two right-handed neutrinos, one neutrino is massless and $M\w L\wdiag$ is not invertible. However the parameterization
+\begin{equation}
+ \MD = \ii \sqrt{M\w H^\text{diag}}R\sqrt{M\w L^\text{diag}}(U^{\text{(Majorana)}}\w{PMNS} )^\dagger
+\end{equation}
+works with\footnote{See, e.g., Ref.~\cite{Brdar:2019iem}. Sho also thanks Kai Schmitz for his note.}
 \begin{align}
@@
 where $z\in\mathbb C$ and $\zeta=\pm1$.
-Note that, in general set-up, $R R^\TT\neq1$.
 

### 2020-10-02 Neutrino.tex
tweak notation

@@
 \begin{equation}
- U\w{PMNS}^{\text{(Dirac)}}=V_e^\dagger V_n.
+ U\w{PMNS}^{\text{Dirac}}=V_e^\dagger V_n.
 \end{equation}
@@
 where $\theta_{ij}\in[0,\pi/2]$ and $\delta\w{CP}\in[0,2\pi]$ as shown in \cref{eq:GeneralUnitary33} and the five phases are removed as done in \cref{sec:SM-CKM}.
-Note that the discussion in \cref{sec:SM-FFdual} holds and $\Theta_W W\tilde W$ term is rotated away.
 
-If the Majorana mass terms are present, $U\w{PMNS}^{\text{(Majorana)}}$, which is defined by \cref{eq:PMNS-defining}, is no longer unitary.
+If the Majorana mass terms are present, $U\w{PMNS}^{\text{Majorana}}$, which is defined by \cref{eq:PMNS-defining}, is no longer unitary.
 However, if $\MN\gg\MD$, it is approximately unitary:
@@
 \begin{align}
- &U\w{PMNS}:=U\w{PMNS}^{\text{(Dirac)}}=
+ &U\w{PMNS}:=U\w{PMNS}^{\text{Dirac}}=
  \pmat{
@@
 };\\
- &U\w{PMNS}^{\text{(Majorana)}} = U\w{PMNS}\pmat{\ee^{\ii\eta_1} \\ & \ee^{\ii\eta_2}\\&&1}
+ &U\w{PMNS}^{\text{Majorana}} = U\w{PMNS}\pmat{\ee^{\ii\eta_1} \\ & \ee^{\ii\eta_2}\\&&1}
 +\Order\left(\frac{\MD}{\MN}\right).
@@
 with the exactness recovered in the Dirac case.
+
+It should be noted that the discussion in \cref{sec:SM-FFdual} holds.
+As far as the baryon number is conserved, we can remove the $\Theta_W W\tilde W$ term by quark rotation.
+Hence, the above-discussed models have CP violation only in the CKM and (extended) PMNS matrices.
 
@@
 \end{equation}
-we can identify $U_1^*\approx U^{\text{(Majorana)}}\w{PMNS}$ and $U_2\approx 1$; the master equation now becomes
+we can identify $U_1^*\approx U^{\text{Majorana}}\w{PMNS}$ and $U_2\approx 1$; the master equation now becomes
 \begin{equation}
  [\ii M\w L\wdiag]^{1/2}[\ii M\w L\wdiag]^{1/2} =
-[(M\w H\wdiag)^{-1/2}\MD U^{\text{(Majorana)}}\w{PMNS}]^\TT
-[(M\w H\wdiag)^{-1/2}\MD U^{\text{(Majorana)}}\w{PMNS}].
+[(M\w H\wdiag)^{-1/2}\MD U^{\text{Majorana}}\w{PMNS}]^\TT
+[(M\w H\wdiag)^{-1/2}\MD U^{\text{Majorana}}\w{PMNS}].
 \end{equation}
@@
 \begin{equation}
- R:=-\ii (M\w H\wdiag)^{-1/2}\MD  U^{\text{(Majorana)}}\w{PMNS}  (M\w L^\text{diag})^{-1/2}
+ R:=-\ii (M\w H\wdiag)^{-1/2}\MD  U^{\text{Majorana}}\w{PMNS}  (M\w L^\text{diag})^{-1/2}
 \quad\Longrightarrow\quad R^\TT R=1.
@@
 \begin{equation}
- \MD = \ii \sqrt{M\w H^\text{diag}}R\sqrt{M\w L^\text{diag}}(U^{\text{(Majorana)}}\w{PMNS} )^\dagger.
+ \MD = \ii \sqrt{M\w H^\text{diag}}R\sqrt{M\w L^\text{diag}}(U^{\text{Majorana}}\w{PMNS} )^\dagger.
 \end{equation}
@@
 Meanwhile, $\tan z=w$ has no solution if and only if $w=\pm \ii$.
-Then, using this fact, one first expresses $R_{i3}$ components by $\zeta\w{A,B,C}=\pm1$ and $\theta\w{A,B}\in\mathbb C$, restricting $0\le\Re\theta\w{A,B}\le\pi/2$ ($\Leftrightarrow\Re{\sin{\theta}}\ge0\land\Re\cos\theta\ge0$), and then get an expression of $R$ with three angles and six signs.
+Then, using this fact, one first expresses $R_{i3}$ components by $\zeta\w{A,B,C}=\pm1$ and $\theta\w{A,B}\in\mathbb C$, restricting $0\le\Re\theta\w{A,B}\le\pi/2$ ($\Leftrightarrow\Re{\sin{\theta}}\ge0\land\Re\cos\theta\ge0$), and then gets an expression of $R$ with three angles and six signs.
 Five signs are absorbed by enlarging $\Re\theta$ and one sign remains, which is $\zeta$.}
@@
 \begin{equation}
- \MD = \ii \sqrt{M\w H^\text{diag}}R\sqrt{M\w L^\text{diag}}(U^{\text{(Majorana)}}\w{PMNS} )^\dagger
+ \MD = \ii \sqrt{M\w H^\text{diag}}R\sqrt{M\w L^\text{diag}}(U^{\text{Majorana}}\w{PMNS} )^\dagger
 \end{equation}

### 2020-10-04 Neutrino.tex
tweak CIP

@@
 Now we successfully parameterized $Y_n$ by a ``complex orthogonal'' matrix $R$.
+The extended PMNS matrix is given by
+\begin{equation}
+ \approx \pmat{
+  U_1^* & \MD^\dagger \MN^{-1}U_2^*\\
+  -\MN^{-1}\MD U_1^* & U_2^*
+}
+\approx
+\pmat{
+ U^{\text{Majorana}}\w{PMNS} & -\ii U^{\text{Majorana}}\w{PMNS} 
+\sqrt{M\w L^\text{diag}}R^\dagger
+(M\w H^\text{diag})^{-1/2}\\
+  -\ii(M\w H^\text{diag})^{-1/2}R\sqrt{M\w L^\text{diag}}
+&1
+}.
+\end{equation}
 
-Such $R$ is given by\footnote{For $w\in\mathbb C$, $\sin z_1=w$ and $\cos z_2=w$ always have solutions $z_{1,2}\in\mathbb C$.
+The parameter matrix $R$ is given by\footnote{For $w\in\mathbb C$, $\sin z_1=w$ and $\cos z_2=w$ always have solutions $z_{1,2}\in\mathbb C$.
 Meanwhile, $\tan z=w$ has no solution if and only if $w=\pm \ii$.

### 2020-10-06 Neutrino.tex
tweak neutrino-spinor notations to avoid confusions

@@
 
-Using Weyl spinor,
+Together with the Dirac mass $\MD$, we introduce left-handed Weyl spinors $\xi$ and $\chi$ to avoid notational confusion:
+\begin{equation}
+ L^1 = \pmat{\xi\\0},\quad
+ N   = \pmat{0\\\bar\chi},\quad
+ \MD := \frac{v}{\sqrt2}Y_n,\quad
+ \MD\wdiag := U_n \MD Y_n^\dagger.
+\end{equation}
+Note that $\xi\equiv\nuL$ but $\bar\chi\equiv\nR$ and $\chi$ is the conjugate of a right-handed fermion $\nR$. Then,
 \begin{align}
-& L^1 = \pmat{\nuL\\0},\quad
- N   = \pmat{0\\\bnR},\quad
- \MD := \frac{v}{\sqrt2}Y_n,\quad
- \MD\wdiag := U_n \MD Y_n^\dagger,\\
-&\label{eq:MajoranaNuLag}
+\label{eq:MajoranaNuLag}
  \mathcal L\w{Y+\nu} 
-\supset
+&\supset
  -\frac{v}{\sqrt2}\overline{N}Y_n \PL L^1 - \frac{1}{2}\overline{N}\MN N^\cc + \text{h.c.}
-=
- -\frac{v}{\sqrt2}\nR Y_n \nuL - \frac{1}{2}\nR \MN\nR + \text{h.c.}\\
-&\phantom{ \mathcal L\w{Y+\nu}}
-=-\frac12\pmat{\nuL & \nR}\pmat{0 & \MD^\TT \\ \MD & \MN} \pmat{\nuL \\ \nR} + \text{h.c.} =: -\frac12\tilde\nu^\TT \tilde M\tilde \nu+\text{h.c.}
+\\&
+=-\frac12\pmat{\xi & \chi}\pmat{0 & \MD^\TT \\ \MD & \MN} \pmat{\xi \\ \chi} + \text{h.c.} =: -\frac12\tilde\nu^\TT \tilde M\tilde \nu+\text{h.c.}
 \end{align}
@@
 \begin{align}
- \pmat{\nuL\\\nR} = \tilde R^*\pmat{\nu_{1\TO3} \\ \nu_{4\TO}}
+ \pmat{\xi\\\chi} \equiv \pmat{\nuL\\(\nR)^\dagger} = \tilde R^*\pmat{\nu_{1\TO3} \\ \nu_{4\TO}}
 \end{align}
@@
 \quad
-\pmat{\nuL\\\nR}
+\pmat{\xi\\\chi}
 =
-\pmat{\begin{smallmatrix}\nu_e\\\nu_\mu\\\nu_\tau\end{smallmatrix}\\\nu_{\text{sterile};i}}
+\pmat{\begin{smallmatrix}\nu_e\\\nu_\mu\\\nu_\tau\end{smallmatrix}\\(\nu_{\text{sterile}})^\dagger_{i}}
 =\tilde R^*\pmat{\nu_{1\TO3}\\\nu_{4\TO}}.
@@
 
-\paragraph{PMNS matrix}\label{sec:PMNS}
+\subsection{PMNS matrix}\label{sec:PMNS}
 In the Dirac neutrino models the PMNS matrix is unitary, which we parameterize
@@
 -\mathcal L\supset \frac12
- \pmat{\nuL & \nR} \pmat{0 & \MD^\TT \\ \MD & \MN} \pmat{\nuL \\ \nR}+\text{h.c.}
-= \frac12\pmat{\nuL[i] & \nR[a]}
+ \pmat{\xi & \chi} \pmat{0 & \MD^\TT \\ \MD & \MN} \pmat{\xi \\ \chi}+\text{h.c.}
+= \frac12\pmat{\xi_i & \chi_a}
 \pmat{0_{ij} & (\MD^\TT)_{ib} \\ \MD[aj] & \MN[ab]}
-\pmat{\nuL[j] \\ \nR[b]}+\text{h.c.}
+\pmat{\xi_j \\ \chi_b}+\text{h.c.}
 \end{equation}

### 2020-10-22 Neutrino.tex
write up neutrino

--- 2020-10-02 Neutrino.tex
@@
 \newcommand\MN[1][]{M_{N#1}}
+\newcommand{\nRHN}{n_{\text{RHN}}}
 
@@
 \clearpage
-\subsection{Convention}
-We extend the SM Yukawa \eqref{eq:SMyukawa} to include the neutrino mass terms:
+\subsection{Convention and Nomenclature}
+We define the neutrino mixing matrix $U$ (or $U_{\alpha i}$) by, following the standard convention,
+\begin{equation}
+ \ket{\nu_\alpha^{\text{flavor}}} = U_{\alpha i}^*\ket{\nu_i^{\text{mass}}},
+\qquad
+ \nu^\text{flavor} = U \nu^{\text{mass}}, \qquad[\alpha=e,\mu,\tau,\mathrm{r}_1,\mathrm{r}_2,\dots].\label{eq:U-defining}
+\end{equation}
+The mass basis are labelled by $i=1,2,\dots$, while $\alpha$ labels the ``flavor'' basis.
+The flavor basis for $\alpha=e,\mu,\tau$ is defined by the charged leptons, i.e., so that the lepton doublets are aligned and $\nu_\alpha l^\pm_\beta W^\mp$ interactions are diagonal.
+Meanwhile, for other ``extra'' neutrinos, the flavor basis can be left arbitrary (undefined) because it is usually not of our interests.
+
+We, in this note, use the term ``PMNS (matrix)'' only if the neutrino mixing matrix is a $3\times3$ unitary matrix.
+
+\subsection{Models}
+\paragraph{Dirac neutrino model}
+The simplest model for neutrino masses are given by extending the SM Yukawa \eqref{eq:SMyukawa} with ($\nRHN$ copies of) right-handed neutrino $N$:
 \begin{align}
-   \mathcal L\w{Y+\nu}&=
-  \overline{U}\Yu H \PL Q
+   \mathcal L\w{Dirac-$\nu$}&=
+\left(  \overline{U}\Yu H \PL Q
 - \overline{D}\Yd H^\dagger \PL Q
@@
 - \overline{E}\Ye H^\dagger \PL L
-- \frac{1}{2}\overline{N}\MN N^\cc
-+ \text{h.c.}
+\right)+ \text{h.c.}
 \\&=
-- \overline{Q^a}\Yu^\dagger \epsilon^{ab}H^{b*} \PR U
+\left(- \overline{Q^a}\Yu^\dagger \epsilon^{ab}H^{b*} \PR U
 - \overline{Q^a}\Yd^\dagger H^{a} \PR D
@@
 - \overline{L^a}\Ye^\dagger H^{a} \PR E
-- \frac{1}{2}\overline{N}\MN N^\cc + \text{h.c.},
+\right) + \text{h.c.}
 \end{align}
-where $\MN$, a \emph{complex symmetric} Majorana mass matrix, may be absent.
-
-As we explicitly describe in \cref{sec:PMNS}, we use the standard convention for the PMNS matrix \GRAY{(Pontecorvo--\japanese{牧}--\japanese{中川}--\japanese{坂田})}:
-\begin{equation}
- \ket{\nu_\alpha^{\text{flavor}}} = [U\w{PMNS}]_{\alpha i}^*\ket{\nu_i^{\text{mass}}},
-\qquad
- \nu^\text{flavor} = U\w{PMNS}\nu^{\text{mass}}, \qquad[\alpha=e,\mu,\tau,\mathrm{r}_1,\mathrm{r}_2,\dots]\label{eq:PMNS-defining}
-\end{equation}
-where the ``flavor basis'' is defined by the charged lepton mass basis for first three elements, while in most cases by the basis that diagonalizes Majorana mass term for the rest.
-
-\paragraph{Dirac neutrino}
-If $\MN$ is absent, neutrinos become Dirac fermions and the discussion goes parallel to the CKM matrix:
+Neutrinos become Dirac fermions and the discussion goes parallel to the CKM matrix:
 \begin{equation}
  \Yn = U_n Y_n\wdiag V_n^\dagger,\qquad
-\{\nuL, \eL, \overline N, \overline E\}^{\text{mass basis}} = \{V_n^\dagger L^1, V_e^\dagger L^2, \overline N U_n, \overline E U_e\}
+\{\nuL, \eL, \overline N, \overline E\}^{\text{mass basis}} = \{V_n^\dagger L^1, V_e^\dagger L^2, \overline N U_n, \overline E U_e\},
 \end{equation}
-and
+where  $\Yn\in\mathbb C^{\nRHN\times3}$, $V_n\in\mathbb U_{\mathbb C}^3$, and $U_n\in\mathbb U_{\mathbb C}^{\nRHN}$.
+The Lagrangian is now given by
 \begin{align}
@@
 \frac{g_2}{\sqrt2}\left[
-\overline{\nuL^{\text{mass}}} [U\w{PMNS}]^\dagger\slashed W^+\eL^{\text{mass}}
-+\overline{\eL^{\text{mass}}}[U\w{PMNS}] \slashed W^-\nuL^{\text{mass}}
-\right].
+\overline{\nuL^{\text{mass}}} U^\dagger\slashed W^+\eL^{\text{mass}}
++\overline{\eL^{\text{mass}}}U\slashed W^-\nuL^{\text{mass}}
+\right],
 \end{align}
-The mass eigenstate Dirac field is given by
+where the definition \eqref{eq:U-defining} provides the equality
+\begin{equation}
+ U=V_e^\dagger V_n\equiv U\w{PMNS}.
+\end{equation}
+According to our terminology, this $3\times 3$ unitary matrix $U$ is called ``the PMNS matrix'' for this model.
+Note that this is given in \emph{the opposite manner} to the CKM matrix ($V\w{CKM}=V_u^\dagger V_d$).
+
+The mass eigenstate Dirac fields are given by
 \begin{equation}
@@
 \end{equation}
-The PMNS matrix is given in \emph{the opposite manner} to the CKM matrix:
+Note that $\nRHN=\{2,3\}$ because models with $\nRHN<3$ yields $3-\nRHN$ massless left-handed neutrinos, while $\nRHN>3$ results in $\nRHN-3$ massless right-handed neutrinos.
+
+\paragraph{Type-I see-saw}
+The type-I see-saw models are given by 
+\begin{align}
+   \mathcal L_{\text{type-I}}&=
+\left(  \overline{U}\Yu H \PL Q
+- \overline{D}\Yd H^\dagger \PL Q
++ \overline{N}\Yn H \PL L
+- \overline{E}\Ye H^\dagger \PL L
+- \frac{1}{2}\overline{N}\MN N^\cc
+\right)+ \text{h.c.}
+\\&=
+\left(- \overline{Q^a}\Yu^\dagger \epsilon^{ab}H^{b*} \PR U
+- \overline{Q^a}\Yd^\dagger H^{a} \PR D
+- \overline{L^a}\Yn^\dagger \epsilon^{ab}H^{b*} \PR N
+- \overline{L^a}\Ye^\dagger H^{a} \PR E
+- \frac{1}{2}\overline{N}\MN N^\cc\right) + \text{h.c.},
+\label{eq:type-1-lagrangian}
+\end{align}
+where $M_N$ is a $\nRHN\times\nRHN$ complex symmetric matrix.
+Models with $\nRHN\ge2$ are viable, because $\nRHN<3$ gives $3-\nRHN$ massless (Weyl) neutrinos and $2\nRHN$ massive (Majorana) neutrinos, while all neutrinos acquire Majorana mass  if $\nRHN\ge 3$.
+
+We introduce left-handed Weyl spinors $\xi$ and $\chi$ to avoid notational confusion and diagonalize the Dirac mass term $\MD$:
 \begin{equation}
- U\w{PMNS}^{\text{Dirac}}=V_e^\dagger V_n.
+ L^1 \equiv \pmat{\xi\\0}\equiv \pmat{\nuL\\0},\quad
+ N   \equiv \pmat{0\\\bar\chi}\equiv \pmat{0\\\nR},\quad
+ \MD := \frac{v}{\sqrt2}Y_n,\quad
+ \MD\wdiag := U_n \MD Y_n^\dagger.
 \end{equation}
-In general, $\Yn\in\mathbb C^{n\times3}$, $V_n\in\mathbb U_{\mathbb C}^3$, $U_n\in\mathbb U_{\mathbb C}^n$, and the Dirac-PMNS matrix is also a $3\times 3$ unitary matrix.
-Models with $n<3$ yields $3-n$ massless left-handed neutrinos, while $n>3$ results in $n-3$ massless right-handed neutrinos.
-
-\paragraph{Majorana neutrino}
-Models with $\MN\neq0$ generate so-called Majorana neutrino masses.
-If $n<3$, the model has $3-n$ massless neutrinos and $2n$ massive neutrinos, while all neutrinos are massive if $n\ge 3$.
-
-Using Weyl spinor,
+Then the mass term is written in matrix form:
 \begin{align}
-& L^1 = \pmat{\nuL\\0},\quad
- N   = \pmat{0\\\bnR},\quad
- \MD := \frac{v}{\sqrt2}Y_n,\quad
- \MD\wdiag := U_n \MD Y_n^\dagger,\\
-&\label{eq:MajoranaNuLag}
- \mathcal L\w{Y+\nu} 
+\label{eq:MajoranaNuLag}
+ \mathcal L_{\text{type-I}}
 \supset
  -\frac{v}{\sqrt2}\overline{N}Y_n \PL L^1 - \frac{1}{2}\overline{N}\MN N^\cc + \text{h.c.}
-=
- -\frac{v}{\sqrt2}\nR Y_n \nuL - \frac{1}{2}\nR \MN\nR + \text{h.c.}\\
-&\phantom{ \mathcal L\w{Y+\nu}}
-=-\frac12\pmat{\nuL & \nR}\pmat{0 & \MD^\TT \\ \MD & \MN} \pmat{\nuL \\ \nR} + \text{h.c.} =: -\frac12\tilde\nu^\TT \tilde M\tilde \nu+\text{h.c.}
+=-\frac12\pmat{\xi & \chi}\pmat{0 & \MD^\TT \\ \MD & \MN} \pmat{\xi \\ \chi} + \text{h.c.} =: -\frac12\tilde\nu^\TT \tilde M\tilde \nu+\text{h.c.},
 \end{align}
-As $\tilde M$ is a complex symmetric $(3+n)\times(3+n)$ matrix, it can be AT-diagonalized:
+where tildes denote ``larger'' objects. We can AT-diagonalize $\tilde M$, which is a symmetric $\mathbb{C}^{(3+\nRHN)\times(3+\nRHN)}$ matrix, as usual:
 \begin{equation}
@@
 \qquad
--\mathcal L\w{Y+\nu}\supset
+-\mathcal L\w{Dirac-$\nu$}\supset
 \frac12\tilde \nu^\TT \tilde M\tilde\nu
-=\frac12(\tilde\nu^{\text{mass}})^\TT \tilde M\wdiag \tilde\nu^{\text{mass}};\qquad \tilde\nu^{\text{mass}} =\tilde R^\TT\tilde\nu
+=\frac12(\tilde\nu^{\text{mass}})^\TT \tilde M\wdiag \tilde\nu^{\text{mass}};\qquad \tilde\nu^{\text{mass}} =\tilde R^\TT\tilde\nu.
 \end{equation}
-and $\tilde M$ gives the neutrino masses.
-The neutrino mixing is then given by
-\begin{align}
- \pmat{\nuL\\\nR} = \tilde R^*\pmat{\nu_{1\TO3} \\ \nu_{4\TO}}
-\end{align}
-and find that the PMNS matrix is now extended to a $3\times n$ matrix $V_e^\dagger [\tilde R^*]\w{upper}$.
+Mass eigenstates $\nu^{\text{mass}}$ are Majorana fermions given by left-handed Weyl spinors.
+The neutrino mixing are given by
+\begin{equation}
+ \pmat{\xi\\\chi} \equiv \pmat{\nuL\\(\nR)^\dagger} = \tilde R^*\pmat{\nu_{1\TO3} \\ \nu_{4\TO}},\qquad
+ \therefore
+   U = \pmat{V_e^\dagger & 0 \\ 0 & X}\tilde R^*,
+\label{eq:type-I-mixing}
+\end{equation}
+where $Y_e$ comes from the charged lepton mass diagonalization, while $X$ is a unitary matrix (but unphysical if, as usual, the right-handed neutrinos are indistinguishable).
 
-Usually the discussion should be start from the basis in which $Y_e$ and $M_N$ are positive diagonal with increasing diagonal elements (``charged lepton mass basis'' combined with ``Majorana mass basis'').
-Then
-\begin{equation}
-\tilde M = \pmat{0 & \MD^\TT \\ \MD & \MN\wdiag},
-\quad
-\pmat{\nuL\\\nR}
-=
-\pmat{\begin{smallmatrix}\nu_e\\\nu_\mu\\\nu_\tau\end{smallmatrix}\\\nu_{\text{sterile};i}}
-=\tilde R^*\pmat{\nu_{1\TO3}\\\nu_{4\TO}}.
-\end{equation}
-The PMNS matrix appears as a submatrices, which is no longer unitary:
-\begin{equation}
-\tilde R^*
-=:
-\pmat{ R_{11}^* & R_{12}^* \\ R_{21}^* & R_{22}^*}
-=:
-\pmat{U\w{PMNS}^{\text{Majorana}} & U_{\text{active-heavy}}\\
-U_{\text{sterile-light}} & U_{\text{sterile-heavy}}
-}.
-\end{equation}
-
-\paragraph{Connection of the above two formalism}
-If we apply AT-diagonalization to the Dirac case (with $n=3$), we obtain
+\paragraph{Connection of the above two models}
+If we set $\MN=0$ in the Type-I see-saw ($\nRHN=3$), the AT-diagonalization becomes
 \begin{align}
@@
   \tilde R=\pmat{V_n^* & \ii V_n^* \\ U_n & -\ii U_n},\quad
-  \tilde M\wdiag=\pmat{\MD\wdiag & 0 \\ 0 & \MD\wdiag}
+  \tilde M\wdiag=\pmat{\MD\wdiag & 0 \\ 0 & \MD\wdiag}.
 \end{align}
-and find that three pairs of degenerate Weyl fermions form three Dirac neutrinos.
+As three pairs of degenerate Weyl fermions form three Dirac neutrinos, the Dirac neutrino model is reproduced as expected.
 
-\paragraph{PMNS matrix}\label{sec:PMNS}
-In the Dirac neutrino models the PMNS matrix is unitary, which we parameterize
+\subsection{Neutrino mixings}
+Experiments have revealed that the upper-left $3\times3$ submatrix is close to a unitary matrix, which we call ``the PMNS matrix'' $U\w{PMNS}$.
+Thus\footnote{Sho thanks Josu Hernandez-Garcia for discussion useful in this whole section.}
+ we decompose the general neutrino mixing angle $U$ to
+\begin{equation}
+ U = U'\pmat{U\w{PMNS} & 0 \\ 0 & U_X}
+  = \pmat{U'_{11} & U'_{12} \\ U'_{21} & U'_{22}}\pmat{U\w{PMNS} & 0 \\ 0 & U_X}
+\end{equation}
+with $U_X$ (and hence $U'$) being unitary.
+Here $U'_{11}\approx1$ and $U'_{12},U'_{21}\approx0$.
+We do not care much about $U'_{22}$ but in type-I see-saw models $U'_{22}U_X$ inherits the redundancy of $X$ and thus we can take, e.g., $U'_{22}\approx 1$ or $U_X=1$.
+
+The matrix $U'$ is theoretically given with some matrix $\Theta$ by~\cite{Blennow:2011vn}\footnote{%
+\TODO{}proof? or obvious if considers $\gSU(6)/(\gSU(3)\times\gSU(3))$?}
+\begin{equation}
+ U'\equiv \exp\pmat{0 & -\Theta^\dagger \\ \Theta & 0}
+=\sum_{n=0}^\infty\pmat{
+  {(-\Theta^\dagger\Theta)^{n}}/{(2n)!} &
+  {-(-\Theta^\dagger\Theta)^{n}\Theta^\dagger}/{(2n+1)!}\\
+  {+\Theta(-\Theta^\dagger\Theta)^{n}}/{(2n+1)!}&
+  {(-\Theta\Theta^\dagger)^{n}}/{(2n)!}
+}
+\end{equation}
+but practically we derive it in perturbative expansion.
+For example, the type-I case is evaluated as\footnote{Cf.~\texttt{calculator/neutrino/diagonalization\_perturbative.wl}}
+\begin{align}
+ \text{(diag.)} = U^\TT\pmat{0 & \MD^\TT \\ \MD & \MN} U;\qquad
+&U'\approx\pmat{
+  1-\frac12 \Theta_0^\dagger \Theta_0&
+  -\Theta_0^\dagger\\
+ \Theta_0&
+  1-\frac12\Theta_0\Theta_0^\dagger
+};\quad \Theta_0:=-\MN^{-1}\MD,
+\label{eq:typeIseesaw-blockdiag-1}\\
+&\text{(block diag.)}=U'^\TT\pmat{0 & \MD^\TT \\ \MD & \MN} U'\approx\pmat{
+  -\MD^\TT\MN^{-1}\MD & 0 \\ 0 & \MN
+},
+\label{eq:typeIseesaw-blockdiag-2}
+\end{align}
+where however note that $\Theta_0$ is close but not equal to $\Theta$.
+
+If we focus only on the upper-left $3\times3$ part and are allowed to neglect $\Theta^4$ (or $\Theta_0^4$) terms, the above decompositions of a unitary matrix into two unitary matrices~\cite{Blennow:2016jkn}
+can be expressed by
+\begin{equation}
+ [U]\w{upper left} \simeq (1-\eta)U\w{PMNS};\qquad
+  \text{$\eta:=\Theta^\dagger\Theta/2$ is a complex positive-semidefinite Hermitian matrix},
+\end{equation}
+where the positive-semidefiniteness yields
+\begin{equation}
+ \eta_{ii}\ge0,\quad |\eta_{ij}|\le\sqrt{\eta_{ii}\eta_{jj}}\le(\eta_{ii}+\eta_{jj})/2.
+\end{equation}
+The QR decomposition is also useful~\cite{Blennow:2016jkn}; together with the polar decomposition, it is summarized by
+\begin{alignat}{2}
+ [U]\w{upper left} &= (1-\alpha)U\w{QR};\qquad&
+  &\text{QR decomposition; $\alpha$ is a complex upper-triangle matrix},\\
+ [U]\w{upper left} &= \eta'U\w{polar};\qquad&
+  &\text{polar decomposition; $\eta'$ is a complex positive-semidefinite Hermitian matrix},
+\end{alignat}
+where  $U\w{QR}\neq U\w{polar}$.
+However, as experiments claim that $[U]\w{upper left} \approx U\w{PMNS}$ and $\eta$ and $\alpha$ are still consistent with zero, they are consistent with $U\w{PMNS}$.
+Note also that, as $[U]\w{upper left}$ seems to be invertible, these decompositions are given uniquely.
+
+\subsection{PMNS matrix}\label{sec:PMNS}
+As experiments have not found any deviation from $[U]\w{upper left}$ being unitary, we express the mixing with the PMNS matrix
+ \GRAY{(Pontecorvo--\japanese{牧}--\japanese{中川}--\japanese{坂田})}
+parameterized by
 \begin{equation}
 \begin{split}
-   U\w{PMNS}^{\text{Dirac}}&=V_e^\dagger V_n
- =\pmat{
- U_{e1} & U_{e2} & U_{e3}\\
- U_{\mu1} & U_{\mu2} & U_{\mu3}\\
- U_{\tau1} & U_{\tau2} & U_{\tau3}
- }
- =
- %
+   U\w{PMNS}&=
  \pmat{1&&\\&\co{23}&\si{23}\\&-\si{23}&\co{23}}
@@
  \pmat{\co{12}&\si{12}&\\-\si{12}&\co{12}&\\&&1}
+\pmat{\ee^{\ii\eta_1} \\ & \ee^{\ii\eta_2}\\&&1}
  %
@@
   \si{12}\si{23} - \co{12} \co{23} \si{13}\ee^{\ii\delta\w{CP}} & -\co{12}\si{23}-\si{12}\co{23}\si{13}\ee^{\ii\delta\w{CP}} & \co{23} \co{13}
-},
+}\pmat{\ee^{\ii\eta_1} \\ & \ee^{\ii\eta_2}\\&&1},
 \end{split}
 \end{equation}
-where $\theta_{ij}\in[0,\pi/2]$ and $\delta\w{CP}\in[0,2\pi]$ as shown in \cref{eq:GeneralUnitary33} and the five phases are removed as done in \cref{sec:SM-CKM}.
-
-If the Majorana mass terms are present, $U\w{PMNS}^{\text{Majorana}}$, which is defined by \cref{eq:PMNS-defining}, is no longer unitary.
-However, if $\MN\gg\MD$, it is approximately unitary:
-\begin{equation}
-   U\w{PMNS}^{\text{Majorana}}=
-\pmat{
- U_{e1} & U_{e2} & U_{e3}\\
- U_{\mu1} & U_{\mu2} & U_{\mu3}\\
- U_{\tau1} & U_{\tau2} & U_{\tau3}
- }
-\approx
-U\w{PMNS}^{\text{Dirac}}
-\pmat{\ee^{\ii\eta_1} \\ & \ee^{\ii\eta_2}\\&&1}.
-\end{equation}
-Here additional two phases are introduced because we can no longer rotate $\nR$; three among $(a,b,c,d,e)$ in \cref{eq:GeneralUnitary33} are removed by rotating $L_i$, and two remains.
-
-Current experiments measure the value of the matrix and the above parameterization still works well (cf.~Ref.~\cite{NUFIT}).
-Thus we hereafter identify the Dirac PMNS matrix as $U\w{PMNS}$:
-\begin{align}
- &U\w{PMNS}:=U\w{PMNS}^{\text{Dirac}}=
- \pmat{
- \co{12} \co{13} & \si{12} \co{13} & \si{13} \ee^{-\ii\delta\w{CP}}\\
- -\si{12} \co{23} - \co{12} \si{23} \si{13} \ee^{\ii\delta\w{CP}}& \co{12} \co{23} - \si{12}\si{23}\si{13} \ee^{\ii\delta\w{CP}}& \si{23}\co{13}\\
-  \si{12}\si{23} - \co{12} \co{23} \si{13}\ee^{\ii\delta\w{CP}} & -\co{12}\si{23}-\si{12}\co{23}\si{13}\ee^{\ii\delta\w{CP}} & \co{23} \co{13}
-};\\
- &U\w{PMNS}^{\text{Majorana}} = U\w{PMNS}\pmat{\ee^{\ii\eta_1} \\ & \ee^{\ii\eta_2}\\&&1}
-+\Order\left(\frac{\MD}{\MN}\right).
-\end{align}
-Accordingly, the components $U_{\alpha i}$ is in general slightly different from $(\alpha,i)$ component of $U\w{PMNS}$, e.g.,
-\begin{equation*}
- U_{e1}\approx \co{12}\co{13}
-\end{equation*}
-with the exactness recovered in the Dirac case.
+where $\theta_{ij}\in[0,\pi/2]$ and $\delta\w{CP}\in[0,2\pi]$ as shown in \cref{eq:GeneralUnitary33}.
+In the Dirac neutrino model, $\eta_1$ and $\eta_2$ (``Majorana phases'') are unphysical because we can rotate $\nR$ to eliminate them, while the rotation is not allowed in presence of the Majorana mass term.
 
@@
 As far as the baryon number is conserved, we can remove the $\Theta_W W\tilde W$ term by quark rotation.
-Hence, the above-discussed models have CP violation only in the CKM and (extended) PMNS matrices.
+Hence, the above-discussed models have CP violation only in the CKM matrix and the neutrino mixing matrix.
 
-\paragraph{PDG and NuFIT convention} The convention agrees with PDG \cite[\S14]{PDG2020} and NuFIT~\cite[v5.0]{NUFIT}.
-Compared with PDG,\footnote{Sho thinks Eq.~(14.9) of PDG2020 lacks 1/2 in the right-most term.}
-\begin{align*}
-& \mathcal L
-\supset
- -\CONV{\bar\nu_s} \CONV{M_D} \nu_L
- -\frac12\CONV{\bar\nu_s} \CONV{M_N} \CONV{\nu^c_s}
- -\overline{L^2}\CONV{M_l}\PR E\quad\text{(14.6)+(14.27)},\\
-&
-\Bigl\{\CONV{(V^{\nu})^\TT \spmat{0&M_D^\TT \\ M_D &M_N} V^\nu}\text{~or~}\CONV{V_R^{\nu\dagger}M_D V^\nu}\Bigr\}=\diag(m_i)\quad
-\text{(Majorana; 14.9)+(Dirac; 14.15)},\\
-&\nu_{L}=\PL\CONV{V^\nu}\nu^{\text{mass}}\quad\text{(14.14)\&(14.18)},\qquad
-\CONV{V^{l\dagger}M_l V^l_R}=\diag(m_e,m_\mu,m_\tau)\quad\text{(14.31)}.
-\end{align*}
-These leads to $\CONV{V^\nu_R}=U_n$ and $\CONV{V^\nu}=V_n$ in Dirac case, $\CONV{M_D} = \Yn v/\sqrt2=\MD$, $\CONV{M_N}=M_N$, and $\CONV{V^\nu}=R^*$ in Majorana case, and $\CONV{V^{l}}=V_e$ and $\CONV{V_R^l}=U_e$ (note $\CONV{M_l}=Y_e^\dagger v/\sqrt{2}$).
-So
-\begin{equation*}
- \CONV{U_{ij}} = \text{(diagonal phases)}\times\CONV{V^{l\dagger}}\CONV{V^{\nu}}\times\text{(diagonal phases)}=\text{(phases)}V_e^\dagger \{V_\nu\text{~or~}R^*\}\text{(phases)}= U\w{PMNS}^{\text{(Dirac/Majorana)}}
-\end{equation*}
+\paragraph{PDG and NuFIT convention}
+Our convention is the same with PDG \cite[\S14]{PDG2020}~\footnote{Sho thinks Eq.~(14.9) of PDG2020 lacks 1/2 in the right-most term.}.
+
+It also agrees with NuFIT~\cite[v5.0]{NUFIT} (their convention is given in Ref.~\cite{Esteban:2018azc}) except for the Majorana phases,
+$
+ \eta_i = \CONV{\alpha_i}.
+$
 
 \subsection{Casas--Ibarra parameterization}
-\paragraph{General basis}
-We start from the neutrino mass matrix
+We start from the approximation given in Eq.~\eqref{eq:typeIseesaw-blockdiag-2}:
 \begin{equation}
--\mathcal L\supset \frac12
- \pmat{\nuL & \nR} \pmat{0 & \MD^\TT \\ \MD & \MN} \pmat{\nuL \\ \nR}+\text{h.c.}
-= \frac12\pmat{\nuL[i] & \nR[a]}
-\pmat{0_{ij} & (\MD^\TT)_{ib} \\ \MD[aj] & \MN[ab]}
-\pmat{\nuL[j] \\ \nR[b]}+\text{h.c.}
+ m\w L \approx -U\w{PMNS}^\TT \MD^\TT \MN^{-1}\MD U\w{PMNS},\qquad
+ m\w H \approx U_X^\TT \MN U_X,
 \end{equation}
-The AT-factorization can be separated into two steps:
+where $m\w L$ and $m\w H$ are diagonal mass matrices for three light neutrinos and heavier neutrinos.
+We then obtain
 \begin{equation}
- \tilde M=\pmat{0 & \MD^\TT \\ \MD & \MN}
-\longrightarrow
-\pmat{M\w L & 0 \\ 0 & M\w H}
-\stackrel{\text{ATF}}\longrightarrow
-\pmat{M\w L\wdiag & 0 \\ 0 & M\w H\wdiag}=\tilde M\wdiag,
+  m\w L
+\approx -U\w{PMNS}^\TT \MD^\TT (U_X^* m\w H U_X^\dagger)^{-1}\MD U\w{PMNS}
+=       -U\w{PMNS}^\TT \MD^\TT U_Xm\w H^{-1}U_X^\TT\MD U\w{PMNS}
 \end{equation}
-where intermediate matrices $M\w L$ and $M\w H$ are complex symmetric.
-This is expressed with unitary matrices $U$, $U_1$, and $U_2$:
+and decompose it as
 \begin{equation}
- \tilde M =
-U \pmat{M\w L & 0 \\ 0 & M\w H} U_1^\TT =
-U \pmat{U_1 M\w L\wdiag U_1^\TT & 0 \\ 0 & U_2 M\w H\wdiag U_2^\TT} U^\TT,
+   [\ii m\w L]^{1/2}[\ii m\w L]^{1/2}
+\approx     \left(\frac{v}{\sqrt2}m\w H^{-1/2}U_X^\TT Y_n U\w{PMNS}\right)^{\TT} \left(\frac{v}{\sqrt2} m\w H^{-1/2}U_X^\TT Y_n U\w{PMNS}\right),
 \end{equation}
-following the convention in \cref{eq:ATF}.
-The first equality is calculated as an expansion in $\MD/\MN$ once we assume the see-saw mechanism:
-\begin{align}
- U&\simeq \pmat{1 & \MD^\TT\MN^{-1} \\ -\MN^{-1}\MD^* & 1}
- - \frac12\pmat{\MD^\TT\MN^{-2}\MD^* & 0 \\ 0 & \MN^{-1}\MD^*\MD^\TT \MN^{-1}} + \cdots,\\
- M\w L&\simeq -\MD^\TT \MN^{-1} \MD+\cdots,\\
- M\w H&\simeq \MN + \frac12 (\MD\MD^\dagger\MN^{-1}+\MN^{-1}\MD^*\MD^\TT)+\cdots,
-\end{align}
-At the leading order,
-\begin{equation*}
- U_1 M\w L\wdiag U_1^\TT = [U^\dagger \tilde M U^*]_{\text{upper left}} \approx -\MD^\TT \MN^{-1}\MD
-\approx -\MD^\TT M\w H^{-1}\MD = -\MD^\TT (U_2 M\w H\wdiag U_2^\TT)^{-1}\MD,
-\end{equation*}
-or
-\begin{equation*}
- -M\w L\wdiag \approx U_1^\dagger \MD^\TT U_2^*(M\w H\wdiag)^{-1}U_2^\dagger\MD U_1^*.
-\end{equation*}
-This is decomposed to
-\begin{equation}
- [\ii M\w L\wdiag]^{1/2}[\ii M\w L\wdiag]^{1/2} = 
-[(M\w H\wdiag)^{-1/2}U_2^\dagger\MD U_1^*]^\TT
-[(M\w H\wdiag)^{-1/2}U_2^\dagger\MD U_1^*].
-\end{equation}
-This is the master equation for Casas-Ibarra parameterization~\cite{Casas:2001sr}.
-
-\paragraph{``Standard'' parameterization}
-Let us assume that we started from the above-discussed $(Y_e,\MN)$-diagonal basis.
-Then, noting
-\begin{equation}
- \tilde R=U\pmat{U_1\\&U_2}\approx \pmat{U_1 & \MD^\TT\MN^{-1}U_2 \\ -\MN^{-1}\MD^* U_1 & U_2},
-\end{equation}
-we can identify $U_1^*\approx U^{\text{Majorana}}\w{PMNS}$ and $U_2\approx 1$; the master equation now becomes
-\begin{equation}
- [\ii M\w L\wdiag]^{1/2}[\ii M\w L\wdiag]^{1/2} =
-[(M\w H\wdiag)^{-1/2}\MD U^{\text{Majorana}}\w{PMNS}]^\TT
-[(M\w H\wdiag)^{-1/2}\MD U^{\text{Majorana}}\w{PMNS}].
-\end{equation}
-We will use this parameterization below.
+which is the master equation for Casas-Ibarra parameterization~\cite{Casas:2001sr}.
+Note that $U_X$ is unphysical, i.e., we can fix the basis of $\nR$ so that $U_X=1$.
+This basis is equivalent (under our approximation) to the basis in which $\MN$ becomes diagonal.
 
@@
 \begin{equation}
- R:=-\ii (M\w H\wdiag)^{-1/2}\MD  U^{\text{Majorana}}\w{PMNS}  (M\w L^\text{diag})^{-1/2}
-\quad\Longrightarrow\quad R^\TT R=1.
+ R:=-\ii m\w H^{-1/2}U_X^\TT \MD  U\w{PMNS}  m\w L^{-1/2}
+\quad\Longrightarrow\quad R^\TT R\approx1.
 \end{equation}
-Conversely, with a matrix $R$ satisfying $R^\TT R=1$, the ``Yukawa matrix'' is given by
+Conversely, with a matrix $R$ satisfying $R^\TT R=1$, the Yukawa matrix is given by
 \begin{equation}
- \MD = \ii \sqrt{M\w H^\text{diag}}R\sqrt{M\w L^\text{diag}}(U^{\text{Majorana}}\w{PMNS} )^\dagger.
+ \frac{v}{\sqrt2}Y_n
+  \approx \ii U_X^*\sqrt{m\w H}R\sqrt{m\w L}U\w{PMNS}^\dagger
 \end{equation}
-Now we successfully parameterized $Y_n$ by a ``complex orthogonal'' matrix $R$.
+or
+\begin{equation}
+ \frac{v}{\sqrt2}Y_n\Big|\w{$\MN$ being diagonal}
+  \approx \ii \sqrt{m\w H}R\sqrt{m\w L}U\w{PMNS}^\dagger.
+\end{equation}
 
-Such $R$ is given by\footnote{For $w\in\mathbb C$, $\sin z_1=w$ and $\cos z_2=w$ always have solutions $z_{1,2}\in\mathbb C$.
+Now we successfully parameterized $Y_n$ by a ``complex orthogonal'' matrix $R$, which can be parameterized by\footnote{For $w\in\mathbb C$, $\sin z_1=w$ and $\cos z_2=w$ always have solutions $z_{1,2}\in\mathbb C$.
 Meanwhile, $\tan z=w$ has no solution if and only if $w=\pm \ii$.
@@
 
+With this parameterization, the neutrino mixing matrix is given by, in the basis with $\MN$ being diagonal,
+\begin{equation}
+ U \approx \pmat{
+  U\w{PMNS}
+  &  \MD^\dagger \MN^{*-1} U_X
+  \\ -\MN^{-1} \MD U\w{PMNS}
+  &  U_X
+}
+\approx\pmat{U\w{PMNS}&0\\0&1}\pmat{
+  1
+  &  -(-\ii{m\w H^{-1/2}}R{m\w L^{1/2}})^\dagger
+  \\ -\ii m\w H^{-1/2}R{m\w L^{1/2}}
+  &  1
+}.
+\end{equation}
+
 \paragraph{Example: two right-handed neutrinos}
@@
 \begin{equation}
- \MD = \ii \sqrt{M\w H^\text{diag}}R\sqrt{M\w L^\text{diag}}(U^{\text{Majorana}}\w{PMNS} )^\dagger
+ \frac{v}{\sqrt2}Y_n
+  \approx \ii U_X^*\sqrt{m\w H}R\sqrt{m\w L}U\w{PMNS}^\dagger
 \end{equation}

### 2021-02-17 lib/minted_formatter.py
configure minted

@@ -1,0 +1,208 @@
+# -*- coding: utf-8 -*-
+from pygments.formatters.latex import LatexFormatter
+import re
+
+
+class Wrapper():
+    def __init__(self, tokensource):
+        self._source = tokensource
+
+    def wrap(self, text):
+        return re.sub(r'\\\[(\w+)\]', lambda m: dictionary.get(m[1], m[0]), text)
+
+    def __iter__(self):
+        for category, token in self._source:
+            yield category, self.wrap(token)
+
+
+class CustomFormatter(LatexFormatter):
+    def __init__(self, **options):
+        super().__init__(**options)
+
+    def format_unencoded(self, tokensource, outfile):
+        super().format_unencoded(Wrapper(tokensource), outfile)
+
+
+dictionary = {
+    "Aleph" : u'ℵ',
+    "Alpha" : u'α',
+    "Angle" : u'∠',
+    "Angstrom": u'Å',
+    "AscendingEllipsis": u'⋰',
+    "Backslash": u'∖',
+    "BeamedEighthNote": u'♫',
+    "BeamedSixteenthNote": u'♬',
+    "Bet": u'ℶ',
+    "Beta": u'β',
+    "Bullet": u'•',
+    "CapitalAlpha": u'Α',
+    "CapitalBeta": u'Β',
+    "CapitalChi": u'Χ',
+    "CapitalDelta": u'Δ',
+    "CapitalDigamma": u'Ϝ',
+    "CapitalEpsilon": u'Ε',
+    "CapitalEta": u'Η',
+    "CapitalGamma": u'Γ',
+    "CapitalIota": u'Ι',
+    "CapitalKappa": u'Κ',
+    "CapitalKoppa": u'Ϟ',
+    "CapitalLambda": u'Λ',
+    "CapitalMu": u'Μ',
+    "CapitalNu": u'Ν',
+    "CapitalOmega": u'Ω',
+    "CapitalOmicron": u'Ο',
+    "CapitalPhi": u'Φ',
+    "CapitalPi": u'Π',
+    "CapitalPsi": u'Ψ',
+    "CapitalRho": u'Ρ',
+    "CapitalSampi": u'Ϡ',
[…]

### 2021-02-17 mathematics.tex
clarify range of angles in unitary matrix
rewrite CIP
more on Spinors and spins

@@
 \subsection{Matrix exponential}
-Excerpted from \S2 and \S5 of Hall 2015 \cite{hall2015}:
+Excerpted from \S2 and \S5 of Hall 2015 \cite{Hall2015}:
 \begin{alignat}{2}
@@
 -\si\theta\ee^{\ii(\alpha+\beta)} & \co\theta\ee^{\ii(\alpha+\gamma)}
-}
+},\qquad
+0\le\theta\le\frac{\pi}2,\quad \alpha,\beta,\gamma\in\mathbb R;
 \end{equation}
@@
 \end{align}
-(see, e.g., Ref.~\cite{Rasin:1997pn}).
+with $0\le \theta_{ij}\le\pi/2$ and $a,b,c,d,e,\delta\in\mathbb R$ (see, e.g., Ref.~\cite{Rasin:1997pn}).
 
@@
  D: \text{non-negative real diagonal matrix}~(D_{ii}\ge0).
+ \label{eq:ATF}
 \end{equation}
@@
 \end{minted}
+
+\subsection{Group theory}\label{sec:group-theory}
+
+\paragraph{Lie Groups}
+This section is mainly based on Ref.~\cite{Hall2015}. Also, $\KK=\mathbb{R}$ or $\mathbb{C}$ and $M^n_\KK=\KK^{n\times n}$.
+
+With $\Omega=\spmat{0&I_{n/2}\\-I_{n/2}}$ and $H=\spmat{I_p&0\\0&-I_q}$,
+\begin{align}
+ \mathop{\mathrm  {GL}}(n;\KK) &= \{g\in M^n_\KK\mathbin|\det g\neq 0\},&
+ \mathop{\mathfrak{gl}}(n;\KK) &= \{a\in M^n_\KK\},\\
+%
+ \mathop{\mathrm  {SL}}(n;\KK) &= \{g\in M^n_\KK\mathbin|\det g=1\},&
+ \mathop{\mathfrak{sl}}(n;\KK) &= \{a\in M^n_\KK\mathbin|\Tr a=0\},\\
+%
+ \mathop{\mathrm  {SU}}(n) &= \{g\in M^n_{\mathbb C}\mathbin|g^\dagger g=1\land\det g=1\},&
+ \mathop{\mathfrak{su}}(n) &= \{a\in M^n_{\mathbb C}\mathbin|a+a^\dagger=0\land\Tr a=0\},\\
+%
+ \mathop{\mathrm  {SO}}(n) &= \{g\in M^n_{\mathbb R}\mathbin|g^\TT g=1\land\det g=1\},&
+ \mathop{\mathfrak{so}}(n) &= \{a\in M^n_{\mathbb R}\mathbin|a+a^\TT=0\},\\
+%
+ \mathop{\mathrm  {SO}}(n;\mathbb C) &= \{g\in M^n_{\mathbb C}\mathbin|g^\TT g=1\land\det g=1\},&
+ \mathop{\mathfrak{so}}(n;\mathbb C) &= \{a\in M^n_{\mathbb C}\mathbin|a+a^\TT=0\},\\
+%
+ \mathop{\mathrm  {SO}}(p,q) &= \{g\in M^{p+q}_{\mathbb R}\mathbin|H g^\TT H=g^{-1}\land\det g=1\},&
+ \mathop{\mathfrak{so}}(p,q) &= \{a\in M^{p+q}_{\mathbb R}\mathbin|H a^\TT H=-a\},\\
+%
+ \mathop{\mathrm  {Sp}}(n;\KK) &= \{g\in M^n_\KK\mathbin|\Omega g^\TT \Omega=-g^{-1}\},&
+ \mathop{\mathfrak{sp}}(n;\KK) &= \{g\in M^n_\KK\mathbin|\Omega a^\TT \Omega=a\},\\
+%
+ \mathop{\mathrm  {Sp}}(n) &= \{g\in M^n_{\mathbb C}\mathbin|\epsilon g^\TT \epsilon=-g^{-1}\land g^\dagger g=1\},&
+ \mathop{\mathfrak{sp}}(n) &= \{g\in M^n_{\mathbb C}\mathbin|\epsilon a^\TT \epsilon=a\land a^\dagger+a=0\}.
+\end{align}
+Of course, $\mathop{\mathrm{U}}(n)$ and $\mathop{\mathrm{O}}(n;\KK)$ (and corresponding Lie algebra) are obtained by removing the condition $\det g=1$ (and $\Tr a=0$).
+
+Important identities are:
+\begin{equation}
+ \gSp(2;\KK)=\gSL(2;\KK),\quad \gSp(2)=\gSU(2),
+\end{equation}
+
+The complexification of a finite-dimensional \emph{real} vector space $V$ is defined by
+\begin{equation}
+ V_{\mathbb C}=V\otimes\mathbb C=\{v_1+\ii v_2\mathbin|v_1,v_2\in V\}\cong V+\ii V.
+\end{equation}
+Then, complexification of a Lie algebra $\mathfrak a$ of a matrix Lie group is defined as follows.
+Since $\mathfrak a$ is a \emph{real} Lie algebra, it can be complexified to $\mathfrak a_{\mathbb C}$, where its bracket operation has a unique extension so that $\mathfrak a_{\mathbb C}$ is a \emph{complex} lie algebra.
+The complex Lie algebra $\mathfrak a_{\mathbb C}$ is called the complexification of $\mathfrak a$.
+In particular,
+\begin{align}\notag
+ &
+ \mathop{\mathfrak{gl}}(n;\mathbb R)_{\mathbb C}
+ \cong \mathop{\mathfrak{u}}(n)_{\mathbb C}
+ \cong \mathop{\mathfrak{gl}}(n;\mathbb C),&
+ &
+ \mathop{\mathfrak{so}}(n)_{\mathbb C}
+ \cong \mathop{\mathfrak{so}}(n;\mathbb C),\\
+ &
+ \mathop{\mathfrak{sl}}(n;\mathbb R)_{\mathbb C}
+ \cong \mathop{\mathfrak{su}}(n)_{\mathbb C}
+ \cong \mathop{\mathfrak{sl}}(n;\mathbb C),&
+ &
+ \mathop{\mathfrak{sp}}(n;\mathbb R)_{\mathbb C}
+ \cong \mathop{\mathfrak{sp}}(n)_{\mathbb C}
+ \cong \mathop{\mathfrak{sp}}(n;\mathbb C).
+\end{align}
+On the other hand, a complex simple Lie algebra $\mathfrak a$ can be, on the other hand, viewed as a real Lie algebra, whose (real) dimension is twice as the original (complex) dimension.
+We describe such real Lie algebra by $\mathfrak a_{\mathbb R}$.
+
+Important identities are:
+\begin{align}
+ &\aSO(3,1)\cong \aSL(2;\mathbb C)_{\mathbb R},&
+ &\aSO(3,1)_{\mathbb C}\cong \aSL(2;\mathbb C)\oplus\aSL(2;\mathbb C).
+\end{align}
+
+\paragraph{Clifford algebra and spin group}\TODO{needs experts...}
+We just provide a little information on spin group $\mathop{\mathrm Spin}(1,3)$ associated to (our) Minkowski metric $\eta=\diag(+,-,-,-)$:~\cite{RauschdeTraubenberg:2005aa,Wikipedia:Clifford_Algebra}.
+To define the Clifford algebra $\mathfrak C(V,Q)$, let
+\begin{itemize}
+ \item[$V$]: a vector space over a field $K$; in our case, $V=\mathbb R^{1,3}$ and $K=\mathbb R$.
+ \item[$Q$]: a quadratic form on $V$,
+ \item[$A$]: %
+  a unital associative algebra over $K$; it means that $(A, +, \ast)$ is a ring and is also a vector space on $K$, it should satisfy
+  $(k a_1)\ast a_2=a_1\ast(ka_2)=k(a_1\ast a_2)$ for $a_i\in A$ and $k\in K$, and it should contain $1_A$ such that $1_A\ast a=a\ast 1_A=a$;
+  here a map $f:K\to A$ with $f(k)=k1_A$ is an injective ring homomorphism, with which $K$ can be included in $A$.
+\end{itemize}
+Then we can define a linear map $i:V\to \mathfrak C(V,Q)$ satisfying $i(v)\ast i(v)=Q(v)1_A$ for all $v\in V$ as follows, which defines the Clifford algebra.
+Given a linear map $j:V\to A$ satisfying $j(v)\ast j(v)=Q(v)1_A$ for all $v\in V$, there is a unique algebra homomorphism $f:\mathfrak C(V,Q)\to A$ such that $f\circ i=j$.
+
+We focus on $\mathbb R^{1,3}$ and $Q(v)=\eta_{\mu\nu}v^\mu v^\nu$ (Minkowski norm). Then, by introducing the basis $e_\mu$ of $V$, the condition yields
+\begin{align}
+& (v^2+2v^\mu w_\mu + w^2)1_A = [j(v)+j(w)]\ast[j(v)+j(w)] = (v^2+w^2)1_A+j(v)\ast j(w)+j(w)\ast j(v),\notag\\
+&\therefore \{j(v),j(w)\}=2v^\mu w_\mu 1_A
+\text{~~and thus~~} \{e_\mu,e_\nu\}=2\eta_{\mu\nu} 1_A.\label{eq:ClifAl}
+\end{align}
+The Clifford algebra $\mathfrak C(\mathbb R^{1,3},\eta)$ is thus seen as the associative algebra generated by $\{1_A, e_0, e_1, e_2, e_3\}$ satisfying \cref{eq:ClifAl}.
+
+Several properties of the Clifford algebra $\mathfrak C_{1,3}$ is in order:
+\begin{itemize}
+ \item It is of dimension $2^4=16$ and the basis is given by
+$\{1, e_\mu, (e_\mu e_\nu)_{\mu<\nu}, (e_\mu e_\nu e_\rho)_{\mu<\nu<\rho}, e_0e_1e_2e_3\}.$
+ \item It contains a subalgebra $\mathfrak C^0_{1,3}$ spanned by $\{1, e_\mu e_\nu, e_0e_1e_2e_3\}$, while the remaining vector space $\mathfrak C^1_{1,3}$ is by $\{e_\mu, e_\mu e_\nu e_\rho\}$.
+ \item We define $\epsilon=e_0e_1e_2e_3$, which satisfies $\epsilon^2=-1$ and $[\epsilon,\mathfrak C^0_{1,3}]=\{\epsilon,\mathfrak C^1_{1,3}\}=0$.
+ \item $\aSO(1,3)\subset\mathfrak C_{1,3}$; explicitly, the generators are given by $S_{\mu\nu}:={[e_\mu,e_\nu]}/{4}$, which satisfy
+\begin{equation}
+ [S_{\mu\nu},e_\rho] = e_\mu\eta_{\nu\rho}-e_\nu\eta_{\mu\rho},\qquad
+ [S_{\mu\nu},S_{\rho\sigma}] = 
+-\eta_{\mu\rho}S_{\nu\sigma}
++\eta_{\nu\rho}S_{\mu\sigma}
++\eta_{\mu\sigma}S_{\nu\rho}
+-\eta_{\nu\sigma}S_{\mu\rho}.
+\end{equation}
+\end{itemize}
+
+The Clifford group is defined by
+\begin{equation}
+\Gamma_{1,3}=\{x\in\mathfrak C_{1,3}\mathbin|\forall v\in \mathbb R^{1,3}, xvx^{-1}\in \mathbb R^{1,3}\},
+\end{equation}
+i.e., $x$ should return a ``vector'' $v^\nu e_\nu\in \mathbb R^{1,3}$ into another ``vector'', which in fact corresponds to the Lorentz transformation:
+$xe_\nu x^{-1} = (\Lambda^{-1})_\nu{}^\mu e_\mu$.
+The correspondence is however one-to-$N$, and thus it should be normalized by the squared-norm $N(x)$
+\begin{equation}
+ N(x)=x\overline{x};\qquad\overline{e_\mu}=e_\mu,\quad\overline{ke_\mu e_\nu e_\rho}=ke_\rho e_\nu e_\mu, \text{etc., where $k\in\mathbb R$}.
+\end{equation}
+Then, we define
+\begin{align*}
+ \mathop{\mathrm{Pin}}(1,3)&=\{x\in\Gamma_{1,3}\mathbin||N(x)|=1\},&
+ \mathop{\mathrm{Spin}}(1,3)&=\mathop{\mathrm{Pin}}(1,3)\cap \mathfrak C^0_{1,3},&
+ \mathop{\mathrm{Spin}}(1,3)^+&=\{x\in\mathop{\mathrm{Spin}}(1,3)\mathbin|N(x)=1\},
+\end{align*}
+whose properties are in order:
+\begin{itemize}
+ \item $\mathop{\mathrm{[S]pin}}(1,3)$ is a double covering of $\mathop{\mathrm{[S]O}}(1,3)$ but it is not connected.
+ \item $\mathop{\mathrm{Spin}}(1,3)^+$ is a connected Lie group and is a double covering of $\gSO(1,3)^+$.
+ \item $\mathop{\mathrm{Spin}}(1,3)^+\cong\gSL(2;\mathbb C)\cong\gSp(2;\mathbb C)$.
+\end{itemize}
+
 \end{document}

### 2021-02-17 spinors.tex
more on Spinors and spins

@@
 \documentclass[CheatSheet]{subfiles}
+
+\newcommand{\LorTr}[2]{\Lambda^{#1}{}_{#2}}
+\newcommand\defeq{\mathrel{\stackrel{\text{def}}=}}
 \begin{document}
@@
 \end{equation}
+
+\subsection{Verbose derivation}
+We here derive the fermion convention under the most generic with signs $h_i=\pm1$, following
+Refs.~\cite{Kugo1}.
+
+\paragraph{Lorentz group and Lorentz tensors}
+The Lorentz transformation $\LorTr\mu\nu$ is defined as a linear transformation $x^\mu\mapsto\LorTr\mu\nu x^\nu$ that conserves $x^2=\eta_{\mu\nu}x^\mu x^\nu$, where $x^\mu$ is a spacetime point and $\eta$ is the Minkowski metric:
+\begin{equation}
+ \eta^{\mu\nu}=\eta_{\mu\nu}\defeq h_\eta\times\diag(+1,-1,-1,-1),\qquad
+ \eta^{\mu\alpha}\eta_{\alpha\nu}=\delta^{\mu}_\nu;\qquad
+ \eta_{\rho\sigma}\defeq\eta_{\mu\nu}\LorTr\mu\rho\LorTr\nu\sigma ~\text{(defining equation)}.
+\label{eq:LorTr}
+\end{equation}
+Its nice to denote its inverse, which satisfies $x_\nu\mapsto x_\mu(\Lambda^{-1})^\mu{}_\nu$, by $\Lambda_\nu{}^\mu$:
+\begin{equation}
+ (\Lambda^{-1})^\mu{}_\nu = \eta^{\mu\alpha}\LorTr\beta\alpha\eta_{\beta\nu}
+\equiv \Lambda_\nu{}^\mu;\qquad
+x_\mu\mapsto \Lambda_\mu{}^\nu x_\nu,\qquad
+\delta^\alpha_\beta
+%=(\Lambda^{-1})^\alpha{}_\mu\LorTr\mu\beta
+%=\LorTr\alpha\mu (\Lambda^{-1})^\mu{}_\beta
+=\Lambda_\mu{}^\alpha\LorTr\mu\beta
+=\LorTr\alpha\mu \Lambda_\beta{}^\mu.
+\end{equation}
+They form a group $L\cong\gO(1,3)$ (Lorentz group), which has four disconnected parts:
+\begin{align}
+ L\w 0  &=\{\Lambda\mathbin|\det\Lambda=+1, \LorTr00\ge1\}\cong \mathop{\mathrm{SO^+}}(1,3),&
+ L\w P  &=\{\Lambda\mathbin|\det\Lambda=-1, \LorTr00\ge1\},\notag\\
+ L\w T  &=\{\Lambda\mathbin|\det\Lambda=+1, \LorTr00\le-1\},&
+ L\w{PT}&=\{\Lambda\mathbin|\det\Lambda=-1, \LorTr00\le-1\}.
+\end{align}
+
+Tensors $T^{\mu_1\mu_2\cdots}_{\nu_1\nu_2\cdots}$ and pseudo-tensors $\tilde T^{\mu_1\mu_2\cdots}_{\nu_1\nu_2\cdots}$ are objects that satisfy
+\begin{equation}
+ T^{\mu_1\mu_2\cdots}_{\nu_1\nu_2\cdots}\mapsto
+\LorTr{\mu_1}{\alpha_1}\cdots
+\Lambda_{\nu_1}{}^{\beta_1}\cdots
+ T^{\alpha_1\alpha_2\cdots}_{\beta_1\beta_2\cdots},
+\qquad
+ \tilde T^{\mu_1\mu_2\cdots}_{\nu_1\nu_2\cdots}\mapsto(\det\Lambda)
+\LorTr{\mu_1}{\alpha_1}\cdots
+\Lambda_{\nu_1}{}^{\beta_1}\cdots
+ \tilde T^{\alpha_1\alpha_2\cdots}_{\beta_1\beta_2\cdots}.
+\end{equation}
+There are two constants that qualify to be (pseudo-)tensors: $\eta^{\mu\nu}$ (and $\eta_{\mu\nu}$, $\delta^{\mu}_\nu$) and the anti-symmetric tensor $\vep^{\mu\nu\rho\sigma}$:
+\begin{equation}
+\eta^{\mu\nu}\mapsto \LorTr\mu\alpha\LorTr\mu\beta\eta^{\alpha\beta}=\eta^{\mu\nu},\qquad
+ \vep^{\mu\nu\rho\sigma}\mapsto \LorTr\mu\alpha\LorTr\nu\beta\LorTr\rho\gamma\LorTr\sigma\delta\vep^{\alpha\beta\gamma\delta}=(\det\Lambda)\vep^{\mu\nu\rho\sigma};\qquad
+ \vep^{0123}\defeq1,\quad \vep_{0123}=-1.
+\end{equation}
+
+\paragraph{Infinitesimal transformation and Lorentz algebra}
+Equation~\eqref{eq:LorTr} gives the representation for a proper orthochronous Lorentz transformation $\Lambda\in L_0$:
+\begin{equation}
+ \Lambda= 1 + \lambda + \Order(\lambda^2);\quad
+\lambda=\spmat{
+ 0 & -\omega_x & -\omega_y & -\omega_z \\
+ -\omega_x & 0        &+\theta_z &-\theta_y\\
+ -\omega_y &-\theta_z & 0        &+\theta_x\\
+ -\omega_z &+\theta_y &-\theta_x & 0
+}
+~\defeq~
+\ii\vc{\omega}\cdot\vc K+\ii\vc\theta\cdot\vc J
+~\defeq~
+\frac{-\ii}{2}d^{\mu\nu}M_{\mu\nu}
+\label{eq:LorTrInf}
+\end{equation}
+where $\theta_i$ is the angle of a \emph{passive} rotation around $i$-axis and $\omega_i$ describes the \emph{passive} boost along $i$-axis with velocity $\beta=\tanh\omega$.
+
+The last definition of Eq.~\eqref{eq:LorTrInf} is used to consider Lorentz algebra $\{M\}$.
+As $[M_{\rho\sigma}]{}^\mu{}_\nu$ should be a tensor, the parameter $d$ should also be a tensor and thus $d^{\mu\nu}=k\cdot\eta^{\mu\rho}\lambda{}^\nu{}_\rho$ ($k$ is a constant), i.e.,
+\begin{equation}
+ M_{\mu\nu}=-M_{\nu\mu}, \qquad
+ d^{\mu\nu}=-d^{\nu\mu}, \qquad
+ \{d^{01},d^{02},d^{03}\}=kh_\eta\times\vc\omega,\qquad
+ \{d^{32},d^{13},d^{21}\}=kh_\eta\times\vc\theta.
+\end{equation}
+Then the element of the Lorentz algebra is given by
+$ (M_{\rho\sigma}){}^\mu{}_\nu=
+(\ii/k)(\delta^\mu_\rho\eta_{\nu\sigma}-\delta^\mu_\sigma \eta_{\nu\rho})
+$, which gives the Lorentz algebra
+\begin{equation}
+[M_{\mu\nu},M_{\rho\sigma}] = 
+(-\ii/k)\left(\eta_{\mu\rho}M_{\nu\sigma}-\eta_{\mu\sigma}M_{\nu\rho}
++\eta_{\nu\sigma}M_{\mu\rho}-\eta_{\nu\rho}M_{\mu\sigma}
+\right).
+\label{eq:LorentzAlgebra}
+\end{equation}
+We will take $k=+1$ to match the notation of Ref.~\cite{Kugo1}.
+
+\paragraph{Rotation and boost}
+The boost and rotation operators, $\vc K$ and $\vc J$, are now given in an abstract form by
+\begin{equation}
+ \vc J = h_\eta (M_{23}, M_{31}, M_{12}), \qquad
+ \vc K = -h_\eta(M_{01}, M_{02}, M_{03}),
+\end{equation}
+so their commutation relation is read from \cref{eq:LorentzAlgebra}:
+\begin{align}
+ [J_i, J_j] &= \ii \epsilon_{ijk}J_k,&
+ [J_i, K_j] &= \ii \epsilon_{ijk}K_k,&
+ [K_i, K_j] &= -\ii \epsilon_{ijk}J_k,
+\end{align}
+which leads to
+\begin{align}
+ \vc A&\defeq\frac{\vc J+\ii \vc K}{2},&
+ \vc B&\defeq\frac{\vc J-\ii \vc K}{2};&
+ [A_i, A_j] &= \ii \epsilon_{ijk}A_k,&
+ [B_i, B_j] &= \ii \epsilon_{ijk}B_k,&
+ [A_i, B_j] &= 0.
+\end{align}
+This means $\aSO(1,3)$ is somewhat similar to $\aSU(2)\oplus \aSU(2)$, or in fact,
+as discussed in \cref{sec:group-theory},
+$\aSO(1,3)_{\mathbb C}\cong \aSU(2)_{\mathbb C}\oplus\aSU(2)_{\mathbb C}$.
+
+\paragraph{More isomorphism}
+Let us see a bit more of  mathematical structure, following the discussion in \cref{sec:group-theory}.
+The isomorphic groups
+$\mathop{\mathrm{Spin}}(1,3)^+\cong\gSL(2,\mathbb C)\cong\gSp(2,\mathbb C)$
+is a double cover of $L_0$;
+in particular, $L_0\cong \mathop{\mathrm{PSL}}(2;\mathbb C)=\gSL(2;\mathbb C)/\mathbb Z_2$.
+
+Meanwhile, the Lorentz algebra $\aSO(1,3)$ is isomorphic to $\aSL(2;\mathbb C)$ viewed as a real Lie albegra~\cite[\S7.8]{Hall2015},
+and its complexification
+$\aSO(1,3)_{\mathbb C}$ is isomorphic to $
+\aSU(2)_{\mathbb C}\oplus\aSU(2)_{\mathbb C}
+=\aSL(2,\mathbb C)\oplus\aSL(2,\mathbb C)$.
+
+\paragraph{Spinors from Clifford algebra}
+First we construct spinors from Clifford algebra $\mathfrak C_{1,3}$ given in \cref{sec:group-theory}. A matrix representation $\gamma^\mu$ of $\mathfrak C_{1,3}$ and spinors $\psi$ are introduced:
+\begin{equation}
+ \overline\psi = \psi^\dagger\beta, \quad\beta\beta=1,\qquad
+\{\gamma^\mu,\gamma^\nu\}=2\eta^{\mu\nu}, \quad
+\gamma_5 = h_5\times\ii\gamma^0\gamma^1\gamma^2\gamma^3; \qquad
+\psi\mapsto T\psi,\quad
+\overline\psi\mapsto \overline\psi\beta T^\dagger \beta;\quad T\in \Gamma_{1,3}.
+\end{equation}
+Thus, in order that $\overline\psi\psi$ and $\overline\psi\gamma^\mu\psi$ be respectively scalar and vector, $T$ should satisfy
+\begin{equation}
+ T^{-1}\gamma^\mu T = \LorTr\mu\nu\gamma^\nu,\quad
+ \beta T^\dagger \beta T=1,
+\end{equation}
+or in infinitesimal form $T=1+(-\ii/2)d^{\mu\nu}S_{\mu\nu}$,
+\begin{equation}
+ (S_{\mu\nu})^\dagger = \beta S_{\mu\nu} \beta,\quad
+[S_{\mu\nu},\gamma^\alpha] = -(M_{\mu\nu})^{\alpha}{}_\beta\gamma^\beta;
+\qquad
+\therefore S_{\mu\nu}=\frac{\ii}{4}[\gamma_\mu,\gamma_\nu];
+\quad
+[\gamma_\mu^\dagger, \gamma_\nu^\dagger]=\beta[\gamma_\mu,\gamma_\nu]\beta;
+\end{equation}
+finally, a representation of the Lorentz group is constructed as
+\begin{equation}
+\Lambda\stackrel{\text{rep}}=
+ \exp\left(\frac{-\ii}{2}d^{\mu\nu}S_{\mu\nu}\right);\qquad
+ [S_{\mu\nu},S_{\rho\sigma}] = -\ii\left(
+\eta_{\mu\rho} S_{\nu\sigma}
+-\eta_{\nu\rho} S_{\mu\sigma}
++\eta_{\mu\sigma} S_{\nu\rho}
+-\eta_{\nu\sigma} S_{\mu\rho}\right),
+\end{equation}
+which is exactly the same as \cref{eq:LorentzAlgebra}.
 

### 2021-02-20 mathematics.tex
more on spinors

@@
 \paragraph{Clifford algebra and spin group}\TODO{needs experts...}
-We just provide a little information on spin group $\mathop{\mathrm Spin}(1,3)$ associated to (our) Minkowski metric $\eta=\diag(+,-,-,-)$:~\cite{RauschdeTraubenberg:2005aa,Wikipedia:Clifford_Algebra}.
+We just provide a little information on spin group $\mathop{\mathrm Spin}(1,3)$ associated to (our) Minkowski metric $\eta=\diag(+,-,-,-)$:~\cite{RauschdeTraubenberg:2005aa,Yamaguchi:spinor,Wikipedia:Clifford_Algebra}.
 To define the Clifford algebra $\mathfrak C(V,Q)$, let

### 2021-02-20 spinors.tex
more on spinors

@@
 
-\changefontsizes{10pt}
+\summarystyle
+
 \section{Spinors}
+\begin{alignat}{2}
+&\text{Gamma matrices}:&\quad&
+\{\gamma^\mu,\gamma^\nu\}=2\eta^{\mu\nu},~
+\gamma_5=\ii\gamma^0\gamma^1\gamma^2\gamma^3;\quad
+\{\gamma^\mu,\gamma_5\}=0,~
+\gamma^5\gamma^5=1.
+\\
+&\text{conjugates}:&&
+\overline{\psi}=\psi^\dagger \beta,\qquad
+\psi^\cc=C(\overline\psi)^\TT
+\end{alignat}
 
-\changefontsizes{9pt}
+\paragraph{chiral notation}
+\begin{align}\label{eq:chiralnotation}
+& \overline{\psi}=\psi^\dagger\gamma^0;~
+ \gamma^\mu=\pmat{0&\sigma^\mu\\\bar\sigma^\mu&0},~
+ \gamma_5=\pmat{-1&0\\0&1};~
+ \PL = \frac{1-\gamma_5}{2},~
+ \PR = \frac{1+\gamma_5}{2}.
+\\&
+(\gamma^\mu)^\dagger = \gamma^0\gamma^\mu\gamma^0,~
+(\gamma^\mu)^*       = \gamma^2\gamma^\mu\gamma^2,~
+(\gamma^\mu)^\TT     = \gamma^0\gamma^2\gamma^\mu\gamma^2\gamma^0,
+\end{align}
+
+\paragraph{(cf.~Dirac notation)}
+\begin{align}\label{eq:diracnotation}
+& %\overline{\psi}=\psi^\dagger\hat\gamma^0;~
+ \hat\gamma^0=\pmat{1&0\\0&-1},~
+ \hat\gamma^i=\pmat{0&\sigma^i\\\bar\sigma^i&0},~
+ \hat\gamma_5=\pmat{0&1\\1&0};~
+ \hat\PL = \frac{1-\hat\gamma_5}{2},~
+ \hat\PR = \frac{1+\hat\gamma_5}{2}.
+\\&
+(\hat\gamma^\mu)^\dagger = \hat\gamma^0\hat\gamma^\mu\hat\gamma^0,~
+(\hat\gamma^\mu)^*       = \hat\gamma^2\hat\gamma^\mu\hat\gamma^2,~
+(\hat\gamma^\mu)^\TT     = \hat\gamma^0\hat\gamma^2\hat\gamma^\mu\hat\gamma^2\hat\gamma^0,
+\end{align}
+
 \begin{equation}
@@
 \end{equation}
+
+\newpage
+\detailstyle
 
@@
 \paragraph{More isomorphism}
-Let us see a bit more of  mathematical structure, following the discussion in \cref{sec:group-theory}.
+Let us see a bit more of  mathematical structure, following the discussion in \cref{sec:group-theory} (cf.~Refs.~\cite{RauschdeTraubenberg:2005aa,Yamaguchi:spinor}).
 The isomorphic groups
@@
 
-\paragraph{Spinors from Clifford algebra}
-First we construct spinors from Clifford algebra $\mathfrak C_{1,3}$ given in \cref{sec:group-theory}. A matrix representation $\gamma^\mu$ of $\mathfrak C_{1,3}$ and spinors $\psi$ are introduced:
+\paragraph{Representation of Clifford algebra}
+As summarized in \cref{sec:group-theory}, the spin group $\mathop{\mathrm{Spin}}(1,3)^+$ generated by Clifford algebra $\mathfrak C_{1,3}$ is a double cover of $L_0$, and thus we can consider a representation of $L_0$ based on $\mathfrak C_{1,3}$.
+
+To construct an irreducible representation of $\mathfrak C_{1,3}$, we utilize the fact that we can form two sets of creation-annihilation operators
+\begin{align}
+ a^\pm = \sqrt{h_\eta}\frac{e^0\pm e^3}{2},\quad
+ b^\pm = \sqrt{h_\eta}\frac{\pm e^2-\ii e^1}{2};\qquad
+ \{a^+,a^-\}=1,~
+ \{b^+,b^-\}=1,~
+ \{\text{(others)}\} = 0.
+\end{align}
+These ladder operator allows us to construct four states starting from $\ket{00}$, which is a non-zero state with $a^-\ket{00}=b^-\ket{00}=0$, and to construct an irreducible representation of $\mathfrak C_{1,3}$ (and, in fact, it is unique for even dimension):
+\begin{equation}
+ \ket{10} = a^+\ket{00},\quad
+ \ket{01} = b^+\ket{00},\quad
+ \ket{11} = a^+b^+\ket{00}\quad\leadsto\quad
+ a^+=\spmat{0&0&1&0\\0&0&0&1\\0&0&0&0\\0&0&0&0},\quad
+ b^+=\spmat{0&-1&0&0\\0&0&0&0\\0&0&0&1\\0&0&0&0},
+\end{equation}
+and $a^-=(a^+)^\dagger$, $b^-=(b^+)^\dagger$.
+We then obtain a representation $\gamma$, which is called ``standard representation.''\footnote{The chiral notation \eqref{eq:chiralnotation} and Dirac notation \eqref{eq:diracnotation} are equivalent to this standard representation, i.e., related by unitary matrices.}
+They are not Hermitian, but as we will see, this non-Hermiticity is solved by amending the inner product by a matrix $\beta$: $(\psi,\gamma^\mu\psi):=\psi^\dagger\beta \gamma^\mu\psi$.
+
+Although $\psi$ forms an irreducible representation $\gamma^\mu$ of $\mathfrak C_{1,3}$, the resulting representation $S_{\mu\nu}$ (see the next paragraph) is a reducible representation of $\mathop{\mathrm {Spin}}(1,3)^+$.
+This is confirmed by
+\begin{equation}
+ \gamma_5\gamma_5=1,\quad
+\{\gamma_5,\gamma^\mu\}=0,\quad
+ [\gamma_5,S_{\mu\nu}]=0;\qquad \gamma_5:=\ii\gamma^0\gamma^1\gamma^2\gamma^3,
+\end{equation}
+and $P{}^{\text L}\w{R} =(1\mp\gamma_5)/2$ works as the projection operators.
+In addition, four state are eigenstates of $J_3=h_\eta S_{12}$ because 
+\begin{equation}
+ [J_3,b^+b^-] = 0,~
+ J_3=b^+b^--\frac12,
+\end{equation}
+which also guarantees that spinors have spin $1/2$.
+In summary,
+\begin{align}
+ &\ket{00} =\ket{-\w L},\quad
+ \ket{10} =:\ket{-\w R},\quad
+ \ket{01} =:\ket{+\w R},\quad
+ \ket{11} =:\ket{+\w L};\\
+&J_3\ket{\pm\w H}=\pm\tfrac12\ket{\pm\w H},\quad
+\PL\ket{\pm\w L}=\ket{\pm\w L},~
+\PR\ket{\pm\w R}=\ket{\pm\w R};\quad
+\PL\ket{\pm\w R}=
+\PR\ket{\pm\w L}=0.
+\end{align}
+For example, in chiral notation with $(+,-,-,-)$, the Lorentz generators $S_{\mu\nu}$ are block diagonal and $\ket{\pm\w L}$ ($\ket{\pm\w R}$) has non-zero component only in the upper (lower) two component:
+\begin{equation}
+ \ket{-\w L} = \spmat{0\\1\\0\\0},~
+ \ket{-\w R} = \spmat{0\\0\\0\\1},~
+ \ket{+\w R} = \spmat{0\\0\\\ii\\0},~
+ \ket{+\w L} = \spmat{\ii\\0\\0\\0}.
+\end{equation}
+
+\paragraph{Four-spinors and Lorentz transformation}
+The above ``theoretical'' discussion can be seen more explicitly, starting from spinors and a matrix representation $\gamma^\mu$ given by
 \begin{equation}
  \overline\psi = \psi^\dagger\beta, \quad\beta\beta=1,\qquad
-\{\gamma^\mu,\gamma^\nu\}=2\eta^{\mu\nu}, \quad
-\gamma_5 = h_5\times\ii\gamma^0\gamma^1\gamma^2\gamma^3; \qquad
+\{\gamma^\mu,\gamma^\nu\}=2\eta^{\mu\nu};\qquad
 \psi\mapsto T\psi,\quad
-\overline\psi\mapsto \overline\psi\beta T^\dagger \beta;\quad T\in \Gamma_{1,3}.
+\overline\psi\mapsto \overline\psi\beta T^\dagger \beta;\quad T\in \mathop{\mathrm{Spin}}(1,3)^+.
 \end{equation}
-Thus, in order that $\overline\psi\psi$ and $\overline\psi\gamma^\mu\psi$ be respectively scalar and vector, $T$ should satisfy
+For $\overline\psi\psi$ and $\overline\psi\gamma^\mu\psi$ to be respectively scalar and vector, $T$ should satisfy
 \begin{equation}
@@
 [\gamma_\mu^\dagger, \gamma_\nu^\dagger]=\beta[\gamma_\mu,\gamma_\nu]\beta;
+\label{eq:betacondition}
 \end{equation}
-finally, a representation of the Lorentz group is constructed as
+the first condition leads to a representation of the Lorentz group
 \begin{equation}
@@
 +\eta_{\mu\sigma} S_{\nu\rho}
--\eta_{\nu\sigma} S_{\mu\rho}\right),
+-\eta_{\nu\sigma} S_{\mu\rho}\right)
 \end{equation}
-which is exactly the same as \cref{eq:LorentzAlgebra}.
+as seen in \cref{eq:LorentzAlgebra}, while the second condition determines what $\beta$ should be, as given in, e.g., \cref{eq:chiralnotation} and \cref{eq:diracnotation}.
+
+\paragraph{Charge conjugation and Majorana spinor}
+The charge conjugation of $\psi$ is something like $\psi^*$ but should obey the same representation as $\psi$ does, i.e., $C'\psi^*$ with $C'$ being a unitary matrix such that $B\psi^*\to TB\psi^*$.
+Or it can be seen that the, the previous procedure with $\psi^*$ may generate another irreducible representation and it should be related to $\gamma^\mu$ by an unitary matrix.
+Anyway, we define the charge conjugation by
+\begin{equation}
+ \psi^\cc = C(\overline\psi)^\TT = C\beta^\TT\psi^*;\quad CC^\dagger = 1.
+\quad
+\therefore
+C^*\beta[\gamma_\mu^\dagger,\gamma_\nu^\dagger]\beta C^\TT =
+[\gamma_\mu^\TT,\gamma_\nu^\TT].
+\end{equation}
+Combining with \cref{eq:betacondition},
+\begin{equation}
+ \beta\beta=1,\quad C C^\dagger = 1,\quad
+ \beta\gamma^\mu\beta=h_\beta(\gamma^\mu)^\dagger,\quad
+ C\gamma_\mu^* C^\dagger = h_C\gamma_\mu^\dagger.
+\end{equation}
+In even dimensions, the expressions have many choices as seen in the signs $h_C$ and $h_\beta$; moreover, the sign $h\w{cc}$ defined in $(\psi^\cc)^\cc=h\w{cc}\psi$ depends on the definition.
+It is thus useful to use a specific notation for further discussion.
+
+For example, the construction of a Majorana spinor $\psi\w M$, which satisfies $(\psi\w M)^\cc=\psi\w M$, is simply done as $\psi\w{M}\propto\psi+\psi^\cc$ if $\eta\w{cc}=1$, but needs some phases if $\eta\w{cc}=-1$.
+
+\paragraph{Weyl spinor}
 

### 2022-05-09 spinors_more.tex
add more on spinors

%#!lualatex
\documentclass{CheatSheet}

\newcommand{\Lubanski}{Luba\'nski\xspace}
\newcommand{\slaP}{}
\DeclareMathOperator\rank{\mathrm{rank}}
\newcommand{\RANK}[1]{\langle #1\rangle}
\bibliographystyle{utphys28mod}
\def\EvenFootPlain{%
This document by Sho Iwamoto is licensed under the Creative Commons CC--BY--NC 4.0 International Public License.}

\begin{document}

\summarystyle

hogehoge

\clearpage

\detailstyle

\subsection{Dirac spinors for massive fermions}
\RefBox{\S2 of \cite{kugo1}; \cite{Kryuchkov:2015acw}}
Dirac equation $(\ii\slashed{\partial}-m)\psi(x)=0$ has plain-wave solutions $\tilde u(p)\ee^{-\ii p x}$, i.e., $(\slashed{p}-m)\tilde u(p)=0$ with $m>0$.
Non-zero solution is available if and only if $\det(\slashed p-m)=0\Leftrightarrow p^2=m^2$.
We hereafter fix $p^0>0$ and consider
\begin{equation}
 u(\vc p)\text{~satisfying~}(\slashed p-m)u(\vc p)=0,
\qquad
 v(\vc p)\text{~satisfying~}(\slashed p+m)v(\vc p)=0,
\end{equation}
where the general solution of the Dirac equation is given by linear combination of $\{u(\vc p)\ee^{-\ii px}, v(\vc p)\ee^{\ii px}\}_{p^0>0}$.

As $\rank(\slashed p\pm m)=2$, each equation has $4-2 = \text{two}$ linear-independent solutions\footnote{%
  Let $X^\pm=\slashed{p}\pm m$. Denote $\det A$ by $|A|$, and $\rank A$ by $\RANK{A}$.
  As Lorentz invariance guarantees $|X^\pm|$ is a function of $p^2$, $|A^+|=|A^-|$.
  Then $|A^+||A^-| = (p^2-m^2)^4$ gives $|A^\pm|=(p^2-m^2)^2$.
  One can show that $\RANK{A^\pm}=2$ if $p^2=m^2$ as follows:
\\%
Frobenius inequality:\qquad $\RANK{AB}+\RANK{BC}\le \RANK B + \RANK{ABC}$\qquad
$\therefore 2\RANK{A^\pm}=\RANK{A^\pm\gamma_5\gamma_5} + \RANK{\gamma_5A^\mp\gamma_5} \le \RANK{\gamma_5}+\RANK{A^\pm\gamma_5\gamma_5A^\mp\gamma_5}=4$,
\\%
rank subadditivity:\qquad$\RANK{A+B}\le\RANK{A}+\RANK{B}$\qquad
$\therefore4\le\RANK{A^+}+\RANK{-A^-}.$

Similar discussion applies for Pauli--\Lubanski operator: as $[\det(\gamma_5\slashed{e}-x)]^2=-e^2-x^2$, it has eigenvalues $\pm1/2$ for $u(\vc p)$ and $v(\vc p)$, and 
$B^\pm=(\gamma_5\slashed e\slashed p\pm m/2)$ both have rank-2.
Furthermore, \TOCHECK as $B^+$ should have solutions in both subspaces spanned by $u(\vc p)$ and $v(\vc p)$, each subspace is spanned by eigenstates with different eigenvalues for the Pauli--\Lubanski operator and thus the label is used to distinguish two $u$s (and $v$s).
}, for which we introduce another label $s=1,2$.
We use the following convention, noting the inner product of this vector space is $(\psi,\psi)=\overline\psi\psi$:
\begin{equation}
 v^s(\vc p)\coloneq C[\overline{u}^s(\vc p)]^\TT,\qquad
 \overline u^s(\vc p)u^t(\vc p)\coloneq 2m\delta^{st}.
\end{equation}
Then $\gamma^\mu v = (h_\beta/h_C)C\beta^\TT(\gamma^\mu u)^*$, for which $v(\vc p)$ satisfies $(\slashed{p}+m)v(\vc p)=0$ as requested.
Orthogonality for $s\neq t$ is guaranteed by its definition given later.
Also, as $u$ and $v$ have different eigenvalues of the matrix $\slashed p$,
\begin{equation}
  \overline v^s(\vc p)v^t(\vc p)=-2m\delta^{st},\qquad
 \overline u^s(\vc p)v^t(\vc p)=\overline v^s(\vc p)u^t(\vc p)=0.
\end{equation}

Several conventions for the label $s$ are available:\vspace{-1em}
\begin{DownPara}
\paragraph{Pauli--\Lubanski pseudovector operator} gives the most generic definition for the spin of a moving particle:
\begin{equation}
 w^\mu = \frac1{2m}\epsilon^{\mu\nu\rho\sigma}P_\nu M_{\rho\sigma}.
\end{equation}
With a reference vector $e^\mu$ such that $e^2=-1$ and $e^\mu p_\mu=0$,
\begin{equation}
 e_\mu w^\mu
= e_\mu \frac1{2m} \epsilon^{\mu\nu\rho\sigma}p_\nu \frac{\ii}{4}[\gamma_\rho,\gamma_\sigma]
= \frac\ii{4m} \epsilon^{\mu\nu\rho\sigma}e_\mu p_\nu \gamma_\rho\gamma_\sigma
= \frac1{2m}\gamma_5 \slashed e \slashed p
\end{equation}
commutes with $(\slashed p\pm m)$ and thus $s=\pm$ may denote the eigenvalue of this operator $\pm1/2$.

\paragraph{Helicity operator}
A simpler option is the eigenvalues $\pm1$ of the helicity operator $h=\vc\sigma\cdot\vc p/\|\vc p\|=:\vc\sigma\cdot\vc n$.
This is in fact a special case of Pauli--\Lubanski operator with $e^\mu= (\|\vc p\|/m,p_0\vc n/m)$, which is verified by
\begin{align}
& h
=\epsilon^{0\nu\rho\sigma}(-n_\nu) S_{\rho\sigma}
=\frac{1}{2\ii}\epsilon^{0\nu\rho\sigma}n_\nu \gamma_\rho\gamma_\sigma
=\frac{\eta^{\mu0}}{2\ii}\epsilon_{\mu\nu\rho\sigma}n^\nu \gamma^\rho\gamma^\sigma
=\frac{\gamma_5}{2}\gamma_5[\slashed{n},\gamma^0]
=\frac{\gamma_5}{\|\vc p\|}({\{\slashed{p},\gamma^0\}}/{2}-\gamma^0\slashed{p});
\\&
w_\mu e^\mu 
=\frac{1}{2\|\vc p\|}\gamma_5(p^0-\gamma^0\slashed{p})\sim \frac{h}{2}.
\end{align}

% Low\[Epsilon][m_,n_,r_,s_]:=-LeviCivitaTensor[4][[m+1,n+1,r+1,s+1]]
% Low\[Epsilon][0,1,2,3]
% Sum[Low\[Epsilon][m,n,r,s]S[m]p[n]GammaMatrix[r,None] . GammaMatrix[s,None],{m,0,3},{n,0,3},{r,0,3},{s,0,3}];
% -2I LorSum[S[#]GammaMatrix[None,#]&] . LorSum[p[#]GammaMatrix[None,#]&] . GammaFive;
% Expand[%%-%]//.{p[0] S[0]->p[1]S[1]+p[2]S[2]+p[3]S[3]}//Simplify

\TODO{Discuss Poincare group before introducing PL operator.: \cite{WeinbergQFT1}, \cite{Kryuchkov:2015acw}}

\TODO{We have not defined the relation between $S_{\mu\nu}$ and $\sigma_i$ etc.; verify for more general case, including the proof of $\epsilon_{\mu\nu\rho\sigma}\gamma^\rho\gamma^\sigma=-\ii\gamma_5[\gamma_\mu,\gamma_\nu]$.}
\end{DownPara}

Projection operators for the subspaces spanned by $u^s$ ($v^s$) are given by
\begin{equation}
 P_u=\frac{m+\slashed p}{2m}, \quad P_v=\frac{m-\slashed p}{2m}.
\end{equation}
As $\{u^1, u^2, v^1, v^2\}$ is a basis of the vector space $\mathbb C^4$ with inner product $(\psi,\psi)=\overline\psi\psi$, we immediately have 
\begin{equation}
 \sum_{s=1,2}u^s(\vc p)\overline u^s(\vc p)=P_u \overline u^s(\vc p)u^s(\vc p) = \slashed p+m,\qquad
 \sum_{s=1,2}v^s(\vc p)\overline v^s(\vc p)=\slashed p-m.
\end{equation}
Similarly, using projection operator $P'_{\pm}=(1\pm\gamma_5\slashed e\slashed p/m)/2$,
\begin{align}
 &u^s(\vc p)\overline u^s(\vc p) = 2m P'_{\pm}P_u = \frac12(\slashed p+m)(1+s\gamma_5\slashed e),\\\
 &v^s(\vc p)\overline v^s(\vc p) = C(u^s\overline u^s)^\TT \beta^*C^\dagger\beta
=\frac12(m+h_C \slashed p)(1-s h_C \gamma_5\slashed e )C\beta^*C^\dagger \beta
\leadsto
\frac12(\slashed p-m)(1+s\gamma_5\slashed e)
\end{align}
for $s=\pm1$.

\subsection{Chiral notation}
In the chiral notation, $\beta=\gamma^0$ and $C=-\ii\gamma^2\gamma^0$, and $h_\beta=1$ and $h_C=-1$, the plain-wave spinors are given by
\begin{equation}
 u^s(\vc p) = \pmat{\sqrt{\sigma\cdot p}\xi^s\\\sqrt{\bar\sigma\cdot p}\xi^s},\quad
 v^s(\vc p) = \pmat{\sqrt{\sigma\cdot p}\eta^s\\-\sqrt{\bar\sigma\cdot p}\eta^s};\qquad
 \eta^s = \pmat{\eta^s_1\\\eta^s_2} \coloneq  \pmat{-(\xi^s_2)^*\\(\xi^s_1)^*},\quad
(\xi^s)^\dagger(\xi^t) = (\eta^s)^\dagger(\eta^t) = \delta^{st},
\end{equation}
where $\{\xi^1,\xi^2\}$ are the orthonormal basis, which fixes the definition of $\eta$, and\footnote{%
The notation is justified by
\begin{equation*}
\sqrt{\sigma\cdot p}\sqrt{\sigma\cdot p} ={\sigma\cdot p},\qquad
\sqrt{\bar\sigma\cdot p}\sqrt{\bar\sigma\cdot p} ={\bar\sigma\cdot p},\qquad
\sqrt{\sigma\cdot p}\sqrt{\bar\sigma\cdot p}=m.
\end{equation*}}
\begin{equation}
  \sqrt{\sigma\cdot p} \coloneq  \frac{m+p^\mu\sigma_\mu}{\sqrt{2(m+p^0)}},
\qquad
 \sqrt{\bar\sigma\cdot p} \coloneq  \frac{m+p^\mu\bar\sigma_\mu}{\sqrt{2(m+p^0)}}.
\end{equation}
\footnote{It is straightforward to verify the above-given equations:
\begin{align}
&
 \overline u u = 2m\delta^{st},\quad
 \overline v v = -2m\delta^{st},\quad
 \overline u v = \overline v u = 0,\quad
 u^\dagger u = v^\dagger v = 2p^0\delta^{st},\quad
\\&
 u^\dagger(\vc p)u(-\vc p)= v^\dagger(\vc p)v(-\vc p)=2m\delta^{st},\quad
 u^\dagger(\vc p)v(-\vc p)= v^\dagger(\vc p)u(-\vc p)=0,\quad
 \overline u(\vc p)u(-\vc p)= -\overline v(\vc p)v(-\vc p)=2p^0\delta^{st}.
\end{align}
}
So that the basis $\xi$ has eigenvalue $\pm1/2$ for the Pauli--\Lubanski operator, it should satisfy
\begin{equation}
 \frac12 \gamma_5\slashed{e} U \pmat{\xi^s\\\xi^s} = \left(\pm\frac12\right)U\pmat{\xi^s\\\xi^s};\qquad
\text{where~}
 U\coloneq \pmat{\sqrt{\sigma\cdot p} & 0 \\ 0 & \sqrt{\bar\sigma\cdot p}},\quad
 U^{-1}=\frac{1}{m}\pmat{\sqrt{\bar\sigma\cdot p} & 0 \\ 0 & \sqrt{\sigma\cdot p}}.
\end{equation}
Thus, by calculating the eigensystem of $U^{-1}\gamma_5\slashed {e} U$, one can get
\begin{equation}
 \xi^+ = \frac{1}{\sqrt{2(1-\phi^3)}}\pmat{\phi^1-\ii \phi^2 \\ 1-\phi^3},\quad
 \xi^- = \frac{1}{\sqrt{2(1-\phi^3)}}\pmat{1-\phi^3\\-\phi^1-\ii \phi^2};\qquad
 \phi^\mu = e^\mu-\frac{e^0p^\mu}{m+p^0}.
\end{equation}
If we use reference vector for the helicity, 
\begin{equation}
 \xi^+ = \frac1{\sqrt{2-2n^3}}\pmat{n^1-\ii n^2\\1-n^3}
       =\pmat{\ee^{-\ii\phi}\cos(\theta/2)\\\sin(\theta/2)},\quad
 \xi^- = \frac1{\sqrt{2-2n^3}}\pmat{1-n^3\\-n^1-\ii n^2}
       =\pmat{\sin(\theta/2)\\-\ee^{\ii\phi}\cos(\theta/2)}
\end{equation}
with $(n^1,n^2,n^3)=(\si\theta\co\phi,\si\theta\si\phi,\co\theta)$ is the direction of the momentum.

\bibliography{CheatSheet}
\end{document}

### 2022-05-09 Neutrino.tex
rewrite with new class

@@
 \documentclass[CheatSheet]{subfiles}
-
-\newcommand\MD[1][]{M_{\mathrm{D}#1}}
-\newcommand\MN[1][]{M_{N#1}}
-\newcommand{\nRHN}{n_{\text{RHN}}}
 
@@
 \paragraph{Dirac neutrino model}
-The simplest model for neutrino masses are given by extending the SM Yukawa \eqref{eq:SMyukawa} with ($\nRHN$ copies of) right-handed neutrino $N$:
+The simplest model for neutrino masses are given by extending the SM Yukawa \eqref{eq:SMyukawa} with ($n\w{RHN}$ copies of) right-handed neutrino $N$:
 \begin{align}
-   \mathcal L\w{Dirac-$\nu$}&=
+   \mathcal L\w*{Dirac-$\nu$}&=
 \left(  \overline{U}\Yu H \PL Q
@@
 \end{equation}
-where  $\Yn\in\mathbb C^{\nRHN\times3}$, $V_n\in\mathbb U_{\mathbb C}^3$, and $U_n\in\mathbb U_{\mathbb C}^{\nRHN}$.
+where  $\Yn\in\mathbb C^{n\w{RHN}\times3}$, $V_n\in\mathbb U_{\mathbb C}^3$, and $U_n\in\mathbb U_{\mathbb C}^{n\w{RHN}}$.
 The Lagrangian is now given by
@@
 \end{equation}
-Note that $\nRHN=\{2,3\}$ because models with $\nRHN<3$ yields $3-\nRHN$ massless left-handed neutrinos, while $\nRHN>3$ results in $\nRHN-3$ massless right-handed neutrinos.
+Note that $n\w{RHN}=\{2,3\}$ because models with $n\w{RHN}<3$ yields $3-n\w{RHN}$ massless left-handed neutrinos, while $n\w{RHN}>3$ results in $n\w{RHN}-3$ massless right-handed neutrinos.
 
@@
 - \overline{E}\Ye H^\dagger \PL L
-- \frac{1}{2}\overline{N}\MN N^\cc
+- \frac{1}{2}\overline{N}M_N  N^\cc
 \right)+ \text{h.c.}
@@
 - \overline{L^a}\Ye^\dagger H^{a} \PR E
-- \frac{1}{2}\overline{N}\MN N^\cc\right) + \text{h.c.},
+- \frac{1}{2}\overline{N}M_N  N^\cc\right) + \text{h.c.},
 \label{eq:type-1-lagrangian}
 \end{align}
-where $M_N$ is a $\nRHN\times\nRHN$ complex symmetric matrix.
-Models with $\nRHN\ge2$ are viable, because $\nRHN<3$ gives $3-\nRHN$ massless (Weyl) neutrinos and $2\nRHN$ massive (Majorana) neutrinos, while all neutrinos acquire Majorana mass  if $\nRHN\ge 3$.
+where $M_N$ is a $n\w{RHN}\times n\w{RHN}$ complex symmetric matrix.
+Models with $n\w{RHN}\ge2$ are viable, because $n\w{RHN}<3$ gives $3-n\w{RHN}$ massless (Weyl) neutrinos and $2n\w{RHN}$ massive (Majorana) neutrinos, while all neutrinos acquire Majorana mass  if $n\w{RHN}\ge 3$.
 
-We introduce left-handed Weyl spinors $\xi$ and $\chi$ to avoid notational confusion and diagonalize the Dirac mass term $\MD$:
+We introduce left-handed Weyl spinors $\xi$ and $\chi$ to avoid notational confusion and diagonalize the Dirac mass term $M\w{D}$:
 \begin{equation}
@@
  N   \equiv \pmat{0\\\bar\chi}\equiv \pmat{0\\\nR},\quad
- \MD := \frac{v}{\sqrt2}Y_n,\quad
- \MD\wdiag := U_n \MD Y_n^\dagger.
+ M\w{D} := \frac{v}{\sqrt2}Y_n,\quad
+ M\w{D}\wdiag := U_n M\w{D} Y_n^\dagger.
 \end{equation}
@@
 \supset
- -\frac{v}{\sqrt2}\overline{N}Y_n \PL L^1 - \frac{1}{2}\overline{N}\MN N^\cc + \text{h.c.}
-=-\frac12\pmat{\xi & \chi}\pmat{0 & \MD^\TT \\ \MD & \MN} \pmat{\xi \\ \chi} + \text{h.c.} =: -\frac12\tilde\nu^\TT \tilde M\tilde \nu+\text{h.c.},
+ -\frac{v}{\sqrt2}\overline{N}Y_n \PL L^1 - \frac{1}{2}\overline{N}M_N  N^\cc + \text{h.c.}
+=-\frac12\pmat{\xi & \chi}\pmat{0 & M\w{D}^\TT \\ M\w{D} & M_N } \pmat{\xi \\ \chi} + \text{h.c.} =: -\frac12\tilde\nu^\TT \tilde M\tilde \nu+\text{h.c.},
 \end{align}
-where tildes denote ``larger'' objects. We can AT-diagonalize $\tilde M$, which is a symmetric $\mathbb{C}^{(3+\nRHN)\times(3+\nRHN)}$ matrix, as usual:
+where tildes denote ``larger'' objects. We can AT-diagonalize $\tilde M$, which is a symmetric $\mathbb{C}^{(3+n\w{RHN})\times(3+n\w{RHN})}$ matrix, as usual:
 \begin{equation}
@@
 \qquad
--\mathcal L\w{Dirac-$\nu$}\supset
+-\mathcal L\w*{Dirac-$\nu$}\supset
 \frac12\tilde \nu^\TT \tilde M\tilde\nu
@@
 \paragraph{Connection of the above two models}
-If we set $\MN=0$ in the Type-I see-saw ($\nRHN=3$), the AT-diagonalization becomes
+If we set $M_N =0$ in the Type-I see-saw ($n\w{RHN}=3$), the AT-diagonalization becomes
 \begin{align}
-\tilde M = \frac12\pmat{0 & (U_n\MD\wdiag V_n^\dagger)^\TT \\ U_n\MD\wdiag V_n^\dagger & 0},\quad
+\tilde M = \frac12\pmat{0 & (U_nM\w{D}\wdiag V_n^\dagger)^\TT \\ U_nM\w{D}\wdiag V_n^\dagger & 0},\quad
   \tilde R=\pmat{V_n^* & \ii V_n^* \\ U_n & -\ii U_n},\quad
-  \tilde M\wdiag=\pmat{\MD\wdiag & 0 \\ 0 & \MD\wdiag}.
+  \tilde M\wdiag=\pmat{M\w{D}\wdiag & 0 \\ 0 & M\w{D}\wdiag}.
 \end{align}
@@
 \begin{align}
- \text{(diag.)} = U^\TT\pmat{0 & \MD^\TT \\ \MD & \MN} U;\qquad
+ \text{(diag.)} = U^\TT\pmat{0 & M\w{D}^\TT \\ M\w{D} & M_N} U;\qquad
 &U'\approx\pmat{
@@
   1-\frac12\Theta_0\Theta_0^\dagger
-};\quad \Theta_0:=-\MN^{-1}\MD,
+};\quad \Theta_0:=-M_N^{-1}M\w{D},
 \label{eq:typeIseesaw-blockdiag-1}\\
-&\text{(block diag.)}=U'^\TT\pmat{0 & \MD^\TT \\ \MD & \MN} U'\approx\pmat{
-  -\MD^\TT\MN^{-1}\MD & 0 \\ 0 & \MN
+&\text{(block diag.)}=U'^\TT\pmat{0 & M\w{D}^\TT \\ M\w{D} & M_N} U'\approx\pmat{
+  -M\w{D}^\TT M_N^{-1}M\w{D} & 0 \\ 0 & M_N
 },
@@
 As experiments have not found any deviation from $[U]\w{upper left}$ being unitary, we express the mixing with the PMNS matrix
- \GRAY{(Pontecorvo--\japanese{牧}--\japanese{中川}--\japanese{坂田})}
+ \GRAY{(Pontecorvo--\JAPANESE{牧}--\JAPANESE{中川}--\JAPANESE{坂田})}
 parameterized by
@@
 $
- \eta_i = \CONV{\alpha_i}.
+ \eta_i = \C{\alpha_i}.
 $
@@
 \begin{equation}
- m\w L \approx -U\w{PMNS}^\TT \MD^\TT \MN^{-1}\MD U\w{PMNS},\qquad
- m\w H \approx U_X^\TT \MN U_X,
+ m\w L \approx -U\w{PMNS}^\TT M\w{D}^\TT M_N^{-1}M\w{D} U\w{PMNS},\qquad
+ m\w H \approx U_X^\TT M_N U_X,
 \end{equation}
@@
   m\w L
-\approx -U\w{PMNS}^\TT \MD^\TT (U_X^* m\w H U_X^\dagger)^{-1}\MD U\w{PMNS}
-=       -U\w{PMNS}^\TT \MD^\TT U_Xm\w H^{-1}U_X^\TT\MD U\w{PMNS}
+\approx -U\w{PMNS}^\TT M\w{D}^\TT (U_X^* m\w H U_X^\dagger)^{-1}M\w{D} U\w{PMNS}
+=       -U\w{PMNS}^\TT M\w{D}^\TT U_Xm\w H^{-1}U_X^\TT M\w{D} U\w{PMNS}
 \end{equation}
@@
 Note that $U_X$ is unphysical, i.e., we can fix the basis of $\nR$ so that $U_X=1$.
-This basis is equivalent (under our approximation) to the basis in which $\MN$ becomes diagonal.
+This basis is equivalent (under our approximation) to the basis in which $M_N$ becomes diagonal.
 
@@
 \begin{equation}
- R:=-\ii m\w H^{-1/2}U_X^\TT \MD  U\w{PMNS}  m\w L^{-1/2}
+ R:=-\ii m\w H^{-1/2}U_X^\TT M\w{D}  U\w{PMNS}  m\w L^{-1/2}
 \quad\Longrightarrow\quad R^\TT R\approx1.
@@
 \begin{equation}
- \frac{v}{\sqrt2}Y_n\Big|\w{$\MN$ being diagonal}
+ \frac{v}{\sqrt2}Y_n\Big|\w*{$M_N$ being diagonal}
   \approx \ii \sqrt{m\w H}R\sqrt{m\w L}U\w{PMNS}^\dagger.
@@
 
-With this parameterization, the neutrino mixing matrix is given by, in the basis with $\MN$ being diagonal,
+With this parameterization, the neutrino mixing matrix is given by, in the basis with $M_N$ being diagonal,
 \begin{equation}
@@
   U\w{PMNS}
-  &  \MD^\dagger \MN^{*-1} U_X
-  \\ -\MN^{-1} \MD U\w{PMNS}
+  &  M\w{D}^\dagger M_N^{*-1} U_X
+  \\ -M_N^{-1} M\w{D} U\w{PMNS}
   &  U_X

### 2022-05-09 cosmology.tex
add notation/convention
rewrite with new class

@@
-% Time-Stamp: <2020-08-12 13:07:42 misho>
+% Time-Stamp: <2022-05-16 15:23:42>
 
 \documentclass[CheatSheet]{subfiles}
-\newcommand{\ddP}[2][3]{\frac{\dd^{#1}\vc#2}{(2\pi)^{#1}}}
 
@@
 \end{equation}\vspace{-2em}
-\begin{align*}
-\text{comoving coordinate~~}& \vc r=(r\sin\theta\cos\phi,r\sin\theta\cos\phi,r\cos\theta),\\
-\text{proper coordinate~~}& \vc x(t)=a(t)\vc r,\\
-\text{comoving distance~~}& \chi_{AB}=\int_{r_A}^{r_B}\frac{\dd r}{\sqrt{1-Kr^2}},\\
-\text{proper distance~~}& d_{AB}(t)=a(t)\chi_{AB}.
-\end{align*}
+\begin{alignat*}{2}
+  &\text{comoving coordinate~~}&& \vc r=(r\si\theta\co\phi,r\si\theta\co\phi,r\co\theta),\\
+  &\text{proper coordinate~~}&& \vc x(t)=a(t)\vc r,\\
+  &\text{comoving distance~~}&& \chi_{AB}=\int_{r_A}^{r_B}\frac{\dd r}{\sqrt{1-Kr^2}},\\
+  &\text{proper distance~~}&& d_{AB}(t)=a(t)\chi_{AB}.
+\end{alignat*}
 \end{minipage}
-\begin{minipage}{0.395\textwidth}
-\hfill\includegraphics[width=0.9\textwidth]{figs/expansion.pdf}
+\begin{minipage}{0.39\textwidth}
+\hfill\includegraphics[width=0.99\textwidth]{figs/expansion.pdf}
 \end{minipage}
@@
 \subsection{FLRW metric}
-Two conventions are known for FLRW \GRAY{(\russian{Фр\'{и}дман}-Lema\^{i}tre-Robertson-Walker)} metric:
+Two conventions are known for FLRW \GRAY{(\RUSSIAN{Фр\'{и}дман}-Lema\^{i}tre-Robertson-Walker)} metric:
 \begin{align}
@@
 \begin{equation}
- \vev{\sigma v}_{AB\to12\cdots n}(T)
+ \mean{\sigma v}_{AB\to12\cdots n}(T)
 =\frac{1}{n_An_B}\int\ddP{k_A}\ddP{k_B}\left(
@@
 \begin{equation}
- \vev{\Gamma}_{A\to12\cdots n}
+ \mean{\Gamma}_{A\to12\cdots n}
 =\frac{1}{n_A}\int\ddP{k_A}
@@
 \begin{align}
- \vev{\sigma v}
+ \mean{\sigma v}
 &=
@@
 \\
- \vev{\Gamma}
+ \mean{\Gamma}
  &=\frac{K_1(m_A/T)}{K_2(m_A/T)}\Gamma.

### 2022-05-09 spinors.tex
rewrite with new class

@@
 
-\newcommand{\LorTr}[2]{\Lambda^{#1}{}_{#2}}
-\newcommand\defeq{\mathrel{\stackrel{\text{def}}=}}
 \begin{document}
@@
 \begin{equation}
- \eta^{\mu\nu}=\eta_{\mu\nu}\defeq h_\eta\times\diag(+1,-1,-1,-1),\qquad
+ \eta^{\mu\nu}=\eta_{\mu\nu}\eqdef h_\eta\times\diag(+1,-1,-1,-1),\qquad
  \eta^{\mu\alpha}\eta_{\alpha\nu}=\delta^{\mu}_\nu;\qquad
- \eta_{\rho\sigma}\defeq\eta_{\mu\nu}\LorTr\mu\rho\LorTr\nu\sigma ~\text{(defining equation)}.
+ \eta_{\rho\sigma}\eqdef\eta_{\mu\nu}\LorTr\mu\rho\LorTr\nu\sigma ~\text{(defining equation)}.
 \label{eq:LorTr}
@@
  \vep^{\mu\nu\rho\sigma}\mapsto \LorTr\mu\alpha\LorTr\nu\beta\LorTr\rho\gamma\LorTr\sigma\delta\vep^{\alpha\beta\gamma\delta}=(\det\Lambda)\vep^{\mu\nu\rho\sigma};\qquad
- \vep^{0123}\defeq1,\quad \vep_{0123}=-1.
+ \vep^{0123}\eqdef1,\quad \vep_{0123}=-1.
 \end{equation}
 
-\paragraph{Infinitesimal transformation and Lorentz algebra}
+\paragraph{Infinitesimal transformation}
 Equation~\eqref{eq:LorTr} gives the representation for a proper orthochronous Lorentz transformation $\Lambda\in L_0$:
@@
 }
-~\defeq~
+~\eqdef~
 \ii\vc{\omega}\cdot\vc K+\ii\vc\theta\cdot\vc J
-~\defeq~
+~\eqdef~
 \frac{-\ii}{2}d^{\mu\nu}M_{\mu\nu}
@@
 \begin{align}
- \vc A&\defeq\frac{\vc J+\ii \vc K}{2},&
- \vc B&\defeq\frac{\vc J-\ii \vc K}{2};&
+ \vc A&\eqdef\frac{\vc J+\ii \vc K}{2},&
+ \vc B&\eqdef\frac{\vc J-\ii \vc K}{2};&
  [A_i, A_j] &= \ii \epsilon_{ijk}A_k,&

### 2024-06-19 calculator/call_pdg.py
update PDG values

@@ -1,0 +1,206 @@
+import pdg
+
+api = pdg.connect("sqlite:///calculator/pdgall-2024-v0.1.0.sqlite", pedantic=True)
+editions = [2024, 2022, 2020]
+
+METER_GEV = 5.0677307e15
+SECOND_GEV = 1.5192674e24
+METER_PER_SECOND = 3.335641e-09
+
+
+def _find_by_class(cls, pid_list=None):
+    if not pid_list:  # look up all
+        for i in api.get_all(edition=editions[0]):
+            if isinstance(i, cls):
+                if s := i.summary_values(summary_table_only=True):
+                    print(s[0].pdgid, s[0].description)
+        return
+    for pid in pid_list:
+        for e in editions:
+            p = api.get_particle_by_mcid(pid, edition=e)
+            for i in p.properties():
+                if isinstance(i, cls):
+                    if s := i.summary_values(summary_table_only=True):
+                        print(
+                            "{:5d} {:5d} {:10s} {:20s} {:10s} {:10s} {:s}".format(
+                                pid,
+                                e,
+                                s[0].pdgid,
+                                s[0].display_value_text,
+                                s[0].get("unit_text", "[No Unit]"),
+                                str(s[0].get("confidence_level")),
+                                s[0].description,
+                            )
+                        )
+                        if cls == pdg.data.PdgLifetime:
+                            print(
+                                convert_sec_to_meter(
+                                    s[0].value,
+                                    s[0]["unit_text"],
+                                    s[0].error_positive,
+                                    s[0].error_negative,
+                                ),
+                                "meter",
+                            )
+
+
+def find(key_list):
+    for key in key_list:
+        for e in editions:
+            if s := api.get(key, edition=e).summary_values():
+                print(
+                    "{:10s} {:5d} {:10s} {:20s} {:10s} {:10s} {:s}".format(
+                        key,
+                        e,
+                        s[0].pdgid,
[…]

### 2024-06-19 calculator/pdg_values.py

@@ -1,0 +1,252 @@
+import re
+
+import jinja2
+import pathlib
+
+
+class EditionValue:
+    #    colors_tex = {"2020": "\\GREEN", "2022": "\\BLUE", "2024": "\\PINK"}
+    colors_tex = {"2020": "\\BLUE", "2022": "\\PINK", "2024": "\\GREEN"}
+    units_tex = {
+        "m": "\\unit{m}",
+        "us": "\\unit{\\mu s}",
+        "um": "\\unit{\\mu m}",
+        "ns": "\\unit{ns}",
+        "nm": "\\unit{nm}",
+        "mm": "\\unit{mm}",
+        "s": "\\unit{s}",
+        "years": "\\unit{yr}",
+        "ecm": "\\ech\\unit{cm}",
+        "mu(N)": "\\mu_N",
+    }
+
+    @classmethod
+    def source(cls):
+        return ", ".join(
+            f"{v}{{\\textbf{{PDG{k}}}}}" for k, v in cls.colors_tex.items()
+        )
+
+    def __init__(self, edition, value, unit, cl):
+        self.edition = edition
+        self.value = value
+        self.unit = unit.replace("@", "")
+        self.cl = cl
+        if m := re.match(r"^(\w+)-(\d+)$", self.unit):
+            self.unit = m.group(1)
+            self.value = f"{self.value})\\EE{{-{m.group(2)}}}".replace(
+                "((", "("
+            ).replace("))", ")")
+
+    def suffix(self, unit=True):
+        if self.cl == "None":
+            cl = ""
+        elif (f := int(float(self.cl) + 0.1)) in [68, 90, 95]:
+            cl = f"\\bound{{{f}}}{{}}"
+        else:
+            raise ValueError(f"Unknown cl: {self.cl}")
+        if unit and self.unit:
+            u = self.units_tex.get(self.unit, "\\" + self.unit)
+            return cl + u
+        else:
+            return cl
+
+    def wrap_by_edition(self, text):
+        color = self.colors_tex[self.edition]
+        if text.startswith(">"):
+            return ">" + color + "{" + text[1:] + "}"
+        elif text.startswith("<"):
+            return "<" + color + "{" + text[1:] + "}"
+        else:
+            return color + "{" + text + "}"
+
+    def __str__(self):
+        v = self.value.replace(" ", "#")
[…]

### 2024-07-16 mathematics.tex
update spinors

@@
 
+\subsection{Mathematical Foundations for Spinors}
+\paragraph{Quadratic form}
+Let~\cite[\S6.3]{Jacobson} $V$ be an $n$-dimensional vector space over a field $K$. A quadratic form $Q$ on $V$ is defined by
+\begin{equation}
+\begin{split}
+   Q\colon V\to K,\qquad
+   &\forall k\in K, v\in V, Q(k v)=k^2Q(v),\\
+   &B\colon (v,w)\mapsto Q(v+w)-Q(v)-Q(w)\text{~is linear in $v$ and $w$.}
+\end{split}
+\end{equation}
+The bilinear form $B$ satisfies $B(x,x)=2Q(x)$. As long as the characteristic of $F$ is not two, $Q$ is determined by $B$.
+
+Equivalently, $Q$ is characterized by a symmetric ``matrix'' $\hat Q$ such that $Q(v)=v^\TT \hat Q v$.
+Focusing on $K=\RR$, $\hat Q$ is a real symmetric matrix and thus diagonalizable.
+Sylvester's law of inertia states that there exists a basis such that $\hat Q=\diag(1,1,\cdots,1,-1,-1,\cdots,-1)$, where the number of $+1$ and $-1$ are respectively $p$ and $q$.
+The pair $(p,q)$ is unique and is called the signature of $Q$, where $\rank \hat Q=p+q$.
+
+Quadratic form induces orthogonality: a basis $\{e_i\}$ of $V$ is called orthogonal [orthonormal] under $Q$ if
+\begin{equation}
+  B(e_i,e_j)\propto \delta_{ij}\qquad\Bigl[B(e_i,e_j)=c_i\delta_{ij}\text{~with~}c_i\in\{2,0,-2\}\Bigr]
+\end{equation}
+and $(V,Q)$ has an orthogonal basis as long as the characteristic of $K$ is not two.
+If $K$ is a ``spin field'' such as $\RR$ or $\CC$ (or, precisely, $\forall \alpha\in K,\exists \beta\in K$ s.t. $\alpha=\beta^2$ or $-\beta^2$), $(V,Q)$ has an orthonormal basis.
+
+\paragraph{Geometric algebra}
+Let~\cite{0907.5356} $V$ be an $n$-dimensional vector space over a field $K$. Let $Q$ be a quadratic form on $V$.
+Tensor algebra on $V$ is defined by
+\begin{equation}
+  T(V) = \bigoplus_{k} V^{\otimes k}
+       = K\oplus V\oplus (V\otimes V)\oplus\cdots.
+\end{equation}
+It has an ideal $K_Q$ generated by $\{v\otimes v - Q(v)\mid v\in V\}$:
+\begin{equation}
+  K_Q = \left\{\sum_{i} a_i\otimes(v_i\otimes v_i-Q(v_i))\otimes b_i\relmiddle| v_i\in V, a_i,b_i \in T(V)\right\}.
+\end{equation}
+Then the Geometric algebra $\GeoA(V,Q)$ is defined by $T(V)/K_Q$. This is isomorphic to $\Cl(E, K, q_E)$ defined below.
+
+The product in $\GeoA(V,Q)$ is written by $vw \deq [v\otimes w]=v\otimes w+K_q$. Particularly, $v^2 = Q(v)$ and $vw+wv=B(v,w)$.
+
+\paragraph{Clifford Algebra}
+For a finite set $X$, a commutative ring $R$ with unit,
+the Clifford algebra $\Cl(X,R,r)$ is defined as the free $R$-module generated by the set $2^X$ of all subset of $X$.
+Here, $r\colon X\to R$ is an arbitrary function thought of as a signature.
+
+The Clifford algebra $\mathrm{Cl}$ has
+addition $\mathrm{Cl}\times\mathrm{Cl}\to\mathrm{Cl}$,
+scalar multiplication $R\times\mathrm{Cl}\to\mathrm{Cl}$, and
+product
+\begin{equation}
+  \mathrm{Cl}\times\mathrm{Cl}\to\mathrm{Cl}\ni AB
+  \deq \tau(A,B)A\setdiff B,
+\end{equation}
+where $A\setdiff B=(A\cup B)\setminus(B\cap A)$,
+defined based on the following map $\tau\colon2^X\times2^X\to R$:
+\begin{equation}
+  \begin{split}
+  &\tau(\{x\},\{x\}) = r(x)\quad\forall x\in X,&
+  &\tau(A,B)\tau(A\setdiff B,C)=\tau(A,B\setdiff C)\tau(B,C)\quad\forall A,B,C\in 2^X,\\
+  &\tau(\{y\},\{x\}) = -\tau(\{x\},\{y\})\quad\forall x,y\in X,x\neq y,&
+  &\tau(\emptyset,A)=\tau(A,\emptyset)=1\quad\forall A\in2^X,\\
+  &\tau(A,B)\in\{-1,1\}\quad \text{if}~A\cap B=\emptyset.
+  \end{split}
+\end{equation}
+
+Several standard operations are defined: for $A\in2^X$ and $k=\binom{|A|}{2}$,
+\begin{equation}
+\text{Grade involution}\quad A^\star\deq(-1)^{|A|}A,\qquad
+\text{Reversion}\quad A^\dagger\deq(-1)^{k}A,\qquad
+\text{Clifford conjugate}\quad A^\mdlgwhtsquare\deq A^{\star\dagger}.
+\end{equation}
+The sign of grade involution (reversion) is negative iff $|A|$ mod 4 is 1 or 3 (2 or 3).
+
+Let $(V,Q)$ be an $n$-dimensional vector space $V$ over a field $K$ equipped with a quadratic form $Q$. Let $E=\{e_i\}$ be an orthogonal basis of $(V,Q)$.
+Then
+\begin{equation}
+  \Cl(E, K, Q|_E) \cong \GeoA(V,Q).
+\end{equation}
+The Clifford algebra $\Cl(X,R,r)$ has subspaces of $k$-vectors $\mathrm{Cl}^k$ and of even and odd vectors $\mathrm{Cl}^\pm$:
+\[
+\Cl(X,R,r)=\mathrm{Cl}^+\oplus\mathrm{Cl}^-=
+\mathrm{Cl}^0\oplus\mathrm{Cl}^1\oplus\mathrm{Cl}^2\oplus\cdots\oplus\mathrm{Cl}^{|X|}.
+\]
+In particular, $\mathrm{Cl}^+$ is a subalgebra.
+
+\paragraph{Groups in Geometric Algebra}
+The following groups are embedded in $\GeoA(V,Q)$,
+where $V^\times=\{v\in V\mid Q(v)\neq 0\}$:
+\begin{align}
+  &\GeoA^\times \deq \{g\in\GeoA\mid \exists g^{-1}\in\GeoA, gg^{-1}=g^{-1}g=1\},
+  \qquad \Gamma\deq\{v_1v_2\cdots v_n\mid v_i\in V^\times\},
+  \qquad \tilde \Gamma\deq \{g\in\GeoA^\times\mid gVg^{-1}\in V\},
+\notag\\
+  &\Pin\deq\{x\in \Gamma\mid xx^\dagger = \pm1\},
+ \qquad\Spin\deq \Pin\cap\GeoA^+,
+ \qquad\Spin^+\deq \{x\in\Spin\mid xx^\dagger=1\},
+\end{align}
+The Lipschitz group $\tilde\Gamma$ is actually equal to the versor group $\Gamma$ and is the smallest group that contains $V^\times$. To summarize,
+\[\GeoA\supset\GeoA^\times\supset \Gamma=\tilde\Gamma\supset V^\times;\qquad\Gamma\supset\Pin\supset\Spin\supset\Spin^+.\]
+
+There are surjective homomorphisms with kernel $\pm1$ for
+\begin{equation}
+  \Pin(s,t)\to\gO(s,t),\qquad
+  \Spin(s,t)\to\gSO(s,t),\qquad
+  \Spin^+(s,t)\to\gSO^+(s,t).
+\end{equation}
+Also, for our interest, there are isomorphisms~\cite{RauschdeTraubenberg:2005aa,Yamaguchi:spinor}
+\begin{equation}
+  \GeoA(\RR^{4,1})\cong \GeoA(\RR^{1,3})\otimes \CC \cong \GeoA(\CC^4)\cong\CC^{4\times4},\qquad
+  \Spin^+(1,3)\cong\gSL(2,\CC)\cong\gSp(2,\CC),\qquad
+  L_0\cong\mathop{\mathrm{PSL}}(2;\CC)=\gSL(2;\CC)/\ZZ_2.
+\end{equation}
+Meanwhile, the Lorentz algebra $\aSO(1,3)$ is isomorphic to $\aSL(2;\mathbb C)$ viewed as a real Lie albegra~\cite[\S7.8]{Hall2015},
+and its complexification
+$\aSO(1,3)_{\mathbb C}$ is isomorphic to $
+\aSU(2)_{\mathbb C}\oplus\aSU(2)_{\mathbb C}
+=\aSL(2,\mathbb C)\oplus\aSL(2,\mathbb C)$.
+
 \paragraph{Clifford algebra and spin group}\TODO{needs experts...}

### 2024-07-16 spinors.tex
update spinors

@@
 \summarystyle
+\section{Spinors}
+\begin{enumerate}[label={}]
+  \item Gamma matrices: \quad
+$\{\gamma^\mu,\gamma^\nu\}=2\eta^{\mu\nu}$,\quad
+$\gamma_5=\ii\gamma^0\gamma^1\gamma^2\gamma^3$;\quad
+$\{\gamma^\mu,\gamma_5\}=0$,~
+$\gamma^5\gamma^5=1$.
+ \item Conjugates: $\overline{\psi} \deq \psi^\dagger \beta,$\qquad $\psi^\cc\deq C(\overline\psi)^\TT$
+\end{enumerate}
 
-\section{Spinors}
-\begin{alignat}{2}
-&\text{Gamma matrices}:&\quad&
-\{\gamma^\mu,\gamma^\nu\}=2\eta^{\mu\nu},~
-\gamma_5=\ii\gamma^0\gamma^1\gamma^2\gamma^3;\quad
-\{\gamma^\mu,\gamma_5\}=0,~
-\gamma^5\gamma^5=1.
-\\
-&\text{conjugates}:&&
-\overline{\psi}=\psi^\dagger \beta,\qquad
-\psi^\cc=C(\overline\psi)^\TT
-\end{alignat}
-
-\paragraph{chiral notation}
+\paragraph{Chiral notation}
 \begin{align}\label{eq:chiralnotation}
@@
  \PL = \frac{1-\gamma_5}{2},~
- \PR = \frac{1+\gamma_5}{2}.
+ \PR = \frac{1+\gamma_5}{2};\qquad C=-\ii\gamma^2\gamma^0.
 \\&
@@
 (\gamma^\mu)^*       = \gamma^2\gamma^\mu\gamma^2,~
-(\gamma^\mu)^\TT     = \gamma^0\gamma^2\gamma^\mu\gamma^2\gamma^0,
+(\gamma^\mu)^\TT     = \gamma^0\gamma^2\gamma^\mu\gamma^2\gamma^0;
+\\&
+C\gamma^\mu C^\dagger = -(\gamma^\mu)^\dagger,~
+C=C^*=-C^{-1}=-C^\dagger=-C^\TT
+\end{align}
+\begin{align}
+  \psi=
 \end{align}
 
-\paragraph{(cf.~Dirac notation)}
+\begin{RemarkBox}Dirac notation is given by
 \begin{align}\label{eq:diracnotation}
@@
 \end{align}
+\end{RemarkBox}
 
@@
 \subsection{Verbose derivation}
-We here derive the fermion convention under the most generic with signs $h_i=\pm1$, following
-Refs.~\cite{Kugo1}.
+We provide a discussion on the fermion convention, introducing various signs $h_i=\pm1$ in order to keep generality as much as possible. We follow Ref.~\cite{Kugo1}.
 
 \paragraph{Lorentz group and Lorentz tensors}
-The Lorentz transformation $\LorTr\mu\nu$ is defined as a linear transformation $x^\mu\mapsto\LorTr\mu\nu x^\nu$ that conserves $x^2=\eta_{\mu\nu}x^\mu x^\nu$, where $x^\mu$ is a spacetime point and $\eta$ is the Minkowski metric:
+The Lorentz transformation $\LorTr\mu\nu$ is defined as a linear transformation $x^\mu\mapsto\LorTr\mu\nu x^\nu$ conserving $x^2=\eta_{\mu\nu}x^\mu x^\nu$, where $x^\mu$ is a spacetime point and $\eta$ is the Minkowski metric:
 \begin{equation}
@@
 \end{equation}
-Its nice to denote its inverse, which satisfies $x_\nu\mapsto x_\mu(\Lambda^{-1})^\mu{}_\nu$, by $\Lambda_\nu{}^\mu$:
+Consequently, $
+  \delta^\beta_\sigma
+  = (\eta_{\alpha\mu}\eta^{\beta\nu}\LorTr\mu\nu)\LorTr\alpha\sigma
+  \eqcolon (\Lambda^{-1})^\beta{}_\alpha\LorTr\alpha\sigma$
+should define the inverse of $\Lambda$:
 \begin{equation}
- (\Lambda^{-1})^\mu{}_\nu = \eta^{\mu\alpha}\LorTr\beta\alpha\eta_{\beta\nu}
-\equiv \Lambda_\nu{}^\mu;\qquad
-x_\mu\mapsto \Lambda_\mu{}^\nu x_\nu,\qquad
+  \Lambda_\beta{}^\alpha
+  \deq (\Lambda^{-1})^\beta{}_\alpha
+  = \eta_{\alpha\mu}\eta^{\beta\nu}\LorTr\mu\nu;
+  \qquad
+  x_\nu
+  \mapsto x_\mu(\Lambda^{-1})^\mu{}_\nu
+  =\Lambda_\mu{}^\nu x_\nu,\qquad
   \delta^\alpha_\beta
@@
   =\Lambda_\mu{}^\alpha\LorTr\mu\beta
-=\LorTr\alpha\mu \Lambda_\beta{}^\mu.
+  =\LorTr\alpha\mu \Lambda_\beta{}^\mu,
 \end{equation}
@@
 \end{equation}
-There are two constants that qualify to be (pseudo-)tensors: $\eta^{\mu\nu}$ (and $\eta_{\mu\nu}$, $\delta^{\mu}_\nu$) and the anti-symmetric tensor $\vep^{\mu\nu\rho\sigma}$:
+In particular, metrics and the antisymmetric tensors are constant (pseudo-)tensors.
+\begin{align}
+&\eta^{\mu\nu}\mapsto \LorTr\mu\alpha\LorTr\mu\beta\eta^{\alpha\beta}=\eta^{\mu\nu},\qquad
+ \vep^{\mu\nu\rho\sigma}\mapsto \LorTr\mu\alpha\LorTr\nu\beta\LorTr\rho\gamma\LorTr\sigma\delta\vep^{\alpha\beta\gamma\delta}=(\det\Lambda)\vep^{\mu\nu\rho\sigma};\\
+&\vep^{0123}\deq1,\quad \vep_{0123}=-1.
+\end{align}
+
+\paragraph{Infinitesimal transformation} In order to consider infinitesimal transformation of a proper orthochronous Lorentz transformation $\Lambda\in L_0$, we identify $\Lambda$ as a matrix:
+\[ \LorTr\mu\nu \circeq \tilde\Lambda\stackrel{\text{def}}\Longleftrightarrow\LorTr\mu\nu=\tilde\Lambda_{\mu+1,\nu+1};\qquad
+\tilde\eta\deq h_\eta\diag(+1,-1,-1,-1).\]
+Then, Eq.~\eqref{eq:LorTr} is read by $\tilde\eta=\tilde\Lambda^\TT\tilde\eta\tilde\Lambda$ and the infinitesimal transformation given by $\tilde\Lambda=1+\tilde\lambda+\Order(\lambda^2)$ must satisfy
 \begin{equation}
-\eta^{\mu\nu}\mapsto \LorTr\mu\alpha\LorTr\mu\beta\eta^{\alpha\beta}=\eta^{\mu\nu},\qquad
- \vep^{\mu\nu\rho\sigma}\mapsto \LorTr\mu\alpha\LorTr\nu\beta\LorTr\rho\gamma\LorTr\sigma\delta\vep^{\alpha\beta\gamma\delta}=(\det\Lambda)\vep^{\mu\nu\rho\sigma};\qquad
- \vep^{0123}\eqdef1,\quad \vep_{0123}=-1.
-\end{equation}
-
-\paragraph{Infinitesimal transformation}
-Equation~\eqref{eq:LorTr} gives the representation for a proper orthochronous Lorentz transformation $\Lambda\in L_0$:
-\begin{equation}
- \Lambda= 1 + \lambda + \Order(\lambda^2);\quad
-\lambda=\spmat{
+\lambda^\mu{}_\nu
+= -\eta_{\nu\alpha}\eta^{\mu\beta} \lambda^\alpha{}_\beta\Longleftrightarrow
+\tilde\lambda^\TT = -\tilde\eta\tilde\lambda\tilde\eta.
+\qquad\therefore\quad
+  \tilde\lambda=\spmat{
     0 & -\omega_x & -\omega_y & -\omega_z \\
@@
     -\omega_z &+\theta_y &-\theta_x & 0
-}
-~\eqdef~
-\ii\vc{\omega}\cdot\vc K+\ii\vc\theta\cdot\vc J
-~\eqdef~
-\frac{-\ii}{2}d^{\mu\nu}M_{\mu\nu}
-\label{eq:LorTrInf}
+  }\eqcolon\ii\vc{\omega}\cdot\vc K+\ii\vc\theta\cdot\vc J,
 \end{equation}
-where $\theta_i$ is the angle of a \emph{passive} rotation around $i$-axis and $\omega_i$ describes the \emph{passive} boost along $i$-axis with velocity $\beta=\tanh\omega$.
-
-The last definition of Eq.~\eqref{eq:LorTrInf} is used to consider Lorentz algebra $\{M\}$.
-As $[M_{\rho\sigma}]{}^\mu{}_\nu$ should be a tensor, the parameter $d$ should also be a tensor and thus $d^{\mu\nu}=k\cdot\eta^{\mu\rho}\lambda{}^\nu{}_\rho$ ($k$ is a constant), i.e.,
+where we can interpret $\theta_i$ as the \Emph{passive} rotation angle about $i$-axis and $\omega_i$ as the \Emph{passive} boost along $i$-axis with velocity $\beta=\tanh\omega$.
+The operators $J_i$ and $K_i$ are defined to be pure-imaginary; $J_i$ is Hermitian, $K_i$ is pseudo-Hermitian, and
 \begin{equation}
- M_{\mu\nu}=-M_{\nu\mu}, \qquad
- d^{\mu\nu}=-d^{\nu\mu}, \qquad
- \{d^{01},d^{02},d^{03}\}=kh_\eta\times\vc\omega,\qquad
- \{d^{32},d^{13},d^{21}\}=kh_\eta\times\vc\theta.
+  [J_i, J_j] = \ii \epsilon_{ijk}J_k,\qquad
+  [J_i, K_j] = \ii \epsilon_{ijk}K_k,\qquad
+  [K_i, K_j] = -\ii \epsilon_{ijk}J_k.
 \end{equation}
-Then the element of the Lorentz algebra is given by
-$ (M_{\rho\sigma}){}^\mu{}_\nu=
-(\ii/k)(\delta^\mu_\rho\eta_{\nu\sigma}-\delta^\mu_\sigma \eta_{\nu\rho})
-$, which gives the Lorentz algebra
+Noticing that $\tilde\eta\tilde\lambda$ is anti-symmetric, we can express the infinitesimal parameters by anti-symmetric coefficients
+\[
+d^{\mu\nu} \deq h_dh_\eta\lambda^\mu{}_\alpha\eta^{\alpha\nu};\qquad d^{\mu\nu}=-d^{\nu\mu},\qquad
+\{d^{01},d^{02},d^{03}\}=h_d\vc\omega,\qquad
+\{d^{32},d^{13},d^{21}\}=h_d\vc\theta.
+\]
+The corresponding generators are then given by anti-symmetric matrices
+\[
+ \{\tilde M_{01},\tilde M_{02},\tilde M_{03}\}=-h_M \vc K,\qquad \{\tilde M_{32},\tilde M_{13},\tilde M_{21}\}=-h_M \vc J;\qquad
+ \tilde M\circeq (M_{\rho\sigma})^\mu{}_\nu\deq h_M h_\eta\cdot\ii(\delta^\mu_\rho\eta_{\nu\sigma}-\delta^\mu_\sigma\eta_{\nu\rho}).
+\]
+The infinitesimal transformation is given by
 \begin{equation}
-[M_{\mu\nu},M_{\rho\sigma}] = 
-(-\ii/k)\left(\eta_{\mu\rho}M_{\nu\sigma}-\eta_{\mu\sigma}M_{\nu\rho}
+\tilde\lambda = -h_Mh_d\frac{\ii}{2} d^{\mu\nu}\tilde M_{\mu\nu}.
+\label{eq:infinitesimal}
+\end{equation}
+The Lorentz algebra is given by
+\begin{equation}
+h_Mh_\eta[M_{\mu\nu},M_{\rho\sigma}] = 
+-\ii\left(\eta_{\mu\rho}M_{\nu\sigma}-\eta_{\mu\sigma}M_{\nu\rho}
 +\eta_{\nu\sigma}M_{\mu\rho}-\eta_{\nu\rho}M_{\mu\sigma}
@@
 \end{equation}
-We will take $k=+1$ to match the notation of Ref.~\cite{Kugo1}.
 
-\paragraph{Rotation and boost}
-The boost and rotation operators, $\vc K$ and $\vc J$, are now given in an abstract form by
-\begin{equation}
- \vc J = h_\eta (M_{23}, M_{31}, M_{12}), \qquad
- \vc K = -h_\eta(M_{01}, M_{02}, M_{03}),
-\end{equation}
-so their commutation relation is read from \cref{eq:LorentzAlgebra}:
-\begin{align}
- [J_i, J_j] &= \ii \epsilon_{ijk}J_k,&
- [J_i, K_j] &= \ii \epsilon_{ijk}K_k,&
- [K_i, K_j] &= -\ii \epsilon_{ijk}J_k,
-\end{align}
-which leads to
+In literature, the sign $h_\eta$ is first determined; $h_M$ is fixed by the Lorentz algebra \eqref{eq:LorentzAlgebra} and then $h_d$ is set by \eqref{eq:infinitesimal}.
+
+%The signs in Ref.~\cite{Kugo1} are $h_d=h_\eta=1$ and $h_M=1$.
+
+\paragraph{Isomorphism of Lorentz Algebra}
+The commutation relations of $\vc J$ and $\vc K$ lead to
 \begin{align}
@@
 $\aSO(1,3)_{\mathbb C}\cong \aSU(2)_{\mathbb C}\oplus\aSU(2)_{\mathbb C}$.
+Explicitly,
+\[
+\tilde M^{\pm}_{\alpha\beta} =
+\frac{\ii}{2}\left(\pm M_{\alpha\beta} + \frac{-1}{2}\eta_{\alpha\mu}\eta_{\beta\nu}\ii\epsilon^{\mu\nu\rho\sigma}M_{\rho\sigma}\right),\qquad
+\{\tilde M^+_{01},\tilde M^+_{02},\tilde M^+_{03}\}=-h_M\vc A,\qquad
+\{\tilde M^-_{01},\tilde M^-_{02},\tilde M^-_{03}\}=-h_M\vc B.
+\]
 
-\paragraph{More isomorphism}
-Let us see a bit more of  mathematical structure, following the discussion in \cref{sec:group-theory} (cf.~Refs.~\cite{RauschdeTraubenberg:2005aa,Yamaguchi:spinor}).
-The isomorphic groups
-$\mathop{\mathrm{Spin}}(1,3)^+\cong\gSL(2,\mathbb C)\cong\gSp(2,\mathbb C)$
-is a double cover of $L_0$;
-in particular, $L_0\cong \mathop{\mathrm{PSL}}(2;\mathbb C)=\gSL(2;\mathbb C)/\mathbb Z_2$.
-
-Meanwhile, the Lorentz algebra $\aSO(1,3)$ is isomorphic to $\aSL(2;\mathbb C)$ viewed as a real Lie albegra~\cite[\S7.8]{Hall2015},
-and its complexification
-$\aSO(1,3)_{\mathbb C}$ is isomorphic to $
-\aSU(2)_{\mathbb C}\oplus\aSU(2)_{\mathbb C}
-=\aSL(2,\mathbb C)\oplus\aSL(2,\mathbb C)$.
+\TODO{reviewed upto here}
 
 \paragraph{Representation of Clifford algebra}
-As summarized in \cref{sec:group-theory}, the spin group $\mathop{\mathrm{Spin}}(1,3)^+$ generated by Clifford algebra $\mathfrak C_{1,3}$ is a double cover of $L_0$, and thus we can consider a representation of $L_0$ based on $\mathfrak C_{1,3}$.
-
 To construct an irreducible representation of $\mathfrak C_{1,3}$, we utilize the fact that we can form two sets of creation-annihilation operators
@@
 
-\paragraph{Weyl spinor}
-
-\subsection[Convention]{Convention \TODO{WIP!}}
-First we prepare a vector $x^\mu$ and a symmetric matrix $\eta^{\mu\nu}$, which we call ``contravariant vector'' $x^\mu$ and the metric $\eta^{\mu\nu}$.
-Then we perform a Lorentz transformation on $x^\mu$ to obtain $(x')^\mu$, with which we can define a matrix $\Lambda(\vc va,\vc \theta)^{\mu}_{\nu}$ through 
-$\Lambda{}^\mu{}_\nu x^{\nu}=(x')^{\mu}$.
-
-We then consider $\Lambda$s for infinitesimal transformations and define $\vc S$, $\vc J$, and $\vc K$ by
-\begin{align}
- \Lambda^\mu{}_\nu
-  \simeq \delta^\mu_\nu-\ii(\vc\theta\cdot\vc J^\mu{}_\nu+\vc\beta\cdot\vc K^\mu{}_\nu)
-  \simeq \delta^\mu_\nu-\frac{\ii}{2}  \left[\Lambda^{\alpha\beta}S_{\alpha\beta}\right]^\mu{}_\nu
-\end{align}
-
-Imposing ``Lorentz condition'' (\TODO{what?}), we get the expression for $S=\ii(\delta\cdots)$ and $[J^i,J^j]=\cdots$; further, we get $\Lambda^{\mu}_{\nu}=\exp(-\ii\vc\theta\cdot\vc J -\ii \vc\xi\cdot\vc K)$, $\vc\theta=(\theta_{23},\theta_{31}, \theta_{12})$, $\vc\xi = \hat{\vc v}\tanh^{-1}\|\vc v\|=(\theta^{10},\theta^{20},\theta^{30})$; $J=(S_{23},S_{31},S_{12})$, $K=(S^{01}, S^{02}, S^{03})$....?
-
-Lorentz transformation with a rotation $\theta$ around an axis $\hat{\vc\theta}$ and a boost $\vc v$ are given by
-\begin{equation}
- \Lambda = \exp\left[-\ii(\vc\theta\cdot\vc J+\vc \beta\cdot \vc K)\right];
-\qquad
-\vc\theta:=\theta\hat{\vc\theta}, \quad \vc\beta := \hat{\vc v}\tanh^{-1}\|\vc v\|,
-\end{equation}
-\TODO{check!}
-
-Lorentz transformation (infinitesimal):
-$\Lambda=\pmat{
-  0 &   & \vc{\beta}^\TT & \\
-    & 0 &-\theta_z & \theta_y\\
-  \vc{\beta} & \theta_z & 0 & -\theta_x \\
-   & -\theta_y & \theta_x & 0
-}$
-
-$[J_{\mu\nu}]^{\alpha}{}_{\beta}=
- \ii(\delta^\alpha_\mu\eta_{\nu\beta}-\delta^\alpha_\nu\eta_{\mu\beta})
-$
-
+\paragraph{fragments}
 \vspace{2em}
@@
 
-In general, metric is symmetric.
+\subsection{Dirac spinors for massive fermions}
+%\RefBox{\S2 of \cite{kugo1}; \cite{Kryuchkov:2015acw}}
+Dirac equation $(\ii\slashed{\partial}-m)\psi(x)=0$ has plain-wave solutions $\tilde u(p)\ee^{-\ii p x}$, i.e., $(\slashed{p}-m)\tilde u(p)=0$ with $m>0$.
+Non-zero solution is available if and only if $\det(\slashed p-m)=0\Leftrightarrow p^2=m^2$.
+We hereafter fix $p^0>0$ and consider
+\begin{equation}
+ u(\vc p)\text{~satisfying~}(\slashed p-m)u(\vc p)=0,
+\qquad
+ v(\vc p)\text{~satisfying~}(\slashed p+m)v(\vc p)=0,
+\end{equation}
+where the general solution of the Dirac equation is given by linear combination of $\{u(\vc p)\ee^{-\ii px}, v(\vc p)\ee^{\ii px}\}_{p^0>0}$.
 
+As $\rank(\slashed p\pm m)=2$, each equation has $4-2 = \text{two}$ linear-independent solutions\footnote{%
+  Let $X^\pm=\slashed{p}\pm m$. Denote $\det A$ by $|A|$, and $\rank A$ by $\RANK{A}$.
+  As Lorentz invariance guarantees $|X^\pm|$ is a function of $p^2$, $|A^+|=|A^-|$.
+  Then $|A^+||A^-| = (p^2-m^2)^4$ gives $|A^\pm|=(p^2-m^2)^2$.
+  One can show that $\RANK{A^\pm}=2$ if $p^2=m^2$ as follows:
+\\%
+Frobenius inequality:\qquad $\RANK{AB}+\RANK{BC}\le \RANK B + \RANK{ABC}$\qquad
+$\therefore 2\RANK{A^\pm}=\RANK{A^\pm\gamma_5\gamma_5} + \RANK{\gamma_5A^\mp\gamma_5} \le \RANK{\gamma_5}+\RANK{A^\pm\gamma_5\gamma_5A^\mp\gamma_5}=4$,
+\\%
+rank subadditivity:\qquad$\RANK{A+B}\le\RANK{A}+\RANK{B}$\qquad
+$\therefore4\le\RANK{A^+}+\RANK{-A^-}.$
+
+Similar discussion applies for Pauli--\Lubanski operator: as $[\det(\gamma_5\slashed{e}-x)]^2=-e^2-x^2$, it has eigenvalues $\pm1/2$ for $u(\vc p)$ and $v(\vc p)$, and 
+$B^\pm=(\gamma_5\slashed e\slashed p\pm m/2)$ both have rank-2.
+Furthermore, \TOCHECK as $B^+$ should have solutions in both subspaces spanned by $u(\vc p)$ and $v(\vc p)$, each subspace is spanned by eigenstates with different eigenvalues for the Pauli--\Lubanski operator and thus the label is used to distinguish two $u$s (and $v$s).
+}, for which we introduce another label $s=1,2$.
+We use the following convention, noting the inner product of this vector space is $(\psi,\psi)=\overline\psi\psi$:
 \begin{equation}
- (\Lambda^{-1})^\mu{}_\nu = \eta_{\nu\rho}\Lambda^{\rho}{}_\sigma(\eta^{-1})^{\sigma\mu}=:\Lambda_\nu{}^\mu
+ v^s(\vc p)\coloneq C[\overline{u}^s(\vc p)]^\TT,\qquad
+ \overline u^s(\vc p)u^t(\vc p)\coloneq 2m\delta^{st}.
+\end{equation}
+Then $\gamma^\mu v = (h_\beta/h_C)C\beta^\TT(\gamma^\mu u)^*$, for which $v(\vc p)$ satisfies $(\slashed{p}+m)v(\vc p)=0$ as requested.
+Orthogonality for $s\neq t$ is guaranteed by its definition given later.
+Also, as $u$ and $v$ have different eigenvalues of the matrix $\slashed p$,
+\begin{equation}
+  \overline v^s(\vc p)v^t(\vc p)=-2m\delta^{st},\qquad
+ \overline u^s(\vc p)v^t(\vc p)=\overline v^s(\vc p)u^t(\vc p)=0.
 \end{equation}
 
+Several conventions for the label $s$ are available:\vspace{-1em}
+\begin{DownPara}
+\paragraph{Pauli--\Lubanski pseudovector operator} gives the most generic definition for the spin of a moving particle:
+\begin{equation}
+ w^\mu = \frac1{2m}\epsilon^{\mu\nu\rho\sigma}P_\nu M_{\rho\sigma}.
+\end{equation}
+With a reference vector $e^\mu$ such that $e^2=-1$ and $e^\mu p_\mu=0$,
+\begin{equation}
+ e_\mu w^\mu
+= e_\mu \frac1{2m} \epsilon^{\mu\nu\rho\sigma}p_\nu \frac{\ii}{4}[\gamma_\rho,\gamma_\sigma]
+= \frac\ii{4m} \epsilon^{\mu\nu\rho\sigma}e_\mu p_\nu \gamma_\rho\gamma_\sigma
+= \frac1{2m}\gamma_5 \slashed e \slashed p
+\end{equation}
+commutes with $(\slashed p\pm m)$ and thus $s=\pm$ may denote the eigenvalue of this operator $\pm1/2$.
+
+\paragraph{Helicity operator}
+A simpler option is the eigenvalues $\pm1$ of the helicity operator $h=\vc\sigma\cdot\vc p/\|\vc p\|=:\vc\sigma\cdot\vc n$.
+This is in fact a special case of Pauli--\Lubanski operator with $e^\mu= (\|\vc p\|/m,p_0\vc n/m)$, which is verified by
+\begin{align}
+& h
+=\epsilon^{0\nu\rho\sigma}(-n_\nu) S_{\rho\sigma}
+=\frac{1}{2\ii}\epsilon^{0\nu\rho\sigma}n_\nu \gamma_\rho\gamma_\sigma
+=\frac{\eta^{\mu0}}{2\ii}\epsilon_{\mu\nu\rho\sigma}n^\nu \gamma^\rho\gamma^\sigma
+=\frac{\gamma_5}{2}\gamma_5[\slashed{n},\gamma^0]
+=\frac{\gamma_5}{\|\vc p\|}({\{\slashed{p},\gamma^0\}}/{2}-\gamma^0\slashed{p});
+\\&
+w_\mu e^\mu 
+=\frac{1}{2\|\vc p\|}\gamma_5(p^0-\gamma^0\slashed{p})\sim \frac{h}{2}.
+\end{align}
+
+% Low\[Epsilon][m_,n_,r_,s_]:=-LeviCivitaTensor[4][[m+1,n+1,r+1,s+1]]
+% Low\[Epsilon][0,1,2,3]
+% Sum[Low\[Epsilon][m,n,r,s]S[m]p[n]GammaMatrix[r,None] . GammaMatrix[s,None],{m,0,3},{n,0,3},{r,0,3},{s,0,3}];
+% -2I LorSum[S[#]GammaMatrix[None,#]&] . LorSum[p[#]GammaMatrix[None,#]&] . GammaFive;
+% Expand[%%-%]//.{p[0] S[0]->p[1]S[1]+p[2]S[2]+p[3]S[3]}//Simplify
+
+\TODO{Discuss Poincare group before introducing PL operator.: \cite{WeinbergQFT1}, \cite{Kryuchkov:2015acw}}
+
+\TODO{We have not defined the relation between $S_{\mu\nu}$ and $\sigma_i$ etc.; verify for more general case, including the proof of $\epsilon_{\mu\nu\rho\sigma}\gamma^\rho\gamma^\sigma=-\ii\gamma_5[\gamma_\mu,\gamma_\nu]$.}
+\end{DownPara}
+
+Projection operators for the subspaces spanned by $u^s$ ($v^s$) are given by
+\begin{equation}
+ P_u=\frac{m+\slashed p}{2m}, \quad P_v=\frac{m-\slashed p}{2m}.
+\end{equation}
+As $\{u^1, u^2, v^1, v^2\}$ is a basis of the vector space $\mathbb C^4$ with inner product $(\psi,\psi)=\overline\psi\psi$, we immediately have 
+\begin{equation}
+ \sum_{s=1,2}u^s(\vc p)\overline u^s(\vc p)=P_u \overline u^s(\vc p)u^s(\vc p) = \slashed p+m,\qquad
+ \sum_{s=1,2}v^s(\vc p)\overline v^s(\vc p)=\slashed p-m.
+\end{equation}
+Similarly, using projection operator $P'_{\pm}=(1\pm\gamma_5\slashed e\slashed p/m)/2$,
+\begin{align}
+ &u^s(\vc p)\overline u^s(\vc p) = 2m P'_{\pm}P_u = \frac12(\slashed p+m)(1+s\gamma_5\slashed e),\\\
+ &v^s(\vc p)\overline v^s(\vc p) = C(u^s\overline u^s)^\TT \beta^*C^\dagger\beta
+=\frac12(m+h_C \slashed p)(1-s h_C \gamma_5\slashed e )C\beta^*C^\dagger \beta
+\leadsto
+\frac12(\slashed p-m)(1+s\gamma_5\slashed e)
+\end{align}
+for $s=\pm1$.
+
+\subsection{Chiral notation}
+In the chiral notation, $\beta=\gamma^0$ and $C=-\ii\gamma^2\gamma^0$, and $h_\beta=1$ and $h_C=-1$, the plain-wave spinors are given by
+\begin{equation}
+ u^s(\vc p) = \pmat{\sqrt{\sigma\cdot p}\xi^s\\\sqrt{\bar\sigma\cdot p}\xi^s},\quad
+ v^s(\vc p) = \pmat{\sqrt{\sigma\cdot p}\eta^s\\-\sqrt{\bar\sigma\cdot p}\eta^s};\qquad
+ \eta^s = \pmat{\eta^s_1\\\eta^s_2} \coloneq  \pmat{-(\xi^s_2)^*\\(\xi^s_1)^*},\quad
+(\xi^s)^\dagger(\xi^t) = (\eta^s)^\dagger(\eta^t) = \delta^{st},
+\end{equation}
+where $\{\xi^1,\xi^2\}$ are the orthonormal basis, which fixes the definition of $\eta$, and\footnote{%
+The notation is justified by
+\begin{equation*}
+\sqrt{\sigma\cdot p}\sqrt{\sigma\cdot p} ={\sigma\cdot p},\qquad
+\sqrt{\bar\sigma\cdot p}\sqrt{\bar\sigma\cdot p} ={\bar\sigma\cdot p},\qquad
+\sqrt{\sigma\cdot p}\sqrt{\bar\sigma\cdot p}=m.
+\end{equation*}}
+\begin{equation}
+  \sqrt{\sigma\cdot p} \coloneq  \frac{m+p^\mu\sigma_\mu}{\sqrt{2(m+p^0)}},
+\qquad
+ \sqrt{\bar\sigma\cdot p} \coloneq  \frac{m+p^\mu\bar\sigma_\mu}{\sqrt{2(m+p^0)}}.
+\end{equation}
+\footnote{It is straightforward to verify the above-given equations:
+\begin{align}
+&
+ \overline u u = 2m\delta^{st},\quad
+ \overline v v = -2m\delta^{st},\quad
+ \overline u v = \overline v u = 0,\quad
+ u^\dagger u = v^\dagger v = 2p^0\delta^{st},\quad
+\\&
+ u^\dagger(\vc p)u(-\vc p)= v^\dagger(\vc p)v(-\vc p)=2m\delta^{st},\quad
+ u^\dagger(\vc p)v(-\vc p)= v^\dagger(\vc p)u(-\vc p)=0,\quad
+ \overline u(\vc p)u(-\vc p)= -\overline v(\vc p)v(-\vc p)=2p^0\delta^{st}.
+\end{align}
+}
+So that the basis $\xi$ has eigenvalue $\pm1/2$ for the Pauli--\Lubanski operator, it should satisfy
+\begin{equation}
+ \frac12 \gamma_5\slashed{e} U \pmat{\xi^s\\\xi^s} = \left(\pm\frac12\right)U\pmat{\xi^s\\\xi^s};\qquad
+\text{where~}
+ U\coloneq \pmat{\sqrt{\sigma\cdot p} & 0 \\ 0 & \sqrt{\bar\sigma\cdot p}},\quad
+ U^{-1}=\frac{1}{m}\pmat{\sqrt{\bar\sigma\cdot p} & 0 \\ 0 & \sqrt{\sigma\cdot p}}.
+\end{equation}
+Thus, by calculating the eigensystem of $U^{-1}\gamma_5\slashed {e} U$, one can get
+\begin{equation}
+ \xi^+ = \frac{1}{\sqrt{2(1-\phi^3)}}\pmat{\phi^1-\ii \phi^2 \\ 1-\phi^3},\quad
+ \xi^- = \frac{1}{\sqrt{2(1-\phi^3)}}\pmat{1-\phi^3\\-\phi^1-\ii \phi^2};\qquad
+ \phi^\mu = e^\mu-\frac{e^0p^\mu}{m+p^0}.
+\end{equation}
+If we use reference vector for the helicity, 
+\begin{equation}
+ \xi^+ = \frac1{\sqrt{2-2n^3}}\pmat{n^1-\ii n^2\\1-n^3}
+       =\pmat{\ee^{-\ii\phi}\cos(\theta/2)\\\sin(\theta/2)},\quad
+ \xi^- = \frac1{\sqrt{2-2n^3}}\pmat{1-n^3\\-n^1-\ii n^2}
+       =\pmat{\sin(\theta/2)\\-\ee^{\ii\phi}\cos(\theta/2)}
+\end{equation}
+with $(n^1,n^2,n^3)=(\si\theta\co\phi,\si\theta\si\phi,\co\theta)$ is the direction of the momentum.
+
 \end{document}

### 2024-07-18 mathematics.tex
add discussion on Clifford Algebra

@@
 \begin{alignat}{2}
- &\ee^X := \sum_{m=0}^\infty \frac{X^m}{m!} \text{~~(converges for any $X$)},
+ &\ee^X \deq \sum_{m=0}^\infty \frac{X^m}{m!} \text{~~(converges for any $X$)},
 \quad&
- &\log X := \sum_{m=1}^\infty (-1)^{m+1}\frac{(A-1)^m}{m} \text{~~(conv.~if $\|A-I\|<1$)}.
+ &\log X \deq \sum_{m=1}^\infty (-1)^{m+1}\frac{(A-1)^m}{m} \text{~~(conv.~if $\|A-I\|<1$)}.
 \end{alignat}
@@
 \begin{equation}
- \text{Hilbert-Schmidt norm}:~\|X\|^2 := \sum_{i,j}|X_{ij}|^2 = {\Tr X^\dagger X}.
+ \text{Hilbert-Schmidt norm:}~\|X\|^2 \deq \sum_{i,j}|X_{ij}|^2 = {\Tr X^\dagger X}.
 \end{equation}
@@
  P\in\mathbb{K}^{n\times n},\quad
- D:\text{diagonal matrix}~(D_{ii}\in\mathbb{K}).
+ D\colon\text{diagonal matrix}~(D_{ii}\in\mathbb{K}).
 \end{equation}
@@
  V\in\mathbb{U}_{\mathbb K}^n,\quad
- D: \text{non-negative real diagonal matrix}~(D_{ii}\ge0).
+ D\colon \text{non-negative real diagonal matrix}~(D_{ii}\ge0).
 \end{equation}
@@
  R\in\mathbb{U}_{\mathbb C}^n,\quad
- D: \text{non-negative real diagonal matrix}~(D_{ii}\ge0).
+ D\colon \text{non-negative real diagonal matrix}~(D_{ii}\ge0).
  \label{eq:ATF}
@@
 \begin{align}
- \mathop{\mathrm  {GL}}(n;\KK) &= \{g\in M^n_\KK\mathbin|\det g\neq 0\},&
+ \mathop{\mathrm  {GL}}(n;\KK) &= \{g\in M^n_\KK\mathbin\mid \det g\neq 0\},&
  \mathop{\mathfrak{gl}}(n;\KK) &= \{a\in M^n_\KK\},\\
 %
- \mathop{\mathrm  {SL}}(n;\KK) &= \{g\in M^n_\KK\mathbin|\det g=1\},&
- \mathop{\mathfrak{sl}}(n;\KK) &= \{a\in M^n_\KK\mathbin|\Tr a=0\},\\
+ \mathop{\mathrm  {SL}}(n;\KK) &= \{g\in M^n_\KK\mathbin\mid \det g=1\},&
+ \mathop{\mathfrak{sl}}(n;\KK) &= \{a\in M^n_\KK\mathbin\mid \Tr a=0\},\\
 %
- \mathop{\mathrm  {SU}}(n) &= \{g\in M^n_{\mathbb C}\mathbin|g^\dagger g=1\land\det g=1\},&
- \mathop{\mathfrak{su}}(n) &= \{a\in M^n_{\mathbb C}\mathbin|a+a^\dagger=0\land\Tr a=0\},\\
+ \mathop{\mathrm  {SU}}(n) &= \{g\in M^n_{\mathbb C}\mathbin\mid g^\dagger g=1\land\det g=1\},&
+ \mathop{\mathfrak{su}}(n) &= \{a\in M^n_{\mathbb C}\mathbin\mid a+a^\dagger=0\land\Tr a=0\},\\
 %
- \mathop{\mathrm  {SO}}(n) &= \{g\in M^n_{\mathbb R}\mathbin|g^\TT g=1\land\det g=1\},&
- \mathop{\mathfrak{so}}(n) &= \{a\in M^n_{\mathbb R}\mathbin|a+a^\TT=0\},\\
+ \mathop{\mathrm  {SO}}(n) &= \{g\in M^n_{\mathbb R}\mathbin\mid g^\TT g=1\land\det g=1\},&
+ \mathop{\mathfrak{so}}(n) &= \{a\in M^n_{\mathbb R}\mathbin\mid a+a^\TT=0\},\\
 %
- \mathop{\mathrm  {SO}}(n;\mathbb C) &= \{g\in M^n_{\mathbb C}\mathbin|g^\TT g=1\land\det g=1\},&
- \mathop{\mathfrak{so}}(n;\mathbb C) &= \{a\in M^n_{\mathbb C}\mathbin|a+a^\TT=0\},\\
+ \mathop{\mathrm  {SO}}(n;\mathbb C) &= \{g\in M^n_{\mathbb C}\mathbin\mid g^\TT g=1\land\det g=1\},&
+ \mathop{\mathfrak{so}}(n;\mathbb C) &= \{a\in M^n_{\mathbb C}\mathbin\mid a+a^\TT=0\},\\
 %
- \mathop{\mathrm  {SO}}(p,q) &= \{g\in M^{p+q}_{\mathbb R}\mathbin|H g^\TT H=g^{-1}\land\det g=1\},&
- \mathop{\mathfrak{so}}(p,q) &= \{a\in M^{p+q}_{\mathbb R}\mathbin|H a^\TT H=-a\},\\
+ \mathop{\mathrm  {SO}}(p,q) &= \{g\in M^{p+q}_{\mathbb R}\mathbin\mid H g^\TT H=g^{-1}\land\det g=1\},&
+ \mathop{\mathfrak{so}}(p,q) &= \{a\in M^{p+q}_{\mathbb R}\mathbin\mid H a^\TT H=-a\},\\
 %
- \mathop{\mathrm  {Sp}}(n;\KK) &= \{g\in M^n_\KK\mathbin|\Omega g^\TT \Omega=-g^{-1}\},&
- \mathop{\mathfrak{sp}}(n;\KK) &= \{g\in M^n_\KK\mathbin|\Omega a^\TT \Omega=a\},\\
+ \mathop{\mathrm  {Sp}}(n;\KK) &= \{g\in M^n_\KK\mathbin\mid \Omega g^\TT \Omega=-g^{-1}\},&
+ \mathop{\mathfrak{sp}}(n;\KK) &= \{g\in M^n_\KK\mathbin\mid \Omega a^\TT \Omega=a\},\\
 %
- \mathop{\mathrm  {Sp}}(n) &= \{g\in M^n_{\mathbb C}\mathbin|\epsilon g^\TT \epsilon=-g^{-1}\land g^\dagger g=1\},&
- \mathop{\mathfrak{sp}}(n) &= \{g\in M^n_{\mathbb C}\mathbin|\epsilon a^\TT \epsilon=a\land a^\dagger+a=0\}.
+ \mathop{\mathrm  {Sp}}(n) &= \{g\in M^n_{\mathbb C}\mathbin\mid \epsilon g^\TT \epsilon=-g^{-1}\land g^\dagger g=1\},&
+ \mathop{\mathfrak{sp}}(n) &= \{g\in M^n_{\mathbb C}\mathbin\mid \epsilon a^\TT \epsilon=a\land a^\dagger+a=0\}.
 \end{align}
@@
 \begin{equation}
- V_{\mathbb C}=V\otimes\mathbb C=\{v_1+\ii v_2\mathbin|v_1,v_2\in V\}\cong V+\ii V.
+ V_{\mathbb C}=V\otimes\mathbb C=\{v_1+\ii v_2\mathbin\mid v_1,v_2\in V\}\cong V+\ii V.
 \end{equation}
@@
 \subsection{Mathematical Foundations for Spinors}
+\subsubsection{Clifford Algebra}
 \paragraph{Quadratic form}
-Let~\cite[\S6.3]{Jacobson} $V$ be an $n$-dimensional vector space over a field $K$. A quadratic form $Q$ on $V$ is defined by
+Let~\cite[\S6.3]{Jacobson1} $V$ be an $n$-dimensional vector space over a field $K$. A quadratic form $Q$ on $V$ is defined by
 \begin{equation}
@@
 Focusing on $K=\RR$, $\hat Q$ is a real symmetric matrix and thus diagonalizable.
-Sylvester's law of inertia states that there exists a basis such that $\hat Q=\diag(1,1,\cdots,1,-1,-1,\cdots,-1)$, where the number of $+1$ and $-1$ are respectively $p$ and $q$.
-The pair $(p,q)$ is unique and is called the signature of $Q$, where $\rank \hat Q=p+q$.
+Sylvester's law of inertia states that there exists a basis such that $\hat Q=\diag(1,1,\cdots,1,-1,-1,\cdots,-1,0,\cdots,0)$, where the number of $+1$ and $-1$ are respectively $p$ and $q$.
+The pair $(p,q)$ is unique and is called the signature of $Q$, where $\rank \hat Q=p+q~~(\le n)$.
+Meanwhile, if $K$ is algebraically closed such as $\CC$, we can take $\hat Q=\diag(1,1,\cdots,1,0,\cdots,0)$ and the algebra is characterized by $\rank\hat Q$.
 
-Quadratic form induces orthogonality: a basis $\{e_i\}$ of $V$ is called orthogonal [orthonormal] under $Q$ if
+Quadratic form induces orthogonality: a basis $\{e_i\}$ of $V$ is called
 \begin{equation}
-  B(e_i,e_j)\propto \delta_{ij}\qquad\Bigl[B(e_i,e_j)=c_i\delta_{ij}\text{~with~}c_i\in\{2,0,-2\}\Bigr]
+  \text{orthogonal under $Q$}\stackrel{\text{def}}\iff B(e_i,e_j)\propto \delta_{ij}\qquad
+  \Bigl[\text{orthonormal under $Q$}\stackrel{\text{def}}\iff B(e_i,e_j)=c_i\delta_{ij}\text{~with~}c_i\in\{2,0,-2\}\Bigr]
 \end{equation}
@@
 \paragraph{Geometric algebra}
-Let~\cite{0907.5356} $V$ be an $n$-dimensional vector space over a field $K$. Let $Q$ be a quadratic form on $V$.
+Let~\cite{Lundholm:2009xd} $V$ be an $n$-dimensional vector space over a field $K$. Let $Q$ be a quadratic form on $V$.
 Tensor algebra on $V$ is defined by
@@
   T(V) = \bigoplus_{k} V^{\otimes k}
-       = K\oplus V\oplus (V\otimes V)\oplus\cdots.
+       = K\oplus V\oplus (V\otimes V)\oplus(V\otimes V\otimes V)\oplus\cdots.
 \end{equation}
@@
 \begin{equation}
-  K_Q = \left\{\sum_{i} a_i\otimes(v_i\otimes v_i-Q(v_i))\otimes b_i\relmiddle| v_i\in V, a_i,b_i \in T(V)\right\}.
+  K_Q = \left\{\sum_{i} a_i\otimes(v_i\otimes v_i-Q(v_i))\otimes b_i\relmiddle| v_i\in V, a_i,b_i \in T(V)\right\},
 \end{equation}
-Then the Geometric algebra $\GeoA(V,Q)$ is defined by $T(V)/K_Q$. This is isomorphic to $\Cl(E, K, q_E)$ defined below.
+and the geometric algebra $\GeoA(V,Q)\deq T(V)/K_Q$.
+The product in $\GeoA(V,Q)$ is written by $vw \deq [v\otimes w]=v\otimes w+K_q$. Particularly, $v^2 = Q(v)\in K$ and $vw+wv=B(v,w)\in K$.
 
-The product in $\GeoA(V,Q)$ is written by $vw \deq [v\otimes w]=v\otimes w+K_q$. Particularly, $v^2 = Q(v)$ and $vw+wv=B(v,w)$.
+The pseudoscalar $I$ is defined by $I\deq e_1e_2\cdots e_n$ with the basis $\{e_i\}$.
+For $\GeoA(\RR^{s,t,u})$, $I^2=(-1)^{n(n-1)/2+t}\delta_{0u}$ and $I$ commutes (anti-commutes) with all $e_i$ if $n$ is even (odd).
 
 \paragraph{Clifford Algebra}
-For a finite set $X$, a commutative ring $R$ with unit,
+For a finite set $X$ and a commutative ring $R$ with unit,
 the Clifford algebra $\Cl(X,R,r)$ is defined as the free $R$-module generated by the set $2^X$ of all subset of $X$.
 Here, $r\colon X\to R$ is an arbitrary function thought of as a signature.
-
 The Clifford algebra $\mathrm{Cl}$ has
@@
   \mathrm{Cl}\times\mathrm{Cl}\to\mathrm{Cl}\ni AB
-  \deq \tau(A,B)A\setdiff B,
+  \deq \tau(A,B)A\setdiff B
 \end{equation}
-where $A\setdiff B=(A\cup B)\setminus(B\cap A)$,
 defined based on the following map $\tau\colon2^X\times2^X\to R$:
@@
   &\tau(\emptyset,A)=\tau(A,\emptyset)=1\quad\forall A\in2^X,\\
-  &\tau(A,B)\in\{-1,1\}\quad \text{if}~A\cap B=\emptyset.
+  &\tau(A,B)\in\{-1,1\}\quad \text{if}~A\cap B=\emptyset,
   \end{split}
+\end{equation}
+where $A\setdiff B=(A\cup B)\setminus(B\cap A)$. Namely,
+\begin{equation}
+  \{x\}\{x\} = r(x)\emptyset,\quad
+  \{x\}\{y\}=-\{y\}\{x\},\quad
+  \emptyset A=A\emptyset=A,\quad
+  (AB)C=A(BC),\quad
+  A\cap B=\emptyset\Rightarrow AB=\pm1(A\cup B).
 \end{equation}
@@
 \end{equation}
-The sign of grade involution (reversion) is negative iff $|A|$ mod 4 is 1 or 3 (2 or 3).
+The sign of grade involution (reversion) is negative iff $|A|$ mod 4 is 1 or 3 (2 or 3). Also, $(xy)^\star=x^\star y^\star$ and $(xy)^\dagger=y^\dagger x^\dagger$.
 
@@
 \end{equation}
+
 The Clifford algebra $\Cl(X,R,r)$ has subspaces of $k$-vectors $\mathrm{Cl}^k$ and of even and odd vectors $\mathrm{Cl}^\pm$:
@@
 
+\paragraph{Isomorphisms}
+There are several algebra-isomorphisms~\cite{Lundholm:2009xd}:
+\begin{align*}
+&\GeoA(\RR^{s,t})\cong \GeoA(\RR^{t+1,s-1}),\quad
+\GeoA^+(\RR^{s,t})\cong \GeoA(\RR^{s,t-1})\cong \GeoA(\RR^{t,s-1})\cong\GeoA^+(\RR^{t,s}),\\
+&\GeoA(\RR^{n+2,0})\cong\GeoA(\RR^{0,n})\otimes\GeoA(\RR^{2,0}),\quad
+\GeoA(\RR^{0,n+2})\cong\GeoA(\RR^{n,0})\otimes\GeoA(\RR^{0,2}),\quad
+\GeoA(\RR^{s+1,t+1})\cong\GeoA(\RR^{s,t})\otimes\GeoA(\RR^{1,1}),\\
+&\GeoA(\RR^{1,0})\cong\RR\oplus\RR,\quad
+\GeoA(\RR^{0,1})\cong\CC,\quad
+\GeoA(\RR^{2,0})\cong\RR^{2\times2},\quad
+\GeoA(\RR^{0,2})\cong\mathbb{H},\\
+&\GeoA(\CC^{s+t})\cong\GeoA(\RR^{s,t})\otimes_\RR\CC,\quad
+\GeoA(\CC^{0})\cong\CC,\quad
+\GeoA(\CC^{1})\cong\CC\oplus\CC,\quad
+\GeoA(\CC^{n+2})\cong\GeoA(\CC^2)\otimes_\CC\GeoA(\CC^n)\cong\GeoA(\CC^n)\otimes_\CC \CC^{2\times2},
+\end{align*}
+\[
+  \text{$s+t$ is odd and $I^2=-1$}\then \GeoA(\RR^{s,t})\cong\GeoA^+(\RR^{t,s})\otimes \CC\cong\GeoA(\CC^{s+t-1}).
+\]
+
+\paragraph{Orthogonal Group}
+The orthogonal (and related) group for $(V,Q)$ is defined by
+\begin{align*}
+&\gO(V,Q)\deq \{f\colon V\to V\colon\text{linear bijection s.t. $Q\circ f=Q$, i.e., $\forall v\in V, Q(f(v))=f(v)$}\},\\
+&\gSO(V,Q)\deq \{f\in\gO(V,Q)\mid \det f=1\},\qquad\gSO(V,Q)\deq\text{the part of $\gSO(V,Q)$ connected to 1},
+\end{align*}
+where the determinant is defined with identifying $V\to V$ as a matrix.
+
 \paragraph{Groups in Geometric Algebra}
-The following groups are embedded in $\GeoA(V,Q)$,
-where $V^\times=\{v\in V\mid Q(v)\neq 0\}$:
+Here we assume $K=\RR$ and $\GeoA$ is non-degenerate, having $\GeoA(\RR^{s,t})$ in mind.
+The geometric algebra $\GeoA(V,Q)$ contains a group
+\begin{equation}
+\GeoA^\times \deq \{g\in\GeoA\mid \exists g^{-1}\in\GeoA, gg^{-1}=g^{-1}g=1\}
+\end{equation}
+with its addition and scalar multiplication being forgotten.
+Furthermore, with $V^\times=\{v\in V\mid Q(v)\neq 0\}$, it contains
 \begin{align}
-  &\GeoA^\times \deq \{g\in\GeoA\mid \exists g^{-1}\in\GeoA, gg^{-1}=g^{-1}g=1\},
-  \qquad \Gamma\deq\{v_1v_2\cdots v_n\mid v_i\in V^\times\},
+  &
+  \Gamma\deq\{v_1v_2\cdots v_n\mid v_i\in V^\times\},
   \qquad \tilde \Gamma\deq \{g\in\GeoA^\times\mid gVg^{-1}\in V\},
@@
  \qquad\Spin\deq \Pin\cap\GeoA^+,
- \qquad\Spin^+\deq \{x\in\Spin\mid xx^\dagger=1\},
+ \qquad\Spin^+\deq \{x\in\Spin\mid xx^\dagger=1\}.
 \end{align}
@@
 
-There are surjective homomorphisms with kernel $\pm1$ for
+Consider a function $R\colon \RR\to\Spin^+(V,Q)$.
+It is known that $R'(0)\in\GeoA^2(V,Q)$, i.e., with considering the rotor group $\Spin^+$ as a Lie group, we see the corresponding Lie algebra $\mathfrak{spin}$ is equal to $\GeoA^2$ (bivector Lie algebra), where the product of $\mathfrak{spin}$ is the commutator in $\GeoA$.
+Namely,
+\begin{equation}
+\aSO(V,Q)\cong\mathfrak{spin}(V,Q)=\GeoA^2(V,Q),\qquad
+\pm\exp\left(\GeoA^2(V,Q)\right)\subset\Spin^+(V,Q).
+\end{equation}
+
+There is an epimorphism with kernel $\RR^\times$
+\begin{equation}
+\tAd\colon \Gamma(V,Q)\to\gO(V,Q),\quad g\to (v\mapsto g^\star v g^{-1}),\qquad(\therefore\quad \Gamma(V,Q)/\RR^\times\cong\gO(V,Q)).
+\end{equation}
+Here, $v\in V$, $g\in \Gamma$, and it is easily shown that $g^\star v g^{-1}\in V$, which concludes $v\mapsto g^\star v g^{-1}$ is a linear bijection $V\to V$.
+
+\medskip
+
+Focusing on $\GeoA(\RR^{s,t})$, we have epimorphisms with kernel $\pm1$:
 \begin{equation}
@@
   \Spin(s,t)\to\gSO(s,t),\qquad
-  \Spin^+(s,t)\to\gSO^+(s,t).
+  \Spin^+(s,t)\to\gSO^+(s,t),
 \end{equation}
+Furthermore, if $s\ge1$ and $t\ge 1$, $\Spin^+$ is of importance because, with any $e^\pm\in V$ s.t. $e^\pm e^\pm=\pm1$ and $e^+e^-+e^-e^+=0$,
+\begin{equation}
+\Pin(s,t)=\Spin^+(s,t)\cdot\{1,e^+,e^-,e^+e^-\},\qquad
+\Spin(s,t)=\Spin^+(s,t)\cdot\{1,e^+e^-\}.
+\end{equation}
+It is also known that 
+\begin{align}
+&\text{$\Spin^+(s,t)$ is simply connected for $(s,t)=(\text{1~or~0},n)$ and $(n,\text{1~or~0})$ with $n\ge3$},\\
+&
+\text{if $s\le1$ or $t\le1$, }\Spin^+({s,t})=\pm\exp\left(\GeoA^2(\RR^{s,t})\right)
+\end{align}
+with the minus part requires only for $(0,0)$, $(1,0\text{--}3)$, and $(0\text{--}3,1)$.
+
+\subsubsection{Representations}
+\paragraph{Primer}
+As given in \eqref{eq:lorentz4parts},
+a Lorentz transformation is an element of $\gO(1,3)$ and the proper orthochronous part $L_0$ is isomorphic to $\gSO^+(1,3)$, which is doubly covered by $\Spin^+(1,3)=\pm\exp(\GeoA^2(\RR^{1,3}))$.
+The corresponding geometric algebra $\GeoA(\RR^{1,3})$ can be embedded in $\GeoA(\CC^4)\cong\CC^{4\times 4}$ and thus has a representation by complex $4\times4$ matrices:
+\begin{align*}
+&L_0\cong \gSO^+(1,3)\cong\Spin^+(1,3)/\ZZ_2\subset L\cong \gO(1,3)\cong\Pin(1,3)/\ZZ_2 \subset \Gamma(\RR^{1,3})\subset \GeoA^\times(\RR^{1,3})\qquad\text{(as groups)},\\
+&\GeoA(\RR^{1,3})\cong\GeoA(\RR^4)\cong\GeoA(\RR^{0,2})\otimes\GeoA(\RR^{2,0})\cong\RR^{2\times2}\otimes\mathbb{H},\qquad
+\GeoA(\RR^{1,3})\otimes_\RR \CC\cong\GeoA(\CC^4)\cong\CC^{4\times4},\\
+&\GeoA(\CC^4)\cong\GeoA(\RR^{1,4})\quad\text{if $(e_0e_1e_2e_3e_4)^2=-1$ }\qquad\text{(as algebras)}.
+\end{align*}
+The above equations are slightly modified for $(\mppp)$-metric. In particular,
+$\GeoA(\RR^{3,1})\cong\GeoA(\RR^{2,0})\otimes\GeoA(\RR^{1,1})\cong\RR^{4\times 4}$ and
+$\GeoA(\RR^{3,1})\otimes_\RR \CC\cong\CC^{4\times4}$.
+
+The isomorphism $\GeoA(\RR^{1,3})\cong\RR^{2\times2}\otimes\mathbb{H}$ motivates the representation
+\begin{equation}
+  g^\mu\req\left\{\spmat{1&0\\0&-1},\spmat{0&\ii\\\ii&0},\spmat{0&\mathrm{j}\\\mathrm{j}&0},\spmat{0&\mathrm{k}\\\mathrm{k}&0}\right\}.
+\end{equation}
+Here and hereafter, $\req$ denotes that the algebraic object in the left-hand side is represented by a matrix in the right-hand side.
+The basis of $\GeoA(\RR^{1,3})$ is represented by, with representing the elementary quaternions $\mathrm{i,j,k}$ by $q$,
+\[
+  1\req\spmat{1&0\\0&1},~~
+  g^\mu\req\spmat{1&0\\0&-1},\spmat{0&q\\q&0},~~
+  \tfrac12[g^0,g^i]\req\spmat{0&q\\-q&0},~~
+  \tfrac12[g^{i+1},g^{i+2}]\req\spmat{q&0\\0&q},~~
+  g^\mu I\req\spmat{0&-1\\-1&0},\spmat{q&0\\0&-q},~~
+  I\req\spmat{0&-1\\1&0},
+\]
+which explicitly shows that they span $\RR^{2\times2}\otimes\mathbb{H}$.
+However, this representation has little connection to the popular ones.
+Although we get $4\times4$ matrices
+$
+  g^\mu=\left\{\spmat{1&0\\0&-1},-\ii\spmat{0&\sigma_i\\-\sigma_i&0}\right\}
+$
+by replacing quaternions by Pauli matrices $\mathrm{\{i,j,k\}}\mapsto-\ii\{\sigma_1,\sigma_2,\sigma_3\}$, we need to perform a basis change, $\hat\gamma^\mu=\hat R^{-1}g^\mu \hat R$ with $\hat R=\diag(1,1,\ii,\ii)$, to reach the Dirac representation, but this is not allowed within $V=\RR$.
+
+\paragraph{Dirac and Chiral representations}
+We instead use the isomorphism $\GeoA(\RR^{2,3})\cong\GeoA(\CC^4)$. By defining $g^5\deq\ii g^0g^1g^2g^3$,
+\[
+\{g^0,g^1,g^2,g^3,g^5\}\req \left\{\spmat{1&0\\0&-1},-\ii\spmat{0&\sigma_i\\-\sigma_i&0},\spmat{0&-\ii\\\ii&0}\right\}
+\]
+forms a basis of $\GeoA(\RR^{2,3})$ and thus the 32 combinations among them form a basis of $\CC^{4\times4}$.
+Then, we are allowed to use the above $\hat R=\diag(1,1,\ii,\ii)$ to get the Dirac representation and the Chiral representation:
+\begin{align}
+&\hat\gamma^0=\hat R^{-1}g^0 \hat R=\spmat{1&0\\0&-1},\quad
+\hat\gamma^i=\hat R^{-1}g^i \hat R=\spmat{0&\sigma^i\\-\sigma^i&0},\quad
+\hat\gamma^5=\hat R^{-1}g^5 \hat R=\ii\hat\gamma^0\hat\gamma^1\hat\gamma^2\hat\gamma^3=\spmat{0&1\\1&0},\\
+&\gamma^0=R^{-1}g^0 R=\spmat{0&1\\1&0},\quad
+\gamma^i=R^{-1}g^i R=\spmat{0&\sigma^i\\-\sigma^i&0},\quad
+\gamma^5=R^{-1}g^5 R=\ii\gamma^0\gamma^1\gamma^2\gamma^3=\spmat{-1&0\\0&1};\quad
+R\deq\spmat{1&1\\-\ii&\ii}.
+\end{align}
+The pseudoscalar is $I=-\ii$ in both representations.
+(Notice that, for $(\mppp)$-metric, we should use $\GeoA(\RR^{4,1})\cong\GeoA(\CC^4)$ by defining $g^5\deq g^0g^1g^2g^3$, where $I=-\ii$; this choice is not included in the following discussion.)
+
+In Chiral or Dirac representation, the conjugate operations are given by, with $A^{\TT*}$ being  the conjugate-transpose of $A$,
+\begin{equation}
+  \text{Grade involution:}\quad -\gamma^2 A^* \gamma^2,\qquad
+  \text{Reversion}\quad \gamma^1\gamma^3 A^\TT \gamma^3\gamma^1,\qquad
+  \text{Clifford conjugate}\quad \gamma^5\gamma^0A^{\TT*}\gamma^0\gamma^5.
+\end{equation}
+Note that $\gamma^a$ and $\hat\gamma^a$ are all unitary (and traceless) and the 32 basis matrices are all unitary.
+
+Let us give the explicit expressions for the spaces. The vector space $V$ is given by
+\begin{equation}
+V=\{v_0\gamma^0+\cdots+v_4\gamma^5\mid v_i\in\RR\},
+\end{equation}
+where all the elements are traceless.
+The inner product and norm are given by
+\begin{equation}
+B(v,w)=(vw+wv)/2=(v_0w_0-v_1w_1-v_2w_2-v_3w_3+v_4w_4)E,\quad
+Q(v)=vv=(v_0^2-v_1^2-v_2^2-v_3^2+v_4^2)E,
+\end{equation}
+where $E$ is the $4\times 4$ identity matrix. Since $v^\dagger=v$, $v^\star=-v$, and $vv^\dagger =Q(v)$, and $\det v=Q(v)^2$ for $v\in V$,
+\begin{equation}
+V^\times=\{v\in V \mid  \det v>0\},\quad
+v^{-1}=v/Q(v).
+\end{equation}
+The versor group $\Gamma$ is characterized by the invertible matrices $v_1,\dots, v_n$, and thus $5n$ real numbers $v_{10},\dots, v_{n4}$:
+\begin{equation}
+  \Gamma(2,3)=\{v_1v_2\cdots v_n\mid v_i\in V, \det v_i>0\}.
+\end{equation}
+For $x=v_1\cdots v_n\in \Gamma$, $xx^\dagger = v_1\cdots v_n v_n^\dagger\cdots v_1^\dagger=\prod Q(v_i)$. Therefore,
+\begin{equation}\begin{split}
+  &\Pin(2,3)=\{v_1v_2\cdots v_n\mid v_i\in V, \det v_i=1\},\\
+  &\Spin(2,3)=\{v_1v_2\cdots v_{2n}\mid v_i\in V, \det v_i=1\},\qquad
+  \Spin^+(2,3)=\{v_1v_2\cdots v_{2n}\mid v_i\in V, Q(v_i)=1\}.
+\end{split}\end{equation}
+The epimorphism $\tAd$, which induces the isomorphism $\Gamma(2,3)/\RR^\times\cong\Pin(2,3)/\ZZ_2\cong \gO(2,3)$, is described as
+\begin{align*}
+&x\in\Gamma,\quad x\req x_1\cdots x_n,\quad x_i\in V, \\
+&v\in V, \quad v \req v_0\gamma^0+\cdots+v_4\gamma^5,\quad v_a\in\RR,\qquad
+\tAd_x(v)=x^\star v x^{-1}\in V
+\end{align*}
+and it preserves the norm: $Q(\tAd_x(v))=\tAd_x(v)\tAd_x(v)=Q(v)=vv$.
+If $s\in\Pin(2,3)$,
+\[
+\tAd_s(v)=s^\star v s^{-1}=(-1)^ns_1\cdots s_n\,v\,s_n\cdots s_1.
+\]
+
+\paragraph{Lorentz Group} Consider $\tAd_w(v)$ with $w=w_a\gamma^a$ limited to $V$.
+If $w_5=0$, $\tAd_w$ does not modify the fifth component.
+So, we can restrict $\Gamma(2,3)$ to $\Gamma(1,3)$ and so on:
+\begin{align}
+  &V_4=\{v_0\gamma^0+\cdots+v_3\gamma^3\mid v_i\in\RR\};\\
+  &\text{for~} x\in \{x_1x_2\cdots x_n\mid x_i\in V_4,\det x_i=1\},\qquad \tAd_x\text{~corresponds to an element in~}\gO(1,3) \quad\text{(double cover)}.
+\end{align}
+Note that, for $v\in V_4$,
+\[
+  v^{-1}=v/Q(v),\quad
+  v^*=-\gamma^2 v^\star \gamma^2=\gamma^2 A\gamma^2,\quad
+  v^\TT=\gamma^1\gamma^3 v\gamma^3\gamma^1,\quad
+  v^{\TT*}=\gamma^0 v\gamma^0.
+\]
+
+The group $\Spin^+(1,3)$ is simply connected and isomorphic to $\pm\exp(\GeoA^2(\RR^{1,3}))$, where
+\begin{equation}
+  \GeoA^2(\RR^{1,3})=\mathop{\mathrm{span}}\left[
+    \gamma^0\gamma^i\req\spmat{-\sigma^i&0\\0&\sigma^i},\quad
+    \gamma^i\gamma^j\req\epsilon^{ijk}\spmat{-\ii\sigma^k&0\\0&-\ii\sigma^k}.
+   \right],
+\end{equation}
+and, in this representation, the isomorphism \TODO{???} is explicit.
+
+\paragraph{Spinor} In Chiral representation, a matrix $v=v_0\gamma^0+\cdots+v_3\gamma^3$ in $V_4$ can be expressed by
+\[
+ v = \psi\psi^{\TT*}\gamma^0;\quad \psi\deq
+ \frac{1}{\sqrt{2(v_0-m)}}
+ \left[(v_0 - m)\gamma^0 + v_1\gamma^1 + v_2\gamma^2 + v_3\gamma^3\right],\quad m\deq\sqrt{v_0^2-v_1^2-v_2^2-v_3^2}.
+\]
+Thus, the operation $\tAd_s(v)$ is written by
+\[
+\tAd_s(v)=s^\star v s^{-1}
+=(-1)^ns_1\cdots s_n \psi\psi^{\TT*}\gamma^0 s_n\cdots s_1
+=(-1)^n(s_1\cdots s_n \psi)(s_1\cdots s_n\psi)^{\TT*}\gamma^0.
+\]
+and each column of $\psi$ is understood as a Dirac spinor.
+With identifying, as usual, $v_\mu$ as the momentum, we have
+\[
+\psi= \frac{1}{\sqrt{2\tilde E}}\pmat{
+  0 & 0 & \tilde E+p_z & \rho^* \\
+  0 & 0 & \rho & \tilde E-p_z \\
+  \tilde E-p_z & -\rho^* & 0 & 0 \\
+  -\rho & \tilde E+p_z & 0 & 0 \\
+};\qquad \rho\deq p_x+\ii p_y\in\CC,\quad \tilde E\deq E-m\in\RR.
+\]
+
+\subsubsection{Fragments}
+\paragraph{For Minkowski spacetime}
 Also, for our interest, there are isomorphisms~\cite{RauschdeTraubenberg:2005aa,Yamaguchi:spinor}
@@
 
-\paragraph{Clifford algebra and spin group}\TODO{needs experts...}
-We just provide a little information on spin group $\mathop{\mathrm Spin}(1,3)$ associated to (our) Minkowski metric $\eta=\diag(+,-,-,-)$:~\cite{RauschdeTraubenberg:2005aa,Yamaguchi:spinor,Wikipedia:Clifford_Algebra}.
-To define the Clifford algebra $\mathfrak C(V,Q)$, let
-\begin{itemize}
- \item[$V$]: a vector space over a field $K$; in our case, $V=\mathbb R^{1,3}$ and $K=\mathbb R$.
- \item[$Q$]: a quadratic form on $V$,
- \item[$A$]: %
-  a unital associative algebra over $K$; it means that $(A, +, \ast)$ is a ring and is also a vector space on $K$, it should satisfy
-  $(k a_1)\ast a_2=a_1\ast(ka_2)=k(a_1\ast a_2)$ for $a_i\in A$ and $k\in K$, and it should contain $1_A$ such that $1_A\ast a=a\ast 1_A=a$;
-  here a map $f:K\to A$ with $f(k)=k1_A$ is an injective ring homomorphism, with which $K$ can be included in $A$.
-\end{itemize}
-Then we can define a linear map $i:V\to \mathfrak C(V,Q)$ satisfying $i(v)\ast i(v)=Q(v)1_A$ for all $v\in V$ as follows, which defines the Clifford algebra.
-Given a linear map $j:V\to A$ satisfying $j(v)\ast j(v)=Q(v)1_A$ for all $v\in V$, there is a unique algebra homomorphism $f:\mathfrak C(V,Q)\to A$ such that $f\circ i=j$.
-
-We focus on $\mathbb R^{1,3}$ and $Q(v)=\eta_{\mu\nu}v^\mu v^\nu$ (Minkowski norm). Then, by introducing the basis $e_\mu$ of $V$, the condition yields
-\begin{align}
-& (v^2+2v^\mu w_\mu + w^2)1_A = [j(v)+j(w)]\ast[j(v)+j(w)] = (v^2+w^2)1_A+j(v)\ast j(w)+j(w)\ast j(v),\notag\\
-&\therefore \{j(v),j(w)\}=2v^\mu w_\mu 1_A
-\text{~~and thus~~} \{e_\mu,e_\nu\}=2\eta_{\mu\nu} 1_A.\label{eq:ClifAl}
-\end{align}
-The Clifford algebra $\mathfrak C(\mathbb R^{1,3},\eta)$ is thus seen as the associative algebra generated by $\{1_A, e_0, e_1, e_2, e_3\}$ satisfying \cref{eq:ClifAl}.
-
-Several properties of the Clifford algebra $\mathfrak C_{1,3}$ is in order:
-\begin{itemize}
- \item It is of dimension $2^4=16$ and the basis is given by
-$\{1, e_\mu, (e_\mu e_\nu)_{\mu<\nu}, (e_\mu e_\nu e_\rho)_{\mu<\nu<\rho}, e_0e_1e_2e_3\}.$
- \item It contains a subalgebra $\mathfrak C^0_{1,3}$ spanned by $\{1, e_\mu e_\nu, e_0e_1e_2e_3\}$, while the remaining vector space $\mathfrak C^1_{1,3}$ is by $\{e_\mu, e_\mu e_\nu e_\rho\}$.
- \item We define $\epsilon=e_0e_1e_2e_3$, which satisfies $\epsilon^2=-1$ and $[\epsilon,\mathfrak C^0_{1,3}]=\{\epsilon,\mathfrak C^1_{1,3}\}=0$.
- \item $\aSO(1,3)\subset\mathfrak C_{1,3}$; explicitly, the generators are given by $S_{\mu\nu}:={[e_\mu,e_\nu]}/{4}$, which satisfy
-\begin{equation}
- [S_{\mu\nu},e_\rho] = e_\mu\eta_{\nu\rho}-e_\nu\eta_{\mu\rho},\qquad
- [S_{\mu\nu},S_{\rho\sigma}] = 
--\eta_{\mu\rho}S_{\nu\sigma}
-+\eta_{\nu\rho}S_{\mu\sigma}
-+\eta_{\mu\sigma}S_{\nu\rho}
--\eta_{\nu\sigma}S_{\mu\rho}.
-\end{equation}
-\end{itemize}
-
-The Clifford group is defined by
-\begin{equation}
-\Gamma_{1,3}=\{x\in\mathfrak C_{1,3}\mathbin|\forall v\in \mathbb R^{1,3}, xvx^{-1}\in \mathbb R^{1,3}\},
-\end{equation}
-i.e., $x$ should return a ``vector'' $v^\nu e_\nu\in \mathbb R^{1,3}$ into another ``vector'', which in fact corresponds to the Lorentz transformation:
-$xe_\nu x^{-1} = (\Lambda^{-1})_\nu{}^\mu e_\mu$.
-The correspondence is however one-to-$N$, and thus it should be normalized by the squared-norm $N(x)$
-\begin{equation}
- N(x)=x\overline{x};\qquad\overline{e_\mu}=e_\mu,\quad\overline{ke_\mu e_\nu e_\rho}=ke_\rho e_\nu e_\mu, \text{etc., where $k\in\mathbb R$}.
-\end{equation}
-Then, we define
-\begin{align*}
- \mathop{\mathrm{Pin}}(1,3)&=\{x\in\Gamma_{1,3}\mathbin||N(x)|=1\},&
- \mathop{\mathrm{Spin}}(1,3)&=\mathop{\mathrm{Pin}}(1,3)\cap \mathfrak C^0_{1,3},&
- \mathop{\mathrm{Spin}}(1,3)^+&=\{x\in\mathop{\mathrm{Spin}}(1,3)\mathbin|N(x)=1\},
-\end{align*}
-whose properties are in order:
-\begin{itemize}
- \item $\mathop{\mathrm{[S]pin}}(1,3)$ is a double covering of $\mathop{\mathrm{[S]O}}(1,3)$ but it is not connected.
- \item $\mathop{\mathrm{Spin}}(1,3)^+$ is a connected Lie group and is a double covering of $\gSO(1,3)^+$.
- \item $\mathop{\mathrm{Spin}}(1,3)^+\cong\gSL(2;\mathbb C)\cong\gSp(2;\mathbb C)$.
-\end{itemize}
+\TODO{SO algebra and Grade-2 Clifford Algebra}
 

### 2025-09-10 cosmology.tex
update particle cosmology

@@
-% Time-Stamp: <2022-05-16 15:23:42>
+% Time-Stamp: <2025-09-10 14:08:10>
 
@@
 \subsection{Particle cosmology}
-The particle number density, pressure, and energy density are calculated from distribution functions:
+\subsubsection{Particle distribution and density}
+In kinematic equilibrium, the particle number density, energy density, and pressure (i.e., (momentum)$\times$(flux per time) against a ``wall'') are calculated from distributions
 \begin{alignat}{3}
@@
  f\w{BE}(\vc k)&=\frac{g}{\ee^{(E-\mu)/T}-1},\qquad&
- f\w{FD}(\vc k)&=\frac{g}{\ee^{(E-\mu)/T}+1};
+ f\w{FD}(\vc k)&=\frac{g}{\ee^{(E-\mu)/T}+1},
 \end{alignat}
@@
  p   &=\int\ddP{k}k_zv_z f(\vc k)
-      =\int\ddP{k}\frac{k^2\cos^2\theta}{E} f(\vc k).
+      =\int\ddP{k}\frac{k^2\cos^2\theta}{E} f(\vc k),
 \end{alignat}
-Note the pressure is (momentum)$\times$(flux per time) on a ``wall''; assuming MB, $p=\rho/3$ for $m\ll T$ and $p=T\rho/m$ for $m\gg T$.
+where we, in principle, take $\mu<m$ for Bose--Einstein statistics.
+It is also useful to define the difference of particle and antiparticle number densities, assuming their chemical potential is opposite in sign:
+\begin{equation}
+  n_\Delta \deq n_+ - n_- = \int\ddP{k}\Bigl[f(\vc k,\mu) - f(\vc k,-\mu)\Bigr].
+\end{equation}
 
+For Maxwell--Boltzmann (MB) distribution, they are given explicitly by, with $\nu\deq \mu/T$ and $x\deq m/T$,
+\begin{equation}
+  n\w{MB} = \frac{gT^3\ee^{\nu}}{2\pi^2}x^2K_2(x),\quad
+  %n\w{\Delta;MB} = g\sinh{\nu}\cdot\frac{T^3}{\pi^2}x^2K_2(x),\quad
+  \rho\w{MB} = \left(3+\frac{xK_1(x)}{K_2(x)}\right)T n\w{MB},\quad
+  p\w{MB} = T n\w{MB},
+\end{equation}
+which reduce in special cases to
+\begin{align}
+  &\text{[massless: $m\ll T$]}
+  &&n\w{MB} = \frac{gT^3\ee^{\nu}}{\pi^2}\left(1-\frac14 x^2+\Order(x^4)\right),\quad
+  \rho\w{MB} = 3Tn\w{MB}\left(1+\frac16x^2+\Order(x^4)\right),
+\\&\text{[low-temperature: $m\gg T$]}
+  &&n\w{MB} \approx \ee^{(\mu-m)/T}\left({\frac{m T}{2\pi}}\right)^{3/2},\quad
+  \rho\w{MB} \approx \left(m+\frac32T\right)n\w{MB}.
+\end{align}
+
+\pagebreak
+
+For Fermi--Dirac or Bose--Einstein distributions, the results are not expressed analytically, even with $\mu=0$. Namely,
+\begin{equation}
+  n=\int_m^\infty\frac{\dd E}{2\pi^2}\frac{gE\sqrt{E^2-m^2}}{\ee^{(E-\mu)/T}\pm1},
+  \quad
+  \rho=\int_m^\infty\frac{\dd E}{2\pi^2}\frac{gE^2\sqrt{E^2-m^2}}{\ee^{(E-\mu)/T}\pm1},
+  \quad
+  p=\frac13\int_m^\infty\frac{\dd E}{2\pi^2}\frac{g(E^2-m^2)^{3/2}}{\ee^{(E-\mu)/T}\pm1}.
+\end{equation}
+Analytical expressions are available for the massless limit as
+\begin{align}
+    &\text{[massless: $m\ll T$]}
+  &&
+  n\w{FD}    = \frac{g L_3^-}{\pi^2}+\Order(x),\quad
+  \rho\w{FD} = \frac{3g L_4^-}{\pi^2}+\Order(x^2),\quad
+  p\w{FD}    = \frac{g L_4^-}{\pi^2}+\Order(x^2),\\
+   &\text{[massless: $m\ll T$; $\mu<0$]}
+  &&
+  n\w{BE}    = \frac{g L_3^+}{\pi^2}+\Order(x),\quad
+  \rho\w{BE} = \frac{3g L_4^+}{\pi^2}+\Order(x^2),\quad
+  p\w{BE}    = \frac{g L_4^+}{\pi^2}+\Order(x^2),
+\end{align}
+where $L^\pm_n=\pm\mathop{\text{\texttt{PolyLog}}}(n,\pm\ee^{\mu/T})$.
+Note we may need thermal masses for proper discussion, especially for the Bose--Einstein case, which is well-defined only for $\mu<m$.
+
+The difference $n_\Delta$ can be expressed under the assumption $\mu/T\ll1$:
+\begin{align}
+ &\text{[near equilibrium: $\mu\ll T$]}
+  &&
+ (n_\Delta{})^{\mathrm{BE}}_{\mathrm{FD}}
+ =\int_{m}^\infty
+ \frac{\dd E}{\pi^2} \frac{gE\sqrt{E^2-m^2}\ee^{E/T}}{(\ee^{E/T}\mp1)^2}\frac{\mu}{T}
+ +\Order\left(\frac{\mu^3}{T^3}\right)
+ =g\mu T^2\xi_\mp(x)
+ +\Order\left(\frac{\mu^3}{T^3}\right),
+\end{align}
+where $\displaystyle
+ \xi_\mp(x)\deq \int_{x}^\infty\frac{k\ee^{k}\sqrt{k^2-x^2}\,\dd k}{(\ee^{k}\mp1)^2\pi^2}
+$ can be calculated numerically and remain finite.
+For fermions,
+\begin{align}
+ \text{[$\mu\ll T$, $m\ll T$]}\qquad
+ (n_\Delta{})_{\mathrm{FD}} = g\mu\left[\frac{T^2}{6} - \frac{m^2}{4\pi^2}
+ +\Order\left(\frac{m^4}{T^4}\right)\right]
+ +\Order\left(\frac{\mu^2}{T^2}\right).
+\end{align}
+For Bose--Einstein case, it results in $\xi_-(0)=1/3$ and hence $n_{\Delta;\mathrm{BE}}\questeq g\mu T^2/3$ with no divergence, but we may discuss it is invalid because of the implicit assumption $\mu<m$.
+
+\subsubsection{Thermal average of cross section and decay rate}
 A thermal average of a cross section $\sigma(s)$ is schematically given by

### 2025-09-22 Neutrino.tex
upgrade cls
neutirno and comosologgasdafafasd

@@
 \paragraph{Connection of the above two models}
-If we set $M_N =0$ in the Type-I see-saw ($n\w{RHN}=3$), the AT-diagonalization becomes
+If we set $M_N =0$ in the Type-I see-saw ($n\w{RHN}=3$), AT-diagonalization results in 
 \begin{align}
@@
   \tilde R=\pmat{V_n^* & \ii V_n^* \\ U_n & -\ii U_n},\quad
-  \tilde M\wdiag=\pmat{M\w{D}\wdiag & 0 \\ 0 & M\w{D}\wdiag}.
+  \tilde M\wdiag=\pmat{M\w{D}\wdiag & 0 \\ 0 & M\w{D}\wdiag}
 \end{align}
-As three pairs of degenerate Weyl fermions form three Dirac neutrinos, the Dirac neutrino model is reproduced as expected.
+and we have three pairs of degenerate Weyl fermions, i.e., three Dirac neutrinos, which is equivalent to the Dirac neutrino model.
+
+\subsection{Weinberg Operator}
+The neutrino masses are often described by the Weinberg operator~\cite{Weinberg:1979sa}, which is obtained by integrating out the right-handed neutrinos.
+It is given by, usually with the convention $C_{ij}=C_{ji}$ because of $(L_i H)(L_j H)=(L_j H)(L_i H)$,
+\begin{equation}
+  \mathcal L\ni \frac{C_{ij}}{2}(L_i H)(L_j H)+\text{H.c.}
+  \equiv
+  \frac{C_{ij}}{2}\epsilon^{ab}\epsilon^{cd}(L_i^a H^b)^\alpha(L_j^c H^d)_\alpha+\text{H.c.}
+  \leadsto
+  \frac{C_{ij}}{2}\mean{H^0}^2(\nu_i\nu_j)+\text{H.c.}
+\end{equation}
+The (Majorana) mass matrix is thus given by $C_{ij}\mean{H^0}^2=C_{ij}v^2/2$ with $v=246\GeV$.
+The overall sign is unphysical here but may appear in the PMNS matrix.
 
@@
 As experiments have not found any deviation from $[U]\w{upper left}$ being unitary, we express the mixing with the PMNS matrix
- \GRAY{(Pontecorvo--\JAPANESE{牧}--\JAPANESE{中川}--\JAPANESE{坂田})}
+ \GRAY{(Pontecorvo--\JA{牧}--\JA{中川}--\JA{坂田})}
 parameterized by
